---
title: "scRNAseq_CITEseq_src"
author: "Laëtitia Racine"
date: "2022-09-02"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

# scRNAseq_CITEseq_R_CreateSeuratObject.Rmd

*OBJECTIVES*      
- load cell ranger output files
- create seurat object with RNA and ADT assays
- update orig.ident with the name of the condition
- visualize nb of cells and features for each condition
- visualize CTRL object

*INPUTS*   
(samples from batch1 are sample_ feature and those from batch2 are filtered_feature)     
- scRNAseq_[DRUG]_sample_feature_bc_matrix_barcodes.tsv.gz  
- scRNAseq_[DRUG]_sample_feature_bc_matrix_features.tsv.gz  
- scRNAseq_[DRUG]_sample_feature_bc_matrix_matrix.mtx.gz  
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_barcodes.tsv.gz  
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_features.tsv.gz    
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_matrix.mtx.gz    
  
*OUTPUTS*  
- list_seurat_obj.rds  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scRNAseq/", "scRNAseq_CITEseq_R_CreateSeuratObject.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scRNAseq_CITEseq_R_CreateSeuratObject.html"))

```

# scRNAseq_CITEseq_R_FilterData.Rmd

*OBJECTIVES*  
Apply qualityt control filters on the dataset : 
- calculate mitochondrial and ribosomal RNA percentage 
-  filter cells based on number of genes detected (between 1000 and 6500)
- filter cells based on mitochondrial and ribosomal RNA percentage (<25% and >5%)
- remove mitochondrial genes 
- create a summary tab on filters (number cells/genes after each filter)
- create QC plots
- create tabs describing the filtered dataset

*INPUTS*        
- list_seurat_obj.rds  
  
*OUTPUTS*      
- plot_feature_filter.png  
- plot_percentMTRB_filter.png  
- filter_tab_seurat_obj.csv  
- summary_cell_tab.csv  
- summary_cond_tab.csv  
- individual_gene_tab.csv  
- tab_UMI.csv  
- list_mat_raw.rds  
- list_seurat_obj_qc.rds  
- list_seurat_obj_qc_each_filter.rds  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scRNAseq/", "scRNAseq_CITEseq_R_FilterData.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scRNAseq_CITEseq_R_FilterData.html"))

```

# scRNAseq_CITEseq_R_NormalizeData.Rmd

*OBJECTIVES*  
- lognormalize RNA assay  
- find 2000 most variable features   
- scale data   
- visualize cell cycle influence   
- create cell cycle plots  
- normalize RNA assay with SCT : create a new assay  
- create a new object with cell cycle regression if needed for later   
- normalize ADT assay with CLR method  
  
*INPUT*    
- list_seurat_obj_qc.rds    
    
*OUTPUTS*    
- plot_2000variablesfeatures.svg  
- plot_cellcyclescoring.svg  
- plot_cellcycle_statUMI.svg  
- tab_cellcycle.csv  
- tab_genes_cellcycle.csv  
- list_seurat_obj_qc_norm_ccr.rds  
- list_seurat_obj_qc_norm.rds  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scRNAseq/", "scRNAseq_CITEseq_R_NormalizeData.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scRNAseq_CITEseq_R_NormalizeData.html"))

```

# scRNAseq_CITEseq_R_BatchEffectCorrection.Rmd

*OBJECTIVES*
- merge seurat objects per batch
- visualize the data per batch (PCA, UMAP) based on sct normalization
- visualize the data per batch (PCA, UMAP) based on log normalization
- merge seurat objects of all batches into one object
- visualize the data (PCA, UMAP) based on sct normalization
- visualize the data (PCA, UMAP) based on log normalization

*INPUT*  
- list_seurat_obj_qc_norm.rds

*OUTPUTS*
- umap_log_batch1.png
- umap_log_batch2.png
- umap_ctrl_log.png
- umap_sct_batch1.png
- umap_sct_batch2.png
- umap_ctrl_sct.png
- umap_sct_bothbatch.png
- umap_sct_bothbatch_split.png
- umap_sct_bothbatch_ctrl.png
- umap_log_bothbatch.png
- umap_log_bothbatch_split.png
- umap_log_bothbatch_ctrl.png

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scRNAseq/", "scRNAseq_CITEseq_R_BatchEffectCorrection.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scRNAseq_CITEseq_R_BatchEffectCorrection.html"))

```


# scRNAseq_CITEseq_R_ReductionDimension.Rmd

*OBJECTIVES*
- perform PCA and determine the optimal dimensionality
- perform UMAP 

*INPUT*
- list_seurat_obj_qc_norm.rds
- list_seurat_obj_qc_norm_ccr.rds  
  
*OUTPUTS*
- dimension_choice_elb_pca.svg
- dimension_choice_pca_genes.svg  
- umap_1-40dim.svg
- umap_1-40dim_split.svg
- umap_1-40dim_split_highlight.svg
- umap_clusters_resolution.svg
- umap_clusters_res_choice.svg
- merged_all_1-40dim.rds
- dimension_choice_elb_pca_cc.svg
- dimension_choice_pca_genes_cc.svg
- umap_1-40dim_cc.svg
- umap_1-40dim_split_cc.svg
- umap_1-40dim_split_highlight_cc.svg
- umap_clusters_resolution_cc.svg
- umap_clusters_res_choice_cc.svg
- merged_cc_1-40dim.rds

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scRNAseq/", "scRNAseq_CITEseq_R_ReductionDimension.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scRNAseq_CITEseq_R_ReductionDimension.html"))

```
