---
title: "scRNAseq_CITEseq_src"
author: "Laëtitia Racine"
date: "2022-09-02"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

This script is a "knit_all" document. It allows to run all the script from pre-processing steps for scRNAseq-CITE-seq dataset in the right order and to produce html files in the *report* folder.    
  

# scRNAseq_CITEseq_R_CreateSeuratObject.Rmd

*OBJECTIVES*        
- load cell ranger output files  
- create seurat object with RNA and ADT assays for each condition  
- update orig.ident with the name of the condition  
- visualize nb of cells and features for each condition  
- visualize CTRL object  
  
*INPUTS*   
(samples from batch1 are sample_feature and those from batch2 are filtered_feature)       
- scRNAseq_[DRUG]_sample_feature_bc_matrix_barcodes.tsv.gz    
- scRNAseq_[DRUG]_sample_feature_bc_matrix_features.tsv.gz    
- scRNAseq_[DRUG]_sample_feature_bc_matrix_matrix.mtx.gz    
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_barcodes.tsv.gz    
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_features.tsv.gz      
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_matrix.mtx.gz    
- functions_constants.R    
    
*OUTPUTS*    
- list_seurat_obj.rds    
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_CreateSeuratObject.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_CreateSeuratObject.html"))

```


# scRNAseq_CITEseq_R_FilterData.Rmd

*OBJECTIVES*    
Apply cumulative quality control filters on the dataset :   
- calculate mitochondrial and ribosomal RNA percentage     
- filter cells based on number of genes detected (keep those between 1000 and 6500)    
- filter cells based on mitochondrial and ribosomal RNA percentage (keep those <25% and >5%)  
- remove mitochondrial genes  
- filter genes detected in less than 3 cells  
- create a summary tab on filters (number cells and genes after each filter)
- create QC plots
- create tabs describing the filtered dataset  
  
*INPUTS*          
- list_seurat_obj.rds (cf. scRNAseq_CITEseq_R_CreateSeuratObject.Rmd)      
- functions_constants.R    
  
*OUTPUTS*        
- plot_feature_filter.png    
- plot_percentMTRB_filter.png    
- filter_tab_seurat_obj.csv  
- list_seurat_obj_qc_each_filter.rds  
- list_mat_raw_proteins.rds  
- list_mat_raw_genes.rds  
- summary_cell_tab.csv    
- summary_cond_tab.cs    
- individual_gene_tab.csv  
- list_seurat_obj_qc.rds  
- list_genes_detected_1cell.rds  
- list_genes_detected_inf_equal_3cells.rds  
- tab_UMI.csv    
- list_mat_detected_genes.rds  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_FilterData.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_FilterData.html"))

```


# scRNAseq_CITEseq_R_NormalizeData.Rmd

*OBJECTIVES*    
- normalize RNA assay with lognormalize + find 2000 most variable features + scale data     
- calculate cell cycle score and create cell cycle influence plots      
- normalize RNA assay with SCT without cell cycle regression      
- create a new object with SCT and cell cycle regression if needed for later      
- normalize ADT assay with CLR method  
- compare LogNorm and SCT normalizations    
      
*INPUTS*        
- list_seurat_obj_qc.rds (cf. scRNAseq_CITEseq_R_FilterData.Rmd)      
- functions_constants.R        
        
*OUTPUTS*        
- plot_2000variablesfeatures.svg    
- plot_cellcyclescoring.svg    
- plot_cellcycle_statUMI.svg    
- tab_cellcycle.csv    
- tab_genes_cellcycle.csv    
- list_seurat_obj_qc_norm_ccr.rds    
- list_seurat_obj_qc_norm.rds    
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_NormalizeData.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_NormalizeData.html"))

```


# scRNAseq_CITEseq_R_BatchEffectCorrection.Rmd

*OBJECTIVES*  
- merge seurat objects per batch (two objects to compare)   
- visualize the data per batch (PCA, UMAP) based on sct normalization  
- visualize the data per batch (PCA, UMAP) based on log normalization  
- merge seurat objects of all batches into one object  
- visualize the data (PCA, UMAP) based on sct normalization  
- visualize the data (PCA, UMAP) based on log normalization  
- determine if batch correction is needed    
    
*INPUTS*    
- list_seurat_obj_qc_norm.rds (cf. scRNAseq_CITEseq_R_NormalizeData.Rmd)   
- functions_constants.R    
    
*OUTPUTS*  
- umap_sct_batch1.png  
- umap_sct_batch2.png  
- umap_log_batch1.png  
- umap_log_batch2.png  
- umap_sct_bothbatch.png  
- umap_sct_bothbatch_split.png  
- umap_sct_bothbatch_ctrl.png  
- umap_log_bothbatch.png  
- umap_log_bothbatch_split.png  
- umap_log_bothbatch_ctrl.png  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_BatchEffectCorrection.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_BatchEffectCorrection.html"))

```


# scRNAseq_CITEseq_R_ReductionDimension.Rmd

*OBJECTIVES*  
- perform PCA and determine the optimal dimensionality with elbow plot  
- perform UMAP and visualize cells distribution per condition      
- perform clusterisation and determine the best resolution  
- prepare seurat object for markers identification  
  
*INPUTS*
- list_seurat_obj_qc_norm.rds (cf. scRNAseq_CITEseq_R_NormalizeData.Rmd)   
- list_seurat_obj_qc_norm_ccr.rds (cf. scRNAseq_CITEseq_R_NormalizeData.Rmd)     
- functions_constants.R      
      
*OUTPUTS*  
- dimension_choice_elb_pca.svg  
- dimension_choice_pca_genes.svg    
- umap_1-40dim.svg  
- umap_1-40dim_split.svg  
- umap_1-40dim_split_highlight.svg  
- umap_clusters_resolution.svg  
- umap_clusters_res_choice.svg  
- merged_all_1-40dim.rds  
- merged_all_prep_1-40dim.rds      
- dimension_choice_elb_pca_cc.svg  
- dimension_choice_pca_genes_cc.svg  
- umap_1-40dim_cc.svg  
- umap_1-40dim_split_cc.svg  
- umap_1-40dim_split_highlight_cc.svg  
- umap_clusters_resolution_cc.svg  
- umap_clusters_res_choice_cc.svg  
- merged_cc_1-40dim.rds    
- merged_cc_prep_1-40dim.rds    
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_ReductionDimension.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_ReductionDimension.html"))

```
