---
title: "src2_scATACseq_R_preprocessing"
author: "Laëtitia"
date: "2022-09-02"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(stringr)

```


# scATACseq_R-SKMK_ReportCreate&Quality = Summary of snakemake scripts

*OBJECTIVE*  
Create a html report with command lines of snakemake rules and html report.     
The two R scripts called in snakemake allow to :   
--> create a usable R object from Cell Ranger outputs  
- load cell ranger output files  
- create seurat object with chromatin assay for each condition  
- add genomic annotations from EnsDB.Hsapiens.v86 database  
- update orig.ident with the name of the condition  
--> check quality of cells and fragments detected and filter the dataset     
- filter on nucleosome banding pattern    
- filter on fragments and blaklist reads    
- filter on TSS enrichment score     
- keep only standard chromosomes    
- create QC plots
The Rmd script will display the code without running it and expose some examples (CTRL objects).  

*INPUTS of Snakemake files*         
- [DRUG]_filtered_peak_bc_matrix.h5      
- [DRUG]_singlecell.csv      
- [DRUG]_fragments.tsv.gz      
- [DRUG]_seurat.rds

*OUTPUTS of Snakemake files*     
- [DRUG]_seurat.rds  
- [DRUG]_seurat_qc.rds    
- [DRUG]_plots_qc_list.rds    
- [DRUG]_seurat_qc_filter.rds    
- [DRUG]_plots_quality_control.svg    
- [DRUG]_df_filter.csv

*INPUTS of Rmd*  
All the outputs of snakemake files.  
  
*OUTPUT of Rmd*    
An html file.  
  
```{r, snakemake report}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R-SKMK_ReportCreateQuality.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R-SKMK_ReportCreateQuality.html"))

```



# scATACseq_R_MergeObjectsCommonPeaks.Rmd

*OBJECTIVE*    
Merge all conditions into one seurat object :  
- load filtered seurat object of each condition (in a list)  
- extract a common list of peaks   
- create a seurat object with merged conditions  

*INPUTS*  
- functions_constants.R  
- [DRUG]_seurat_qc_filter.rds **(cf. scATACseq_R-SKMK_ReportCreate&Quality)**      
- [DRUG]_fragments.tsv.gz      
  
*OUTPUT*     
- merged_seurat_qc_filter.rds    
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_MergeObjectsCommonPeaks.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_MergeObjectsCommonPeaks.html"))

```


# annotationsFiles_createGrange.Rmd

*OBJECTIVE*  
Extract hg38 and hg19 annotations informations (annotatr package) in R objects.    
  
*OUTPUTS*   
- hg19_annotations_gr.rds  
- hg38_annotations_gr.rds  
  
```{r, annotation_grange}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/", "annotationsFiles_createGrange.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "annotationsFiles_createGrange.html"))

```

# scATACseq_R_AddAnnotations

*OBJECTIVE*    
Add genomic annotations in each element of the seurat object list :  
- add annotations from EnsDb database in assay/peaks/annotation  
- add peak name (chr-start-end) in the metadata  
- associate peaks with hg38 annotations from annotatr package in the metadata  
  
*INPUTS*    
- functions_constants.R    
- merged_seurat_qc_filter.rds **(cf.scATACseq_R_MergeObjectsCommonPeaks)**    
- hg38_annotations_gr.rds **(cf. annotationsFiles_createGrange)**    
    
*OUTPUT*       
- merged_seurat_qc_filter_annot.rds  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_AddAnnotations.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_AddAnnotations.html"))

```

# scATACseq_R_NormalizationReduction.Rmd

*OBJECTIVE*  
Normalize and reduced the dataset :  
- load filtered merged object based on a common list of peaks  
- normalize it with TFIDF and SVD  
- choose working dimensions  
- visualize clustering with different resolutions  
- umap plots with individual conditions   
- choose best resolutions to keep   
  
*INPUTS* 
- functions_constants.R  
- merged_seurat_qc_filter_annot.rds **(cf.scATACseq_R_NormalizationReduction)**  
  
*OUTPUTS*  
- plot_dimension_choice.svg  
- plot_resolution.svg  
- plot_umap_allconditions_merged.svg  
- plot_umap_individual.svg  
- plot_umap_clusters.svg  
- merged_seurat_qc_filter_annot_normreduc.rds  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_NormalizationReduction.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_NormalizationReduction.html"))

```

