---
title: "scRNAseq_CITEseq_DWNS_src"
author: "Laëtitia Racine"
date: "2023-01-11"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

This script is a "knit_all" document. It allows to run all the scripts from downstream analysis for scRNAseq-CITE-seq dataset and produce the html report in the *report* folder. There is no order between the scripts except for CorrelationADTRNA and STEMNET. Others can be run in any order. The pre-processing steps must be completed before to launch this document. See src1_scRNAseq_R_preprocessing.RmD. 
  
# scRNAseq_CITEseq_GenomePosition_GeneNameCountMatrix.Rmd

*OBJECTIVE*  
- create a correspondence tab with gene symbol, ENsembl_ID and gene position (molecular and cytogenetic informations)  

*INPUTS*    
- scRNAseq_CTRL_sample_feature_bc_matrix_features.tsv.gz (cf. data)
- Homo_sapiens.gene_info.gz (cf.data)

*OUTPUT*  
- scRNAseq_correspondence_tab_EnsemblID_geneName_genomePos.csv  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_GenomePosition_GeneNameCountMatrix.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_GenomePosition_GeneNameCountMatrix.html"))

```


# scRNAseq_CITEseq_R_DWNS_ClustersStatistics.Rmd

*OBJECTIVES*        
- visualize cell's distribution in clusters depending on their culture condition  
- create table with percentage of cells from each condition in each cluster    
- create table with percentage of cells from each cluster in each condition  
- create bar plots and pie plots  
  
*INPUTS*       
- merged_all-1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)  
- functions_constants.R    
    
*OUTPUTS*    
- tab_number_cells_clusters.csv  
- tab_number_norm_cells_clusters.csv  
- bp_clust.svg  
- bp_cond.svg  
- bp_number.svg  
- pie_cond.svg  
- pie_clust.svg  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_DWNS_ClustersStatistics.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_DWNS_ClustersStatistics.html"))

```


# scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA.Rmd

*OBJECTIVES*  
 - visualize the proteins distribution in the cell population to identify outliers  
 - visualize RNA and ADT expression on UMAP per condition to compare  
 - define groups of cells with high/low level of expression of those proteins  
 - define groups of cells with high/medium/low level of expression of those proteins
 - visualize groups of cells on UMAP  
 - visualize groups of cells in the clusters and in the conditions  
  
*INPUTS*       
- merged_all_1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)      
- functions_constants.R  
    
*OUTPUTS*  
- tab_raw_cond_two_groups.csv  
- tab_norm_cond_two_groups.csv  
- tab_percent_cond_two_groups.csv  
- tab_raw_cond_three_groups.csv  
- tab_norm_cond_three_groups.csv  
- tab_percent_cond_three_groups.csv  
- tab_raw_clusters_two_groups.csv  
- tab_norm_clusters_two_groups.csv  
- tab_percent_clusters_two_groups.csv  
- tab_raw_clusters_three_groups.csv  
- tab_norm_clusters_three_groups.csv  
- tab_percent_clusters_three_groups.csv  
- adt_distri_expr.svg  
- umap_adt_cd34.svg  
- umap_adt_cd133.svg  
- umap_rna_cd34.svg  
- umap_rna_cd133.svg  
- umap_adt_cd34_twogroups.svg  
- umap_adt_cd133_twogroups.svg  
- umap_adt_cd34_threegroups.svg  
- umap_adt_cd133_threegroups.svg  
- scatter_plot_adt.svg  
- umap_adt_clusters_twogroups.svg  
- umap_adt_clusters_threegroups.svg  
- vln_adt_clusters.svg  
- merged_all_CITEgroups.rds  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA.html"))

```


# scRNAseq_CITEseq_R_DWNS_CompareConditions_GO.Rmd

*OBJECTIVES*          
- transform gene expression matrix into binary matrix (on/off analysis)    
- extract list of genes lost across conditions    
- extract list of genes conserved across conditions    
- extract list of genes expressed in all conditions simultaneously     
- extract list of genes never expressed simultaneously  
- extract list of genes expressed only in CTRL  
- extract list of genes expressed only in AOA  
- extract list of genes expressed only in 2DG  
- perform GO on the lists and plot the results  
- extract visual outliers with qualitative analysis  
- extract list of genes with FindMarkers qualitative analysis  
- perform GO on list of genes from qualitative analysis and plot the results  

*INPUTS*      
- merged_all_prep_1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)   
- functions_constants.R  
- scRNAseq_correspondence_tab_EnsemblID_geneName_genomePos.csv (cf. scRNAseq_CITEseq_R_GenomePosition folder)  
- scRNAseq_CTRL_sample_feature_bc_matrix_features.tsv.gz (cf. data folder)  
        
*OUTPUTS*  
- list_binary_mat.rds  
- list_compCond_quali.rds  
- list_compCond_quali_genes.xlsx  
- list_compCond_quali_ggo.xlsx  
- list_compCond_quali_ego.xlsx  
- [INFORMATION_LIST_GENES]_barplot.svg  
- [INFORMATION_LIST_GENES]_dotplot.svg  
- list_CompCond_quanti.rds  
- list_CompCond_quanti_genes.xlsx  
- list_CompCond_quanti_ggo.xlsx  
- list_CompCond_quanti_ego.xlsx  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_DWNS_CompareConditions_GO.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_DWNS_CompareConditions_GO.html"))

```


# scRNAseq_CITEseq_R_DWNS_PathwayAnalysis.Rmd

*OBJECTIVES*            
- extract pathway database and gene associated      
- define list of interesting pathways to analyze    
- pathway analysis with ReactomeGSA per cluster and per condition    
- pathway analysis with SCPA per condition  
- use AddModuleScore on interesting pathways    
- visualize expression of selected genes on UMAP and dotplot  
  
*INPUTS*          
- merged_all_prep_1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)    
- functions_constants.R    
    
*OUTPUTS*      
- genes_msigdbr.rds  
- genes_reactome.rds  
- pathways_database_reactome.csv    
- pathways_genes_reactome.csv    
- pathways_database_reactome_format_scpa.rds    
- pathways_interest_reactome.csv    
- pathways_genes_interest_reactome.csv
- pathways_db_ms.csv  
- RGSA_gsva_result_clusters.rds    
- RGSA_gsva_result_condition.rds    
- RGSA_pathway_clusters.csv  
- RGSA_pathway_conditions.csv    
- RGSA_foldchange_conditions.csv    
- RGSA_foldchange_clusters.csv    
- RGSA_outliers_clusters.svg    
- RGSA_outliers_conditions.svg    
- RGSA_relevant_path_clust.svg    
- RGSA_relevant_path_cond.svg    
- RGSA_heatmap_clust.pdf    
- RGSA_heatmap_cond.pdf    
- RGSA_interest_path_clust.svg    
- RGSA_interest_path_cond.svg    
- RGSA_heatmap_interest_clust.pdf    
- RGSA_heatmap_interest_cond.pdf    
- SCPA_list_scpa_conditions.rds    
- SCPA_heatmap_interest_path.pdf    
- SCPA_enrich_cond.svg    
- SCPA_interest_enrich_cond.svg    
- SCPA_rank_cond.svg    
- AMS_modulescore_clust.svg    
- AMS_list_modulescore_cond.rds    
- AMS_modulescore_clust.rds  
- AMS_dotplot_list_genes_clust.svg    
- AMS_dotplot_list_genes_cond.svg    
- AMS_dotplot_list_genes_clust_cond_1.svg    
- AMS_dotplot_list_genes_clust_cond_2.svg    
- AMS_dotplot_list_genes_clust_cond_3.svg  
- GL_dotplot_list_genes_clust.svg  
- GL_dotplot_list_genes_clust_cond1.svg  
- GL_dotplot_list_genes_clust_cond2.svg
- GL_dotplot_list_genes_clust_cond3.svg
- GL_dotplot_list_genes_cond.svg  
- GL_violinplot_list_genes_clust_cond_1.svg
- GL_violinplot_list_genes_clust_cond_2.svg
- GL_violinplot_list_genes_clust_cond_3.svg
- GL_violinplot_list_genes_clust_cond_4.svg
- GL_violinplot_list_genes_clust_cond_5.svg

**Note : the first part of the code is to big to be supported in knit. Run it by hand in external cluster and then launch script n2.**

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory,
                                 "/bin/scRNAseq/",
                                 "scRNAseq_CITEseq_R_DWNS_PathwaysAnalysis_n2.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(),
                                       '_',
                                       "scRNAseq_CITEseq_R_DWNS_PathwaysAnalysis_n2.html"))

```


# scRNAseq_CITEseq_R_DWNS_ClustersInvestigation_GO.Rmd

*OBJECTIVES*          
- FindAllMarkers on clusters  
- select relevant markers and apply GO analysis  
- visualize GO results on heatmap, dotplot, violinplot and feature plot  
- FindMarkers for specific clusters comparisons  

*INPUTS*      
- merged_all_prep_1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)      
- functions_constants.R    
- scRNAseq_CTRL_sample_feature_bc_matrix_features.tsv.gz (cf.data folder)    
- scRNAseq_correspondence_tab_EnsemblID_geneName_genomePos.csv (cf. scRNAseq_CITEseq_R_GenomePosition folder)    
      
*OUTPUTS*        
- clust_markers.rds      
- clust_markers.csv  
- top20_pos_markers.csv  
- top20_pos_filtered_markers.csv  
- list_GO_clust.rds  
- [INFORMATION_CLUSTER_MARKERS]_GO_barplot.svg  
- [INFORMATION_CLUSTER_MARKERS]_GO_dotplot.svg 
- pos_markers_heatmap.svg  
- pos_markers_dotplot.svg  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_DWNS_ClustersInvestigation_GO.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_DWNS_ClustersInvestigation_GO.html"))

```

