---
title: "src2_scATACseq_R_preprocessing"
author: "Laëtitia"
date: "2022-09-02"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
library(stringr)

```




# scATACseq_R-SKMK_ReportCreate&Quality = Summary of snakemake scripts

*OBJECTIVE*    
Create a html report with command lines of snakemake rules.       
The two R scripts called in snakemake allow to :     
--> create a usable R object from Cell Ranger outputs    
- load cell ranger output files    
- create seurat object with chromatin assay for each condition    
- add genomic annotations from EnsDB.Hsapiens.v86 (hg38) database    
- update orig.ident with the name of the condition    
- add peak name in grange metadata    
--> check quality of cells and fragments detected and filter the dataset       
- filter on nucleosome banding pattern        
- filter on fragments and blacklist reads        
- filter on TSS enrichment score       
- keep only standard chromosomes      
- create QC plots  
The Rmd script will display the code without running it and expose some examples.      
  
*script scATACseq_SKMK_CreateSeuratObject.R*    
*INPUTS*    
- [DRUG]_filtered_peak_bc_matrix.h5        
- [DRUG]_singlecell.csv        
- [DRUG]_fragments.tsv.gz        
*OUTPUT*       
- [DRUG]_seurat.rds : Seurat object with two filters and database annotations info    
    
*script scATACseq_SKMK_QualityControl.R*  
*INPUT*    
- [DRUG]_seurat.rds (cf. scATACseq_SKMK_CreateSeuratObject.R)    
*OUTPUTS*     
- [DRUG]_seurat_qc.rds : Seurat object with QC parameters in metadata but no new filter applied       
- [DRUG]_plots_qc_list.rds        
- [DRUG]_seurat_qc_filter.rds : Seurat object with QC parameters and filters applied        
- [DRUG]_plots_quality_control.svg      
- [DRUG]_df_filter.csv  
  
*script scATACseq_R-SKMK_ReportCreate&Quality.Rmd*  
*INPUTS*  
All the outputs of snakemake files.    
*OUTPUTS*      
An html and .csv files with list of peaks for each condition after QC filters.    
    
```{r, snakemake report}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R-SKMK_ReportCreateQuality.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R-SKMK_ReportCreateQuality.html"))

```





# scATACseq_R_MergeObjectsCommonPeaks.Rmd

*OBJECTIVE*    
Merge all conditions into one seurat object :  
- load filtered seurat object of each condition (in a list)  
- extract a common list of peaks   
- create a seurat object with merged conditions  

*INPUTS*  
- functions_constants.R  
- [DRUG]_seurat_qc_filter.rds **(cf. scATACseq_R-SKMK_ReportCreate&Quality)**      
- [DRUG]_fragments.tsv.gz      
  
*OUTPUT*     
- merged_seurat_qc_filter.rds    
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_MergeObjectsCommonPeaks.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_MergeObjectsCommonPeaks.html"))

```


# annotationsFiles_createGrange.Rmd

*OBJECTIVE*  
Extract hg38 and hg19 annotations information (annotatr package) in R objects.    
  
*OUTPUTS*   
- hg19_annotations_gr.rds  
- hg38_annotations_gr.rds  
  
```{r, annotation_grange}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/", "annotationsFiles_createGrange.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "annotationsFiles_createGrange.html"))

```

# scATACseq_R_AddAnnotations

*OBJECTIVE*    
Add genomic annotations in each element of the seurat object list :  
- add annotations from EnsDb database in assay/peaks/annotation  
- add peak name (chr-start-end) in the metadata  
- associate peaks with hg38 annotations from annotatr package in the metadata  
  
*INPUTS*    
- functions_constants.R    
- merged_seurat_qc_filter.rds **(cf.scATACseq_R_MergeObjectsCommonPeaks)**    
- hg38_annotations_gr.rds **(cf. annotationsFiles_createGrange)**    
    
*OUTPUT*       
- merged_seurat_qc_filter_annot.rds  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_AddAnnotations.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_AddAnnotations.html"))

```

# scATACseq_R_NormalizationReduction.Rmd

*OBJECTIVE*  
Normalize and reduced the dataset :  
- load filtered merged object based on a common list of peaks  
- normalize it with TFIDF and SVD  
- choose working dimensions  
- visualize clustering with different resolutions  
- umap plots with individual conditions   
- choose best resolutions to keep   
  
*INPUTS* 
- functions_constants.R  
- merged_seurat_qc_filter_annot.rds **(cf.scATACseq_R_NormalizationReduction)**  
  
*OUTPUTS*  
- plot_dimension_choice.svg  
- plot_resolution.pdf  
- plot_umap_allconditions_merged.svg
- plot_umap_allconditions_split.svg 
- plot_umap_individual.svg  
- plot_umap_clusters.svg  
- merged_seurat_qc_filter_annot_normreduc.rds  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_NormalizationReduction.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_NormalizationReduction.html"))

```

