---
title: "src3_scATACseq_R_dwns"
author: "Laëtitia"
date: "2022-09-02"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```



# scATAcseq_R_DWNS_CellsPeaksStatistics

*OBJECTIVE*  
Qualitative analysis of the peaks detection :  
- for each cell of each condition, the number of peaks/region in which the cell have at least one fragment inside
- for each condition, the number of peaks/region with cells containing fragments inside it  Quantitative analysis of the peaks detection :  
- min/max/med/mean of the number of fragments in the peaks detected for each cell  
- min/max/med/mean of the number of fragments in the peaks for each cell that detected this peak  
  
*INPUTS*    
- functions_constants.R  
- merged_seurat_qc_filter_annot_normreduc.rds (cf. scATACseq_R_NormalizationReduction)  

*OUTPUTS*  
- list_tab_nbcells_peak.rds  
- list_tab_nbpeaks_percell.rds  
- list_binary_matrix.rds  
- global_tab_cell.csv  
- statistic_tab_cell.csv  
- nbpeaks_per_cell_quali_plot.svg  
- global_tab_peak.csv  
- statistic_tab_peak.csv  
- nbcell_sharing_peaks_quali_plot.svg  
- mat_df.rds  
- mat_df_cut.rds  
- stat_nbfragments_perpeak_percell.csv  
- nbpeaks_per_cell_quanti_plot.svg  
- stat_nbfragments_percell_perpeaks.csv  
- nbcell_sharing_peaks_quanti_plot.svg  
  
!!! Note : part binary count of the first script makes Rstudio crashes each time with knitr. Do it by hand before to run the entire knitr. !!!    
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATAcseq_R_DWNS_CellsPeaksStatistics_part1.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATAcseq_R_DWNS_CellsPeaksStatistics_part1.html"))

.rs.restartR()

```

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATAcseq_R_DWNS_CellsPeaksStatistics_part2.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATAcseq_R_DWNS_CellsPeaksStatistics_part2.html"))

.rs.restartR()

```


# scATACseq_R_DWNS_ClustersInvestigation_Stat

*OBJECTIVE*    
Count the number of cells of each condition per cluster.  
Normalize it by the number of initial cells per condition.  
Visualize the cluster compositions of cells with tables and plots (barplot, pieplot).  
Associate peaks with their genome annotation and count the number of peaks per annotation per cluster.  

*INPUTS*    
- functions_constants.R  
- merged_seurat_qc_filter_annot_normreduc.rds (cf. scATACseq_R_NormalizationReduction)  
  
*OUTPUTS*     
- nbcell_perclust_percond.csv  
- nbcell_perclust_percond_norm.csv  
- percent_cond_inclust.csv  
- percent_clust_incond.csv  
- barplot_clust_cellcomp.svg  
- pieplot1_clust_cellcomp.svg  
- pieplot2_clust_cellcomp.svg  
- tab_dim2-50_res.0.25_bulkpeaks.csv  
- tab_dim2-50_res.0.25_distripeaks.csv  
- tab_dim2-50_res.0.25_distripeaks_stat.csv  
- stat_peakscomp.svg  
- distri_peakscomp.svg  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_DWNS_ClustersInvestigation_Stat.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_DWNS_ClustersInvestigation_Stat.html"))

.rs.restartR()

```


# scATACseq_R_DWNS_ClustersInvestigation_DA

*OBJECTIVE*    
Extract list of differentially accessible peaks between cell's cluster.  
Associate the genomic annotation based on hg38 database from annotatr.   
Extract the most relevant DA peaks based on pvalue and difference in pct populations.  
Draw heatmap on the most relevant DA peaks. 
Proceed GO on list of DA peaks for each cluster.  
Proceed motif enrichment analysis on list of DA peaks for each cluster.  

*INPUTS*    
- functions_constants.R  
- merged_seurat_qc_filter_annot_normreduc.rds (cf. scATACseq_R_NormalizationReduction)  
=> Note : first part of the code to run on IFB because it requires  a lot of memory.  
- clust_markers_[X]_20230417.rds  

*OUTPUTS*     
- clust_markers.rds  
- tab_clust_DApeaks_annotation.csv  
- top20_pos.csv  
- top20_pos_filtered.csv  
- heatmap_DApeaks_clust_all.png  
- heatmap_DApeaks_clust_genes.png  
- list_GO_clust.rds  
- GO_DApeaks_clust.svg  
- tab_GO_DApeaks_clust.csv  
- seurat_motif.rds  
- enriched_motifs_clust_plot.png  
- enriched_motifs_clust_tab.csv  
- unique_enriched_motif_clust.csv  
- common_enriched_motif_clust.csv  
- unique_enriched_motif_clust_plot1.png  
- unique_enriched_motif_clust_plot2.png  
- common_enriched_motif_clust_plot1.png
- common_enriched_motif_clust_plot2.png
- common_enriched_motif_clust_plot3.png
- common_enriched_motif_clust_plot4.png
- common_enriched_motif_clust_plot5.png
- common_enriched_motif_clust_plot6.png
- common_enriched_motif_clust_plot7.png
- common_enriched_motif_clust_plot8.png
- common_enriched_motif_clust_plot9.png

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_DWNS_ClustersInvestigation_DA.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_DWNS_ClustersInvestigation_DA.html"))

.rs.restartR()

```


# scATACseq_R_DWNS_CompareConditions_DA

*OBJECTIVE*    
Extract list of differentially accessible peaks between conditions.    
Associate the genomic annotation based on hg38 database from annotatr.   
Extract the most relevant DA peaks based on pvalue and difference in pct populations.  
Draw heatmap on the most relevant DA peaks. 
Proceed GO on list of DA peaks for each condition  
Proceed motif enrichment analysis on list of DA peaks for each condition    
  
*INPUTS*    
- functions_constants.R  
- merged_seurat_qc_filter_annot_normreduc.rds (cf. scATACseq_R_NormalizationReduction)  
=> Note : first part of the code to run on IFB because it requires  a lot of memory.  
- cond_markers_FAM_20230416.rds  
- seurat_motif.rds (cf. scATACseq_R_DWNS_ClustersInvestigation) => facultative, to save time  
  
*OUTPUTS*     
- cond_markers.rds  
- tab_cond_DApeaks_annotation.csv  
- top20_pos.csv  
- top20_pos_filtered.csv  
- heatmap_DApeaks_cond_all.png  
- heatmap_DApeaks_cond_genes.png  
- list_GO_cond.rds  
- GO_DApeaks_cond.svg  
- tab_GO_DApeaks_GO.csv  
- seurat_motif.rds  
- enriched_motifs_cond_plot.png  
- enriched_motifs_cond_tab.csv  
- unique_enriched_motif_cond.csv  
- common_enriched_motif_cond.csv  
- unique_enriched_motif_cond_plot1.png  
- unique_enriched_motif_cond_plot2.png  
- unique_enriched_motif_cond_plot3.png  
- common_enriched_motif_cond_plot1.png
- common_enriched_motif_cond_plot2.png
- common_enriched_motif_cond_plot3.png
- common_enriched_motif_cond_plot4.png
- common_enriched_motif_cond_plot5.png

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_DWNS_CompareConditions_DA.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_DWNS_CompareConditions_DA.html"))

.rs.restartR()

```


# scATACseq_R_DWNS_GeneActivity_GeneCoverage

*OBJECTIVE*    
Visualize the putative activity of markers genes based on chromatine accessibility (gene activity).  
Draw coverage plot for list of specific genes.  
Visualize reads peaks distribution on chromosomes.  

*INPUTS*  
- functions_constants.R  
- merged_seurat_qc_filter_annot_normreduc.rds (cf. scATACseq_R_NormalizationReduction)  
- hg38_annotations_gr.rds (cf. AnnotationsFiles_createGrange)  
- mat.rds (cf. scATACseq_R_NormalizationReduction)  

*OUTPUTS*  
- gene.activities.rds  
- plot_geneactivity_TF.png  
- plot_geneactivity_varfeature.png  
- plot_geneactivity_glutglyc.png  
- plot_geneactivity_mettca.png  
- [GENE]_coverage_allcond.svg  
- [GENE]_coverage_tile_allcond.svg  
- list_plot_chr.rds  
- list_graph.rds  
- list_graph_stat.rds  
- chromosome_distribution_nbfrag.svg  
- chromosome_distribution_stat.svg  
=> Note : part to run on IFB because require lots of memory. 
- coverage_kpbam_chr[X].png  
    
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, "/bin/scATACseq/", "scATACseq_R_DWNS_GeneActivity_GeneCoverage.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), '_', "scATACseq_R_DWNS_GeneActivity_GeneCoverage.html"))

```


