---
title: "scRNAseq_CITEseq_src_all"
author: "Laëtitia Racine"
date: "2023-03-20"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

This script is a "knit_all" document. It's a combination of src1_preprocessing and src2_downs 
It allows to run all the script from pre-processing steps AND downstream analysis for scRNAseq-CITE-seq dataset in the right order and to produce html files in the *report* folder.    

*****************************
src1_scRNAseq_R_preprocessing
*****************************
  

# scRNAseq_CITEseq_R_CreateSeuratObject.Rmd

*OBJECTIVES*        
- load cell ranger output files  
- create seurat object with RNA and ADT assays for each condition  
- update orig.ident with the name of the condition  
- visualize nb of cells and features for each condition  
- visualize CTRL object  
  
*INPUTS*   
(samples from batch1 are sample_feature and those from batch2 are filtered_feature)       
- scRNAseq_[DRUG]_sample_feature_bc_matrix_barcodes.tsv.gz    
- scRNAseq_[DRUG]_sample_feature_bc_matrix_features.tsv.gz    
- scRNAseq_[DRUG]_sample_feature_bc_matrix_matrix.mtx.gz    
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_barcodes.tsv.gz    
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_features.tsv.gz      
- scRNAseq_[DRUG]_filtered_feature_bc_matrix_matrix.mtx.gz    
- functions_constants.R    
    
*OUTPUTS*    
- list_seurat_obj.rds    
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_CreateSeuratObject.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_CreateSeuratObject.html"))

```


# scRNAseq_CITEseq_R_FilterData.Rmd

*OBJECTIVES*    
Apply cumulative quality control filters on the dataset :   
- calculate mitochondrial and ribosomal RNA percentage     
- filter cells based on number of genes detected (keep those between 1000 and 6500)    
- filter cells based on mitochondrial and ribosomal RNA percentage (keep those <25% and >5%)  
- remove mitochondrial genes  
- filter genes detected in less than 3 cells  
- create a summary tab on filters (number cells and genes after each filter)
- create QC plots
- create tabs describing the filtered dataset  
  
*INPUTS*          
- list_seurat_obj.rds (cf. scRNAseq_CITEseq_R_CreateSeuratObject.Rmd)      
- functions_constants.R    
  
*OUTPUTS*        
- plot_feature_filter.png    
- plot_percentMTRB_filter.png    
- filter_tab_seurat_obj.csv  
- list_seurat_obj_qc_each_filter.rds  
- list_mat_raw_proteins.rds  
- list_mat_raw_genes.rds  
- summary_cell_tab.csv    
- summary_cond_tab.cs    
- individual_gene_tab.csv  
- list_seurat_obj_qc.rds  
- list_genes_detected_1cell.rds  
- list_genes_detected_inf_equal_3cells.rds  
- tab_UMI.csv    
- list_mat_detected_genes.rds  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_FilterData.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_FilterData.html"))

```


# scRNAseq_CITEseq_R_NormalizeData.Rmd

*OBJECTIVES*    
- normalize RNA assay with lognormalize + find 2000 most variable features + scale data     
- calculate cell cycle score and create cell cycle influence plots      
- normalize RNA assay with SCT without cell cycle regression      
- create a new object with SCT and cell cycle regression if needed for later      
- normalize ADT assay with CLR method  
- compare LogNorm and SCT normalizations    
      
*INPUTS*        
- list_seurat_obj_qc.rds (cf. scRNAseq_CITEseq_R_FilterData.Rmd)      
- functions_constants.R        
        
*OUTPUTS*        
- plot_2000variablesfeatures.svg    
- plot_cellcyclescoring.svg    
- plot_cellcycle_statUMI.svg    
- tab_cellcycle.csv    
- tab_genes_cellcycle.csv    
- list_seurat_obj_qc_norm_ccr.rds    
- list_seurat_obj_qc_norm.rds    
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_NormalizeData.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_NormalizeData.html"))

```


# scRNAseq_CITEseq_R_BatchEffectCorrection.Rmd

*OBJECTIVES*  
- merge seurat objects per batch (two objects to compare)   
- visualize the data per batch (PCA, UMAP) based on sct normalization  
- visualize the data per batch (PCA, UMAP) based on log normalization  
- merge seurat objects of all batches into one object  
- visualize the data (PCA, UMAP) based on sct normalization  
- visualize the data (PCA, UMAP) based on log normalization  
- determine if batch correction is needed    
    
*INPUTS*    
- list_seurat_obj_qc_norm.rds (cf. scRNAseq_CITEseq_R_NormalizeData.Rmd)   
- functions_constants.R    
    
*OUTPUTS*  
- umap_sct_batch1.png  
- umap_sct_batch2.png  
- umap_log_batch1.png  
- umap_log_batch2.png  
- umap_sct_bothbatch.png  
- umap_sct_bothbatch_split.png  
- umap_sct_bothbatch_ctrl.png  
- umap_log_bothbatch.png  
- umap_log_bothbatch_split.png  
- umap_log_bothbatch_ctrl.png  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_BatchEffectCorrection.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_BatchEffectCorrection.html"))

```


# scRNAseq_CITEseq_R_ReductionDimension.Rmd

*OBJECTIVES*  
- perform PCA and determine the optimal dimensionality with elbow plot  
- perform UMAP and visualize cells distribution per condition      
- perform clusterisation and determine the best resolution  
- prepare seurat object for markers identification  
  
*INPUTS*
- list_seurat_obj_qc_norm.rds (cf. scRNAseq_CITEseq_R_NormalizeData.Rmd)   
- list_seurat_obj_qc_norm_ccr.rds (cf. scRNAseq_CITEseq_R_NormalizeData.Rmd)     
- functions_constants.R      
      
*OUTPUTS*  
- dimension_choice_elb_pca.svg  
- dimension_choice_pca_genes.svg    
- umap_1-40dim.svg  
- umap_1-40dim_split.svg  
- umap_1-40dim_split_highlight.svg  
- umap_clusters_resolution.svg  
- umap_clusters_res_choice.svg  
- merged_all_1-40dim.rds  
- merged_all_prep_1-40dim.rds      
- dimension_choice_elb_pca_cc.svg  
- dimension_choice_pca_genes_cc.svg  
- umap_1-40dim_cc.svg  
- umap_1-40dim_split_cc.svg  
- umap_1-40dim_split_highlight_cc.svg  
- umap_clusters_resolution_cc.svg  
- umap_clusters_res_choice_cc.svg  
- merged_cc_1-40dim.rds    
- merged_cc_prep_1-40dim.rds    
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_ReductionDimension.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_ReductionDimension.html"))

```



********************
src2_scRNAseq_R_dwns
********************


# scRNAseq_CITEseq_GenomePosition_GeneNameCountMatrix.Rmd

*OBJECTIVE*  
- create a correspondence tab with gene symbol, ENsembl_ID and gene position (molecular and cytogenetic informations)  

*INPUTS*    
- scRNAseq_CTRL_sample_feature_bc_matrix_features.tsv.gz (cf. data)
- Homo_sapiens.gene_info.gz (cf.data)

*OUTPUT*  
- scRNAseq_correspondence_tab_EnsemblID_geneName_genomePos.csv  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_GenomePosition_GeneNameCountMatrix.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_GenomePosition_GeneNameCountMatrix.html"))

```


# scRNAseq_CITEseq_R_DWNS_ClustersStatistics.Rmd

*OBJECTIVES*        
- visualize cell's distribution in clusters depending on their culture condition  
- create table with percentage of cells from each condition in each cluster    
- create table with percentage of cells from each cluster in each condition  
- create bar plots and pie plots  
  
*INPUTS*       
- merged_all-1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)  
- functions_constants.R    
    
*OUTPUTS*    
- tab_number_cells_clusters.csv  
- tab_number_norm_cells_clusters.csv  
- bp_clust.svg  
- bp_cond.svg  
- bp_number.svg  
- pie_cond.svg  
- pie_clust.svg  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_DWNS_ClustersStatistics.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_DWNS_ClustersStatistics.html"))

```


# scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA.Rmd

*OBJECTIVES*  
 - visualize the proteins distribution in the cell population to identify outliers  
 - visualize RNA and ADT expression on UMAP per condition to compare  
 - define groups of cells with high/low level of expression of those proteins  
 - define groups of cells with high/medium/low level of expression of those proteins
 - visualize groups of cells on UMAP  
 - visualize groups of cells in the clusters and in the conditions  
  
*INPUTS*       
- merged_all_1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)      
- functions_constants.R  
    
*OUTPUTS*  
- tab_raw_cond_two_groups.csv  
- tab_norm_cond_two_groups.csv  
- tab_percent_cond_two_groups.csv  
- tab_raw_cond_three_groups.csv  
- tab_norm_cond_three_groups.csv  
- tab_percent_cond_three_groups.csv  
- tab_raw_clusters_two_groups.csv  
- tab_norm_clusters_two_groups.csv  
- tab_percent_clusters_two_groups.csv  
- tab_raw_clusters_three_groups.csv  
- tab_norm_clusters_three_groups.csv  
- tab_percent_clusters_three_groups.csv  
- adt_distri_expr.svg  
- umap_adt_cd34.svg  
- umap_adt_cd133.svg  
- umap_rna_cd34.svg  
- umap_rna_cd133.svg  
- umap_adt_cd34_twogroups.svg  
- umap_adt_cd133_twogroups.svg  
- umap_adt_cd34_threegroups.svg  
- umap_adt_cd133_threegroups.svg  
- scatter_plot_adt.svg  
- umap_adt_clusters_twogroups.svg  
- umap_adt_clusters_threegroups.svg  
- vln_adt_clusters.svg  
- merged_all_CITEgroups.rds  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA.html"))

```


# scRNAseq_CITEseq_R_DWNS_CompareConditions_GO.Rmd

*OBJECTIVES*          
- transform gene expression matrix into binary matrix (on/off analysis)    
- extract list of genes lost across conditions    
- extract list of genes conserved across conditions    
- extract list of genes expressed in all conditions simultaneously     
- extract list of genes never expressed simultaneously  
- extract list of genes expressed only in CTRL  
- extract list of genes expressed only in AOA  
- extract list of genes expressed only in 2DG  
- perform GO on the lists and plot the results  
- extract visual outliers with qualitative analysis  
- extract list of genes with FindMarkers qualitative analysis  
- perform GO on list of genes from qualitative analysis and plot the results  

*INPUTS*      
- merged_all_prep_1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)   
- functions_constants.R  
- scRNAseq_correspondence_tab_EnsemblID_geneName_genomePos.csv (cf. scRNAseq_CITEseq_R_GenomePosition folder)  
- scRNAseq_CTRL_sample_feature_bc_matrix_features.tsv.gz (cf. data folder)  
        
*OUTPUTS*  
- list_binary_mat.rds  
- list_compCond_quali.rds  
- list_compCond_quali_genes.xlsx  
- list_compCond_quali_ggo.xlsx  
- list_compCond_quali_ego.xlsx  
- [INFORMATION_LIST_GENES]_barplot.svg  
- [INFORMATION_LIST_GENES]_dotplot.svg  
- list_CompCond_quanti.rds  
- list_CompCond_quanti_genes.xlsx  
- list_CompCond_quanti_ggo.xlsx  
- list_CompCond_quanti_ego.xlsx  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_DWNS_CompareConditions_GO.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_DWNS_CompareConditions_GO.html"))

```


# scRNAseq_CITEseq_R_DWNS_PathwayAnalysis.Rmd

*OBJECTIVES*            
- extract pathway database and gene associated      
- define list of interesting pathways to analyze    
- pathway analysis with ReactomeGSA per cluster and per condition    
- pathway analysis with SCPA per condition  
- use AddModuleScore on interesting pathways    
- visualize expression of selected genes on UMAP and dotplot  
  
*INPUTS*          
- merged_all_prep_1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)    
- functions_constants.R    
    
*OUTPUTS*      
- genes_msigdbr.rds  
- genes_reactome.rds  
- pathways_database_reactome.csv    
- pathways_genes_reactome.csv    
- pathways_database_reactome_format_scpa.rds    
- pathways_interest_reactome.csv    
- pathways_genes_interest_reactome.csv
- pathways_db_ms.csv  
- RGSA_gsva_result_clusters.rds    
- RGSA_gsva_result_condition.rds    
- RGSA_pathway_clusters.csv  
- RGSA_pathway_conditions.csv    
- RGSA_foldchange_conditions.csv    
- RGSA_foldchange_clusters.csv    
- RGSA_outliers_clusters.svg    
- RGSA_outliers_conditions.svg    
- RGSA_relevant_path_clust.svg    
- RGSA_relevant_path_cond.svg    
- RGSA_heatmap_clust.pdf    
- RGSA_heatmap_cond.pdf    
- RGSA_interest_path_clust.svg    
- RGSA_interest_path_cond.svg    
- RGSA_heatmap_interest_clust.pdf    
- RGSA_heatmap_interest_cond.pdf    
- SCPA_list_scpa_conditions.rds    
- SCPA_heatmap_interest_path.pdf    
- SCPA_enrich_cond.svg    
- SCPA_interest_enrich_cond.svg    
- SCPA_rank_cond.svg    
- AMS_modulescore_clust.svg    
- AMS_list_modulescore_cond.rds    
- AMS_modulescore_clust.rds  
- AMS_dotplot_list_genes_clust.svg    
- AMS_dotplot_list_genes_cond.svg    
- AMS_dotplot_list_genes_clust_cond_1.svg    
- AMS_dotplot_list_genes_clust_cond_2.svg    
- AMS_dotplot_list_genes_clust_cond_3.svg  
- GL_dotplot_list_genes_clust.svg  
- GL_dotplot_list_genes_clust_cond1.svg  
- GL_dotplot_list_genes_clust_cond2.svg
- GL_dotplot_list_genes_clust_cond3.svg
- GL_dotplot_list_genes_cond.svg  
- GL_violinplot_list_genes_clust_cond_1.svg
- GL_violinplot_list_genes_clust_cond_2.svg
- GL_violinplot_list_genes_clust_cond_3.svg
- GL_violinplot_list_genes_clust_cond_4.svg
- GL_violinplot_list_genes_clust_cond_5.svg

**Note : the first part of the code is to big to be supported in knit. Run it by hand in external cluster and then launch script n2.**

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory,
                                 "/bin/scRNAseq/",
                                 "scRNAseq_CITEseq_R_DWNS_PathwaysAnalysis_n2.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(),
                                       '_',
                                       "scRNAseq_CITEseq_R_DWNS_PathwaysAnalysis_n2.html"))

```


# scRNAseq_CITEseq_R_DWNS_ClustersInvestigation_GO.Rmd

*OBJECTIVES*          
- FindAllMarkers on clusters  
- select relevant markers and apply GO analysis  
- visualize GO results on heatmap, dotplot, violinplot and feature plot  
- FindMarkers for specific clusters comparisons  

*INPUTS*      
- merged_all_prep_1-40dim.rds (cf. scRNAseq_CITEseq_R_ReductionDimension.Rmd)      
- functions_constants.R    
- scRNAseq_CTRL_sample_feature_bc_matrix_features.tsv.gz (cf.data folder)    
- scRNAseq_correspondence_tab_EnsemblID_geneName_genomePos.csv (cf. scRNAseq_CITEseq_R_GenomePosition folder)    
      
*OUTPUTS*        
- clust_markers.rds      
- clust_markers.csv  
- top20_pos_markers.csv  
- top20_pos_filtered_markers.csv  
- list_GO_clust.rds  
- [INFORMATION_CLUSTER_MARKERS]_GO_barplot.svg  
- [INFORMATION_CLUSTER_MARKERS]_GO_dotplot.svg 
- pos_markers_heatmap.svg  
- pos_markers_dotplot.svg  

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/scRNAseq/", 
                                 "scRNAseq_CITEseq_R_DWNS_ClustersInvestigation_GO.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "scRNAseq_CITEseq_R_DWNS_ClustersInvestigation_GO.html"))

```


