---
title: "bulkATACseq_DWNS_bloc3_src"
author: "Laëtitia Racine"
date: "2022-09-02"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

This script is a "knit_all" document. It allows to run all the script from downstream analysis steps for bulkATACseq dataset and to produce html files in the *report* folder.    
  

# bulkATACseq_R_ChangeAnnotation.Rmd

*OBJECTIVES*        
Change the annotations metadata of the Granges object output of bloc2.
Use the annotation file obtain in annotationsFiles_createGrange.Rmd with annotatr package.
Add peak_name in metadata of granges.  
Change name aK in CTRLaK for grange.  
   
*INPUTS*  
- [DRUG]_[TIME]_[DONORS]_threshold_10_ann.gr.rds (cf. bloc2 of bulkATACseq analysis)  
- functions_constants.R 
- hg19_annotations_gr.rds (cf. annotationsFiles_createGrange.Rmd)  

*OUTPUTS*    
-[DRUG]_[TIME]_[DONORS]_threshold_10_update_annot.gr.rds
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/bulkATACseq/", 
                                 "bulkATACseq_R_ChangeAnnotation.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "bulkATACseq_R_ChangeAnnotation.html"))

```


# bulkATACseq_R_NumberPeaksReport.Rmd

*OBJECTIVES*    
 - Count the total number of peaks  
 - Count the number of peaks per annotation  
 - Create a summary table  
 - Create bar plots, line plots and upset plots
  
*INPUTS*          
- [DRUG]_[TIME]_[DONORS]_threshold_10_update_annot.gr.rds (cf. bulkATACseq_R_ChangeAnnotation.Rmd)
- functions_constants.R
    
*OUTPUTS*    
- tab_nbpeaks_percond_perannot.csv
- plot_nbpeaks_pertime.svg
- plot_nbpeaks_percond.svg
- plot_nbpeaks_pertime_filter.svg
- plot_nbpeaks_percond_filter.svg
- plot_nbpeaks_annotation_pertime.svg
- plot_nbpeaks_annotation_percond.svg
- plot_nbpeaks_annotation_pertime_filter.svg
- plot_nbpeaks_annotation_percond_filter.svg
- plot_nbpeaks_annotation_chrono.svg
- plot_upset_[DRUG]_[TIME].svg  
- plot_upset_zoom_[DRUG]_[TIME].svg  
- plot_upset_setsize_[DRUG]_[TIME].png   
- plot_upset_zoom_setsize_[DRUG]_[TIME].png  
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/bulkATACseq/", 
                                 "bulkATACseq_R_NumberPeaksReport.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "bulkATACseq_R_NumberPeaksReport.html"))

```


# bulkATACseq_GenomeDistribution_GeneCoverage.Rmd

*OBJECTIVES*    
 - Create chromosome accessibility plots  
 - Create gene coverage plots  
      
*INPUTS*        
- functions_constants.R       
- [DRUG]_[TIME]_[DONORS]_threshold_10_update_annot.gr.rds (cf. bulkATACseq_R_ChangeAnnotation.Rmd)
- [DRUG]_[TIME]_[DONORS]_downsampled.bam (cf. data downsampled_bam_merged from bloc2)
- hg19_annotations_gr.rds (cf. annotationsFiles_createGrange.Rmd)  

*OUTPUTS*        
- plot_allchr_[DRUG]_[TIME].png  
- plot_allsamples_chr[X].png
- ranges_genes_interest.csv  
- [DRUG]_[GENE]_coverage_plot_alltime.pdf  
- list_chr_track.rds
- list_track_samples.rds 
- [TIME]_[GENE]_coverage_plot_allcond.pdf  
- list_iso_prom_track.rds
  
```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/bulkATACseq/", 
                                 "bulkATACseq_R_GenomeDistribution_GeneCoverage.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "bulkATACseq_R_GenomeDistribution_GeneCoverage.html"))

```


# bulkATACseq_R_Multiparametric_DifferentialAccessibility.Rmd

*OBJECTIVES*  
- Create a common set of regions to perform differential accessibility analysis  
- Use featureCounts to create a matrix with the number of reads per region for each condition  
- Perform differential accessibility analysis for all conditions (pca, heatmap, pathway, GO)
- Perform differential accessibility analysis two conditions by two conditions (volcano plot, pathway, GO)
- Plot peaks dynamics over time  
    
*INPUTS*    
- functions_constants.R  
- [DRUG]_[TIME]_[DONORS]_threshold_10_update_annot.gr.rds (cf. bulkATACseq_R_ChangeAnnotation.Rmd)
- [DRUG]_[TIME]_[DONORS]_downsampled.bam (cf. data downsampled_bam_merged from bloc2)
- [DRUG]_[TIME]_[DONOR]_downsampled.bam (cf. data downsampled_bam_separated from bloc2)
- hg19_annotations_gr.rds (cf. annotationsFiles_createGrange.Rmd)  

*OUTPUTS*  
- [DRUG1]_[TIME1]_vs_[DRUG2]_[TIME2]_volcano.svg  
- [DRUG]_all_peaks_dyn.png  
- [DRUG]_hg19_genes_intergenic_peaks_dyn.png  
- [DRUG]_hg19_genes_promoters_peaks_dyn.png
- [DRUG1]_[TIME1]_vs_[DRUG2]_[TIME2]_dotplot_GO_Up.svg
- [DRUG1]_[TIME1]_vs_[DRUG2]_[TIME2]_dotplot_GO_Down.svg
- [DRUG1]_[TIME1]_vs_[DRUG2]_[TIME2]_dotplot_pathway_Up.svg
- [DRUG1]_[TIME1]_vs_[DRUG2]_[TIME2]_dotplot_pathway_Down.svg 
- ATAC_pca_allsamples.png  
- common_regions_union_allsamples.csv  
- DEseq_readcount_df_union_allsamples.csv  
- DEseq_readcount_union_allsamples.rds  
- peaks_close_open_conditions.csv  
- readcount_df_peaks_union_allsamples.csv  
- readcount_union_allsamples.rds  
- union_gr_allsamples.gr.rds

```{r}

directory = getwd()

rmarkdown::render(input = paste0(directory, 
                                 "/bin/bulkATACseq/", 
                                 "bulkATACseq_R_Multiparametric_DifferentialAccessibility.Rmd"),
                  output_format = "html_document",
                  output_dir = paste0(directory, "/report/"),
                  output_file = paste0(Sys.Date(), 
                                       '_', 
                                       "bulkATACseq_R_Multiparametric_DifferentialAccessibility.html"))

```

