---
title: "scATACseq_bulkATACseq_R_Overlap"
author: "Laëtitia Racine"
date: "2023-03-22"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: false
    toc_float: false
    theme: journal
---


<style>
body {text-align: justify}
div.beige {background-color:#FFE0B2; border-radius: 5px; padding: 8px;}
</style>


```{r, Setup, include=F}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(knitr)
library(stringr)
library(GenomicRanges)
library(tibble)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(patchwork)

```

```{r, Working directory}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
directory_data = paste0(directory, "data/scATACseq/")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scATACseq_bulkATACseq_R_Overlap/"))
dir.create(path = paste0(directory, "exp/scATACseq_bulkATACseq_R_Overlap/", current_date))
directory_output = paste0(directory, "exp/scATACseq_bulkATACseq_R_Overlap/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

<br><br><br>

# Overview

1)  scATACseq data : .bed file from CellRanger output folder\
    initially mapped on hg38 genome\
    lift over hg19 genome with <https://genome.ucsc.edu/cgi-bin/hgLiftOver> website\
    transform into **granges** and add annotations in scATACseq_bulkATACseq_R\_SwitchGenome.Rmd script

2)  bulkATACseq data : **granges** from genomic ranges folder from bloc2 analysis\
    mapped on hg19 genome\
    update annotations to use the same annotation file as scATACseq in bulkATACseq_ChangeAnnotation.Rmd script

```{r, Input loading}

dir_sc = pic_last_dir(paste0(directory, "exp/", "scATACseq_bulkATACseq_R_SwitchGenome/"))
sc_gr = readRDS(paste0(dir_sc, "/", "list_hg19_gr_ann.rds"))
  
dir_bulk = paste0(pic_last_dir(paste0(directory, "exp/", "bulkATACseq_R_ChangeAnnotation/")), "/")
bulk_gr = list.files(dir_bulk, pattern = "update_annot.gr.rds")
bulk_gr = paste0(dir_bulk, bulk_gr)  
cond_time = str_extract(bulk_gr, pattern = "(?<=/)[:alnum:]+_[:digit:]{2}h(?=_D)")
bulk_gr = lapply(bulk_gr, readRDS)
names(bulk_gr) = cond_time

```

<br>

**Isolate common conditions**\
We have 15 conditions in bulk ATACseq and 4 conditions in scATACseq.\
We need to isolate the common conditions to do the comparison (DON, 2DG, AOA, CTRL at 24h).\
Note : sometimes CTRL is indicated with MP

```{r}

# Keep only common condition
cond_to_keep = paste0(names(sc_gr), "_24h")
cond_to_keep = c(cond_to_keep, "MP_24h")
bulk_gr = bulk_gr[names(bulk_gr) %in% cond_to_keep]

# Create table
sc_df = lapply(sc_gr, as_tibble)
bulk_df = lapply(bulk_gr, as_tibble)

```

<br>

**Calculate the overlap between peaks from bulk and peaks from single cell**\
Questions we ask : how many peaks of bulk are present in sc ? How many peaks of sc are present in bulk ?\
Note : when we calculate the overlap, one peak from one method can overlap multiple peaks in the other method.

```{r, function definition}

overlap_method = function(peaks_gr1, peaks_gr2, tab1, tab2, title1, title2, cond) {

  # Compute the overlap between peaks from method 1 and method 2
  overlap = findOverlaps(
    query = peaks_gr1,
    subject = peaks_gr2
  )
  
  # Retrieve peaks from method1 dataset positive for an overlap with the peaks in method2 dataset
  index_method1 = unique(queryHits(overlap))
  method1_overlap = as.data.frame(peaks_gr1[index_method1,])
  method1_overlap = distinct(method1_overlap)
  method1_overlap$peak_name = str_c(method1_overlap$seqnames,
                                    method1_overlap$start,
                                    method1_overlap$end,
                                    sep = "_")
  method1_overlap = method1_overlap$peak_name
  
  # Retrieve peaks from method2 dataset positive for an overlap with the peaks in method1 dataset
  index_method2 = subjectHits(overlap)
  method2_overlap = as.data.frame(peaks_gr2[index_method2,])
  method2_overlap = distinct(method2_overlap)
  method2_overlap$peak_name = str_c(method2_overlap$seqnames,
                                    method2_overlap$start,
                                    method2_overlap$end, sep = "_")
  method2_overlap = method2_overlap$peak_name
  
  # Add peak_name in the table
  tab1$peak_name = str_c(tab1$seqnames, tab1$start, tab1$end, sep = "_")
  tab2$peak_name = str_c(tab2$seqnames, tab2$start, tab2$end, sep = "_")
  tab2$test = str_c(tab2$seqnames, tab2$start, tab2$end, sep = "_")
  
  # Add a column to indicate if the peak overlap with the other method
  tab1$overlap = ifelse(tab1$peak_name %in% method1_overlap, "Shared peaks", "Not shared peaks")
  tab2$overlap = ifelse(tab2$peak_name %in% method2_overlap, "Shared peaks", "Not shared peaks") 

  ### All peaks 
  
  # Summarize the results
  tab1_summary = as.data.frame(table(tab1$overlap)) %>%
    dplyr::rename("Peak_category" = "Var1", "NumberOfPeaks" = "Freq") %>%
    dplyr::mutate(PercentagePeak = (NumberOfPeaks/nrow(tab1))*100)
  tab2_summary = as.data.frame(table(tab2$overlap)) %>%
    dplyr::rename("Peak_category" = "Var1", "NumberOfPeaks" = "Freq") %>%
    dplyr::mutate(PercentagePeak = (NumberOfPeaks/nrow(tab2))*100)
  
  # Plot the results
  plot1 = ggplot(tab1, aes(x = overlap)) +
    geom_bar(fill = c("#D5269D", "#A9D23D")) +
    geom_text(stat='count', aes(label=..count..), vjust = 0) +
    ggtitle(paste(title1, cond, "- All peaks"))
  plot2 = ggplot(tab2, aes(x = overlap)) +
    geom_bar(fill = c("#D5269D", "#A9D23D")) +
    geom_text(stat='count', aes(label=..count..), vjust = 0) +
    ggtitle(paste(title2, cond, "- All peaks"))
  
  ### Promoters peaks
  tab1_prom = tab1 %>% dplyr::filter(hg19_genes_promoters == TRUE)
  tab2_prom = tab2 %>% dplyr::filter(hg19_genes_promoters == TRUE)
  
   # Summarize the results
  tab1_summary_prom = as.data.frame(table(tab1_prom$overlap)) %>%
    dplyr::rename("Peak_category" = "Var1", "NumberOfPeaks" = "Freq") %>%
    dplyr::mutate(PercentagePeak = (NumberOfPeaks/nrow(tab1_prom))*100)
  tab2_summary_prom = as.data.frame(table(tab2_prom$overlap)) %>%
    dplyr::rename("Peak_category" = "Var1", "NumberOfPeaks" = "Freq") %>%
    dplyr::mutate(PercentagePeak = (NumberOfPeaks/nrow(tab2_prom))*100)
  
  # Plot the results
  plot1_prom = ggplot(tab1_prom, aes(x = overlap)) +
    geom_bar(fill = c("#D5269D", "#A9D23D")) +
    geom_text(stat='count', aes(label=..count..), vjust = 0) +
    ggtitle(paste(title1, cond, "- Promoter peaks"))
  plot2_prom = ggplot(tab2_prom, aes(x = overlap)) +
    geom_bar(fill = c("#D5269D", "#A9D23D")) +
    geom_text(stat='count', aes(label=..count..), vjust = 0) +
    ggtitle(paste(title2, cond, "- Promoter peaks"))
  
  ### Intergenic peaks
  tab1_inter = tab1 %>% dplyr::filter(hg19_genes_intergenic == TRUE)
  tab2_inter = tab2 %>% dplyr::filter(hg19_genes_intergenic == TRUE)
  
   # Summarize the results
  tab1_summary_inter = as.data.frame(table(tab1_inter$overlap)) %>%
    dplyr::rename("Peak_category" = "Var1", "NumberOfPeaks" = "Freq") %>%
    dplyr::mutate(PercentagePeak = (NumberOfPeaks/nrow(tab1_inter))*100)
  tab2_summary_inter = as.data.frame(table(tab2_inter$overlap)) %>%
    dplyr::rename("Peak_category" = "Var1", "NumberOfPeaks" = "Freq") %>%
    dplyr::mutate(PercentagePeak = (NumberOfPeaks/nrow(tab2_inter))*100)
  
  # Plot the results
  plot1_inter = ggplot(tab1_inter, aes(x = overlap)) +
    geom_bar(fill = c("#D5269D", "#A9D23D")) +
    geom_text(stat='count', aes(label=..count..), vjust = 0) +
    ggtitle(paste(title1, cond, "- Intergenic peaks"))
  plot2_inter = ggplot(tab2_inter, aes(x = overlap)) +
    geom_bar(fill = c("#D5269D", "#A9D23D")) +
    geom_text(stat='count', aes(label=..count..), vjust = 0) +
    ggtitle(paste(title2, cond, "- Intergenic peaks"))
  
  # Output of the function
  results = list(tab1 = tab1,
                 tab2 = tab2,
                 tab1_summary = tab1_summary,
                 tab2_summary = tab2_summary,
                 plot1 = plot1,
                 plot2 = plot2,
                 tab1_summary_prom = tab1_summary_prom,
                 tab2_summary_prom = tab2_summary_prom,
                 plot1_prom = plot1_prom,
                 plot2_prom = plot2_prom,
                 tab1_summary_inter = tab1_summary_inter,
                 tab2_summary_inter = tab2_summary_inter,
                 plot1_inter = plot1_inter,
                 plot2_inter = plot2_inter)
  return(results)
  
}

```

```{r, call function}

CTRL_overlap = overlap_method(peaks_gr1 = bulk_gr[["MP_24h"]], 
                              peaks_gr2 = sc_gr[["CTRL"]], 
                              tab1 = bulk_df[["MP_24h"]], 
                              tab2 = sc_df[["CTRL"]], 
                              title1 = "bulkATACseq dataset", 
                              title2 = "scATACseq dataset",
                              cond = "CTRL")

DON_overlap = overlap_method(peaks_gr1 = bulk_gr[["DON_24h"]], 
                             peaks_gr2 = sc_gr[["DON"]], 
                             tab1 = bulk_df[["DON_24h"]], 
                             tab2 = sc_df[["DON"]], 
                             title1 = "bulkATACseq dataset", 
                             title2 = "scATACseq dataset",
                             cond = "DON")

DDG_overlap = overlap_method(peaks_gr1 = bulk_gr[["2DG_24h"]], 
                             peaks_gr2 = sc_gr[["2DG"]], 
                             tab1 = bulk_df[["2DG_24h"]], 
                             tab2 = sc_df[["2DG"]], 
                             title1 = "bulkATACseq dataset", 
                             title2 = "scATACseq dataset",
                             cond = "2DG")

AOA_overlap = overlap_method(peaks_gr1 = bulk_gr[["AOA_24h"]], 
                             peaks_gr2 = sc_gr[["AOA"]], 
                             tab1 = bulk_df[["AOA_24h"]], 
                             tab2 = sc_df[["AOA"]], 
                             title1 = "bulkATACseq dataset", 
                             title2 = "scATACseq dataset",
                             cond = "AOA")

```

```{r, all peaks, fig.width = 16, fig.height = 4, results = "hold"}

# Visualize summary
cat("Peaks of bulkATACseq")
CTRL_overlap$tab1_summary %>%
  kable(caption = "CTRL - All peaks") %>%
  kable_styling()
DON_overlap$tab1_summary %>%
  kable(caption = "DON - All peaks") %>%
  kable_styling()
DDG_overlap$tab1_summary %>%
  kable(caption = "2DG - All peaks") %>%
  kable_styling()
AOA_overlap$tab1_summary %>%
  kable(caption = "AOA - All peaks") %>%
  kable_styling()

cat("Peaks of scATACseq")
CTRL_overlap$tab2_summary %>%
  kable(caption = "CTRL - All peaks") %>%
  kable_styling()
DON_overlap$tab2_summary %>%
  kable(caption = "DON - All peaks") %>%
  kable_styling()
DDG_overlap$tab2_summary %>%
  kable(caption = "2DG - All peaks") %>%
  kable_styling()
AOA_overlap$tab2_summary %>%
  kable(caption = "AOA - All peaks") %>%
  kable_styling()

# Show plots
CTRL_overlap$plot1 | DON_overlap$plot1 | DDG_overlap$plot1 | AOA_overlap$plot1
CTRL_overlap$plot2 | DON_overlap$plot2 | DDG_overlap$plot2 | AOA_overlap$plot2

```

```{r, promoter peaks, fig.width = 16, fig.height = 4, results = "hold"}

# Visualize summary
cat("Peaks of bulkATACseq")
CTRL_overlap$tab1_summary_prom %>%
  kable(caption = "CTRL - Promoters peaks") %>%
  kable_styling()
DON_overlap$tab1_summary_prom %>%
  kable(caption = "DON - Promoters peaks") %>%
  kable_styling()
DDG_overlap$tab1_summary_prom %>%
  kable(caption = "2DG - Promoters peaks") %>%
  kable_styling()
AOA_overlap$tab1_summary_prom %>%
  kable(caption = "AOA - Promoters peaks") %>%
  kable_styling()

cat("Peaks of scATACseq")
CTRL_overlap$tab2_summary_prom %>%
  kable(caption = "CTRL - Promoters peaks") %>%
  kable_styling()
DON_overlap$tab2_summary_prom %>%
  kable(caption = "DON - Promoters peaks") %>%
  kable_styling()
DDG_overlap$tab2_summary_prom %>%
  kable(caption = "2DG - Promoters peaks") %>%
  kable_styling()
AOA_overlap$tab2_summary_prom %>%
  kable(caption = "AOA - Promoters peaks") %>%
  kable_styling()

# Show plots
CTRL_overlap$plot1_prom | DON_overlap$plot1_prom | DDG_overlap$plot1_prom | AOA_overlap$plot1_prom
CTRL_overlap$plot2_prom | DON_overlap$plot2_prom | DDG_overlap$plot2_prom | AOA_overlap$plot2_prom

```

```{r, intergenic peaks, fig.width = 16, fig.height = 4, results = "hold"}

# Visualize summary
cat("Peaks of bulkATACseq")
CTRL_overlap$tab1_summary_inter %>%
  kable(caption = "CTRL - Intergenic peaks") %>%
  kable_styling()
DON_overlap$tab1_summary_inter %>%
  kable(caption = "DON - Intergenic peaks") %>%
  kable_styling()
DDG_overlap$tab1_summary_inter %>%
  kable(caption = "2DG - Intergenic peaks") %>%
  kable_styling()
AOA_overlap$tab1_summary_inter %>%
  kable(caption = "AOA - Intergenic peaks") %>%
  kable_styling()

cat("Peaks of scATACseq")
CTRL_overlap$tab1_summary_inter %>%
  kable(caption = "CTRL - Intergenic peaks") %>%
  kable_styling()
DON_overlap$tab1_summary_inter %>%
  kable(caption = "DON - Intergenic peaks") %>%
  kable_styling()
DDG_overlap$tab1_summary_inter %>%
  kable(caption = "2DG - Intergenic peaks") %>%
  kable_styling()
AOA_overlap$tab1_summary_inter %>%
  kable(caption = "AOA - Intergenic peaks") %>%
  kable_styling()

# Show plots
CTRL_overlap$plot1_inter | DON_overlap$plot1_inter | DDG_overlap$plot1_inter | AOA_overlap$plot1_inter
CTRL_overlap$plot2_inter | DON_overlap$plot2_inter | DDG_overlap$plot2_inter | AOA_overlap$plot2_inter

```

```{r, save outputs, fig.show="hide"}

plot_bulk = (CTRL_overlap$plot1 | DON_overlap$plot1 | DDG_overlap$plot1 | AOA_overlap$plot1) /
  (CTRL_overlap$plot1_prom | DON_overlap$plot1_prom | DDG_overlap$plot1_prom | AOA_overlap$plot1_prom) /
  (CTRL_overlap$plot1_inter | DON_overlap$plot1_inter | DDG_overlap$plot1_inter | AOA_overlap$plot1_inter)

plot_sc = (CTRL_overlap$plot2 | DON_overlap$plot2 | DDG_overlap$plot2 | AOA_overlap$plot2) /
  (CTRL_overlap$plot2_prom | DON_overlap$plot2_prom | DDG_overlap$plot2_prom | AOA_overlap$plot2_prom) /
  (CTRL_overlap$plot2_inter | DON_overlap$plot2_inter | DDG_overlap$plot2_inter | AOA_overlap$plot2_inter)

ggsave(plot = plot_bulk, width = 18, height = 12,
       filename = paste0(directory_output, "plot_compare_bulkpeaks.svg"))
ggsave(plot = plot_sc, width = 18, height = 12,
       filename = paste0(directory_output, "plot_compare_singlecellpeaks.svg"))
rm(plot_bulk, plot_sc)

bk_all = bind_rows(CTRL_overlap$tab1_summary %>% dplyr::mutate(Condition = "CTRL"),
                   DON_overlap$tab1_summary %>% dplyr::mutate(Condition = "DON"),
                   DDG_overlap$tab1_summary %>% dplyr::mutate(Condition = "2DG"),
                   AOA_overlap$tab1_summary %>% dplyr::mutate(Condition = "AOA")
) %>% dplyr::mutate(Method = "bulk", Peaks_category = "all peaks")

sc_all = bind_rows(CTRL_overlap$tab2_summary %>% dplyr::mutate(Condition = "CTRL"),
                   DON_overlap$tab2_summary %>% dplyr::mutate(Condition = "DON"),
                   DDG_overlap$tab2_summary %>% dplyr::mutate(Condition = "2DG"),
                   AOA_overlap$tab2_summary %>% dplyr::mutate(Condition = "AOA")
) %>% dplyr::mutate(Method = "single cell", Peaks_category = "all peaks")

bk_prom = bind_rows(CTRL_overlap$tab1_summary_prom %>% dplyr::mutate(Condition = "CTRL"),
                    DON_overlap$tab1_summary_prom %>% dplyr::mutate(Condition = "DON"),
                    DDG_overlap$tab1_summary_prom %>% dplyr::mutate(Condition = "2DG"),
                    AOA_overlap$tab1_summary_prom %>% dplyr::mutate(Condition = "AOA")
) %>% dplyr::mutate(Method = "bulk", Peaks_category = "promoter peaks")

sc_prom = bind_rows(CTRL_overlap$tab2_summary_prom %>% dplyr::mutate(Condition = "CTRL"),
                    DON_overlap$tab2_summary_prom %>% dplyr::mutate(Condition = "DON"),
                    DDG_overlap$tab2_summary_prom %>% dplyr::mutate(Condition = "2DG"),
                    AOA_overlap$tab2_summary_prom %>% dplyr::mutate(Condition = "AOA")
) %>% dplyr::mutate(Method = "single cell", Peaks_category = "promoter peaks")

bk_inter = bind_rows(CTRL_overlap$tab1_summary_inter %>% dplyr::mutate(Condition = "CTRL"),
                     DON_overlap$tab1_summary_inter %>% dplyr::mutate(Condition = "DON"),
                     DDG_overlap$tab1_summary_inter %>% dplyr::mutate(Condition = "2DG"),
                     AOA_overlap$tab1_summary_inter %>% dplyr::mutate(Condition = "AOA")
) %>% dplyr::mutate(Method = "bulk", Peaks_category = "intergenic peaks")

sc_inter = bind_rows(CTRL_overlap$tab2_summary_inter %>% dplyr::mutate(Condition = "CTRL"),
                     DON_overlap$tab2_summary_inter %>% dplyr::mutate(Condition = "DON"),
                     DDG_overlap$tab2_summary_inter %>% dplyr::mutate(Condition = "2DG"),
                     AOA_overlap$tab2_summary_inter %>% dplyr::mutate(Condition = "AOA")
) %>% dplyr::mutate(Method = "single cell", Peaks_category = "intergenic peaks")

tab_glob = bind_rows(bk_all, sc_all, bk_prom, sc_prom, bk_inter, sc_inter)
rm(bk_all, sc_all, bk_prom, sc_prom, bk_inter, sc_inter)
write.csv2(x = tab_glob, row.names = FALSE,
           file = paste0(directory_output, "tab_percentage_peaks_shared.csv"))

```


<br>

<div class = "beige">
Most of bulkATACseq peaks have an equivalent in the scATACseq dataset (80% of all peaks and more than 90% of promoter peaks).\
However, when we look at the scATACseq peaks we have more not shared peaks than shared peaks except for promoters where shared peaks are predominant. We want to investigate a little bit further on scATACseq peaks to understand the particularity of the peaks that we can't detect with the bulk method.
</div>

<br>

**Investigate not shared peaks for scATACseq**\
How many cells of scATACseq are sharing the peaks that we can't detect in bulk ATACseq ? Those peaks correspond to which genomic annotation ?
```{r, eval = FALSE}

### !!! Extremely long !!!

# Load the matrix of counts
dir =  pic_last_dir(paste0(directory, "exp/scATACseq_bulkATACseq_R_SwitchGenome/"))
list_mtx_hg19 = readRDS(paste0(dir, "/", "list_mtx_hg19.rds"))

# Transform matrices into a binary one : detected/no detected
list_mtx_hg19_bin = list()
for (i in 1:length(list_mtx_hg19)) {
  mtx = list_mtx_hg19[[i]]
  mtx[mtx>0] = 1
  list_mtx_hg19_bin[[i]] = mtx
}
names(list_mtx_hg19_bin) = names(list_mtx_hg19)
rm(list_mtx_hg19)
saveRDS(list_mtx_hg19_bin, file = paste0(directory_output, "list_mtx_hg19_binary.rds"))

```

```{r}

list_mtx_hg19_bin = readRDS(paste0(directory_output, "list_mtx_hg19_binary.rds"))

```

```{r, fig.width = 16, fig.height = 4}

# Function definition : How much the peaks are shared among cells ?
peaks_nbcells_fun = function(bin_mtx, cond) {
  # Summing rows to count number of cells sharing a peak
  df = as.data.frame(rowSums(bin_mtx)) 
  colnames(df) = "nb_cells_sharing_this_peak"
  df$peak_name = rownames(df)
  # Histogram of number of peaks (= freq) shared by n cells
  plot = ggplot(data = df, aes(x = nb_cells_sharing_this_peak)) +
    geom_histogram(bins = 50) +
    ggtitle(label = paste(cond, "Distribution - Nb cell sharing a peak"))
  # Function outputs
  return(list(df = df, plot = plot))
}

# Call the function
CTRL_peaks_shared = peaks_nbcells_fun(list_mtx_hg19_bin[["CTRL"]], "CTRL")
DON_peaks_shared = peaks_nbcells_fun(list_mtx_hg19_bin[["DON"]], "DON")
DDG_peaks_shared = peaks_nbcells_fun(list_mtx_hg19_bin[["2DG"]], "2DG")
AOA_peaks_shared = peaks_nbcells_fun(list_mtx_hg19_bin[["AOA"]], "AOA")

# Visualize the plots
CTRL_peaks_shared$plot | DON_peaks_shared$plot | DDG_peaks_shared$plot | AOA_peaks_shared$plot

```

```{r, fig.width = 16, fig.height = 16}

# Create summary table (scATACseq peaks are in tab2)
CTRL_over_tab = full_join(x = CTRL_overlap$tab2, y = CTRL_peaks_shared$df, by = "peak_name")
DON_over_tab = full_join(x = DON_overlap$tab2, y = DON_peaks_shared$df, by = "peak_name")
DDG_over_tab = full_join(x = DDG_overlap$tab2, y = DDG_peaks_shared$df, by = "peak_name")
AOA_over_tab = full_join(x = AOA_overlap$tab2, y = AOA_peaks_shared$df, by = "peak_name")

# Function definition : number of cells sharing the peaks depending if they are shared or not
plot_shared_cells_fun = function(tab_recap, annotation, cond) {
  # Subset tab if necessary
  if (!annotation %in% c("all", "promoter", "intergenic")) { 
    print("annotation not recognize")
  } else {
    if (annotation == "all") { tab_recap = tab_recap }
    if (annotation == "promoter") { tab_recap = tab_recap %>% dplyr::filter(hg19_genes_promoters == TRUE) }
    if (annotation == "intergenic") { tab_recap = tab_recap %>% dplyr::filter(hg19_genes_intergenic == TRUE)} 
    }
  # Calculate the median of nb cells sharing a peak per group
  tab_recap = tab_recap %>%
    dplyr::group_by(overlap) %>%
    dplyr::mutate(sharing_cell_med = median(nb_cells_sharing_this_peak))
  # plot the graph
  median_shared = unique((tab_recap %>% dplyr::filter(overlap == "Shared peaks"))$sharing_cell_med)
  median_noshared = unique((tab_recap %>% dplyr::filter(overlap == "Not shared peaks"))$sharing_cell_med)
  plot = ggplot(tab_recap, aes(x = nb_cells_sharing_this_peak, color = overlap)) +
    geom_histogram(aes(fill = overlap), alpha = 0.3, position = "identity", binwidth = 300) +
    geom_vline(aes(xintercept=sharing_cell_med, color=overlap),linetype="dashed") +
    scale_discrete_manual(aesthetics = c("fill", "col"), values = c('Shared peaks'= "#D5269D", 'Not shared peaks'="#A9D23D")) +
    ggtitle(label = paste(cond, "Distribution - Nb cells sharing peaks"),
            subtitle = paste("Total number of peaks :", nrow(tab_recap), "\n",
                           "Peaks genomic category :", annotation, "\n",
                           "Median peaks common bulk/single cell :", median_shared, "cells sharing the peak \n",
                           "Median peaks only single cell :", median_noshared, "cells sharing the peak \n"))
  # output function
  return(plot)
}

# Call the function
CTRL_all = plot_shared_cells_fun(CTRL_over_tab, "all", "CTRL")
CTRL_prom = plot_shared_cells_fun(CTRL_over_tab, "promoter", "CTRL")
CTRL_inter = plot_shared_cells_fun(CTRL_over_tab, "intergenic", "CTRL")
DON_all = plot_shared_cells_fun(CTRL_over_tab, "all", "DON")
DON_prom = plot_shared_cells_fun(CTRL_over_tab, "promoter", "DON")
DON_inter = plot_shared_cells_fun(CTRL_over_tab, "intergenic", "DON")
DDG_all = plot_shared_cells_fun(CTRL_over_tab, "all", "2DG")
DDG_prom = plot_shared_cells_fun(CTRL_over_tab, "promoter", "2DG")
DDG_inter = plot_shared_cells_fun(CTRL_over_tab, "intergenic", "2DG")
AOA_all = plot_shared_cells_fun(CTRL_over_tab, "all", "AOA")
AOA_prom = plot_shared_cells_fun(CTRL_over_tab, "promoter", "AOA")
AOA_inter = plot_shared_cells_fun(CTRL_over_tab, "intergenic", "AOA")

# Show plot 
plot_grid = (CTRL_all | CTRL_prom | CTRL_inter) /
  (DON_all | DON_prom | DON_inter) /
  (DDG_all | DDG_prom | DDG_inter) /
  (AOA_all | AOA_prom | AOA_inter) 
plot_grid

```

```{r, fig.show = "hide"}

ggsave(plot = (CTRL_peaks_shared$plot | DON_peaks_shared$plot | 
                 DDG_peaks_shared$plot | AOA_peaks_shared$plot),
       width = 18, height = 8,
       filename = paste0(directory_output, "plots_scATAC_shared_peaks.svg"))

ggsave(plot = plot_grid, width = 18, height = 16,
       filename = paste0(directory_output, "plots_overlap_shared_peaks.svg"))

```


<br>

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```
