---
title: "bulkATACseq_R_Multiparametric_wotVPA"
author: "Laëtitia Racine"
date: "2023-03-27"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=F}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(knitr)
library(stringr)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(DESeq2)
library(Rsubread)
library(tidyverse) # column_to_rownames() function
library(pheatmap)

```

```{r, Working directory}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/bulkATACseq_R_Multiparametric_wotVPA/"))
dir.create(path = paste0(directory, "exp/bulkATACseq_R_Multiparametric_wotVPA/", 
                         current_date))
directory_output = paste0(directory, "exp/bulkATACseq_R_Multiparametric_wotVPA/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

Analysis based on : https://tobiasrausch.com/courses/atac/atac-seq-data-analysis.html  
Same analysis as in the first part of bulkATACseq_R_Multiparametric_DifferentialAccessibility.Rmd.  
We want to calculate the PCA without the VPA sample.  
We use the count_df table as input and we remove VPA information.  

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/bulkATACseq_R_Multiparametric_DifferentialAccessibility/"))
count_df = read.csv2(paste0(dir, "/readcount_df_peaks_union_allsamples.csv"))
count_df = count_df %>% dplyr::select(!str_subset(colnames(count_df), "VPA"))
# automatix X is added when object is loaded
colnames(count_df) = str_replace_all(string = colnames(count_df), pattern = "X2DG", replacement = "2DG")

```


<br><br><br>


# Perform DataNormalization with DEseq

https://www.rdocumentation.org/packages/DESeq2/versions/1.12.3/topics/DESeqDataSet-class    
https://tobiasrausch.com/courses/atac/atac-seq-data-analysis.html#data-normalization    
```{r}

# Extract the matrix with number of reads in each merged donors
cols_to_keep = str_subset(string = colnames(count_df), 
                          pattern = "D[:digit:].D[:digit:]_downsampled.bam")
counts = count_df[ , cols_to_keep]

## Extract information in the colnames
coldata = data.frame(file = cols_to_keep, 
                     name = unlist(str_extract_all(cols_to_keep, 
                                                   pattern=".+(?=_downsampled.bam)"))) %>%
  dplyr::mutate(donors = str_extract(name, pattern = "D[:digit:].+")) %>%
  dplyr::mutate(condtime = paste0(str_extract(name, pattern = ".+(?=h)"), "h")) %>% 
  tidyr::separate(col = condtime, into = c("condition", "time"), remove = TRUE) 
coldata = coldata %>%
  tidyr::unite(condition_time, condition, time, sep = "_", remove = FALSE) %>%
  remove_rownames %>%
  column_to_rownames(var = "file") %>%
  dplyr::select(-name)

# Data normalization with DEseq
dds = DESeqDataSetFromMatrix(
  countData = counts, 
  colData = coldata, 
  design = ~ condition)
dds = DESeq(dds)
saveRDS(dds, paste0(directory_output, "DEseq_readcount_union_allsamples.rds"))

# Organize DEseq results in a table
cm = data.frame(counts(dds, normalized=TRUE)) %>%
  tibble::rownames_to_column("region") %>%
  tidyr::separate(region, c("seqnames", "start", "end"), remove = FALSE) %>%
  column_to_rownames("region")
colnames(cm) = str_replace_all(colnames(cm), pattern = "X2DG", replacement = "2DG")
write.csv2(x = cm, row.names = FALSE,
           file = paste0(directory_output, "DEseq_readcount_df_union_allsamples.csv"))

# Visualize the table 
cm[1:20, ] %>%
  kable(caption = "First 20 rows of the 'cm' table") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "250px")

```


# Principal Component Analysis (PCA)

https://tobiasrausch.com/courses/atac/atac-seq-data-analysis.html#principal-component-analysis-pca  

<br>

**Choice of PC for the plot**  
We want to select the PCs that explain the most the variance.   
```{r}

pca = prcomp(t(cm[,-c(1:3)]))
print(summary(pca))
pcaData = as.data.frame(pca$x)
pcaData$sample=rownames(pcaData)
coldata = coldata %>% rownames_to_column(var = "sample")
pcaData=merge(pcaData, coldata)
percentVar = round(100*(pca$sdev^2/sum(pca$sdev^2)))

varexp = data.frame(x=1:length(percentVar), y=percentVar)
varexp$x = factor(varexp$x)
ggplot(data=varexp, aes(x=x, y=y)) + 
  geom_bar(stat="identity") + 
  xlab("Principal Component") + 
  ylab("Proportion of variation (%)")

```
We choose PC1 and PC2.

<br>

**Plot the PCA**
```{r}
 
plot_pca = ggplot(data=pcaData, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = condition, shape = time), size=6) +
  scale_color_manual(values = color_code, limits = force) +
  scale_shape_manual(values=c(15, 16, 17, 7)) +
  ggtitle("PCA on normalized read counts from union of all samples") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_text(vjust = 1 ,size = 14),
        axis.text.x = element_text(size = 11, colour = "black"),
        axis.ticks = element_blank()) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) 
plot_pca

ggsave(plot = plot_pca,
       filename = paste0(directory_output, "ATAC_pca_samplesWoT_VPA.png"),
       width = 16*0.75, height = 9*0.75)

```

Comment : Time point 00h, 03h and 12h are separated by peaks from PC1. For those time points, the different conditions are really close to each other. The conditions start to differ at 24h, they are separated on the plot by peaks from PC2 mostly but also PC1.

<br>

**Loadings from each PCs**  
Which peaks contribute the most to the separation of the samples ?  
```{r}

# Peaks contribution for each PC
loadings = abs(pca$rotation)
contribution = as.data.frame(sweep(loadings, 2, colSums(loadings), "/"))

# 500 peaks with the highest contribution for PC1
pc1_500peaks = contribution %>% dplyr::arrange(desc(PC1))
pc1_500peaks = rownames(pc1_500peaks[1:500,])

# 500 peaks with the highest contribution for PC2
pc2_500peaks = contribution %>% dplyr::arrange(desc(PC2))
pc2_500peaks = rownames(pc2_500peaks[1:500,])

# list of interesting peaks for all samples
pc1_pc2_peaks = intersect(pc1_500peaks, pc2_500peaks)
pc1_only_peaks = setdiff(pc1_500peaks, pc2_500peaks)
pc2_only_peaks = setdiff(pc1_500peaks, pc2_500peaks)

```

<br>

## Peaks investigation 

**Heatmap on the 500 peaks with the highest loading for PC1 and PC2**
```{r, fig.height=16, fig.width = 8, results="hold"}

# Prepare dataframe for heatmap
cm_heatmap = cm[, -c(1:3)]
colnames(cm_heatmap) = str_replace_all(colnames(cm_heatmap),"(.D[:digit:])+_downsampled.bam","")
cm_heatmap = cm_heatmap %>%
  dplyr::select(Xvivo_00h, MP_03h, MP_12h, MP_24h, DON_03h,
                DON_12h, DON_24h, '2DG_03h', '2DG_12h', '2DG_24h',
                AOA_03h, AOA_12h, AOA_24h, aK_24h)
# heatmap on peaks from PC1
mat_pc1 = cm_heatmap %>% dplyr::filter(row.names(cm_heatmap) %in% pc1_only_peaks)
mat_pc1 = mat_pc1 - rowMeans(mat_pc1)
pheatmap(mat_pc1, scale="row", cluster_rows = FALSE, cluster_cols = FALSE, fontsize_row = 5)

# heatmap on peaks from PC2
mat_pc2 = cm_heatmap %>% dplyr::filter(row.names(cm_heatmap) %in% pc2_only_peaks)
mat_pc2 = mat_pc2 - rowMeans(mat_pc2)
pheatmap(mat_pc2, scale="row", cluster_rows = FALSE, cluster_cols = FALSE, fontsize_row = 5)

# heatmap on peaks from PC1 and PC2
mat_pc1_pc2 = cm_heatmap %>% dplyr::filter(row.names(cm_heatmap) %in% pc1_pc2_peaks)
mat_pc1_pc2 = mat_pc1_pc2 - rowMeans(mat_pc1_pc2)
pheatmap(mat_pc1_pc2, scale="row", cluster_rows=FALSE, cluster_cols=FALSE, fontsize_row=5)

```


<br><br><br>



```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric(end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

