---
title: "bulkATACseq_R_Multiparametric_DifferentialAccessibility"
author: "Laëtitia Racine"
date: "2023-03-27"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=F}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(knitr)
library(stringr)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(DESeq2)
library(Rsubread)
library(tidyverse) # column_to_rownames() function
library(pheatmap)
library(gridExtra)
library(patchwork)
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(org.Hs.eg.db)
library(ReactomePA)
library(clusterProfiler)

```

```{r, Working directory}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
# current_date = format(Sys.time(), "%Y%m%d")
current_date = "20230403"
dir.create(path = paste0(directory, "exp/bulkATACseq_R_Multiparametric_DifferentialAccessibility/"))
dir.create(path = paste0(directory, "exp/bulkATACseq_R_Multiparametric_DifferentialAccessibility/", current_date))
directory_output = paste0(directory, "exp/bulkATACseq_R_Multiparametric_DifferentialAccessibility/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

# Load grange objects 
dir = pic_last_dir(paste0(directory, "exp/bulkATACseq_R_ChangeAnnotation/"))
gr_ann_list = list.files(path = dir, pattern = "update_annot.gr.rds", full.names = TRUE)
name_list = str_extract(gr_ann_list, pattern = "(?<=[:digit:]{8}/)[:graph:]+(?=_update_annot.gr.rds)")
gr_ann_list = lapply(gr_ann_list, readRDS)
names(gr_ann_list) = name_list
  
# Extract bam file list for merged donors
dir_data = paste0(directory, "data/bulkATACseq/downsampled_bam_merged")
bam_merged_list = list.files(path = dir_data, pattern = "_downsampled.bam$", full.names = TRUE)

# Extract bam file for separated donors
# We use the files from bloc2 snakefile_separated_without06h_commita38dc
dir_data = paste0(directory, "data/bulkATACseq/downsampled_bam_separated")
bam_sep_list = list.files(path = dir_data, pattern = "_downsampled.bam$", full.names = TRUE)

# Merge the list
bam_list = c(bam_merged_list, bam_sep_list)

# Load grange with hg19 annotations
dir_annot = pic_last_dir(paste0(directory, "exp/", "AnnotationsFiles_createGrange/"))
hg19_annotations = readRDS(paste0(dir_annot, "/", "hg19_annotations_gr.rds"))

```


Analysis based on : https://tobiasrausch.com/courses/atac/atac-seq-data-analysis.html  


<br><br><br>



# Create a common set of regions for all sample

<br>

**Common list of regions for all samples**  
To be able to compare the peaks from one sample to another, we need to create a big Grange with all the regions detected in all sample.  
Each "peak" in the df_gr_union table is a region in which we can find at least one peak among all the samples  .  
https://stackoverflow.com/questions/28545688/understand-the-reduce-function    
https://bioconductor.org/packages/devel/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesIntroduction.html    
```{r}

# Create a grange with the union of all peaks from all sample
gr_union = Reduce(GenomicRanges::union, gr_ann_list)

# Add annotation on grange union
annotations_types = levels(factor(hg19_annotations$type))
metadata = matrix(FALSE, ncol = length(annotations_types), nrow = length(gr_union))
colnames(metadata) = annotations_types
mcols(gr_union) = metadata
for (i in 1:ncol(metadata)){
    sub_annot = hg19_annotations[hg19_annotations$type == annotations_types[i]]
    overlaps = findOverlaps(gr_union, sub_annot)
    mcols(gr_union)[queryHits(overlaps),i] = TRUE
}
saveRDS(gr_union, file=paste0(directory_output, "union_gr_allsamples.gr.rds"))

# Annotation table for featureCounts function
df_gr_union = data.frame(gr_union) %>%
      tibble::rownames_to_column(var = "number") %>%
      dplyr::mutate(GeneID = paste0("peak_", number), .keep = "unused") %>%
      dplyr::rename(Chr = seqnames, Start = start, End = end, Strand = strand) %>%
      dplyr::select(GeneID, Chr, Start, End, Strand)
write.csv2(x = df_gr_union, row.names = FALSE, 
           file=paste0(directory_output, "common_regions_union_allsamples.csv") )

# Annotation table with genomic location
df_gr_union_annot = data.frame(gr_union) %>%
      tibble::rownames_to_column(var = "number") %>%
      dplyr::mutate(region = paste0("region_", number), .keep = "unused", .before = seqnames)

# Visualize table
df_gr_union[1:20,] %>%
  kable(caption = "First 20 rows of the df_gr_union table") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "250px")

```

<br>

**Number of reads detected per sample for each region of the union**  
We use the merged donors files and the separated donors files (needed for DEseq2 analysis later).  
```{r}

# Count the number of reads, for each sample, present in the region resulting from the union
readCount <- featureCounts(files = bam_list,
                           annot.ext = df_gr_union,
                           isPairedEnd = TRUE,
                           nthreads = 1,
                           countChimericFragments = FALSE,
                           countMultiMappingReads = TRUE)
saveRDS(readCount, file = paste0(directory_output, "readcount_union_allsamples.rds"))

```

```{r, eval = FALSE}
readCount = readRDS(paste0(directory_output, "readcount_union_allsamples.rds"))
```

<br>

**Matching region - peak**   
For each sample (merged donors), we indicate if the region of the union was detected as a peak or not.  
```{r}

gr_readcount = makeGRangesFromDataFrame(df = df_gr_union)

peak_corr = list()
for(i in 1:length(gr_ann_list)) {
  temp = as.data.frame(as.logical(countOverlaps(gr_readcount, gr_ann_list[[i]])))
  name = str_extract(names(gr_ann_list)[i], "[:alnum:]+_[:digit:]{2}h+_[:graph:]+")
  colnames(temp) = paste0("peaks_in_", name)
  peak_corr[[i]] = temp
}
peak_corr = do.call("cbind", peak_corr) %>%
      tibble::rownames_to_column(var = "region") %>%
      dplyr::mutate(region = paste0("region_", region))

```

<br>

**Global table with list of regions, number of reads per sample and peaks detected or not for that condition**  
```{r}

# Create the table
count_df = left_join(
  x = data.frame(readCount$annotation) %>% dplyr::select(-Strand, -Length),
  y = tibble::rownames_to_column(data.frame(readCount$counts), "GeneID"),
  by = "GeneID") %>% 
  dplyr::mutate(region = str_replace(GeneID, "peak", "region"), .before = Chr) %>%
  dplyr::rename("seqnames" = "Chr", "start" = "Start", "end" = "End") %>%
  dplyr::select(-GeneID)
count_df = left_join(count_df, peak_corr, by = "region")
colnames(count_df) = str_replace_all(string = colnames(count_df), pattern = "X2DG", replacement = "2DG")
rownames(count_df) = paste0(count_df$seqnames, '_', count_df$start, '_', count_df$end)

# Save the table
write.table(count_df, sep = ";", row.names = FALSE, 
            file = paste0(directory_output, "readcount_df_peaks_union_allsamples.csv"))    

# Visualize the table
count_df[1:20, ] %>%
  kable(caption = "First 20 rows of the 'count_df' table") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "250px")

```

<br>

**Table with list of regions and annotations for pathway analysis**
```{r}

gr = makeGRangesFromDataFrame(count_df, keep.extra.columns = T)
peakAnno = annotatePeak(gr, tssRegion=c(-1000, 1000), 
                        TxDb=TxDb.Hsapiens.UCSC.hg19.knownGene, 
                        annoDb="org.Hs.eg.db")
dfPA = as.data.frame(peakAnno) 
dfPA = dfPA  %>%
  dplyr::select(-str_subset(colnames(dfPA), "downsampled")) %>% 
  dplyr::select(-str_subset(colnames(dfPA), "peaks_in"))

```

```{r}

list_to_keep = c("start_time", "directory_output", "color_code",
                 "count_df", "dfPA", "df_gr_union_annot", "gr_ann_list")
rm(list = setdiff(ls(), list_to_keep))

```


<br><br><br>



# Differential accessibility - All samples comparison

For this part of the analysis, we use the number of reads detected in merged donors (one element per time point).  

## Perform DataNormalization with DEseq

https://www.rdocumentation.org/packages/DESeq2/versions/1.12.3/topics/DESeqDataSet-class    
https://tobiasrausch.com/courses/atac/atac-seq-data-analysis.html#data-normalization    
```{r}

# Extract the matrix with number of reads in each merged donors
cols_to_keep = str_subset(string = colnames(count_df), pattern = "D[:digit:].D[:digit:]_downsampled.bam")
counts = count_df[ , cols_to_keep]

## Extract information in the colnames
coldata = data.frame(file = cols_to_keep, name = unlist(str_extract_all(cols_to_keep, pattern = ".+(?=_downsampled.bam)"))) %>%
  dplyr::mutate(donors = str_extract(name, pattern = "D[:digit:].+")) %>%
  dplyr::mutate(condtime = paste0(str_extract(name, pattern = ".+(?=h)"), "h")) %>% 
  tidyr::separate(col = condtime, into = c("condition", "time"), remove = TRUE) 
coldata = coldata %>%
  tidyr::unite(condition_time, condition, time, sep = "_", remove = FALSE) %>%
  remove_rownames %>%
  column_to_rownames(var = "file") %>%
  dplyr::select(-name)
  
# Data normalization with DEseq
dds = DESeqDataSetFromMatrix(
  countData = counts, 
  colData = coldata, 
  design = ~ condition)
dds = DESeq(dds)
saveRDS(dds, paste0(directory_output, "DEseq_readcount_union_allsamples.rds"))

# Organize DEseq results in a table
cm = data.frame(counts(dds, normalized=TRUE)) %>%
  tibble::rownames_to_column("region") %>%
  tidyr::separate(region, c("seqnames", "start", "end"), remove = FALSE) %>%
  column_to_rownames("region")
colnames(cm) = str_replace_all(colnames(cm), pattern = "X2DG", replacement = "2DG")
write.csv2(x = cm, row.names = FALSE,
           file = paste0(directory_output, "DEseq_readcount_df_union_allsamples.csv"))

# Visualize the table 
cm[1:20, ] %>%
  kable(caption = "First 20 rows of the 'cm' table") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "250px")

```


## Principal Component Analysis (PCA)

https://tobiasrausch.com/courses/atac/atac-seq-data-analysis.html#principal-component-analysis-pca  

<br>

**Choice of PC for the plot**  
We want to select the PCs that explain the most the variance.   
```{r}

pca = prcomp(t(cm[,-c(1:3)]))
print(summary(pca))
pcaData = as.data.frame(pca$x)
pcaData$sample=rownames(pcaData)
coldata = coldata %>% rownames_to_column(var = "sample")
pcaData=merge(pcaData, coldata)
percentVar = round(100*(pca$sdev^2/sum(pca$sdev^2)))

varexp = data.frame(x=1:length(percentVar), y=percentVar)
varexp$x = factor(varexp$x)
ggplot(data=varexp, aes(x=x, y=y)) + 
  geom_bar(stat="identity") + 
  xlab("Principal Component") + 
  ylab("Proportion of variation (%)")

```
We choose PC1 and PC2.

<br>

**Plot the PCA**
```{r}
 
plot_pca = ggplot(data=pcaData, aes(x = PC1, y = PC2)) +
  geom_point(aes(colour = condition, shape = time), size=6) +
  scale_color_manual(values = color_code, limits = force) +
  scale_shape_manual(values=c(15, 16, 17, 7)) +
  ggtitle("PCA on normalized read counts from union of all samples") +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_text(vjust = 1 ,size = 14),
        axis.text.x = element_text(size = 11, colour = "black"),
        axis.ticks = element_blank()) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) 
plot_pca

ggsave(plot = plot_pca,
       filename = paste0(directory_output, "ATAC_pca_allsamples.png"),
       width = 16*0.75, height = 9*0.75)

```

Comment : Time point 00h, 03h and 12h are separated by peaks from PC1. For those time points, the different conditions are really close to each other. The conditions start to differ at 24h, they are separated on the plot by peaks from PC2 mostly but also PC1.

<br>

**Loadings from each PCs**  
Which peaks contribute the most to the separation of the samples ?  
```{r}

# Peaks contribution for each PC
loadings = abs(pca$rotation)
contribution = as.data.frame(sweep(loadings, 2, colSums(loadings), "/"))

# 500 peaks with the highest contribution for PC1
pc1_500peaks = contribution %>% dplyr::arrange(desc(PC1))
pc1_500peaks = rownames(pc1_500peaks[1:500,])

# 500 peaks with the highest contribution for PC2
pc2_500peaks = contribution %>% dplyr::arrange(desc(PC2))
pc2_500peaks = rownames(pc2_500peaks[1:500,])

# list of interesting peaks for all samples
pc1_pc2_peaks = intersect(pc1_500peaks, pc2_500peaks)
pc1_only_peaks = setdiff(pc1_500peaks, pc2_500peaks)
pc2_only_peaks = setdiff(pc1_500peaks, pc2_500peaks)

```

<br>

## Peaks investigation 

**Heatmap on the 500 peaks with the highest loading for PC1 and PC2**
```{r, fig.height=16, fig.width = 8, results="hold"}

# Prepare dataframe for heatmap
cm_heatmap = cm[, -c(1:3)]
colnames(cm_heatmap) = str_replace_all(colnames(cm_heatmap), "(.D[:digit:])+_downsampled.bam", "")
cm_heatmap = cm_heatmap %>%
  dplyr::select(Xvivo_00h, MP_03h, MP_12h, MP_24h, DON_03h,
                DON_12h, DON_24h, '2DG_03h', '2DG_12h', '2DG_24h',
                AOA_03h, AOA_12h, AOA_24h, aK_24h, VPA_24h)
# heatmap on peaks from PC1
mat_pc1 = cm_heatmap %>% dplyr::filter(row.names(cm_heatmap) %in% pc1_only_peaks)
mat_pc1 = mat_pc1 - rowMeans(mat_pc1)
pheatmap(mat_pc1, scale="row", cluster_rows = FALSE, cluster_cols = FALSE, fontsize_row = 5)

# heatmap on peaks from PC2
mat_pc2 = cm_heatmap %>% dplyr::filter(row.names(cm_heatmap) %in% pc2_only_peaks)
mat_pc2 = mat_pc2 - rowMeans(mat_pc2)
pheatmap(mat_pc2, scale="row", cluster_rows = FALSE, cluster_cols = FALSE, fontsize_row = 5)

# heatmap on peaks from PC1 and PC2
mat_pc1_pc2 = cm_heatmap %>% dplyr::filter(row.names(cm_heatmap) %in% pc1_pc2_peaks)
mat_pc1_pc2 = mat_pc1_pc2 - rowMeans(mat_pc1_pc2)
pheatmap(mat_pc1_pc2, scale="row", cluster_rows = FALSE, cluster_cols = FALSE, fontsize_row = 5)

```

<br>

**Pathway analysis and GO enrichment analysis on those genes** 
```{r, fig.width = 20, fig.height = 8}

# Extract interesting peaks
dfP = dfPA
rownames(dfP) = paste0(dfP$seqnames, '_', dfP$start, '_', dfP$end)
selpeaks_pc1 = dfP[rownames(dfP) %in% pc1_only_peaks,]
selpeaks_pc1 = selpeaks_pc1[abs(selpeaks_pc1$distanceToTSS) < 5000,]$geneId
selpeaks_pc2 = dfP[rownames(dfP) %in% pc2_only_peaks,]
selpeaks_pc2 = selpeaks_pc2[abs(selpeaks_pc2$distanceToTSS) < 5000,]$geneId
selpeaks_pc1pc2 = dfP[rownames(dfP) %in% pc1_pc2_peaks,]
selpeaks_pc1pc2 = selpeaks_pc1pc2[abs(selpeaks_pc1pc2$distanceToTSS) < 5000,]$geneId

# Perform pathway  enrichment analysis
pathway_pc1 = enrichPathway(selpeaks_pc1)
pathway_pc2 = enrichPathway(selpeaks_pc2)
pathway_pc1pc2 = enrichPathway(selpeaks_pc1pc2)

# Perform GO analysis
ego_pc1 = enrichGO(gene = selpeaks_pc1, 
                   OrgDb = org.Hs.eg.db,
                   ont = "ALL", 
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.2,
                   readable = TRUE,
                   keyType = "ENTREZID")
ego_pc2 = enrichGO(gene = selpeaks_pc2, 
                   OrgDb = org.Hs.eg.db,
                   ont = "ALL", 
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.2,
                   readable = TRUE,
                   keyType = "ENTREZID")
ego_pc1pc2 = enrichGO(gene = selpeaks_pc1pc2, 
                   OrgDb = org.Hs.eg.db,
                   ont = "ALL", 
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.2,
                   readable = TRUE,
                   keyType = "ENTREZID")

dotplot(ego_pc1, showCategory = 25) + 
  facet_grid(ONTOLOGY ~ ., scales="free") +
  ggtitle("Gene associated with peaks from PC1") |
  dotplot(ego_pc2, showCategory = 25) + 
  facet_grid(ONTOLOGY ~ ., scales="free") +
  ggtitle("Gene associated with peaks from PC2") |
  dotplot(ego_pc1pc2, showCategory = 25) + 
  facet_grid(ONTOLOGY ~ ., scales="free") +
  ggtitle("Gene associated with peaks from PC1 and PC2")

```

No pathway enrichment for any list tested (peaks from PC1 only, peaks from PC2 only and peaks from PC1 and PC2).  
Some GO enrichment but not really interesting. 

```{r}

list_to_keep = c("start_time", "directory_output", "color_code",
                 "count_df", "dfPA", "df_gr_union_annot", "gr_ann_list")
rm(list = setdiff(ls(), list_to_keep))

```

<br><br><br>



# Differential accessibility - Dual comparison

## Choose comparisons

```{r}

combination = list(c("MP_03h", "MP_12h"), c("MP_12h", "MP_24h"),
                   c("DON_03h", "DON_12h"), c("DON_12h", "DON_24h"),
                   c("2DG_03h", "2DG_12h"),c("2DG_12h", "2DG_24h"),
                   c("AOA_03h", "AOA_12h"),c("AOA_12h", "AOA_24h"),
                   c("MP_03h","DON_03h"),c("MP_12h","DON_12h"), c("MP_24h","DON_24h"), 
                   c("MP_03h","2DG_03h"),c("MP_12h","2DG_12h"), c("MP_24h","2DG_24h"),
                   c("MP_03h","AOA_03h"),c("MP_12h","AOA_12h"), c("MP_24h","AOA_24h"),
                   c("MP_24h", "VPA_24h"), c("MP_03h", "Xvivo_00h"))

```


## Visualize peaks evolution

Note : we work with the list of regions from the union of peaks of all samples. Some peaks from individual samples were merged during this operation. So when we count the number of peaks for each sample among this list of regions, it's normal that we don't find the same number as the initial number of peaks per sample.   
https://stackoverflow.com/questions/27197617/filter-data-frame-by-character-column-name-in-dplyr

```{r}

df_type_peak = data.frame()

for(i in 1:length(combination)) {
  
  # Extract conditions information
  cond1 = combination[[i]][1]
  cond2 = combination[[i]][2] 
  name = paste0(cond1, "_vs_", cond2)
  col_cond1 = str_subset(string = colnames(count_df), pattern = paste0("peaks_in_", cond1))
  col_cond2 = str_subset(string = colnames(count_df), pattern = paste0("peaks_in_", cond2))

  # Classify peaks in category
  temp = data.frame(
    differential = name,
    nb_peaks_cond1_in_union = nrow(count_df %>% dplyr::filter(!!as.symbol(col_cond1) == TRUE)), 
    nb_peaks_cond2_in_union = nrow(count_df %>% dplyr::filter(!!as.symbol(col_cond2) == TRUE)),
    persistant_peaks_cond1_cond2 = nrow(count_df %>% dplyr::filter(!!as.symbol(col_cond1) == TRUE & 
                                                            !!as.symbol(col_cond2) == TRUE)),
    close_open_peaks_cond1_cond2 = nrow(count_df %>% dplyr::filter(!!as.symbol(col_cond1) == FALSE & 
                                                            !!as.symbol(col_cond2) == TRUE)),
    open_close_peaks_cond1_cond2 = nrow(count_df %>% dplyr::filter(!!as.symbol(col_cond1) == TRUE & 
                                                            !!as.symbol(col_cond2) == FALSE))
    )
  df_type_peak = rbind(df_type_peak, temp)
  
}

write.csv2(x = df_type_peak, row.names = FALSE,
           file = paste0(directory_output, "peaks_close_open_conditions.csv"))

```


## Volcano plot

Note : for DEseq2 comparison, we need duplicate samples. We work with the individual donors files to perform it. 
Are the peaks bigger (more reads) or smaller between the two conditions ?  
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pvaluesNA  
https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#indfilttheory  
"If a row is filtered by automatic independent filtering, for having a low mean normalized count, then only the adjusted p value will be set to NA." "The goal of independent filtering is to filter out those tests from the procedure that have no, or little chance of showing significant evidence, without even looking at their test statistic."  
Based on this explanation, we remove peaks with NA values in padj column.  

```{r, Define a function}

volcano_DEseq_fun = function(comb, tab_count, tab_annot) {
  
  # Extract conditions information
  cond1 = comb[1]
  cond2 = comb[2] 
  name = paste0(cond1, "_vs_", cond2)
  col_peak_cond1 = str_subset(string = colnames(tab_count), pattern = paste0("peaks_in_", cond1))
  col_peak_cond2 = str_subset(string = colnames(tab_count), pattern = paste0("peaks_in_", cond2))
  col_read_cond1 = str_subset(string = colnames(tab_count), pattern = paste0(cond1, "_D[:digit:]_downsampled.bam"))
  col_read_cond2 = str_subset(string = colnames(tab_count), pattern = paste0(cond2, "_D[:digit:]_downsampled.bam"))
  
  # Extract the read counts for the two conditions
  matrix_peaks = tab_count
  ## put region number as rownames
  rownames(matrix_peaks) = matrix_peaks$region
  ## keep only the regions that correspond to peaks in both conditions
  matrix_peaks = matrix_peaks %>% 
    dplyr::filter(!!as.symbol(col_peak_cond1) == TRUE & !!as.symbol(col_peak_cond2) == TRUE) 
  ## keep only the reads from the two conditions
  matrix_peaks = matrix_peaks[colnames(matrix_peaks) %in% c(col_read_cond1, col_read_cond2)]
  
  # DEseq2 parameters
  coldata <- data.frame(sample = colnames(matrix_peaks)) %>%
    dplyr::mutate(condition = ifelse(str_detect(sample, pattern = cond1), "before", "after")) %>%
    dplyr::mutate(type = "paired-end")
  
  # Normalize data with DEseq
  dds <- DESeqDataSetFromMatrix(
    countData = matrix_peaks,
    colData = coldata,
    design = ~ condition)
  dds$condition <- relevel(dds$condition, ref = "before") 
  dds <- DESeq(dds)
  
  # Extract matrix count from deseq
  cm = data.frame(counts(dds, normalized=TRUE))
  
  # Extract statistics from deseq
  res = as_tibble(results(dds)) 
  res = res %>% 
    dplyr::mutate(region = rownames(cm), .before = baseMean) %>%
    dplyr::mutate(regulation = case_when(pvalue < 0.01 ~ "significative", 
                                         pvalue > 0.01 ~ "non-significative")) %>%
    dplyr::filter(!is.na(padj))
  
  # Add genomic annotations for the regions
  DEseq_results_annotated = right_join(df_gr_union_annot, res, by = "region")
  start = grep("baseMean", colnames(DEseq_results_annotated))
  end = grep("padj", colnames(DEseq_results_annotated))
  DEseq_results_annotated[,start:end] = lapply(DEseq_results_annotated[,start:end], as.numeric)
  
  # Pivot table to draw the plot
  DEseq_results_long = DEseq_results_annotated %>% 
    pivot_longer(cols = str_subset(colnames(DEseq_results_annotated), "hg19"), 
                 names_to = "feature", 
                 values_to = "feature_overlap") 
  
  # Plot volcano
  df_plot = DEseq_results_long %>% 
    dplyr::filter(feature == "hg19_genes_promoters" & feature_overlap == TRUE | 
                  feature == "hg19_genes_intergenic" & feature_overlap == TRUE |
                  feature == "hg19_genes_introns" & feature_overlap == TRUE |
                  feature == "hg19_enhancers_fantom" & feature_overlap == TRUE)
  
  plot = ggplot(df_plot, aes(x = log2FoldChange, y = -1 * log10(pvalue), colour = feature)) +
    geom_point(size = 3, alpha = 0.5, fill = NA, shape = 21, stroke = 2) +
    geom_hline(aes(yintercept = 2), colour = "red", linetype = "dashed") +
    geom_text(label = "p-value = 0.01", colour = "red", aes(x = -5, y = 3), size = 3) +
    coord_cartesian(xlim = c(-7, 7), ylim = c(NA, 10)) + 
    labs(subtitle = name, x = "log2(FoldChange)", y = "-log10(Pvalue)") +
    facet_wrap(facets = .~feature) +
    theme(axis.line.y = element_line(color = "black"),
          axis.line.x = element_line(color = "black"),
          legend.position = "none",
          axis.title = element_text(size = 10),
          plot.subtitle = element_text(size = 14, hjust = 0.5))

  ## save plot 
  ggsave(plot, width = 8, height = 8,
         filename = paste0(directory_output, name, "_volcano.svg"))
  
  # Output function
  return(list(volcano_plot = plot,
              deseq_res = res,
              matrix_count = cm))
  
}

```

```{r, Apply function on each combination}

list_deseq_volcano = list()
for(i in 1:length(combination)) {
  list_name = paste0(combination[[i]][1], "_vs_", combination[[i]][2])
  list_deseq_volcano[[list_name]] = volcano_DEseq_fun(comb = combination[[i]], 
                                                      tab_count = count_df, 
                                                      tab_annot = df_gr_union_annot)
}

```

```{r, Show volcano plots, fig.width = 37, fig.height = 30}

list_volcano = unlist(list_deseq_volcano, recursive=FALSE)
list_volcano = list_volcano[str_detect(names(list_volcano), "volcano_plot")]
do.call("grid.arrange", c(list_volcano, ncol = 4))

```


## Pathway enrichment analysis

We extracted up regulated and down regulated peaks between two conditions and we perform enrichment pathway analysis.
```{r, Show pathway enrichment, fig.width = 8, fig.height=12}

for(i in 1:length(list_deseq_volcano)) {
  
  print(i)
  tp = list_deseq_volcano[[i]]
  tp_res = tp$deseq_res
  tp_cm = tp$matrix_count
  comp = names(list_deseq_volcano)[i]
  
  # Up regulated peaks in the first condition
  region_up = rownames(tp_cm[which(tp_res$padj < 0.1 & tp_res$log2FoldChange>0),])
  selpeaks_up = dfPA %>% dplyr::filter(region %in% region_up)
  selpeaks_up = selpeaks_up[abs(selpeaks_up$distanceToTSS) < 5000,]$geneId
  if(length(selpeaks_up) != 0) {
    pathwayUp = enrichPathway(selpeaks_up)
    tab_u = data.frame(pathwayUp)
    if(nrow(tab_u) != 0) {tab_u = tab_u  %>% dplyr::mutate(Peak_Category = "Up_regulated", .before = "ID")}
    ego_up = enrichGO(gene = selpeaks_up, 
                   OrgDb = org.Hs.eg.db,
                   ont = "ALL", 
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.2,
                   readable = TRUE,
                   keyType = "ENTREZID")

  } else {
    tab_u = data.frame()
  }
  
  # Down regulated peaks in the first condition
  region_down = rownames(tp_cm[which(tp_res$padj < 0.1 & tp_res$log2FoldChange<0),])
  selpeaks_down = dfPA %>% dplyr::filter(region %in% region_down)
  selpeaks_down = selpeaks_down[abs(selpeaks_down$distanceToTSS) < 5000,]$geneId
  if(length(selpeaks_down) != 0) {
    pathwayDown = enrichPathway(selpeaks_down)
    tab_d = data.frame(pathwayDown)
    if(nrow(tab_d) != 0) {tab_d = tab_d  %>% dplyr::mutate(Peak_Category = "Down_regulated", .before = "ID")}
    ego_down = enrichGO(gene = selpeaks_down, 
                   OrgDb = org.Hs.eg.db,
                   ont = "ALL", 
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.2,
                   readable = TRUE,
                   keyType = "ENTREZID")
  } else {
    tab_d = data.frame()
  }
  
  # Results pathway analysis
  tab = rbind(tab_u, tab_d)
  
  if (nrow(tab) != 0) {
    
    write.csv2(tab, paste0(directory_output, comp, "_pathway_UpDown.csv"))
    
    if (nrow(tab_d != 0)) {
      plot_d = dotplot(pathwayDown) + 
        ggtitle("Down_regulated") + 
        plot_annotation(title = comp)
      ggsave(plot_d, height = 6, width = 14,
             filename = paste0(directory_output, comp, "_dotplot_pathway_Down.svg"))
      print(plot_d)
    }
  
    if (nrow(tab_u != 0)) {
      plot_u = dotplot(pathwayUp) + 
        ggtitle("Up_regulated") + 
        plot_annotation(title = comp)
      ggsave(plot_u, height = 6, width = 14,
             filename = paste0(directory_output, comp, "_dotplot_pathway_Up.svg"))
      print(plot_u)
    }
    
  } else { print("No pathway enrichment in up and down regulated peaks") }
    
  if (exists("ego_up") | exists("ego_down")) {
    
    if (exists("ego_up")) {
      if (!is.null(ego_up)) {
        if (nrow(ego_up@result) != 0) {
          dot_up = dotplot(ego_up, showCategory = 25) + 
          facet_grid(ONTOLOGY ~ ., scales="free") +
          ggtitle(paste0(comp, "Gene associated to peaks upregulated in cond1 compared to cond2"))
        print(dot_up)
        ggsave(dot_up, height = 15, width = 8,
               filename = paste0(directory_output, comp, "_dotplot_GO_Up.svg"))
        }
      }
    }
    
    if (exists("ego_down")) {
      if (!is.null(ego_down)) {
        if (nrow(ego_down@result) != 0) {
        dot_down = dotplot(ego_down, showCategory = 25) + 
          facet_grid(ONTOLOGY ~ ., scales="free") +
          ggtitle(paste0(comp, "Gene associated to peaks downregulated in cond1 compared to cond2"))
        ggsave(dot_down, height = 15, width = 8,
               filename = paste0(directory_output, comp, "_dotplot_GO_Down.svg"))
        }
      }
    }
    
  } else { print("No GO enrichment in up and down regulated peaks") }

  rm(tab_u, tab_d, tab, ego_up, ego_down)
  
}

```

```{r}

list_to_keep = c("start_time", "directory_output", "color_code",
                 "count_df", "gr_ann_list", "df_gr_union_annot")
rm(list = setdiff(ls(), list_to_keep))

```



<br><br><br>


# Visualize peaks dynamics 

```{r}

peaks_dynamics_fun = function(df_readcount, df_peaks_in, choice_annot, condition) {
  
  if (!choice_annot %in% c(str_subset(colnames(df_gr_union_annot), pattern = "hg19"), "all")) {
    
    print("error in choice_annot")
    
  } else {
    
    # Select columns from the condition and keep only merge donors info
    df_readcount = df_readcount %>%
      dplyr::select(region, seqnames, start, end, 
                    str_subset(colnames(df_readcount), pattern = condition),
                    - str_subset(colnames(df_readcount), pattern = "[:alnum:]+_[:digit:]{2}h_D[:digit:]_downsampled.bam"))
    
    # Keep only peaks corresponding to the wanted genomic annotation
    df_readcount = left_join(df_readcount, df_peaks_in, by = c("region", "seqnames", "start", "end"))
    if (!choice_annot == "all") { 
      df_readcount = df_readcount %>% dplyr::filter(!!as.symbol(choice_annot) == TRUE) }
    df_readcount = df_readcount %>% 
      dplyr::select(-str_subset(colnames(df_readcount), "hg19"), 
                    -width, -strand, -seqnames, -start, -end)
    
    # Extract combination
    cols <- sapply(df_readcount, is.logical)
    df_readcount[,cols] <- lapply(df_readcount[,cols], as.numeric)
    df_readcount = df_readcount %>% 
      tidyr::unite(str_subset(colnames(df_readcount), "peaks_in"), 
                   col = "peaks_evolution", remove = TRUE, sep = "_")
    df_readcount = df_readcount %>% 
      dplyr::mutate(peaks_category = case_when(
        peaks_evolution %in% c("1_1_0", "1_0_0") ~ "closing_peaks",
        peaks_evolution %in% c("0_0_1", "0_1_1") ~ "opening_peaks",
        peaks_evolution %in% c("1_1_1", "0_0_0") ~ "stable_peaks",
        peaks_evolution %in% c("0_1_0", "1_0_1") ~ "others")) %>%
      pivot_longer(cols = str_subset(colnames(df_readcount), pattern = "downsampled"),
                   names_to = "condition",
                   values_to = "nbreads") %>% 
      tidyr::separate(condition, c("condition", "time", "donors", "temp"), 
                      sep = "_", remove = TRUE) %>% 
      dplyr::select(-temp, -donors)
    df_readcount = df_readcount %>%
      dplyr::group_by(peaks_evolution, time) %>%
      dplyr::mutate(mean_nbreads = mean(nbreads)) %>%
      dplyr::ungroup()
    
    # Plot the peaks dynamics
    plot = ggplot(df_readcount, aes(x = time, y = nbreads, group = region)) +
      geom_line(alpha = 0.2, col = color_code[condition]) +
      geom_point(col = color_code[condition]) +
      geom_line(aes(x = time, y = mean_nbreads), colour = "red", linetype = "dashed") +
      ggtitle(paste0(condition, "-", choice_annot)) +
      facet_wrap(facets = peaks_category~peaks_evolution, scales = "free") 
    
    # Save plot
    ggsave(plot, width = 12, height = 12,
           file = paste0(directory_output, condition, "_", choice_annot, "_peaks_dyn.png"))
       
    # Function's output 
    return(list(df_peaks_dyn = df_readcount,
                plot_dyn = plot))
    
  }}

```

```{r}

# Keep only conditions with multiple time points
cond_chrono = data.frame(temp = str_extract(names(gr_ann_list), pattern = "[:alnum:]+_[:digit:]{2}h"))
cond_chrono = tidyr::separate(cond_chrono, temp, into = c("cond", "time"), remove = TRUE)
cond_chrono = cond_chrono %>% group_by(cond) %>% dplyr::filter(n() > 1) %>% dplyr::pull(cond)
cond_chrono = unique(cond_chrono)

# Call the function for each condition
list_dyn_peaks = list()

for(i in 1:length(cond_chrono)){
  
  list_dyn_peaks[[cond_chrono[i]]]$prom = peaks_dynamics_fun(
    df_readcount = count_df,
    df_peaks_in =  df_gr_union_annot,
    choice_annot = "hg19_genes_promoters",
    condition = cond_chrono[i]
  )
  
   list_dyn_peaks[[cond_chrono[i]]]$intergenic = peaks_dynamics_fun(
    df_readcount = count_df,
    df_peaks_in =  df_gr_union_annot,
    choice_annot = "hg19_genes_intergenic",
    condition = cond_chrono[i]
  )
   
  list_dyn_peaks[[cond_chrono[i]]]$all = peaks_dynamics_fun(
    df_readcount = count_df,
    df_peaks_in =  df_gr_union_annot,
    choice_annot = "all",
    condition = cond_chrono[i]
  )
  
}

```

Here, we present one peaks dynamic plot. To see all the plots, refer to the exp folder.

```{r, fig.width = 20, fig.height = 20}
 
plots =  unlist(unlist(list_dyn_peaks,recursive=FALSE),recursive=FALSE)
plots = plots[str_detect(names(plots), "plot")]
plots[[1]]

```



<br><br><br>



```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric(end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

