---
title: "bulkATACseq_R_GenomeDistribution_GeneCoverage"
author: "Laëtitia Racine"
date: "2023-03-24"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>


```{r, Setup, include=F}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(knitr)
library(stringr)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(karyoploteR)
library(Gviz)
library(ggplotify)
library(gridExtra)

```

```{r, Working directory}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/bulkATACseq_R_GenomeDistribution_GeneCoverage/"))
dir.create(path = paste0(directory, "exp/bulkATACseq_R_GenomeDistribution_GeneCoverage/", current_date))
directory_output = paste0(directory, "exp/bulkATACseq_R_GenomeDistribution_GeneCoverage/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

# Load grange objects 
dir = pic_last_dir(paste0(directory, "exp/bulkATACseq_R_ChangeAnnotation/"))
gr_ann_list = list.files(path = paste0(dir, "/"), pattern = "update_annot.gr.rds")
name_list = str_extract(gr_ann_list, pattern = "[:graph:]+(?=_threshold_10_update_annot.gr.rds)")
gr_ann_list = paste0(dir, "/", gr_ann_list)
gr_ann_list = lapply(gr_ann_list, readRDS)
names(gr_ann_list) = name_list

# Extract bam file list 
dir_data = paste0(directory, "data/bulkATACseq/downsampled_bam_merged/")
bam_list = list.files(path = dir_data, pattern = "_downsampled.bam$")
bam_list = paste0(dir_data, bam_list)

# Extract promoter localisation from hg19 annotation gr (see annotationsFiles_createGrange.Rmd)
dir = pic_last_dir(paste0(directory, "exp/AnnotationsFiles_createGrange/"))
annot_hg19 = readRDS(paste0(dir, "/", "hg19_annotations_gr.rds"))
prom_hg19 = annot_hg19[annot_hg19$type == "hg19_genes_promoters"]
prom_hg19 = prom_hg19[!is.na(prom_hg19$symbol)]

# Select list of interesting genes to study
list_gene = unique(c(list_genes_metabo,
                     list_genes_metabo_glut,
                     list_genes_metabo_glycolysis,
                     list_genes_metabo_tca,
                     list_genes_TF_hemato))

```



<br><br><br>



# Peaks distribution across chromosomes

Using karyoplotR package :   
https://bernatgel.github.io/karyoploter_tutorial/  
https://github.com/bernatgel/karyoploteR/issues/51  

<br>

## One plot per sample with all chromosomes

```{r, fig.show = "hide"}

list_sample_chr = list()

for( i in 1:length(gr_ann_list)) {
  
    gr = gr_ann_list[[i]]
    name_sample = str_extract(string = names(gr_ann_list)[i], 
                            pattern = "[:alnum:]+_[:digit:]{2}h")
    name_cond = str_extract(string = names(gr_ann_list)[i], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
    
    # visualize on html file
    list_sample_chr[[i]] = as.ggplot(expression(
      plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
        kpAddMainTitle(main = paste0("Peaks distribution on hg19 genome from ", name_sample), cex=1.5) %>%
        kpPlotDensity(data=gr, data.panel = 2)
      ))
  
   # save the plot
   png(filename = paste0(directory_output, "plot_allchr_", name_sample, ".png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
     kpPlotRegions(data=gr, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
     kpAddMainTitle(main = paste0("Peaks distribution on hg19 genome from ", name_sample), cex=1.5) %>%
     kpPlotDensity(data=gr, data.panel = 2)
  dev.off()
  
}

```

```{r, fig.width = 30, fig.height = 65}

do.call("grid.arrange", c(list_sample_chr, ncol = 3))

```

<br>

## One plot per chromosome with all samples

```{r, fig.width = 16, fig.height = 10}

chromosomes = c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", 
                "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", 
                "chr16", "chr17", "chr18", "chr19", "chr20", "chr21", "chr22", 
                "chrX", "chrY")

# Display settings: https://bernatgel.github.io/karyoploter_tutorial//Tutorial/PlotParams/PlotParams.html 
pp = getDefaultPlotParams(plot.type=1)
pp$data1height = 10
pp$ideogramheight = 6
pp$bottommargin = 6
pp$data1inmargin = 6

for (i in 1:length(chromosomes)) { 

  # visualize on html file
    kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
    for (j in 1:length(gr_ann_list)) {
      name_cond = str_extract(string = names(gr_ann_list)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
      name_sample = str_extract(string = names(gr_ann_list)[j], 
                                pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_ann_list[[j]], border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_ann_list[[j]], data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6)
    }
  
  # save plot
  png(paste0(directory_output, "plot_allsamples_",chromosomes[i], ".png"), 
      width = 2000, height = 1000)
  kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
  for (j in 1:length(gr_ann_list)) {
    name_cond = str_extract(string = names(gr_ann_list)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
    name_sample = str_extract(string = names(gr_ann_list)[j], 
                            pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_ann_list[[j]], border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_ann_list[[j]], data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6) 
    }
  dev.off()
  
}

```

```{r}

list_to_keep = c("start_time", "directory_output", "gr_ann_list", "bam_list", 
                 "prom_hg19", "color_code", "list_gene", "directory")
rm(list = setdiff(ls(), list_to_keep))

```



<br><br><br>



# Gene coverage 

https://www.bioconductor.org/packages/devel/bioc/vignettes/Gviz/inst/doc/Gviz.html   
http://www.sthda.com/english/wiki/visualize-ngs-data-with-r-and-bioconductor  
Note !!! Need .bai file in the folder with the .bam files  

<br>

## Extract genomic location for each studied gene = df_region_genes

```{r}

# Extract and organize gene information from genesymbol via Gviz
data(genesymbol, package = "biovizBase")
df_region_genes = tibble()

for(i in 1:length(list_gene)) {
  region = genesymbol[list_gene[i]]
  region = range(region, ignore.strand = TRUE)
  region = region + 5000
  region = keepStandardChromosomes(region)
  df_region_genes[i,"gene"] = list_gene[i]
  df_region_genes[i,"chr"] = as.character(seqnames(region))
  df_region_genes[i,"start"] = start(region)
  df_region_genes[i,"end"] = end(region)
}

# Save output
write.csv2(x = df_region_genes, row.names = FALSE,
            file = paste0(directory_output, "ranges_genes_interest.csv"))

# Visualize output
df_region_genes %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "250px")

```

<br>

## Create *isoform and promotor track* for each studied gene (ensGenes + promoterTrack)

```{r}

# # Define function to create the gene track with isoform and promoters
# iso_prom_tracking_fun = function(df_region, gene_index, prom_grange) {
# 
#   # Isoform tracking for the studied gene
#   # *************************************
# 
#   ensGenes <- UcscTrack(genome = "hg19",
#                         chromosome = df_region$chr[gene_index],
#                         background.title = "transparent",
#                         alpha.title = 1,
#                         col.title = "black",
#                         cex.title = 1.5,
#                         track = "ensGene",
#                         from = df_region$start[gene_index],
#                         to = df_region$end[gene_index],
#                         trackType = "GeneRegionTrack",
#                         rstarts = "exonStarts",
#                         rends = "exonEnds",
#                         transcript = "name",
#                         strand = "strand",
#                         fill = "lightgrey",
#                         name = df_region$gene[gene_index])
# 
#   # Promoters tracking for the studied gene
#   # *************************************
# 
#   # Extract ranges of the gene promoters
#   prom_peak = subsetByOverlaps(prom_grange,
#                                GRanges(seqnames = df_region$chr[gene_index],
#                                        ranges = IRanges(start = df_region$start[gene_index],
#                                                         end = df_region$end[gene_index])))
#   # Store promoters limits into a dataframe
#   df_prom_retained = tibble("start" = start(prom_peak),
#                             "end" = end(prom_peak),
#                             "chr" = as.character(seqnames(prom_peak)))
# 
#   # if we found promoters for the gene in our list, we add them on the track
#   if(nrow(df_prom_retained) != 0){
# 
#     # Add promoters on the isoform tracking
#     promoterTrack = HighlightTrack(trackList = ensGenes,
#                                    start = df_prom_retained$start,
#                                    end = df_prom_retained$end,
#                                    chromosome = df_region$chr[gene_index],
#                                    col = "green", fill = "green")
#     return(promoterTrack)
# 
#   # if we didn't find promoters for the gene in our list, we keep only the isoform track
#   } else {
# 
#     print("No promoter found")
#     return(ensGenes)
#   }
# }
# 
# # Call function for each gene of the studied list
# list_iso_prom_track = list()
# for (i in 1:length(list_gene)) {
#   list_iso_prom_track[i] = list(
#     iso_prom_tracking_fun(df_region = df_region_genes,
#                           gene_index = i,
#                           prom_grange = prom_hg19)
#     )
# }
# names(list_iso_prom_track) = list_gene
# 
# # Save the list of tracks
# saveRDS(list_iso_prom_track, file = paste0(directory_output, "list_iso_prom_track.rds"))

## really hard to reach the database sometimes
## we saved the object in the data folder for further analysis
list_iso_prom_track = readRDS(paste0(directory, "data/bulkATACseq/list_iso_prom_track.rds"))

```

<br>

## Create *chromosome track*, axis scale and blank space

```{r}

list_chr_track = list()
chr_list = unique(df_region_genes$chr)

for (i in 1:length(chr_list)){
list_chr_track[[chr_list[i]]] = IdeogramTrack(genome="hg19", chromosome= chr_list[i]) 
}
  
saveRDS(list_chr_track, paste0(directory_output, "list_chr_track.rds"))

```

```{r}

axTrack <- GenomeAxisTrack()
blank_space = GenomeAxisTrack(col = "white",fill = "white", fontcolor = "white")

```

<br>

## Create *peaks track* for each studied gene for each sample

```{r}

# Function to draw the track of a sample for a specific gene
track_sample_fun = function(var_name_sample, var_col, var_bam_path, var_grange_sample, var_ranges_gene) {
  
   # create track of fragments for the sample
    dt = DataTrack(type = "histogram",
                   name = var_name_sample,
                   background.title = "transparent",
                   fill.histogram = "#0072B2",
                   col.histogram = "#0072B2",
                   alpha.title = 1,
                   col.title = var_col,
                   col.axis = "black",
                   cex.title = 1,
                   range = var_bam_path,
                   genome = "hg19",
                   window = -1,
                   chromosome = var_ranges_gene$chr)
  
    # determine if peaks are detected in this gene for this sample
    peaks_in = subsetByOverlaps(var_grange_sample,
                                GRanges(seqnames = var_ranges_gene$chr,
                                        ranges = IRanges(start = var_ranges_gene$start, 
                                                         end = var_ranges_gene$end)))
    
    # if no peak detected, just the graph of the fragments for this sample for this gene
    if (length(peaks_in) == 0){  
      
      print(paste(var_name_sample, "no_peaks_detected"))
      return(dt)
      
    # if peaks detected, add highlighted peaks on the fragments track
      } else {  
        
      df_peaks_retained = tibble("start" = start(peaks_in),
                                 "end" = end(peaks_in),
                                 "chr" = as.character(seqnames(peaks_in)))
      
      ht = HighlightTrack(trackList = dt,
                          start = df_peaks_retained$start,
                          end = df_peaks_retained$end,
                          chromosome = var_ranges_gene$chr)
      return(ht)
      
      }

}


# Call function for sample and each gene of the studied list
list_track_samples = list()

for(i in 1:length(gr_ann_list)) { # track for each sample
  
  name_sample = names(gr_ann_list)[i]
  var_bam_path = ifelse(str_detect(name_sample, pattern="aK") == FALSE, name_sample, 
                        str_replace(name_sample, pattern ="CTRL", replacement =""))
  var_bam_path = str_subset(bam_list, pattern = var_bam_path)
    
  list_temp = list()
  
  for(j in 1:nrow(df_region_genes)){  # track for each gene

    track = track_sample_fun(var_name_sample = name_sample, 
                             var_col = color_code[str_extract(name_sample, 
                                          pattern ="[:alnum:]+(?=_[:digit:]{2}h)")],
                             var_bam_path = var_bam_path,
                             var_grange_sample = gr_ann_list[[name_sample]],
                             var_ranges_gene = df_region_genes[j,])
    list_temp[[df_region_genes[j,]$gene]] = track
    
  }
  
  list_track_samples[[name_sample]] = list_temp
  
}

saveRDS(list_track_samples, paste0(directory_output, "list_track_samples.rds"))

```

```{r}

list_to_keep = c("start_time", "directory_output", "gr_ann_list",
                 "list_gene", "list_chr_track", "list_iso_prom_track",
                 "list_track_samples", "blank_space", "axTrack",
                 "color_code", "df_region_genes")
rm(list=setdiff(ls(), list_to_keep))

```

<br>

## Plot the final graphs

A tracking plot is composed of (from top to bottom) :  
- one chromosome track (list_chr_track)  
- one axis scale (adxTrack)    
- one isoform/promoters track for the gene studied (list_iso_prom_track)    
- one or multiple samples tracks for the gene studied (list_track_samples)    
- blank space between element for clearness (blank_space)    
  
All those elements were already created in the code above. Here we will just assemble the graph as we desire.  

<br>

### One coverage plot with all time point of the same condition per gene

```{r}

conditions = unique(str_extract(string = names(gr_ann_list), 
                                pattern = "[:alnum:]+(?=_[:digit:]{2}h)"))

list_cov_genes_cond = list()

for(i in 1:length(conditions)){

  # extract tracks of the samples for the condition
  samples_track = list_track_samples[str_detect(names(list_track_samples), conditions[i])]
  samples_track = unlist(samples_track,recursive=FALSE)
  
  for(j in 1:length(list_gene)) {

    # extract gene_info
    gene_info = df_region_genes %>% dplyr::filter(gene == list_gene[j])
    # extract track of chromosome for the gene
    chr_track = list_chr_track[[str_subset(names(list_chr_track), paste0(gene_info$chr, "$"))]]
    # extract track of iso/prom for the gene
    isoprom_track = list_iso_prom_track[[str_subset(names(list_iso_prom_track), gene_info$gene)]]
    # extract list of samples track for the gene
    peaks_track = samples_track[str_detect(names(samples_track), gene_info$gene)]
    # assemble the plots 
    list_to_plot = list(chr_track, axTrack, blank_space, isoprom_track)
    for(k in 1:length(peaks_track)) { list_to_plot = c(list_to_plot, blank_space, peaks_track[k])}
    # visualize graph on html
    list_cov_genes_cond = append(x = list_cov_genes_cond, 
                            values = list(as.ggplot(expression(
                              plotTracks(trackList = list_to_plot, 
                                         from = gene_info$start, 
                                         to = gene_info$end, 
                                         showTitle = TRUE, 
                                         add = TRUE)))))
    # save graph 
    pdf(file =  paste0(directory_output, conditions[i], "_", 
                       gene_info$gene, "_coverage_plot_alltime.pdf"), 
        width = 24, height = 20)
    plotTracks(list_to_plot, from = gene_info$start, to = gene_info$end, showTitle = TRUE)
    dev.off()  
  }

}

```

Here, we present some gene coverage plots per condition. To see all the plots, refer to the exp folder.

```{r, fig.width = 37, fig.height = 35}

do.call("grid.arrange", c(list_cov_genes_cond[1:12], ncol = 3))

```

<br>

### One coverage plot with all conditions of the same time point

https://stackoverflow.com/questions/14748557/skipping-error-in-for-loop

```{r}

time = unique(str_extract(string = names(gr_ann_list), 
                          pattern = "[:digit:]{2}h"))

list_cov_genes_time = list()

for(i in 1:length(time)){

  # extract tracks of the samples for this time 
  samples_track = list_track_samples[str_detect(names(list_track_samples), time[i])]
  samples_track = unlist(samples_track,recursive=FALSE)
  
  for(j in 1:length(list_gene)) {
    
    skip_to_next <- FALSE
    # extract gene_info
    gene_info = df_region_genes %>% dplyr::filter(gene == list_gene[j])
    # extract track of chromosome for the gene
    chr_track = list_chr_track[[str_subset(names(list_chr_track), paste0(gene_info$chr, "$"))]]
    # extract track of iso/prom for the gene
    isoprom_track = list_iso_prom_track[[str_subset(names(list_iso_prom_track), gene_info$gene)]]
    # extract list of samples track for the gene
    peaks_track = samples_track[str_detect(names(samples_track), gene_info$gene)]
    # assemble the plots 
    list_to_plot = list(chr_track, axTrack, blank_space, isoprom_track)
    for(k in 1:length(peaks_track)) { list_to_plot = c(list_to_plot, blank_space, peaks_track[k])}
    # visualize graph
    tryCatch({
   list_cov_genes_time = append(x = list_cov_genes_time, 
                                 values = list(as.ggplot(expression(
                                   plotTracks(trackList = list_to_plot, 
                                              from = gene_info$start, 
                                              to = gene_info$end, 
                                              showTitle = TRUE, 
                                              add = TRUE)))))
    }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
    # save graph 
    pdf(file =  paste0(directory_output, time[i], "_", gene_info$gene, "_coverage_plot_allcond.pdf"), 
        width = 24, height = 40)
    plotTracks(list_to_plot, from = gene_info$start, to = gene_info$end, showTitle = TRUE)
    dev.off()  
  }

}

```

Here, we present gene coverage plots per time. 

```{r, fig.width = 37, fig.height = 35}
do.call("grid.arrange", c(list_cov_genes_time[1:8], ncol = 3))
do.call("grid.arrange", c(list_cov_genes_time[9:12], ncol = 3))
```

```{r, fig.width = 37, fig.height = 20}
do.call("grid.arrange", c(list_cov_genes_time[13:16], ncol = 3))
```



<br><br><br>



```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```


