---
title: "bulkATACseq_R_NumberPeaksReport"
author: "Laëtitia Racine"
date: "2023-03-24"

subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: false
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=F}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(knitr)
library(stringr)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(gridExtra) # grid.arrange() function
library(tidyverse) # pivot_longer() function
library(ggupset) # scale_x_upset() function
library(grid) # grid.text() function
library(UpSetR)

```

```{r, Working directory }

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/bulkATACseq_R_NumberPeaksReport/"))
dir.create(path = paste0(directory, "exp/bulkATACseq_R_NumberPeaksReport/", current_date))
directory_output = paste0(directory, "exp/bulkATACseq_R_NumberPeaksReport/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

# Load grange objects 
dir = pic_last_dir(paste0(directory, "exp/bulkATACseq_R_ChangeAnnotation/"))
gr_ann_list = list.files(path = paste0(dir, "/"), pattern = "update_annot.gr.rds")
name_list = str_extract(gr_ann_list, pattern = "[:graph:]+(?=_update_annot.gr.rds)")
gr_ann_list = paste0(dir, "/", gr_ann_list)
gr_ann_list = lapply(gr_ann_list, readRDS)
names(gr_ann_list) = name_list
  
```


<br><br><br>


# Overview

This code aims at performing a quantitative analysis of bulkATACseq dataset by counting the total number of peaks or the number of peaks per annotation depending on the condition and the time point. It takes as input files the *grange* files from bloc2 (preprocessing steps with threshold 10 on minimum number of reads per peak) and annotated in bulkATACseq_R_ChangeAnnotation.   
*Outputs* : \
- global table with number of peaks per condition and per annotation \
- bar plots with total number of peaks per condition per time point \
- bar plots with number of peaks per annotation per condition per time point \
- line plot with number of peaks per annotation per condition over time \
- upset plots with combination of annotations \
- upset plots focus on the 50 more important categories \
NB : copy and adaptation of the previous code ATAC_plot_peaks_annotation.Rmd (also compatible with snakemake) 

<br><br><br> 


**Create global table with number of peaks per condition and per annotation**  
```{r}

df_peaks_annotation = data.frame()

for (i in 1:length(gr_ann_list)) {
  temp = as.data.frame(gr_ann_list[[i]])
  temp = temp %>% dplyr::select(-seqnames, -start, -end, -width, -strand) %>%
    dplyr::mutate(sample = names(gr_ann_list)[[i]])
  temp2 = temp %>% 
    dplyr::group_by(sample) %>%
    dplyr::summarise(across(.cols = 1:(ncol(temp)-2), .fns = sum)) %>%
    tidyr::separate(col = sample, into = c("condition","time","donor"), sep = "_", remove = TRUE) %>%
    dplyr::mutate(condition = str_replace(condition, "MP", "CTRL")) %>%
    dplyr::mutate(total_nbpeaks = nrow(temp), .after = time)
  df_peaks_annotation = rbind(df_peaks_annotation, temp2)
}

# Add levels to define order legend in plots
df_peaks_annotation$condition = factor(df_peaks_annotation$condition, levels = c("Xvivo", "CTRL", "2DG", "DON", "AOA", "CTRLaK", "VPA"))
df_peaks_annotation$time = factor(df_peaks_annotation$time, levels = c("00h", "03h", "12h", "24h"))

df_peaks_annotation %>%
  kable() %>% 
  kable_styling(bootstrap_options = "striped") %>% 
  scroll_box(width = "100%", height = "250px")

write.csv2(x = df_peaks_annotation, row.names = FALSE,
           file = paste0(directory_output, "tab_nbpeaks_percond_perannot.csv")) 


# Remove aK and VPA conditions for some plots
df_peaks_annotation_filter = df_peaks_annotation %>% 
  dplyr::filter(condition != "CTRLaK" & condition != "VPA")

# Pivot du tableau
df_peaks_annotation_l = df_peaks_annotation %>% dplyr::select(-donor) 
df_peaks_annotation_l = df_peaks_annotation_l %>%
  tidyr::pivot_longer(cols = 3:ncol(df_peaks_annotation_l), 
                      names_to = "annotation", 
                      values_to = "count")

df_peaks_annotation_l_filter = df_peaks_annotation_l %>% 
  dplyr::filter(condition != "CTRLaK" & condition != "VPA")

```

<br>

**Display total number of peaks per condition - bar plots**  
```{r, out.width="100%", fig.width=18, fig.height=4, results='hold'}

# Plot with time on x axis
nbpeaks_pertime = ggplot(df_peaks_annotation, aes(x = time,  y = total_nbpeaks, label = total_nbpeaks, fill = condition)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  geom_text(position = position_dodge2(width = 0.9, preserve = "single"), angle = 90, hjust=1.5) +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black"),
        axis.ticks = element_blank()) +
  scale_fill_manual(values = color_code, 
                    breaks = unique(df_peaks_annotation$condition)) +
  labs(y = "Number of peaks detected")

# Plot with condition on x axis
nbpeaks_percond = ggplot(df_peaks_annotation,aes(x = condition,  y = total_nbpeaks, label = total_nbpeaks, fill = time)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  geom_text(position = position_dodge2(width = 0.9, preserve = "single"), angle = 90, hjust=1.5) +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black"),
        axis.ticks = element_blank()) +
  labs(y = "Number of peaks detected")

# Show plots
grid.arrange(grobs = list(nbpeaks_percond, nbpeaks_pertime), ncol = 2, nrow = 1)

# Save plots
ggsave(plot = nbpeaks_pertime, width = 16*0.75, height = 9*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_pertime.svg"))
ggsave(plot = nbpeaks_percond, width = 16*0.75, height = 9*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_percond.svg"))

```

```{r, out.width="100%", fig.width=18, fig.height=4, results = 'hold'}

# Plot with time on x axis
nbpeaks_pertime_filter = ggplot(df_peaks_annotation_filter, aes(x = time,  y = total_nbpeaks, label = total_nbpeaks, fill = condition)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  geom_text(position = position_dodge2(width = 0.9, preserve = "single"), angle = 90, hjust=1.5) +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black"),
        axis.ticks = element_blank()) +
  scale_fill_manual(values = color_code, 
                    breaks = unique(df_peaks_annotation$condition)) +
  labs(y = "Number of peaks detected")

# Plot with condition on x axis
nbpeaks_percond_filter = ggplot(df_peaks_annotation_filter ,aes(x = condition,  y = total_nbpeaks, label = total_nbpeaks, fill = time)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  geom_text(position = position_dodge2(width = 0.9, preserve = "single"), angle = 90, hjust=1.5) +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black"),
        axis.ticks = element_blank()) +
  labs(y = "Number of peaks detected")

# Show plots
grid.arrange(grobs = list(nbpeaks_percond_filter, nbpeaks_pertime_filter), ncol = 2, nrow = 1)

# save plots
ggsave(plot = nbpeaks_pertime, width = 16*0.75, height = 9*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_pertime_filter.svg"))
ggsave(plot = nbpeaks_percond, width = 16*0.75, height = 9*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_percond_filter.svg"))

```

<br>

**Display total number of peaks per condition per annotation - bar plots**  
```{r, out.width="100%", fig.width=18, fig.height=14, results = 'hold'}

# Plot with time on x axis
annot_pertime = ggplot(df_peaks_annotation_l, aes(x = time,  y = count, fill = condition)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black")) +
  scale_fill_manual(values = color_code, 
                    breaks=unique(df_peaks_annotation$condition)) +
  facet_wrap(annotation~., scale = "free_x")

# Plot with condition on x axis 
annot_percond = ggplot(df_peaks_annotation_l,aes(x = condition,  y = count, fill = time)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black")) +
  facet_wrap(annotation~., scale = "free_x")

# Show plots
grid.arrange(grobs = list(annot_percond, annot_pertime), ncol = 1, nrow = 2)

# Save plots
ggsave(plot = annot_pertime, width = 18*0.75, height = 12*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_annotation_pertime.svg"))
ggsave(plot = annot_percond, width = 22*0.75, height = 14*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_annotation_percond.svg"))

```

```{r, out.width="100%", fig.width=18, fig.height=14, results='hold'}

# Plot with time on x axis
annot_pertime_filter = ggplot(df_peaks_annotation_l_filter, aes(x = time,  y = count, fill = condition)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black")) +
  scale_fill_manual(values = color_code, 
                    breaks=unique(df_peaks_annotation$condition)) +
  facet_wrap(annotation~., scale = "free_x")

# Plot with condition on x axis 
annot_percond_filter = ggplot(df_peaks_annotation_l_filter,aes(x = condition,  y = count, fill = time)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme(legend.position = "right", 
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black")) +
  facet_wrap(annotation~., scale = "free_x")

# Show plots
grid.arrange(grobs = list(annot_percond_filter, annot_pertime_filter), ncol = 1, nrow = 2)

# Save plots
ggsave(plot = annot_pertime_filter, width = 16*0.75, height = 9*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_annotation_pertime_filter.svg"))
ggsave(plot = annot_percond_filter, width = 16*0.75, height = 9*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_annotation_percond_filter.svg"))

```

<br>

**Display chronological modification in number of peaks per annotation - line plot**
```{r, out.width="100%", fig.width=18, fig.height=14, results='hold'}

# Keep only conditions with multiple time points
df_chrono = data_frame()  
for (i in 1:length(unique(df_peaks_annotation_l$condition))) {
  df_temp = df_peaks_annotation_l %>% dplyr::filter(condition == unique(df_peaks_annotation_l$condition)[i])
  if (length(unique(df_temp$time))>1) df_chrono = rbind(df_temp, df_chrono) 
}

line_report = ggplot(df_chrono, aes(x = time, y = count, group = condition)) +
  geom_line(aes(color = condition), linewidth = 1)+
  geom_point(aes(color = condition), size = 2)+
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold", colour= "black", size = 16),
        strip.text.x = element_text(size=11, color="black", face="bold.italic"),
        axis.line.y = element_line(colour = "black"),
        axis.text.y = element_text(size = 11, colour = "black"),
        axis.ticks.y = element_line() ,
        axis.title.y = element_text(vjust = 1 ,size = 14),
        axis.title.x = element_blank(),
        axis.text.x = element_text(vjust = -0.5, size = 11, colour = "black"),
        axis.ticks = element_blank()) +
  facet_grid(condition~annotation, scales = "free_x") +
  labs(title = "Overtime progression") +
  ylab(label = "Number of peaks detected") +
  scale_discrete_manual(aesthetics = c("fill", "colour"), 
                        values = color_code, 
                        breaks=unique(df_chrono$condition))
line_report

ggsave(plot = line_report, width = 28*0.75, height = 14*0.75,
       filename = paste0(directory_output, "plot_nbpeaks_annotation_chrono.svg"))

```

<br>

**Display peaks annotation distribution - upset plot**  
```{r, out.width="100%", fig.width=18, fig.height=12, results="hide"}

for (i in 1:length(gr_ann_list)){
  
  # Extract Grange and name
  gr_file = gr_ann_list[[i]]
  name_sample = str_extract(string = names(gr_ann_list)[i], 
                            pattern = "[:alnum:]+_[:digit:]{2}h")
  name_cond = str_extract(string = names(gr_ann_list)[i], 
                            pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
  
  # Organize dataframe for plotting
  peaks_ann = as.data.frame(gr_file) %>% 
    dplyr::select(-seqnames, -start, -end, -width, -strand, -peak_name) %>%
    tibble::rownames_to_column(var = "peakID") 
  
  peaks_pivot = pivot_longer(data = peaks_ann, 
                             cols = colnames(peaks_ann)[-1], 
                             names_to = "annotation", 
                             values_to = "value") %>%
    dplyr::filter(value == TRUE) %>%
    dplyr::select(-value)

  peaks_group_anno = peaks_pivot %>%
    dplyr::group_by(peakID) %>%
    dplyr::summarize(annotations = list(annotation)) 

  # simple upset plot
  plot_glob = ggplot(peaks_group_anno, aes(x = annotations)) +
    geom_bar(fill = color_code[name_cond], color = "black") +
    ggtitle(paste0("Peaks distribution in genomic features - ", name_sample)) +
    ylab("Number of peaks") +
    scale_x_upset()
  
  # minimize maximum number of intersections displayed
  plot_zoom = ggplot(peaks_group_anno, aes(x = annotations)) +
    geom_bar(fill = color_code[name_cond], color = "black") +
    ggtitle(paste0("Peaks distribution in genomic features - ", name_sample)) +
    ylab("Number of peaks") +
    scale_x_upset(n_intersection = 50)
  
  print(grid.arrange(grobs = list(plot_glob, plot_zoom), ncol = 1, nrow = 2))
  
  ggsave(plot = plot_glob, width = 22*0.75, height = 9*0.75,
         filename = paste0(directory_output, "plot_upset_", name_sample, ".svg"))
  ggsave(plot = plot_zoom, width = 16*0.75, height = 9*0.75,
         filename = paste0(directory_output, "plot_upset_zoom_", name_sample, ".svg"))

  # upset plot with size 
  png(filename = paste0(directory_output, "plot_upset_setsize_", name_sample, ".png"), 
      width = 2000, height = 1000)
  print(peaks_pivot %>%
          unnest(cols = annotation) %>%
          mutate(annotMember=1) %>%
          pivot_wider(names_from = annotation, 
                      values_from = annotMember, 
                      values_fill = list(annotMember = 0)) %>%
          as.data.frame() %>%
          UpSetR::upset(sets = colnames(peaks_ann)[-1],
                        empty.intersections = NULL,
                        nintersects = NA,
                        order.by = "freq",
                        main.bar.color = color_code[name_cond])
  )
  grid.text(paste0("Peaks distribution in genomic features - ", name_sample), 
            x = 0.65, y=0.95, gp=gpar(fontsize=20))
  dev.off()
  
  png(filename = paste0(directory_output, "plot_upset_zoom_setsize_", name_sample, ".png"), 
      width = 2000, height = 1000)
  print(peaks_pivot %>%
          unnest(cols = annotation) %>%
          mutate(annotMember=1) %>%
          pivot_wider(names_from = annotation, 
                      values_from = annotMember, 
                      values_fill = list(annotMember = 0)) %>%
          as.data.frame() %>%
          UpSetR::upset(sets = colnames(peaks_ann)[-1],
                        empty.intersections = NULL,
                        nintersects = 50,
                        order.by = "freq",
                        main.bar.color = color_code[name_cond])
  )
  grid.text(paste0("Peaks distribution in genomic features - ", name_sample), 
            x = 0.65, y=0.95, gp=gpar(fontsize=20))
  dev.off()
  
}

```


<br><br><br>


```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```
