---
title: "bulkATACseq_ChangeAnnotation"
author: "Laëtitia Racine"
date: "2023-03-22"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=F}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(knitr)
library(stringr)
library(GenomicRanges)

```

```{r, Working directory }

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
directory_data = paste0(directory, "data/bulkATACseq/genomic_ranges/")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/bulkATACseq_R_ChangeAnnotation/"))
dir.create(path = paste0(directory, "exp/bulkATACseq_R_ChangeAnnotation/", current_date))
directory_output = paste0(directory, "exp/bulkATACseq_R_ChangeAnnotation/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

<br>

The annotation of bulkATACseq files was performed using the old annotation file (from MARS ATAC article).  
As I couldn't find the origin of this document, I recreate a new file (see annotationsFiles_createGrange.Rmd).  
In this code, we transform the bulkATACseq grange from bloc2 analysis by changing the annotation.  
It's an alternative to avoid rerunning all the bloc2 analysis but the better would be to change it directly in bloc2.  
For now, we will use the modified grange to perform the comparison between bulkATACseq and scATACseq.  
We can change the bloc2 analysis code latter.  

<br>

```{r, Input loading}

# Load grange objects 
gr_list = list.files(path = directory_data, pattern = "threshold_10_ann.gr.rds")
gr_list = paste0(directory_data, gr_list)
gr_list = lapply(gr_list, readRDS)
  
# Load grange with hg19 annotations
dir_annot = pic_last_dir(paste0(directory, "exp/", "AnnotationsFiles_createGrange/"))
hg19_annotations = readRDS(paste0(dir_annot, "/", "hg19_annotations_gr.rds"))

```

```{r}

gr_list_update = list()

for (i in 1:length(gr_list)) {
  
  # First a matrix is created filled with FALSE and added to the Grange
  grange = gr_list[[i]]
  annotations_types = levels(factor(hg19_annotations$type))
  metadata = matrix(FALSE, ncol = length(annotations_types), nrow = length(names(grange)))
  colnames(metadata) = annotations_types
  mcols(grange) = metadata
  
  # For each of the annotations types an overlap is calculated and 
  # used to assigned the peak as TRUE when overlapping with the annotation
  for (i in 1:ncol(metadata)){
    sub_annot = hg19_annotations[hg19_annotations$type == annotations_types[i]]
    overlaps = findOverlaps(grange, sub_annot)
    mcols(grange)[queryHits(overlaps),i] = TRUE
  }

  # Add peak_name in metadata
  peak_name = data.frame(grange) %>%
    dplyr::mutate(peak_name = str_c(seqnames, start, end, sep = "_")) %>%
    dplyr::select(peak_name)
  mcols(grange) = cbind(mcols(grange), peak_name)
  
  # Save 
 gr_list_update = c(gr_list_update, grange)

}

```

```{r}

for (i in 1:length(gr_list_update)) {
  
  gr = gr_list_update[[i]]
  name = str_extract(string = names(gr)[1], pattern = "[:alnum:]+_[:digit:]{2}h_[:graph:]+(?=_peak)")
  if(str_detect(name, "aK") == TRUE) {
    name = paste0("CTRL", name) 
    print(name)
  } else {
    print(name) 
  }
  saveRDS(gr, paste0(directory_output, name, "_threshold_10_update_annot.gr.rds"))
  
}

```


```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

