---
title: "bulkATACseq_R_ConditionSpecificPeaks"
author: "Laëtitia Racine"
date: "2023-03-27"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=F}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(knitr)
library(stringr)
library(dplyr)
library(ggplot2)
library(tibble)
library(kableExtra)
library(org.Hs.eg.db)
library(clusterProfiler)
library(ChIPseeker)
library(ReactomePA)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(GenomicRanges)
library(ggplotify)
library(karyoploteR)
library(gridExtra)

```

```{r, Working directory}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/bulkATACseq_R_ConditionSpecificPeaks/"))
dir.create(path = paste0(directory, "exp/bulkATACseq_R_ConditionSpecificPeaks/", 
                         current_date))
directory_output = paste0(directory, "exp/bulkATACseq_R_ConditionSpecificPeaks/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading, results="hide"}

dir = pic_last_dir(paste0(directory, "exp/bulkATACseq_R_Multiparametric_DifferentialAccessibility/"))
# count_df = read.csv2(dir, "readcount_df_peaks_union_allsamples.csv")
count_df = read.csv2("~/Documents/Git_Differentiation_Metabolism/exp/bulkATACseq_exp_commit_0c92b243ed627213e241ac1e46a5d30e975d2730/bulkATACseq_R_Multiparametric_DifferentialAccessibility/20230403/readcount_df_peaks_union_allsamples.csv")

rownames(count_df) = paste(count_df$seqnames, count_df$start, count_df$end, sep = "_")

# Prepare dataframe for pathway analysis
gr = makeGRangesFromDataFrame(count_df, keep.extra.columns = T)
peakAnno = annotatePeak(gr, tssRegion=c(-1000, 1000), 
                        TxDb=TxDb.Hsapiens.UCSC.hg19.knownGene, 
                        annoDb="org.Hs.eg.db")
dfPA = as.data.frame(peakAnno) 
dfPA = dfPA %>%
  dplyr::select(-str_subset(colnames(dfPA), "downsampled")) %>%
  dplyr::select(-str_subset(colnames(dfPA), "peaks_in"))
rownames(dfPA) = paste0(dfPA$seqnames, '_', dfPA$start, '_', dfPA$end)

```

```{r, Function definition}

path_go_fun = function(df_annot, list_peaks, cond) {

    # Extract interesting peaks
    selpeaks = df_annot[rownames(df_annot) %in% list_peaks,]
    selpeaks = selpeaks[abs(selpeaks$distanceToTSS) < 5000,]$geneId
    
    # Perform pathway enrichment analysis
    pathway = enrichPathway(selpeaks)

    # Perform GO analysis
    ego = enrichGO(gene = selpeaks,
                   OrgDb = org.Hs.eg.db,
                   ont = "ALL",
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.2,
                   readable = TRUE,
                   keyType = "ENTREZID")
    
    df_ego = as.data.frame(ego@result)
    
    if (nrow(df_ego) != 0) {
      
      plot = dotplot(ego, showCategory = 25) +
        facet_grid(ONTOLOGY ~ ., scales="free") +
        ggtitle(cond)
      
      results = list(pathway = pathway@result,
                     ego = ego,
                     dotplot = plot)
    } else {
      
      results = list(pathway = pathway@result,
                     ego = ego)
    }
    
  return(results)
    
}

```


<br><br><br>


# Overview

Using the data frame containing the common list of peaks for all the bulk conditions, with number of fragments in each .bam and information about "is a peak/is not" for each condition, we want to :  
- identify the peaks detected in only one condition  
- identify the peaks detected in all condition simultaneously  
- run a pathway and GO analysis on those peaks  
- localize those peaks on chromosomes  
We use as *input* the "readcount_df_peaks_union_allsamples.csv" from bulkATACseq_R_Multiparametric_DifferentialAccessibility.

```{r}

col_names = c(str_subset(colnames(count_df), "peaks_in"))
df = count_df %>% dplyr::select(all_of(col_names)) 

```


<br><br><br>


# Peaks present in only one condition

## Extract list of specific peaks (all conditions : MP/DON/2DG/AOA/MPaK/VPA)

```{r}

# Keep the peaks that are present in only one sample
df_spe = df %>% 
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) %>%
  dplyr::filter(Sum == 1)

# Count the number of specific peaks for each condition  
tab = data.frame(colSums(df_spe)) %>%
  rownames_to_column(var = "Condition") %>%
  dplyr::filter(Condition != "Sum") %>%
  dplyr::mutate(Condition = str_extract(Condition, "(?<=peaks_in_)[:graph:]+_[:digit:]{2}h"))
colnames(tab) = c("Condition", "NbPeaks_Specific")

tab %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")

# Extract and store list of specific peaks 
spe_peaks_all = list(
  "AOA_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_03h_D1.D2.D3_threshold_10 == TRUE)),
  "AOA_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_12h_D1.D2.D3_threshold_10 == TRUE)), 
  "AOA_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_24h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_03h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_12h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_24h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_03h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_12h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_24h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_03h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_12h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_24h_D1.D2.D3.D5.D6_threshold_10 == TRUE)),
  "MPaK_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_CTRLaK_24h_D4.D5.D6_threshold_10 == TRUE)),
  "VPA_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_VPA_24h_D4.D5.D6_threshold_10 == TRUE)),
  "Xvivo_00h" = rownames(df_spe %>% dplyr::filter(peaks_in_Xvivo_00h_D1.D2.D3_threshold_10 == TRUE))
)

spe_regions_all = list(
  "AOA_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["AOA_03h"]]))$region,
  "AOA_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["AOA_12h"]]))$region, 
  "AOA_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["AOA_24h"]]))$region,
  "DON_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["DON_03h"]]))$region,
  "DON_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["DON_12h"]]))$region,
  "DON_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["DON_24h"]]))$region,
  "2DG_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["2DG_03h"]]))$region,
  "2DG_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["2DG_12h"]]))$region,
  "2DG_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["2DG_24h"]]))$region,
  "MP_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["MP_03h"]]))$region,
  "MP_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["MP_12h"]]))$region,
  "MP_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["MP_24h"]]))$region,
  "MPaK_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["MPaK_24h"]]))$region,
  "VPA_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["VPA_24h"]]))$region,
  "Xvivo_00h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["Xvivo_00h"]]))$region
)

```

## Extract list of specific peaks (just MP/DON/2DG/AOA)

```{r}

# Remove the unwanted condition and keep only the specific peaks
to_remove = str_subset(colnames(df), pattern = "CTRLaK|VPA|Xvivo")
df_spe_filt = df %>% 
  dplyr::select(-all_of(to_remove)) %>%
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) %>%
  dplyr::filter(Sum == 1)

# Count the number of specific peaks for each condition  
tab_filt = data.frame(colSums(df_spe_filt)) %>%
  rownames_to_column(var = "Condition") %>%
  dplyr::filter(Condition != "Sum") %>%
  dplyr::mutate(Condition = str_extract(Condition, "(?<=peaks_in_)[:graph:]+_[:digit:]{2}h"))
colnames(tab_filt) = c("Condition", "NbPeaks_Specific")

tab_filt %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")

# Extract and store list of specific peaks 
spe_peaks_filt = list(
  "AOA_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_03h_D1.D2.D3_threshold_10 == TRUE)),
  "AOA_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_12h_D1.D2.D3_threshold_10 == TRUE)), 
  "AOA_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_24h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_03h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_12h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_24h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_03h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_12h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_24h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_03h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_12h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_24h_D1.D2.D3.D5.D6_threshold_10 == TRUE))
)

spe_regions_filt = list(
  "AOA_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["AOA_03h"]]))$region,
  "AOA_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["AOA_12h"]]))$region, 
  "AOA_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["AOA_24h"]]))$region,
  "DON_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["DON_03h"]]))$region,
  "DON_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["DON_12h"]]))$region,
  "DON_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["DON_24h"]]))$region,
  "2DG_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["2DG_03h"]]))$region,
  "2DG_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["2DG_12h"]]))$region,
  "2DG_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["2DG_24h"]]))$region,
  "MP_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["MP_03h"]]))$region,
  "MP_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["MP_12h"]]))$region,
  "MP_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["MP_24h"]]))$region,
  "MPaK_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["MPaK_24h"]]))$region,
  "VPA_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["VPA_24h"]]))$region,
  "Xvivo_00h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["Xvivo_00h"]]))$region
)

```

## Visualize chromosome localisation

### All chromosomes per condition

**All conditions**
```{r}

list_sample_chr = list()

for( i in 1:length(spe_peaks_all)) {
  
  # filter grange to keep only peaks of interest 
  gr_temp = subset(gr, region %in% spe_regions_all[[i]])
  name_cond = str_extract(string = names(spe_regions_all)[i], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
  
  list_sample_chr[[i]] = as.ggplot(expression(
      plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
        kpAddMainTitle(main = paste0("Peaks only detected in  ", names(spe_regions_all)[i]), cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
      )) 
  
   # save the plot
   png(filename = paste0(directory_output, "plot_allchr_specificpeaks_", names(spe_regions_all)[i], ".png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
     kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
     kpAddMainTitle(main = paste0("Peaks only detected in ", names(spe_regions_all)[i]), cex=1.5) %>%
     kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()
  
}

```

```{r, fig.width = 30, fig.height = 65}

do.call("grid.arrange", c(list_sample_chr, ncol = 3))

```

**Conditions without MPaK, VPA and Xvivo**
```{r}

list_sample_chr = list()

for( i in 1:length(spe_peaks_filt)) {
  
  # filter grange to keep only peaks of interest 
  gr_temp = subset(gr, region %in% spe_regions_filt[[i]])
  name_cond = str_extract(string = names(spe_regions_filt)[i], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
  
  list_sample_chr[[i]] = as.ggplot(expression(
      plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
        kpAddMainTitle(main = paste0("Peaks only detected in  ", names(spe_regions_filt)[i]), cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
      )) 
  
   # save the plot
   png(filename = paste0(directory_output, "plot_allchr_filtspecificpeaks_", names(spe_regions_filt)[i], ".png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
     kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
     kpAddMainTitle(main = paste0("Peaks only detected in  ", names(spe_regions_filt)[i]), cex=1.5) %>%
     kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()
  
}

```

```{r, fig.width = 30, fig.height = 65}

do.call("grid.arrange", c(list_sample_chr, ncol = 3))

```

### All conditions per chromosome

```{r}

chromosomes = c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", 
                "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", 
                "chr16", "chr17", "chr18", "chr19", "chr20", "chr21", "chr22", 
                "chrX", "chrY")

# Display settings: https://bernatgel.github.io/karyoploter_tutorial//Tutorial/PlotParams/PlotParams.html 
pp = getDefaultPlotParams(plot.type=1)
pp$data1height = 10
pp$ideogramheight = 6
pp$bottommargin = 6
pp$data1inmargin = 6

```

**All conditions**
```{r}

for (i in 1:length(chromosomes)) { 

  # visualize on html file
    kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
    for (j in 1:length(spe_peaks_all)) {
      gr_temp = subset(gr, region %in% spe_regions_all[[j]])
      name_cond = str_extract(string = names(spe_peaks_all)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
      name_sample = str_extract(string = names(spe_peaks_all)[j], 
                                pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6)
    }
  
  # save plot
  png(paste0(directory_output, "plot_allsamples_specificpeaks_",chromosomes[i], ".png"), 
      width = 2000, height = 1000)
  kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
  for (j in 1:length(spe_peaks_all)) {
    gr_temp = subset(gr, region %in% spe_regions_all[[j]])
    name_cond = str_extract(string = names(spe_peaks_all)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
    name_sample = str_extract(string = names(spe_peaks_all)[j], 
                            pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6) 
    }
  dev.off()
  
}

```

**Conditions without MPaK, VPA and Xvivo**
```{r}

for (i in 1:length(chromosomes)) { 

  # visualize on html file
    kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
    for (j in 1:length(spe_peaks_filt)) {
      gr_temp = subset(gr, region %in% spe_regions_filt[[j]])
      name_cond = str_extract(string = names(spe_peaks_filt)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
      name_sample = str_extract(string = names(spe_peaks_filt)[j], 
                                pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6)
    }
  
  # save plot
  png(paste0(directory_output, "plot_filtsamples_specificpeaks_",chromosomes[i], ".png"), 
      width = 2000, height = 1000)
  kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
  for (j in 1:length(spe_peaks_all)) {
    gr_temp = subset(gr, region %in% spe_regions_all[[j]])
    name_cond = str_extract(string = names(spe_peaks_filt)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
    name_sample = str_extract(string = names(spe_peaks_filt)[j], 
                            pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6) 
    }
  dev.off()
  
}

```

## Perform pathway and GO analysis 

```{r}

pathgo_spe_peaks_all = list()
for (i in 1:length(spe_peaks_all)) {
  name = names(spe_peaks_all)[i]
  pathgo_spe_peaks_all[[name]] = path_go_fun(dfPA, spe_peaks_all[[i]], name)
}

pathgo_spe_peaks_filt = list()
for (i in 1:length(spe_peaks_filt)) {
  pathgo_spe_peaks_filt[[i]] = path_go_fun(dfPA, spe_peaks_filt[[i]], name)
}

saveRDS(pathgo_spe_peaks_all, file = paste0(directory_output, "pathgo_spe_peaks_all.rds"))
saveRDS(pathgo_spe_peaks_filt, file = paste0(directory_output, "pathgo_spe_peaks_filt.rds"))

```

```{r, fig.width = 20, fig.height=40}

list_all = unlist(pathgo_spe_peaks_all, recursive=FALSE)
list_all_pathway = list_all[str_detect(names(list_all), "pathway")]
list_all_plot = list_all[str_detect(names(list_all), "dotplot")]
list_all_plot = lapply(list_all_plot, function(x) {x= x+theme(axis.text.y=element_text(size=8))})

pathway_tab = do.call("rbind", list_all_pathway) 
pathway_tab = pathway_tab %>%
  dplyr::mutate(Condition = str_extract(rownames(pathway_tab), "[:alnum:]+_[:digit:]{2}h(?=.pathway)"), .before = "ID")
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_spe_all_peaks.csv"))

pathway_short = pathway_tab %>%
  dplyr::group_by(Condition) %>%
  dplyr::top_n(10)

pathway_short %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

do.call("grid.arrange", c(list_all_plot, ncol = 3))

```

```{r, fig.width = 20, fig.height=30}

list_filt = unlist(pathgo_spe_peaks_filt, recursive=FALSE)
list_filt_pathway = list_filt[str_detect(names(list_filt), "pathway")]
list_filt_plot = list_filt[str_detect(names(list_filt), "dotplot")]
list_filt_plot = lapply(list_filt_plot, function(x) {x= x+theme(axis.text.y=element_text(size=8))})

pathway_tab = do.call("rbind", list_filt_pathway) 
pathway_tab = pathway_tab %>%
  dplyr::mutate(Condition = str_extract(rownames(pathway_tab), "[:alnum:]+_[:digit:]{2}h(?=.pathway)"), .before = "ID")
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_spe_filt_peaks.csv"))

pathway_short = pathway_tab %>%
  dplyr::group_by(Condition) %>%
  dplyr::top_n(10)

pathway_short %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

do.call("grid.arrange", c(list_filt_plot, ncol = 3))

```

<br><br><br>



# Peaks present in all conditions simultaneously

## Extract list of peaks shared by all samples 

```{r}

# Keep the peaks that are present in all sample (all conditions : MP/DON/2DG/AOA/MPaK/VPA) at the same time
df_all = df %>% 
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) %>%
  dplyr::filter(Sum == ncol(df))
cat("Number of shared peaks :", nrow(df_all))
peaks_all = rownames(df_all)
regions_all = (count_df %>% dplyr::filter(rownames(count_df) %in% peaks_all))$region
  
# Remove the unwanted condition and keep only the peaks that are present in samples MP/DON/2DG/AOA at the same time
to_remove = str_subset(colnames(df), pattern = "CTRLaK|VPA|Xvivo")
df_all_filt = df %>% 
  dplyr::select(-all_of(to_remove)) %>%
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) 
df_all_filt = df_all_filt %>%
  dplyr::filter(Sum == (ncol(df_all_filt) -1))
cat("Number of shared peaks :", nrow(df_all_filt))
peaks_all_filt = rownames(df_all_filt)
regions_all_filt = (count_df %>% dplyr::filter(rownames(count_df) %in% peaks_all_filt))$region

```

## Visualize chromosome localisation

**All conditions**
```{r}

gr_temp = subset(gr, region %in% regions_all)

plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE) %>%
        kpAddMainTitle(main = "Peaks detected in all conditions", cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)

# save the plot
   png(filename = paste0(directory_output, "plot_allcondpeaks.png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE) %>%
        kpAddMainTitle(main = "Peaks detected in all conditions", cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()

```

**Conditions without MPaK, VPA and Xvivo**
```{r}

gr_temp = subset(gr, region %in% regions_all_filt)

plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE) %>%
        kpAddMainTitle(main = "Peaks detected in all conditions", cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)

# save the plot
   png(filename = paste0(directory_output, "plot_allfiltcondpeaks.png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE) %>%
        kpAddMainTitle(main = "Peaks detected in MP/DON/2DG/AOA simulatenously", cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()


```

## Perform pathway and GO analysis 

```{r, fig.width = 6, fig.height = 10}

pathgo_all_peaks = path_go_fun(dfPA, peaks_all, "Peaks in all conditions simultaneously")
saveRDS(pathgo_all_peaks, file = paste0(directory_output, "pathgo_all_peaks.rds"))

pathway_tab = pathgo_all_peaks$pathway
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_sim_all_peaks.csv"))

pathway_tab %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

pathgo_all_peaks$dotplot +
  theme(axis.text.y=element_text(size=8))

```

```{r, fig.width = 6, fig.height = 10}

pathgo_filt_peaks = path_go_fun(dfPA, peaks_all_filt, "Peaks in MP/DON/AOA/2DG simultaneously")
saveRDS(pathgo_filt_peaks, file = paste0(directory_output, "pathgo_filt_peaks.rds"))

pathway_tab = pathgo_filt_peaks$pathway
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_sim_filt_peaks.csv"))

pathway_tab %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

pathgo_filt_peaks$dotplot +
  theme(axis.text.y=element_text(size=8))

```



<br><br><br>

# Conclusion

The chromosome representation are too condensed, we can't see interesting thing and the "unique" peaks seem to be at the same spot (which is not true).  
If we want to use this kind of representation, we need to zoom on a special window of the chromosome like an interesting gene for example.  



<br><br><br>

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

