---
title: "bulkATACseq_R_ConditionSpecificPeaks"
author: "Laëtitia Racine"
date: "2023-03-27"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=F}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(knitr)
library(stringr)
library(dplyr)
library(ggplot2)
library(tibble)
library(kableExtra)
library(org.Hs.eg.db)
library(clusterProfiler)
library(ChIPseeker)
library(ReactomePA)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(GenomicRanges)
library(ggplotify)
library(karyoploteR)
library(gridExtra)

```

```{r, Working directory}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/bulkATACseq_R_ConditionSpecificPeaks/"))
dir.create(path = paste0(directory, "exp/bulkATACseq_R_ConditionSpecificPeaks/", 
                         current_date))
directory_output = paste0(directory, "exp/bulkATACseq_R_ConditionSpecificPeaks/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading, results="hide"}

dir = pic_last_dir(paste0(directory, "exp/bulkATACseq_R_Multiparametric_DifferentialAccessibility/"))
# count_df = read.csv2(dir, "readcount_df_peaks_union_allsamples.csv")
count_df = read.csv2("~/Documents/Git_Differentiation_Metabolism/exp/bulkATACseq_exp_commit_0c92b243ed627213e241ac1e46a5d30e975d2730/bulkATACseq_R_Multiparametric_DifferentialAccessibility/20230403/readcount_df_peaks_union_allsamples.csv")

rownames(count_df) = paste(count_df$seqnames, count_df$start, count_df$end, sep = "_")

# Prepare dataframe for pathway analysis
gr = makeGRangesFromDataFrame(count_df, keep.extra.columns = T)
peakAnno = annotatePeak(gr, tssRegion=c(-1000, 1000), 
                        TxDb=TxDb.Hsapiens.UCSC.hg19.knownGene, 
                        annoDb="org.Hs.eg.db")
dfPA = as.data.frame(peakAnno) 
dfPA = dfPA %>%
  dplyr::select(-str_subset(colnames(dfPA), "downsampled")) %>%
  dplyr::select(-str_subset(colnames(dfPA), "peaks_in"))
rownames(dfPA) = paste0(dfPA$seqnames, '_', dfPA$start, '_', dfPA$end)

```

```{r, Function definition}

path_go_fun = function(df_annot, list_peaks, cond) {

    # Extract interesting peaks
    selpeaks = df_annot[rownames(df_annot) %in% list_peaks,]
    selpeaks = selpeaks[abs(selpeaks$distanceToTSS) < 5000,]$geneId
    
    # Perform pathway enrichment analysis
    pathway = enrichPathway(selpeaks)

    # Perform GO analysis
    ego = enrichGO(gene = selpeaks,
                   OrgDb = org.Hs.eg.db,
                   ont = "ALL",
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.2,
                   readable = TRUE,
                   keyType = "ENTREZID")
    
    df_ego = as.data.frame(ego@result)
    
    if (nrow(df_ego) != 0) {
      
      plot = dotplot(ego, showCategory = 25) +
        facet_grid(ONTOLOGY ~ ., scales="free") +
        ggtitle(cond)
      
      results = list(pathway = pathway@result,
                     ego = ego,
                     dotplot = plot)
    } else {
      
      results = list(pathway = pathway@result,
                     ego = ego)
    }
    
  return(results)
    
}

```


<br><br><br>


# Overview

Using the data frame containing the common list of peaks for all the bulk conditions, with number of fragments in each .bam and information about "is a peak/is not" for each condition, we want to :  
- identify the peaks detected in only one condition  
- identify the peaks detected in all condition simultaneously  
- run a pathway and GO analysis on those peaks  
- localize those peaks on chromosomes  
We use as *input* the "readcount_df_peaks_union_allsamples.csv" from bulkATACseq_R_Multiparametric_DifferentialAccessibility.

```{r}

col_names = c(str_subset(colnames(count_df), "peaks_in"))
df = count_df %>% dplyr::select(all_of(col_names)) 

```


<br><br><br>


# Peaks present in only one condition

## Isolate specific peaks

<br>

### Extract list of specific peaks (all conditions : MP/DON/2DG/AOA/MPaK/VPA/Xvivo)

```{r}

# Keep the peaks that are present in only one sample
df_spe = df %>% 
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) %>%
  dplyr::filter(Sum == 1)

# Count the number of specific peaks for each condition  
tab = data.frame(colSums(df_spe)) %>%
  rownames_to_column(var = "Condition") %>%
  dplyr::filter(Condition != "Sum") %>%
  dplyr::mutate(Condition = str_extract(Condition, "(?<=peaks_in_)[:graph:]+_[:digit:]{2}h"))
colnames(tab) = c("Condition", "NbPeaks_Specific")

tab %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")

# Extract and store list of specific peaks 
spe_peaks_all = list(
  "AOA_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_03h_D1.D2.D3_threshold_10 == TRUE)),
  "AOA_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_12h_D1.D2.D3_threshold_10 == TRUE)), 
  "AOA_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_AOA_24h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_03h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_12h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_DON_24h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_03h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_12h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_2DG_24h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_03h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_03h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_12h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_12h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_MP_24h_D1.D2.D3.D5.D6_threshold_10 == TRUE)),
  "MPaK_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_CTRLaK_24h_D4.D5.D6_threshold_10 == TRUE)),
  "VPA_24h" = rownames(df_spe %>% dplyr::filter(peaks_in_VPA_24h_D4.D5.D6_threshold_10 == TRUE)),
  "Xvivo_00h" = rownames(df_spe %>% dplyr::filter(peaks_in_Xvivo_00h_D1.D2.D3_threshold_10 == TRUE))
)

spe_regions_all = list(
  "AOA_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["AOA_03h"]]))$region,
  "AOA_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["AOA_12h"]]))$region, 
  "AOA_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["AOA_24h"]]))$region,
  "DON_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["DON_03h"]]))$region,
  "DON_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["DON_12h"]]))$region,
  "DON_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["DON_24h"]]))$region,
  "2DG_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["2DG_03h"]]))$region,
  "2DG_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["2DG_12h"]]))$region,
  "2DG_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["2DG_24h"]]))$region,
  "MP_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["MP_03h"]]))$region,
  "MP_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["MP_12h"]]))$region,
  "MP_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["MP_24h"]]))$region,
  "MPaK_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["MPaK_24h"]]))$region,
  "VPA_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["VPA_24h"]]))$region,
  "Xvivo_00h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_all[["Xvivo_00h"]]))$region
)

```

<br>

### Extract list of specific peaks (just MP/DON/2DG/AOA)

```{r}

# Remove the unwanted condition and keep only the specific peaks
to_remove = str_subset(colnames(df), pattern = "CTRLaK|VPA|Xvivo")
df_spe_filt = df %>% 
  dplyr::select(-all_of(to_remove)) %>%
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) %>%
  dplyr::filter(Sum == 1)

# Count the number of specific peaks for each condition  
tab_filt = data.frame(colSums(df_spe_filt)) %>%
  rownames_to_column(var = "Condition") %>%
  dplyr::filter(Condition != "Sum") %>%
  dplyr::mutate(Condition = str_extract(Condition, "(?<=peaks_in_)[:graph:]+_[:digit:]{2}h"))
colnames(tab_filt) = c("Condition", "NbPeaks_Specific")

tab_filt %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")

# Extract and store list of specific peaks 
spe_peaks_filt = list(
  "AOA_03h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_AOA_03h_D1.D2.D3_threshold_10 == TRUE)),
  "AOA_12h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_AOA_12h_D1.D2.D3_threshold_10 == TRUE)), 
  "AOA_24h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_AOA_24h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_03h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_DON_03h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_12h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_DON_12h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_24h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_DON_24h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_03h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_2DG_03h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_12h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_2DG_12h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_24h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_2DG_24h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_03h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_MP_03h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_12h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_MP_12h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_24h" = rownames(df_spe_filt %>% dplyr::filter(peaks_in_MP_24h_D1.D2.D3.D5.D6_threshold_10 == TRUE))
)

spe_regions_filt = list(
  "AOA_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["AOA_03h"]]))$region,
  "AOA_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["AOA_12h"]]))$region, 
  "AOA_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["AOA_24h"]]))$region,
  "DON_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["DON_03h"]]))$region,
  "DON_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["DON_12h"]]))$region,
  "DON_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["DON_24h"]]))$region,
  "2DG_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["2DG_03h"]]))$region,
  "2DG_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["2DG_12h"]]))$region,
  "2DG_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["2DG_24h"]]))$region,
  "MP_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["MP_03h"]]))$region,
  "MP_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["MP_12h"]]))$region,
  "MP_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filt[["MP_24h"]]))$region
)

```

<br>

### Extract list of specific peaks (just Xvivo/MP/DON/2DG/AOA)

```{r}

# Remove the unwanted condition and keep only the specific peaks
to_remove = str_subset(colnames(df), pattern = "CTRLaK|VPA")
df_spe_filtx = df %>% 
  dplyr::select(-all_of(to_remove)) %>%
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) %>%
  dplyr::filter(Sum == 1)

# Count the number of specific peaks for each condition  
tab_filtx = data.frame(colSums(df_spe_filtx)) %>%
  rownames_to_column(var = "Condition") %>%
  dplyr::filter(Condition != "Sum") %>%
  dplyr::mutate(Condition = str_extract(Condition, "(?<=peaks_in_)[:graph:]+_[:digit:]{2}h"))
colnames(tab_filtx) = c("Condition", "NbPeaks_Specific")

tab_filtx %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")

# Extract and store list of specific peaks 
spe_peaks_filtx = list(
  "AOA_03h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_AOA_03h_D1.D2.D3_threshold_10 == TRUE)),
  "AOA_12h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_AOA_12h_D1.D2.D3_threshold_10 == TRUE)), 
  "AOA_24h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_AOA_24h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_03h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_DON_03h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_12h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_DON_12h_D1.D2.D3_threshold_10 == TRUE)),
  "DON_24h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_DON_24h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_03h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_2DG_03h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_12h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_2DG_12h_D1.D2.D3_threshold_10 == TRUE)),
  "2DG_24h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_2DG_24h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_03h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_MP_03h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_12h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_MP_12h_D1.D2.D3_threshold_10 == TRUE)),
  "MP_24h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_MP_24h_D1.D2.D3.D5.D6_threshold_10 == TRUE)),
  "Xvivo_00h" = rownames(df_spe_filtx %>% dplyr::filter(peaks_in_Xvivo_00h_D1.D2.D3_threshold_10 == TRUE))
)

spe_regions_filtx = list(
  "AOA_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["AOA_03h"]]))$region,
  "AOA_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["AOA_12h"]]))$region, 
  "AOA_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["AOA_24h"]]))$region,
  "DON_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["DON_03h"]]))$region,
  "DON_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["DON_12h"]]))$region,
  "DON_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["DON_24h"]]))$region,
  "2DG_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["2DG_03h"]]))$region,
  "2DG_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["2DG_12h"]]))$region,
  "2DG_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["2DG_24h"]]))$region,
  "MP_03h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["MP_03h"]]))$region,
  "MP_12h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["MP_12h"]]))$region,
  "MP_24h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["MP_24h"]]))$region,
  "Xvivo_00h" = (count_df %>% dplyr::filter(rownames(count_df) %in% spe_peaks_filtx[["Xvivo_00h"]]))$region
)

```

<br>


## Visualize chromosome localisation

<br>

### All chromosomes per condition

**All conditions**
```{r}

list_sample_chr = list()

for( i in 1:length(spe_peaks_all)) {
  
  # filter grange to keep only peaks of interest 
  gr_temp = subset(gr, region %in% spe_regions_all[[i]])
  name_cond = str_extract(string = names(spe_regions_all)[i], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
  
  list_sample_chr[[i]] = as.ggplot(expression(
      plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
        kpAddMainTitle(main = paste0("Peaks only detected in  ", names(spe_regions_all)[i]), cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
      )) 
  
   # save the plot
   png(filename = paste0(directory_output, "plot_allchr_specificpeaks_", names(spe_regions_all)[i], ".png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
     kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
     kpAddMainTitle(main = paste0("Peaks only detected in ", names(spe_regions_all)[i]), cex=1.5) %>%
     kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()
  
}

```

```{r, fig.width = 30, fig.height = 65}

do.call("grid.arrange", c(list_sample_chr, ncol = 3))

```

<br>

**Conditions without MPaK, VPA and Xvivo**
```{r}

list_sample_chr = list()

for( i in 1:length(spe_peaks_filt)) {
  
  # filter grange to keep only peaks of interest 
  gr_temp = subset(gr, region %in% spe_regions_filt[[i]])
  name_cond = str_extract(string = names(spe_regions_filt)[i], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
  
  list_sample_chr[[i]] = as.ggplot(expression(
      plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
        kpAddMainTitle(main = paste0("Peaks only detected in  ", names(spe_regions_filt)[i]), cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
      )) 
  
   # save the plot
   png(filename = paste0(directory_output, "plot_allchr_filtspecificpeaks_", names(spe_regions_filt)[i], ".png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
     kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
     kpAddMainTitle(main = paste0("Peaks only detected in  ", names(spe_regions_filt)[i]), cex=1.5) %>%
     kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()
  
}

```

```{r, fig.width = 30, fig.height = 65}

do.call("grid.arrange", c(list_sample_chr, ncol = 3))

```

<br>

**Conditions without MPaK and VPA**
```{r}

list_sample_chr = list()

for( i in 1:length(spe_peaks_filtx)) {
  
  # filter grange to keep only peaks of interest 
  gr_temp = subset(gr, region %in% spe_regions_filtx[[i]])
  name_cond = str_extract(string = names(spe_regions_filtx)[i], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
  
  list_sample_chr[[i]] = as.ggplot(expression(
      plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
        kpAddMainTitle(main = paste0("Peaks only detected in  ", names(spe_regions_filtx)[i]), cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
      )) 
  
   # save the plot
   png(filename = paste0(directory_output, "plot_allchr_specificpeaks_", names(spe_regions_filtx)[i], ".png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
     kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE, col = color_code[name_cond]) %>%
     kpAddMainTitle(main = paste0("Peaks only detected in ", names(spe_regions_filtx)[i]), cex=1.5) %>%
     kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()
  
}

```

```{r, fig.width = 30, fig.height = 65}

do.call("grid.arrange", c(list_sample_chr, ncol = 3))

```

<br>

### All conditions per chromosome

<br>

```{r}

chromosomes = c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", 
                "chr9", "chr10", "chr11", "chr12", "chr13", "chr14", "chr15", 
                "chr16", "chr17", "chr18", "chr19", "chr20", "chr21", "chr22", 
                "chrX", "chrY")

# Display settings: https://bernatgel.github.io/karyoploter_tutorial//Tutorial/PlotParams/PlotParams.html 
pp = getDefaultPlotParams(plot.type=1)
pp$data1height = 10
pp$ideogramheight = 6
pp$bottommargin = 6
pp$data1inmargin = 6

```

**All conditions**
```{r}

for (i in 1:length(chromosomes)) { 

  # visualize on html file
    kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
    for (j in 1:length(spe_peaks_all)) {
      gr_temp = subset(gr, region %in% spe_regions_all[[j]])
      name_cond = str_extract(string = names(spe_peaks_all)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
      name_sample = str_extract(string = names(spe_peaks_all)[j], 
                                pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6)
    }
  
  # save plot
  png(paste0(directory_output, "plot_allsamples_specificpeaks_",chromosomes[i], ".png"), 
      width = 2000, height = 1000)
  kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
  for (j in 1:length(spe_peaks_all)) {
    gr_temp = subset(gr, region %in% spe_regions_all[[j]])
    name_cond = str_extract(string = names(spe_peaks_all)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
    name_sample = str_extract(string = names(spe_peaks_all)[j], 
                            pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6) 
    }
  dev.off()
  
}

```

**Conditions without MPaK, VPA and Xvivo**
```{r}

for (i in 1:length(chromosomes)) { 

  # visualize on html file
    kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
    for (j in 1:length(spe_peaks_filt)) {
      gr_temp = subset(gr, region %in% spe_regions_filt[[j]])
      name_cond = str_extract(string = names(spe_peaks_filt)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
      name_sample = str_extract(string = names(spe_peaks_filt)[j], 
                                pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6)
    }
  
  # save plot
  png(paste0(directory_output, "plot_filtsamples_specificpeaks_",chromosomes[i], ".png"), 
      width = 2000, height = 1000)
  kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
  for (j in 1:length(spe_peaks_filt)) {
    gr_temp = subset(gr, region %in% spe_regions_filt[[j]])
    name_cond = str_extract(string = names(spe_peaks_filt)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
    name_sample = str_extract(string = names(spe_peaks_filt)[j], 
                            pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6) 
    }
  dev.off()
  
}

```

**Conditions without MPaK and VPA**
```{r}

for (i in 1:length(chromosomes)) { 

  # visualize on html file
    kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
    for (j in 1:length(spe_peaks_filtx)) {
      gr_temp = subset(gr, region %in% spe_regions_filtx[[j]])
      name_cond = str_extract(string = names(spe_peaks_filtx)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
      name_sample = str_extract(string = names(spe_peaks_filtx)[j], 
                                pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6)
    }
  
  # save plot
  png(paste0(directory_output, "plot_filtXsamples_specificpeaks_",chromosomes[i], ".png"), 
      width = 2000, height = 1000)
  kp = plotKaryotype(plot.type = 1, chromosomes=chromosomes[i], plot.params = pp)
  for (j in 1:length(spe_peaks_filtx)) {
    gr_temp = subset(gr, region %in% spe_regions_filtx[[j]])
    name_cond = str_extract(string = names(spe_peaks_filtx)[j], 
                              pattern = "[:alnum:]+(?=_[:digit:]{2}h)")
    name_sample = str_extract(string = names(spe_peaks_filtx)[j], 
                            pattern = "[:alnum:]+_[:digit:]{2}h")
      kpPlotRegions(kp, data=gr_temp, border=color_code[name_cond], r0=j-1, r1=j-0.6)
      kpPlotDensity(kp, data=gr_temp, data.panel = 1, r0=j-0.6, r1=j-0.2)
      kpAddLabels(kp, labels=name_sample, data.panel = 1, r0=j-1, r1=j-0.6) 
    }
  dev.off()
  
}

```

<br>

## Perform pathway and GO analysis 

<br>

```{r}

pathgo_spe_peaks_all = list()
for (i in 1:length(spe_peaks_all)) {
  name = names(spe_peaks_all)[i]
  pathgo_spe_peaks_all[[name]] = path_go_fun(dfPA, spe_peaks_all[[i]], name)
}

pathgo_spe_peaks_filt = list()
for (i in 1:length(spe_peaks_filt)) {
  name = names(spe_peaks_filt)[i]
  pathgo_spe_peaks_filt[[i]] = path_go_fun(dfPA, spe_peaks_filt[[i]], name)
}

pathgo_spe_peaks_filtx = list()
for (i in 1:length(spe_peaks_filtx)) {
  name = names(spe_peaks_filtx)[i]
  pathgo_spe_peaks_filtx[[i]] = path_go_fun(dfPA, spe_peaks_filtx[[i]], name)
}

saveRDS(pathgo_spe_peaks_all, file = paste0(directory_output, "pathgo_spe_peaks_all.rds"))
saveRDS(pathgo_spe_peaks_filt, file = paste0(directory_output, "pathgo_spe_peaks_filt.rds"))
saveRDS(pathgo_spe_peaks_filtx, file = paste0(directory_output, "pathgo_spe_peaks_filtx.rds"))

```

<br>

**All conditions**
```{r, fig.width = 20, fig.height=40}

list_all = unlist(pathgo_spe_peaks_all, recursive=FALSE)
list_all_pathway = list_all[str_detect(names(list_all), "pathway")]
list_all_plot = list_all[str_detect(names(list_all), "dotplot")]
list_all_plot = lapply(list_all_plot, function(x) {x= x+theme(axis.text.y=element_text(size=8))})

pathway_tab = do.call("rbind", list_all_pathway) 
pathway_tab = pathway_tab %>%
  dplyr::mutate(Condition = str_extract(rownames(pathway_tab), "[:alnum:]+_[:digit:]{2}h(?=.pathway)"), .before = "ID")
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_spe_all_peaks.csv"))

pathway_short = pathway_tab %>%
  dplyr::group_by(Condition) %>%
  dplyr::top_n(10)

pathway_short %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

do.call("grid.arrange", c(list_all_plot, ncol = 3))

```

<br>

**Conditions MP, DON, 2DG and AOA**
```{r, fig.width = 20, fig.height=30}

list_filt = unlist(pathgo_spe_peaks_filt, recursive=FALSE)
list_filt_pathway = list_filt[str_detect(names(list_filt), "pathway")]
list_filt_plot = list_filt[str_detect(names(list_filt), "dotplot")]
list_filt_plot = lapply(list_filt_plot, function(x) {x= x+theme(axis.text.y=element_text(size=8))})

pathway_tab = do.call("rbind", list_filt_pathway) 
pathway_tab = pathway_tab %>%
  dplyr::mutate(Condition = str_extract(rownames(pathway_tab), "[:alnum:]+_[:digit:]{2}h(?=.pathway)"), .before = "ID")
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_spe_filt_peaks.csv"))

pathway_short = pathway_tab %>%
  dplyr::group_by(Condition) %>%
  dplyr::top_n(10)

pathway_short %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

do.call("grid.arrange", c(list_filt_plot, ncol = 3))

```

<br>

**Conditions Xvivo, MP, DON, 2DG and AOA**
```{r, fig.width = 20, fig.height=30}

list_filtx = unlist(pathgo_spe_peaks_filtx, recursive=FALSE)
list_filtx_pathway = list_filt[str_detect(names(list_filtx), "pathway")]
list_filtx_plot = list_filt[str_detect(names(list_filtx), "dotplot")]
list_filtx_plot = lapply(list_filtx_plot, function(x) {x= x+theme(axis.text.y=element_text(size=8))})

pathway_tab = do.call("rbind", list_filtx_pathway) 
pathway_tab = pathway_tab %>%
  dplyr::mutate(Condition = str_extract(rownames(pathway_tab), "[:alnum:]+_[:digit:]{2}h(?=.pathway)"), .before = "ID")
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_spe_filtX_peaks.csv"))

pathway_short = pathway_tab %>%
  dplyr::group_by(Condition) %>%
  dplyr::top_n(10)

pathway_short %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

do.call("grid.arrange", c(list_filtx_plot, ncol = 3))

```



<br><br><br>



# Peaks present in all conditions simultaneously

## Extract list of peaks shared by all samples 

```{r}

# Keep the peaks that are present in all samples (all conditions : Xvivo/MP/DON/2DG/AOA/MPaK/VPA) at the same time
df_all = df %>% 
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) %>%
  dplyr::filter(Sum == ncol(df))
cat("Number of shared peaks :", nrow(df_all), "\n")
peaks_all = rownames(df_all)
regions_all = (count_df %>% dplyr::filter(rownames(count_df) %in% peaks_all))$region
  
# Remove the unwanted condition and keep only the peaks that are present in samples Xvivo/MP/DON/2DG/AOA at the same time
to_remove = str_subset(colnames(df), pattern = "CTRLaK|VPA")
df_all_filtx = df %>% 
  dplyr::select(-all_of(to_remove)) %>%
  dplyr::mutate(Sum = rowSums(., na.rm=TRUE)) 
df_all_filtx = df_all_filtx %>%
  dplyr::filter(Sum == (ncol(df_all_filtx) -1))
cat("Number of shared peaks :", nrow(df_all_filtx), "\n")
peaks_all_filtx = rownames(df_all_filtx)
regions_all_filtx = (count_df %>% dplyr::filter(rownames(count_df) %in% peaks_all_filtx))$region

```
<br>

## Visualize chromosome localisation

<br>

**All conditions**
```{r}

gr_temp = subset(gr, region %in% regions_all)

plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE) %>%
        kpAddMainTitle(main = "Peaks detected in all conditions", cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)

# save the plot
   png(filename = paste0(directory_output, "plot_allcondpeaks.png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE) %>%
        kpAddMainTitle(main = "Peaks detected in all conditions", cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()

```

<br>

**Conditions without MPaK and VPA**
```{r}

gr_temp = subset(gr, region %in% regions_all_filtx)

plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE) %>%
        kpAddMainTitle(main = "Peaks detected in all conditions", cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)

# save the plot
   png(filename = paste0(directory_output, "plot_allfiltXpeaks.png"),
       width = 1000, height = 1400)
   kp = plotKaryotype(genome = "hg19", plot.type = 2) %>%
        kpPlotRegions(data=gr_temp, avoid.overlapping = FALSE) %>%
        kpAddMainTitle(main = "Peaks detected in Xvivo/MP/DON/2DG/AOA simulatenously", cex=1.5) %>%
        kpPlotDensity(data=gr_temp, data.panel = 2)
  dev.off()


```

<br>

## Perform pathway and GO analysis 

<br>

**All conditions**
```{r, fig.width = 6, fig.height = 10}

pathgo_all_peaks = path_go_fun(dfPA, peaks_all, "Peaks in all conditions simultaneously")
saveRDS(pathgo_all_peaks, file = paste0(directory_output, "pathgo_all_peaks.rds"))

pathway_tab = pathgo_all_peaks$pathway
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_sim_all_peaks.csv"))

pathway_tab %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

pathgo_all_peaks$dotplot +
  theme(axis.text.y=element_text(size=8))

```

<br>

**Conditions Xvivo/MP/DON/2DG/AOA**
```{r, fig.width = 6, fig.height = 10}

pathgo_filtx_peaks = path_go_fun(dfPA, peaks_all_filtx, "Peaks in Xvivo/MP/DON/AOA/2DG simultaneously")
saveRDS(pathgo_filtx_peaks, file = paste0(directory_output, "pathgo_filtx_peaks.rds"))

pathway_tab = pathgo_filtx_peaks$pathway
rownames(pathway_tab) = NULL
write.csv2(x = pathway_tab, row.names = FALSE,
           file = paste0(directory_output, "pathway_tab_sim_filtx_peaks.csv"))

pathway_tab %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width="100%", height="500px")

pathgo_filtx_peaks$dotplot +
  theme(axis.text.y=element_text(size=8))

```



<br><br><br>

# Peaks specific and common (Xvivo/MP/DON/2DG/AOA) : chance or real localisation signification ?

<br>

Create the data.frame with windows on chromosome parts  
http://www.insilicase.com/Web/Chromlen.aspx  
```{r}

chr_end = c("chr1"=247249719, "chr2"=242951149, "chr3"=199501827,
            "chr4"=191273063, "chr5"=180857866,"chr6"=170899992, 
            "chr7"=158821424,"chr8"=146274826,"chr9"=140273252,
            "chr10"=135374737,"chr11"=134452384,"chr12"=132349534,
            "chr13"=114142980,"chr14"=106368585,"chr15"=100338915,
            "chr16"=88827254,"chr17"=78774742,"chr18"=76117153,
            "chr19"=63811651,"chr20"=62435964,"chr21"=46944323,
            "chr22"=49691432,"chrX"=154913754,"chrY"=57772954)	

window_fun = function(chr, chr_end, length_window) {
  size_chr = as.integer(chr_end[chr])
  windows = round(seq(from=0, to=size_chr, by=length_window))
  start = as.integer(windows)
  start = start[-length(start)]
  start = c(start[1], start[-1]+1) 
  end = as.integer(windows[-1])
  tab = data.frame(chromosome = chr,
                 start_w = start,
                 end_w = end,
                 window = paste0(start,"-", end))
  return(tab)
}

df_windows = data.frame()
for (i in 1:length(chr_end)) {
  temp = window_fun(names(chr_end)[i], chr_end, 100000)
  df_windows = rbind(df_windows, temp)                  
}

```

<br>

Create a table with the list of specific peaks for each condition and the list of common peaks.  
```{r}

spe_peaks_df = data.frame()
for (i in 1:length(spe_peaks_filtx)) {
  temp = data.frame(peak = spe_peaks_filtx[[i]],
                    condition = names(spe_peaks_filtx)[i]) %>%
    tidyr::separate(col = peak, into = c("seqnames", "start", "end"), sep ="_")
  temp$condition = paste0("only_", temp$condition)
  spe_peaks_df = rbind(spe_peaks_df, temp)
}

common_peaks_df = data.frame(peak = peaks_all_filtx,
                             condition = "common_allcond")  %>%
    tidyr::separate(col = peak, into = c("seqnames", "start", "end"), sep ="_")

peaks_df = rbind(spe_peaks_df, common_peaks_df)
peaks_df$start = as.integer(peaks_df$start)
peaks_df$end = as.integer(peaks_df$end)

```

<br>

For each window, count the number of peaks that are specific to one condition or common to all.
```{r}

df_peaks_window = data.frame()

for (i in 1:nrow(df_windows)) {
  
  # extract parameters of the window
  chr_row = df_windows$chromosome[i]
  start_row = df_windows$start[i]
  end_row = df_windows$end[i]
  
  # total number of peaks in the window
  df_tot = count_df %>% dplyr::filter(seqnames == chr_row & 
                                        start >= start_row & 
                                        end <= end_row)
  totpeaks = data.frame(total_nbpeaks = nrow(df_tot))
  
  # number of peaks specific to one condition in the window
  df_spe_com = peaks_df %>% dplyr::filter(seqnames == chr_row & 
                                        start >= start_row & 
                                        end <= end_row)

  nbpeaks = data.frame("only_2DG_03h" = integer(), "only_2DG_12h"= integer(), 
                       "only_2DG_24h"= integer(), "only_AOA_03h"= integer(), 
                       "only_AOA_12h"= integer(), "only_AOA_24h"= integer(), 
                       "common_allcond"= integer(), "only_DON_03h"= integer(), 
                       "only_DON_12h"= integer(), "only_DON_24h"= integer(), 
                       "only_MP_03h"= integer(), "only_MP_12h"= integer(), 
                       "only_MP_24h"= integer(), "only_Xvivo_00h"= integer())
  
  temp = data.frame(table(df_spe_com$condition)) 
  if (nrow(temp) != 0) {
    temp = temp %>% tidyr::pivot_wider(names_from = "Var1", values_from = "Freq")
    nbpeaks = rbind.fill(nbpeaks, temp)
  } else {
    nbpeaks[1,] = rep(0, 14)
  }

  # merge all the information into one data.frame
  temp = cbind(df_windows[i,], totpeaks, nbpeaks)
  df_peaks_window = rbind(df_peaks_window, temp)
}

df_peaks_window = df_peaks_window %>% replace(is.na(.), 0)

write.csv2(df_peaks_window, paste0(directory_output, "df_peaks_windows.csv"))

```

<br>

https://stackoverflow.com/questions/37299206/divide-all-columns-by-the-value-from-the-2nd-column-apply-for-all-rows  
```{r}

# Calcul ratio of peaks per category / total peak in the window
df_ratio = df_peaks_window
df_ratio[, -(1:4)] = sweep(df_ratio[, -(1:4)], 1, df_ratio[, 5], "/")
df_ratio[, -(1:4)] = df_ratio[, -(1:4)]*100
df_ratio[, -(1:4)] = round(df_ratio[, -(1:4)], 3)
df_ratio = df_ratio %>% dplyr::select(-total_nbpeaks)

write.csv2(df_ratio, paste0(directory_output, "df_ratio.csv"))

# Statistical test 
df_pval = data.frame()
for (i in 1:nrow(df_ratio)) {
  temp = df_ratio[i,]
  common = temp$common_allcond
  temp_pval = data.frame("pval_2DG_03h_VS_common" = ifelse(temp$only_2DG_03h == 0, NA, wilcox.test(temp$only_2DG_03h, common)$p.value),
                       "pval_2DG_12h_VS_common" = ifelse(temp$only_2DG_12h == 0, NA, wilcox.test(temp$only_2DG_12h, common)$p.value),
                       "pval_2DG_24h_VS_common" = ifelse(temp$only_2DG_24h == 0, NA, wilcox.test(temp$only_2DG_24h, common)$p.value),
                       "pval_AOA_03h_VS_common" = ifelse(temp$only_AOA_03h == 0, NA, wilcox.test(temp$only_AOA_03h, common)$p.value),
                       "pval_AOA_12h_VS_common" = ifelse(temp$only_AOA_12h == 0, NA, wilcox.test(temp$only_AOA_12h, common)$p.value),
                       "pval_AOA_24h_VS_common" = ifelse(temp$only_AOA_24h == 0, NA, wilcox.test(temp$only_AOA_24h, common)$p.value),
                       "pval_DON_03h_VS_common" = ifelse(temp$only_DON_03h == 0, NA, wilcox.test(temp$only_DON_03h, common)$p.value),
                       "pval_DON_12h_VS_common" = ifelse(temp$only_DON_12h == 0, NA, wilcox.test(temp$only_DON_12h, common)$p.value),
                       "pval_DON_24h_VS_common" = ifelse(temp$only_DON_24h == 0, NA, wilcox.test(temp$only_DON_24h, common)$p.value),
                       "pval_MP_03h_VS_common" = ifelse(temp$only_MP_03h == 0, NA, wilcox.test(temp$only_MP_03h, common)$p.value),
                       "pval_MP_12h_VS_common" = ifelse(temp$only_MP_12h == 0, NA, wilcox.test(temp$only_MP_12h, common)$p.value),
                       "pval_MP_24h_VS_common" = ifelse(temp$only_MP_24h == 0, NA, wilcox.test(temp$only_MP_24h, common)$p.value),
                       "pval_Xvivo_00h_VS_common" =ifelse(temp$only_Xvivo_00h == 0, NA, wilcox.test(temp$only_Xvivo_00h, common)$p.value))
  temp_pval = cbind(df_ratio[i,1:4], temp_pval)
  df_pval = rbind(df_pval, temp_pval)
}


```

<br>



<br>

```{r}

# Representation pour un chromosome 24hrs : specifique pour chaque cond et commun pour le chromosome 21 par exemple

```


<br><br><br>

# Conclusion

The chromosome representation are too condensed, we can't see interesting thing and the "unique" peaks seem to be at the same spot (which is not true).  
If we want to use this kind of representation, we need to zoom on a special window of the chromosome like an interesting gene for example.  



<br><br><br>

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

