---
title: "scATACseq_R_DWNS_GeneActivity_GeneCoverage.Rmd"
author: "Laëtitia"
date: "24/08/2022"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(Signac)
library(knitr)
library(stringr)
library(ggplot2)
library(Gviz)
library(tibble)
library(karyoploteR)
library(ggplotify)
library(gridExtra)
library(tidyr)

```

```{r, Working directories and external script}

directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_GeneActivity_GeneCoverage/"))
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_GeneActivity_GeneCoverage/", 
                         current_date))
directory_output = paste0(directory,"exp/","scATACseq_R_DWNS_GeneActivity_GeneCoverage/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scATACseq_R_NormalizationReduction/"))
seurat_obj = readRDS(paste0(dir, "/", "merged_seurat_qc_filter_annot_normreduc.rds"))
# mat =readRDS("/home/lracine/Documents/Git_Differentiation_Metabolism/exp/scATACseq_R_DWNS_GeneActivity_GeneCoverage/20230412/mat.rds")

# Extract localisation of hg38 annotation (see annotationsFiles_createGrange.Rmd)
dir = pic_last_dir(paste0(directory, "exp/AnnotationsFiles_createGrange/"))
annot_hg38 = readRDS(paste0(dir, "/", "hg38_annotations_gr.rds"))

```



<br><br><br>



# Overview

As a reminder : 

```{r, fig.width=16, fig.height=5}

seurat_obj = SetIdent(object = seurat_obj, value = "orig.ident")
ind_umap = list() 

for (i in 1:length(unique(seurat_obj$orig.ident))) {
  
  # Select condition to highlight
  highlight_cond = unique(seurat_obj$orig.ident)[i]
  color = unname(noquote(color_code[highlight_cond]))
  others_cond = setdiff(unique(seurat_obj$orig.ident), highlight_cond)
  others_cond_group = paste(others_cond, collapse=", ")
  
  # Sort cell according to their condition
  cell_highlight = unlist(CellsByIdentities(object = seurat_obj, idents = highlight_cond))
  cell_nohighlight = unlist(CellsByIdentities(object = seurat_obj, idents = others_cond))
  
  # Plot 
  plot_umap = DimPlot(object = seurat_obj,
                      group.by = "orig.ident",
                      reduction ="umap",
                      pt.size = 0.1,
                      cells.highlight = list(cell_highlight, cell_nohighlight),
                      cols.highlight = c(color,"grey")) +
    ggtitle(highlight_cond) +
    theme(legend.position = "none")
  
  ind_umap[[i]] = plot_umap
  
}

clust_umap = DimPlot(seurat_obj, 
                     reduction = "umap", 
                     group.by = "peaks_snn_res.0.25") + 
  ggtitle("Resolution_0.25") 

clust_umap | ind_umap[[1]] | ind_umap[[2]] | ind_umap[[3]] | ind_umap[[4]]

```



<br><br><br>



# Gene activity matrix

https://stuartlab.org/signac/articles/pbmc_vignette.html#create-a-gene-activity-matrix  
Note : "The activities will be much noisier than scRNAseq measurements. This is because they represent measurements from sparse chromatin data, and because they assume a general correspondence between gene body/promoter accessibility and gene expression which may not always be the case."    
It is just an indicative measurement.  

**Create the gene activity matrix**  
"To create a gene activity matrix, we extract gene coordinates and extend them to include the 2 kb upstream region (as promoter accessibility is often correlated with gene expression). We then count the number of fragments for each cell that map to each of these regions."  
```{r, Really slow, eval = FALSE}

gene.activities <- GeneActivity(seurat_obj)
saveRDS(gene.activities, paste0(directory_output, "gene.activities.rds"))

```

```{r}
gene.activities = readRDS(paste0(directory_output, "gene.activities.rds"))
```

```{r}

seurat_obj[['RNA']] <- CreateAssayObject(counts = gene.activities)

seurat_obj <- NormalizeData(
  object = seurat_obj,
  assay = 'RNA',
  normalization.method = 'LogNormalize',
  scale.factor = median(seurat_obj$nCount_RNA)
)

```

<br>

**Visualize the putative activity of marker genes - list of characteristic genes**
```{r, results='hold', fig.width=16, fig.height=12, message=TRUE}

DefaultAssay(seurat_obj) <- 'RNA'

plot_TF = FeaturePlot(
  object = seurat_obj,
  features = list_genes_TF_hemato,
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3,
  cols = c('#1873CC','#1798E5','#00BFFF','#4AC596','#00CC00',
           '#A2E700','#FFFF00','#FFD200','#FFA500')
)

plot_gl = FeaturePlot(
  object = seurat_obj,
  features = c(list_genes_metabo_glut, list_genes_metabo_glycolysis),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3,
  cols = c('#1873CC','#1798E5','#00BFFF','#4AC596','#00CC00',
           '#A2E700','#FFFF00','#FFD200','#FFA500')
)

plot_mettca = FeaturePlot(
  object = seurat_obj,
  features = c(list_genes_metabo, list_genes_metabo_tca),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3,
  cols = c('#1873CC','#1798E5','#00BFFF','#4AC596','#00CC00',
           '#A2E700','#FFFF00','#FFD200','#FFA500')
)

plot_TF
plot_gl
plot_mettca

```

<br>

**Visualize the putative activity of marker genes - list of more variable genes between cells**
```{r, fig.width=16, fig.height=22}

seurat_obj <- FindVariableFeatures(object = seurat_obj, assay = 'RNA', nfeatures = 100)

plot_varfeature = FeaturePlot(
  object = seurat_obj,
  features = seurat_obj[['RNA']]@var.features[1:20],
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3,
    cols = c('#1873CC','#1798E5','#00BFFF','#4AC596','#00CC00',
           '#A2E700','#FFFF00','#FFD200','#FFA500')
)
plot_varfeature

```

<br>

```{r, Save, fig.show="hide"}

ggsave(plot = plot_varfeature, 
       filename = paste0(directory_output, "plot_geneactvity_varfeature.png"), 
       width = 18, height = 22)

ggsave(plot = plot_TF, 
       filename = paste0(directory_output,"plot_geneactivity_TF.png"), 
       width = 18, height = 20)

ggsave(plot = plot_gl, 
       filename = paste0(directory_output,"plot_geneactivity_glutglyc.png"), 
       width = 18, height = 20)

ggsave(plot = plot_mettca, 
       filename = paste0(directory_output,"plot_geneactivity_mettca.png"), 
       width = 18, height = 20)

```

```{r}

list_to_keep = c("directory_output", "start_time", "color_code", "seurat_obj",
                 "list_genes_metabo", "list_genes_metabo_glut", "list_genes_TF_hemato",
                 "list_genes_metabo_glycolysis","list_genes_metabo_tca", "annot_hg38")
rm(list = setdiff(ls(), list_to_keep))
gc()

```



<br><br><br>



# Gene coverage

For each peak, we look for the closest feature manually based on hg38 annotations table.  
Then, we can do coverage plot on peaks related to interesting genes. 
https://stuartlab.org/signac/reference/coverageplot
Note : we can't use the closestfeature function here because it uses the info stored in *annotation slot*. Those annotations only indicate cds, gap, exons or cds but we don't have promoters localisation. To study promoter, we use the annotations from annotatr package and so we do it manually.  

```{r}

list_gene = unique(c(list_genes_metabo, list_genes_metabo_glut, 
                     list_genes_metabo_glycolysis, list_genes_metabo_tca,
                     list_genes_TF_hemato), "PKM")
# PKM and PKM2 same gene
annot = annot_hg38[annot_hg38$symbol %in% list_gene]

```

```{r, results="hold", fig.width=14, fig.height=8}

for(i in 1:length(list_gene)) {
  
  # Extract region limits for the studied gene
  annot_genes = annot[annot$symbol == list_gene[i]]
  if(length(annot_genes) != 0) {
    limits = c(min(start(annot_genes@ranges)), max(end(annot_genes@ranges)))
    chr = as.character(unique(seqnames(annot_genes)))
    region = paste(chr, paste(limits, collapse="-"), sep="-")
    
    # Keep cds and promoter info
    annot1 = annot_genes[annot_genes$type %in% c("hg38_genes_promoters", "hg38_genes_cds")]
    plot1 = CoveragePlot(
      object = seurat_obj,
      assay = "peaks",
      region = region,
      extend.upstream = 40000,
      extend.downstream = 20000,
      peaks = TRUE,
      annotation = FALSE,
      ranges = annot1,
      ranges.group.by = "type",
      ranges.title = "annotation",
      tile = FALSE)
    
    plot1_tile = CoveragePlot(
      object = seurat_obj,
      assay = "peaks",
      region = region,
      extend.upstream = 40000,
      extend.downstream = 20000,
      peaks = TRUE,
      annotation = FALSE,
      ranges = annot1,
      ranges.group.by = "type",
      ranges.title = "annotation",
      tile = TRUE)
    
    # Keep exons and introns info
    annot2 = annot_genes[annot_genes$type %in% c("hg38_genes_exons", "hg38_genes_introns")]
    plot2 = CoveragePlot(
      object = seurat_obj,
      assay = "peaks",
      region = region,
      extend.upstream = 40000,
      extend.downstream = 20000,
      peaks = TRUE,
      annotation = FALSE,
      ranges = annot2,
      ranges.group.by = "type",
      ranges.title = "annotation",
      tile = FALSE)
    
    plot2_tile = CoveragePlot(
      object = seurat_obj,
      assay = "peaks",
      region = region,
      extend.upstream = 40000,
      extend.downstream = 20000,
      peaks = TRUE,
      annotation = FALSE,
      ranges = annot2,
      ranges.group.by = "type",
      ranges.title = "annotation",
      tile = TRUE)
    
    # show and save
    print(plot1|plot2)
    print(plot1_tile|plot2_tile)
    ggsave(plot = plot1|plot2, width = 12, height = 6,
           filename = paste0(directory_output, list_gene[i], "_coverage_allcond.svg"))
     ggsave(plot = plot1_tile|plot2_tile, width = 12, height = 8,
           filename = paste0(directory_output, list_gene[i], "_coverage_tile_allcond.svg"))
    
  }

}

```



<br><br><br>



# Peaks distribution on chromosomes

Really slow part, require a lot of memory. If possible, run it on cluster or big computer. 
```{r}

dir = pic_last_dir(paste0(directory, "exp/", "scATACseq_R_NormalizationReduction/"))
mat = readRDS(paste0(dir, "/", "mat_fragments.rds"))

```

<br>

We sample 500 cells of each condition to plot.  
```{r}

list_graph = list()
list_graph_stat = list()

for (i in 1:length(unique(seurat_obj$orig.ident))) {
  
  cond = unique(seurat_obj$orig.ident)[i]
  cells_to_keep = sample(str_subset(colnames(mat), cond), 500)
  mat_cond = mat %>% dplyr::select(all_of(cells_to_keep))
  mat_cond = mat_cond %>% rownames_to_column("peak_name")
  mat_cond = mat_cond %>% separate(col = peak_name, 
                                   into = c("seqname", "start", "end")) 
  mat_cond = mat_cond %>% pivot_longer(cols = c(5:ncol(mat_cond)), 
                                       names_to = "cell", 
                                       values_to = "nbfragments")
  mat_cond$condition = cond

  plot_nbfragment = ggplot(mat_cond, aes(x=condition, y=nbfragments, col=condition)) +
    geom_violin() +
    geom_point(position = "jitter", alpha = 0.5) +
    facet_wrap(seqname~.) +
    scale_color_manual(values = color_code, limits = force) +
    theme(legend.position = "none")
  
  list_graph[[cond]] = plot_nbfragment
  
  mat_stat = mat_cond %>% 
    dplyr::group_by(condition, cell, seqname) %>%
    dplyr::mutate(median_fragment = median(nbfragments)) %>%
    dplyr::mutate(mean_fragment = mean(nbfragments)) %>%
    dplyr::mutate(min_fragment = min(nbfragments)) %>%
    dplyr::mutate(max_fragment = max(nbfragments)) %>%
    dplyr::select(-start, -end, -nbfragments)
  mat_stat = mat_stat %>%
    pivot_longer(cols = c("median_fragment", "mean_fragment", 
                          "min_fragment", "max_fragment"),
                 names_to = "stat", values_to = "stat_value")
  
  plot_stat_nbfragment = ggplot(mat_stat, aes(x=condition, y=stat_value, col=condition)) +
    geom_violin() +
    geom_point(position = "jitter", alpha = 0.5) +
    facet_grid(seqname~stat) +
    scale_color_manual(values = color_code, limits = force) +
    theme(legend.position = "none")
  
  list_graph_stat[[cond]] = plot_stat_nbfragment
  
}

```

```{r}



```



<br><br><br>



```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results = "hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

