---
title: "scATACseq_R_DWNS_CellsPeaksStatistics_part2"
author: "Laëtitia Racine"
date: "2022-09-06"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(ggplot2)
library(Seurat)
library(dplyr)
library(tidyverse)
library(kableExtra)
library(knitr)
library(gridExtra)

```

```{r, Working directories and external script}

directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_CellsPeaksStatistics/"))
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_CellsPeaksStatistics/", 
                         current_date))
directory_output = paste0(directory,"exp/","scATACseq_R_DWNS_CellsPeaksStatistics/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scATACseq_R_NormalizationReduction/"))
seurat_obj = readRDS(paste0(dir, "/", "merged_seurat_qc_filter_annot_normreduc.rds"))

number_conditions = length(unique(seurat_obj$orig.ident))
conditions = unique(seurat_obj$orig.ident)

matrix_seurat = GetAssayData(seurat_obj, slot = "data")
  
```


<br><br><br>


# Overview

This code aims at determining the statistics of the peaks among conditions.    
We work with the object containing all the conditions merged and normalized with a common set of regions.      
  
As a reminder :    
- In scATACseq, for each condition, the individual cells are pooled together to determine a list of peaks for the condition. All the paks detected are included in the matrix count and for each cell, the number of fragments detected in those peaks are indicated. Some fragments are removed because they didn't overlap with an identified peak, they are considered as noise. Overall, we can't say that a cell presents a peak but we can say if a cell presents fragments in an identified peak in the population.   
- The peak calling were performed individually for each condition. As a result, list of peaks from one condition to the other are different even thought some peaks can corresponds to the same genomic region (but start and end slightly different). To be able to compare the conditions, we decided to merge them and we created a common set of regions. This common set of region corresponds to the union of all the peaks from all the conditions. Ultimately, we obtain a matrix with a list of region as rows and the list of all the cells from all the conditions as columns. Then, for each cell, we know if it contains fragments in the region.    
  

<br><br><br>


# Quantitative analysis : number of fragments inside each peaks 

**WARNING : the quantitative analysis with all the cells require a huge amount of memory and is impossible to run on our computer.**    
Thus, we decided to pick randomly 1000 cells of each condition to perform it with a sample of the cells.   

```{r, results="hide"}

dir = pic_last_dir(paste0(directory, "exp/", "scATACseq_R_NormalizationReduction/"))
mat_df = readRDS(paste0(dir, "/mat.rds"))

# select randomly 1000 cells for each condition
col_to_keep = c()
for(i in 1:number_conditions){
  cond = conditions[i]
  col_list = sample(str_subset(colnames(mat_df), pattern = cond), 1000)
  col_to_keep = c(col_to_keep, col_list)
}
mat_df_cut = mat_df %>% dplyr::select(all_of(col_to_keep))
rm(mat_df)
saveRDS(mat_df_cut, paste0(directory_output, "mat_df_cut.rds"))

gc()

```

<br>

## Number of fragments in the peaks detected for each cell

For one condition, we extract all the associated cells and we look at the mean/median/min/max of number of fragments per peak detected (peak detected => we previously removed peaks with 0 fragment). The plot contains x values with x the number of cells. Example of interpretation for one value : for this cell, in mean/median/... there are y fragments in the peaks that are detected.  

```{r}

fragments_cell_fun = function(tab_matrix, var_cond, color_code) {
  
  tab_matrix = tab_matrix %>% 
    dplyr::select(str_subset(colnames(tab_matrix), var_cond)) %>%
    tidyr::pivot_longer(cols = everything(), 
                        names_to = "cell", 
                        values_to = "nbfragments")
  
  tab_matrix = tab_matrix %>%
    dplyr::filter(nbfragments != 0) %>%
    dplyr::group_by(cell) %>%
    dplyr::mutate(mean_nbfragments = mean(nbfragments),
                  median_nbfragments = median(nbfragments),
                  min_nbfragments = min(nbfragments),
                  max_nbfragments = max(nbfragments)) %>%
    dplyr::ungroup() %>% 
    dplyr::select(-nbfragments) %>%
    dplyr::distinct() 
  
 tab_matrix = tab_matrix %>%
    tidyr::pivot_longer(cols = c(2:ncol(tab_matrix)), 
                        names_to = "statistic", 
                        values_to = "nbfragments") %>%
    dplyr::mutate(condition = var_cond) 
  
  plot = ggplot(tab_matrix, aes(x=nbfragments, fill = condition)) +
    geom_histogram(binwidth = 0.1) +
    scale_fill_manual(values = color_code, limits = force) +
    xlab("Number of fragments in the peaks detected") +
    ylab("Number of cells") +
    scale_x_continuous(breaks=seq(0,max(tab_matrix$nbfragments),0.2)) +
    facet_wrap(facets = .~statistic) +
    ggtitle(label = paste0(var_cond, " (binwidth = 0.1)")) +
    theme(legend.position = "none", axis.text.x = element_text(angle=90)) 
  
  return(list(cell_stat = tab_matrix,
              cell_plot = plot))
}


list_quanti_cell = list()
for (i in 1:number_conditions) {
  list_quanti_cell[[conditions[i]]] = fragments_cell_fun(tab_matrix = mat_df_cut,
                                                         var_cond = conditions[i],
                                                         color_code = color_code)
  gc()
}

```

```{r, fig.width = 24, fig.height = 16}

plots_cell = unlist(list_quanti_cell, recursive = FALSE)
plots_cell = plots_cell[str_detect(names(plots_cell), pattern = "plot")]
do.call("grid.arrange", c(plots_cell, ncol=2, nrow=2))

```

```{r, fig.show = "hide"}

saveRDS(list_quanti_cell, paste0(directory_output, "stat_nbfragments_perpeak_percell.csv"))
ggsave(plot = do.call("grid.arrange", c(plots_cell, ncol=2, nrow=2)), width = 25, height = 15,
       paste0(directory_output, "nbpeaks_per_cell_quanti_plot.svg"))

rm(plots_cell, list_quanti_cell)

```

<br>


## Number of fragments in the cells that detected those peaks

For one condition, we extract all the associated peaks and we look at the mean/median/min/max of number of fragments in the cells for those peaks. The plot contains x values with x the number of peaks. Example of interpretation for one value : for this peak, cells have in mean/median/... y fragments in this region (the cells with 0 were removed).   

```{r}

fragments_peaks_fun = function(tab_matrix, var_cond, color_code) {
  
  tab_matrix = tab_matrix %>% 
    dplyr::select(str_subset(colnames(tab_matrix), var_cond)) %>%
    tibble::rownames_to_column("region") 
  
 tab_matrix = tab_matrix %>%
    tidyr::pivot_longer(cols = c(2:ncol(tab_matrix)), 
                        names_to = "cell", 
                        values_to = "nbfragments") %>%
    dplyr::filter(nbfragments != 0) 
  
  tab_matrix = tab_matrix %>%
    dplyr::group_by(region) %>%
    dplyr::mutate(mean_nbfragments = mean(nbfragments),
                  median_nbfragments = median(nbfragments),
                  min_nbfragments = min(nbfragments),
                  max_nbfragments = max(nbfragments)) %>%
    dplyr::ungroup() %>% 
    dplyr::select(-nbfragments, -cell) %>%
    dplyr::distinct() 
  
  tab_matrix = tab_matrix %>%
    tidyr::pivot_longer(cols = c(2:ncol(tab_matrix)), 
                        names_to = "statistic", 
                        values_to = "nbfragments") %>%
    dplyr::mutate(condition = var_cond) 
  
  plot = ggplot(tab_matrix, aes(x=nbfragments, fill=condition)) +
    geom_histogram(binwidth = 0.1) +
    scale_fill_manual(values = color_code, limits = force) +
    xlab("Number of fragments for the cells that detected this peak") +
    ylab("Number of peaks") +
    facet_wrap(facets = .~statistic) +
    scale_x_continuous(breaks=seq(0,max(tab_matrix$nbfragments),0.2)) +
    facet_wrap(facets = .~statistic) +
    ggtitle(label = paste0(var_cond, " (binwidth = 0.1)")) +
    theme(legend.position = "none", axis.text.x = element_text(angle=90)) 
  
  return(list(peak_stat = tab_matrix,
              peak_plot = plot))
}


list_quanti_peaks = list()
for (i in 1:number_conditions) {
  list_quanti_peaks[[conditions[i]]] = fragments_peaks_fun(tab_matrix = mat_df_cut,
                                                          var_cond = conditions[i],
                                                          color_code = color_code)
  gc()
}

```

```{r, fig.width = 24, fig.height = 16}

plots_peaks = unlist(list_quanti_peaks, recursive = FALSE)
plots_peaks = plots_peaks[str_detect(names(plots_peaks), pattern = "plot")]
do.call("grid.arrange", c(plots_peaks, ncol=2, nrow=2))

```

```{r, fig.show="hide"}

saveRDS(list_quanti_peaks, paste0(directory_output, "stat_nbfragments_percell_perpeaks.csv"))
ggsave(plot = do.call("grid.arrange", c(plots_peaks, ncol=2, nrow=2)), width = 25, height = 15,
       paste0(directory_output, "nbcell_sharing_peaks_quanti_plot.svg"))

```


<br><br><br>


```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results="hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

