---
title: "scATACseq_R_DWNS_GeneActivity_GeneCoverage.Rmd"
author: "Laëtitia"
date: "2023-03-28"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(Signac)
library(knitr)
library(stringr)
library(ggplot2)
library(Gviz)
library(tibble)
library(karyoploteR)
library(ggplotify)
library(gridExtra)
library(tidyr)

```

```{r, Working directories and external script}

directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_GeneActivity_GeneCoverage/"))
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_GeneActivity_GeneCoverage/", 
                         current_date))
directory_output = paste0(directory,"exp/","scATACseq_R_DWNS_GeneActivity_GeneCoverage/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scATACseq_R_NormalizationReduction/"))
seurat_obj = readRDS(paste0(dir, "/", "merged_seurat_qc_filter_annot_normreduc.rds"))

# Extract localisation of hg38 annotation (see annotationsFiles_createGrange.Rmd)
dir = pic_last_dir(paste0(directory, "exp/AnnotationsFiles_createGrange/"))
annot_hg38 = readRDS(paste0(dir, "/", "hg38_annotations_gr.rds"))

```



<br><br><br>



# Overview

As a reminder : 

```{r, fig.width=16, fig.height=5}

seurat_obj = SetIdent(object = seurat_obj, value = "orig.ident")
ind_umap = list() 

for (i in 1:length(unique(seurat_obj$orig.ident))) {
  
  # Select condition to highlight
  highlight_cond = unique(seurat_obj$orig.ident)[i]
  color = unname(noquote(color_code[highlight_cond]))
  others_cond = setdiff(unique(seurat_obj$orig.ident), highlight_cond)
  others_cond_group = paste(others_cond, collapse=", ")
  
  # Sort cell according to their condition
  cell_highlight = unlist(CellsByIdentities(object = seurat_obj, idents = highlight_cond))
  cell_nohighlight = unlist(CellsByIdentities(object = seurat_obj, idents = others_cond))
  
  # Plot 
  plot_umap = DimPlot(object = seurat_obj,
                      group.by = "orig.ident",
                      reduction ="umap",
                      pt.size = 0.1,
                      cells.highlight = list(cell_highlight, cell_nohighlight),
                      cols.highlight = c(color,"grey")) +
    ggtitle(highlight_cond) +
    theme(legend.position = "none")
  
  ind_umap[[i]] = plot_umap
  
}

clust_umap = DimPlot(seurat_obj, 
                     reduction = "umap", 
                     group.by = "peaks_snn_res.0.25") + 
  ggtitle("Resolution_0.25") 

clust_umap | ind_umap[[1]] | ind_umap[[2]] | ind_umap[[3]] | ind_umap[[4]]

```



<br><br><br>



# Gene activity matrix

https://stuartlab.org/signac/articles/pbmc_vignette.html#create-a-gene-activity-matrix  
Note : "The activities will be much noisier than scRNAseq measurements. This is because they represent measurements from sparse chromatin data, and because they assume a general correspondence between gene body/promoter accessibility and gene expression which may not always be the case."    
It is just an indicative measurement.  

**Create the gene activity matrix**  
"To create a gene activity matrix, we extract gene coordinates and extend them to include the 2 kb upstream region (as promoter accessibility is often correlated with gene expression). We then count the number of fragments for each cell that map to each of these regions."  
```{r, Really slow}

gene.activities <- GeneActivity(seurat_obj)
saveRDS(gene.activities, paste0(directory_output, "gene.activities.rds"))

```

```{r, eval=FALSE}
gene.activities = readRDS(paste0(directory_output, "gene.activities.rds"))
```

```{r}

seurat_obj[['RNA']] <- CreateAssayObject(counts = gene.activities)

seurat_obj <- NormalizeData(
  object = seurat_obj,
  assay = 'RNA',
  normalization.method = 'LogNormalize',
  scale.factor = median(seurat_obj$nCount_RNA)
)

```

<br>

**Visualize the putative activity of marker genes - list of characteristic genes**
```{r, results='hold', fig.width=16, fig.height=12, message=TRUE}

DefaultAssay(seurat_obj) <- 'RNA'

plot_TF = FeaturePlot(
  object = seurat_obj,
  features = list_genes_TF_hemato,
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3,
  cols = c('#1873CC','#1798E5','#00BFFF','#4AC596','#00CC00',
           '#A2E700','#FFFF00','#FFD200','#FFA500')
)

plot_gl = FeaturePlot(
  object = seurat_obj,
  features = c(list_genes_metabo_glut, list_genes_metabo_glycolysis),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3,
  cols = c('#1873CC','#1798E5','#00BFFF','#4AC596','#00CC00',
           '#A2E700','#FFFF00','#FFD200','#FFA500')
)

plot_mettca = FeaturePlot(
  object = seurat_obj,
  features = c(list_genes_metabo, list_genes_metabo_tca),
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3,
  cols = c('#1873CC','#1798E5','#00BFFF','#4AC596','#00CC00',
           '#A2E700','#FFFF00','#FFD200','#FFA500')
)

plot_TF
plot_gl
plot_mettca

```

<br>

**Visualize the putative activity of marker genes - list of more variable genes between cells**
```{r, fig.width=16, fig.height=22}

seurat_obj <- FindVariableFeatures(object = seurat_obj, assay = 'RNA', nfeatures = 100)

plot_varfeature = FeaturePlot(
  object = seurat_obj,
  features = seurat_obj[['RNA']]@var.features[1:20],
  pt.size = 0.1,
  max.cutoff = 'q95',
  ncol = 3,
    cols = c('#1873CC','#1798E5','#00BFFF','#4AC596','#00CC00',
           '#A2E700','#FFFF00','#FFD200','#FFA500')
)
plot_varfeature

```

<br>

```{r, Save, fig.show="hide"}

ggsave(plot = plot_varfeature, 
       filename = paste0(directory_output, "plot_geneactvity_varfeature.png"), 
       width = 18, height = 22)

ggsave(plot = plot_TF, 
       filename = paste0(directory_output,"plot_geneactivity_TF.png"), 
       width = 18, height = 20)

ggsave(plot = plot_gl, 
       filename = paste0(directory_output,"plot_geneactivity_glutglyc.png"), 
       width = 18, height = 20)

ggsave(plot = plot_mettca, 
       filename = paste0(directory_output,"plot_geneactivity_mettca.png"), 
       width = 18, height = 20)

```

```{r, results="hide"}

list_to_keep = c("directory_output", "start_time", "color_code", "seurat_obj",
                 "list_genes_metabo", "list_genes_metabo_glut", "list_genes_TF_hemato",
                 "list_genes_metabo_glycolysis","list_genes_metabo_tca", 
                 "annot_hg38", "directory")
rm(list = setdiff(ls(), list_to_keep))
gc()

```



<br><br><br>



# Gene coverage

For each peak, we look for the closest feature manually based on hg38 annotations table.  
Then, we can do coverage plot on peaks related to interesting genes. 
https://stuartlab.org/signac/reference/coverageplot
Note : we can't use the closestfeature function here because it uses the info stored in *annotation slot*. Those annotations only indicate cds, gap, exons or cds but we don't have promoters localisation. To study promoter, we use the annotations from annotatr package and so we do it manually.  

```{r}

list_gene = unique(c(list_genes_metabo, list_genes_metabo_glut, 
                     list_genes_metabo_glycolysis, list_genes_metabo_tca,
                     list_genes_TF_hemato), "PKM")
# PKM and PKM2 same gene
annot = annot_hg38[annot_hg38$symbol %in% list_gene]

```

```{r, results="hold", fig.width=14, fig.height=8}

for(i in 1:length(list_gene)) {
  
  # Extract region limits for the studied gene
  annot_genes = annot[annot$symbol == list_gene[i]]
  if(length(annot_genes) != 0) {
    limits = c(min(start(annot_genes@ranges)), max(end(annot_genes@ranges)))
    chr = as.character(unique(seqnames(annot_genes)))
    region = paste(chr, paste(limits, collapse="-"), sep="-")
    
    # Keep cds and promoter info
    annot1 = annot_genes[annot_genes$type %in% c("hg38_genes_promoters", "hg38_genes_cds")]
    plot1 = CoveragePlot(
      object = seurat_obj,
      assay = "peaks",
      region = region,
      extend.upstream = 40000,
      extend.downstream = 20000,
      peaks = TRUE,
      annotation = FALSE,
      ranges = annot1,
      ranges.group.by = "type",
      ranges.title = "annotation",
      tile = FALSE)
    
    plot1_tile = CoveragePlot(
      object = seurat_obj,
      assay = "peaks",
      region = region,
      extend.upstream = 40000,
      extend.downstream = 20000,
      peaks = TRUE,
      annotation = FALSE,
      ranges = annot1,
      ranges.group.by = "type",
      ranges.title = "annotation",
      tile = TRUE)
    
    # Keep exons and introns info
    annot2 = annot_genes[annot_genes$type %in% c("hg38_genes_exons", "hg38_genes_introns")]
    plot2 = CoveragePlot(
      object = seurat_obj,
      assay = "peaks",
      region = region,
      extend.upstream = 40000,
      extend.downstream = 20000,
      peaks = TRUE,
      annotation = FALSE,
      ranges = annot2,
      ranges.group.by = "type",
      ranges.title = "annotation",
      tile = FALSE)
    
    plot2_tile = CoveragePlot(
      object = seurat_obj,
      assay = "peaks",
      region = region,
      extend.upstream = 40000,
      extend.downstream = 20000,
      peaks = TRUE,
      annotation = FALSE,
      ranges = annot2,
      ranges.group.by = "type",
      ranges.title = "annotation",
      tile = TRUE)
    
    # show and save
    print(plot1|plot2)
    print(plot1_tile|plot2_tile)
    ggsave(plot = plot1|plot2, width = 12, height = 6,
           filename = paste0(directory_output, list_gene[i], "_coverage_allcond.svg"))
     ggsave(plot = plot1_tile|plot2_tile, width = 12, height = 8,
           filename = paste0(directory_output, list_gene[i], "_coverage_tile_allcond.svg"))
    
  }

}

```

```{r, results="hide"}

list_to_keep = c("directory_output", "start_time", "color_code", 
                 "seurat_obj", "directory")
rm(list = setdiff(ls(), list_to_keep))
gc()

```



<br><br><br>



# Peaks distribution on chromosomes

<br>

We extracted the size of each chromosome (https://genome.ucsc.edu/cgi-bin/hgTracks?chromInfoPage=).   

```{r}

chrom_gr = data.frame(seqnames = c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", 
                                   "chr7", "chr8", "chr9", "chr10", "chr11", 
                                   "chr12", "chr13", "chr14", "chr15", "chr16", 
                                   "chr17", "chr18", "chr19", "chr20", "chr21", 
                                   "chr22", "chrX", "chrY"),
                      start = rep(0, 24),
                      end = c(248956422, 242193529, 198295559, 190214555,
                              181538259,170805979,159345973,145138636,138394717,
                              133797422, 135086622,133275309,114364328,107043718,
                              101991189,90338345,83257441,80373285,58617616,
                              64444167,46709983,50818468,156040895,57227415))

```

<br>

**Possibility 1 : use karyoplotR**    
We first tried to use karyoplotR as we did for bulkATACseq. The problem is that it uses only the list of peaks inside the Grange of the seurat object and not the matrix count. Here, we have the exact same list of peaks for all the conditions. In bulkATACseq, it was not a problem because every condition had its own set of peaks, the common list of peaks was done after for multiparametric and differential analysis only. For scATACseq with common list of peaks, we need to use the fragments repartition  so we try the *kpPlotBAMDensity* function.  
https://bernatgel.github.io/karyoploter_tutorial//Tutorial/PlotBAMCoverage/PlotBAMCoverage.html  
https://rdrr.io/bioc/karyoploteR/man/kpPlotBAMDensity.html  
"This function plots the actual base-level coverage and so it’s use is limited to relatively small regions. By default it will error if we try to plot a region larger than 1Mb, but this is configurable and with sufficient time and memory there should be no hard limit the size of the plot region, we simply should change the max.valid.region.size to the size of the region we want to plot. However, if we want to plot a global view of coverage along the whole genome, for example, we should consider using kpPlotBAMDensity instead, which is much faster."    
=> Not possible on our local computer, we can run it on IFb external cluster.  

```{r, karyoploteR IFB, eval=FALSE}

### code for IFB cluster

path_bam_2DG = paste0(directory, "data/2DG_possorted_bam.bam")
path_bam_AOA = paste0(directory, "data/AOA_possorted_bam.bam")
path_bam_CTRL = paste0(directory, "data/CTRL_possorted_bam.bam")
path_bam_DON = paste0(directory, "data/DON_possorted_bam.bam")

pp = getDefaultPlotParams(plot.type=1)
pp$ideogramheight = 20
pp$data1height = 500
pp$topmargin = 20

for(i in 1:nrow(chrom_gr)){
  
  chr = chrom_gr$seqnames[i]

  # find max value of peak to put the same axe on all coverages
  kp <- plotKaryotype(genome = "hg38", chromosomes = chr)
  temp <- kpPlotBAMDensity(kp, data=path_bam_2DG, window.size = 2e7)
  ymax_axis = temp$latest.plot$computed.values$max.density
  temp <- kpPlotBAMDensity(kp, data=path_bam_AOA, window.size = 2e7)
  if (ymax_axis < temp$latest.plot$computed.values$max.density) {ymax_axis = temp$latest.plot$computed.values$max.density}
  temp <- kpPlotBAMDensity(kp, data=path_bam_DON, window.size = 2e7)
  if (ymax_axis < temp$latest.plot$computed.values$max.density) {ymax_axis = temp$latest.plot$computed.values$max.density}
  temp <- kpPlotBAMDensity(kp, data=path_bam_CTRL, window.size = 2e7)
  if (ymax_axis < temp$latest.plot$computed.values$max.density) {ymax_axis = temp$latest.plot$computed.values$max.density}
  rm(kp)

  # draw coverage
  png(filename = paste0(directory_output, "coverage_kpbam_", chr, ".png"),
      width = 1000, height = 1400)
  
  kp <- plotKaryotype(genome = "hg38", chromosomes = chr, plot.type=1, plot.params = pp)
  kpAddBaseNumbers(kp, tick.dist=10000000, add.units=TRUE, cex=0.5)
  
  kp <- kpPlotBAMDensity(kp, data=path_bam_2DG, window.size = 5e5, col = color_code["2DG"], r0=0,  r1=0.23)
  kp <- kpPlotBAMDensity(kp, data=path_bam_AOA, window.size = 5e5, col = color_code["AOA"], r0=0.27, r1=0.48)
  kp <- kpPlotBAMDensity(kp, data=path_bam_DON, window.size = 5e5, col = color_code["DON"], r0=0.52,  r1=0.73)
  kp <- kpPlotBAMDensity(kp, data=path_bam_CTRL, window.size = 5e5, col = color_code["CTRL"], r0=0.77,  r1=1)
  
  kpAddLabels(kp, "2DG", label.margin = 0.01, side="right", col = color_code["2DG"], r0=0,  r1=0.15, cex=1)
  kpAddLabels(kp, "AOA", label.margin = 0.01, side="right", col = color_code["AOA"], r0=0.27, r1=0.4, cex=1)
  kpAddLabels(kp, "DON", label.margin = 0.01, side="right", col = color_code["DON"], r0=0.52,  r1=0.6, cex=1)
  kpAddLabels(kp, "CTRL", label.margin = 0.01, side="right", col = color_code["CTRL"], r0=0.77,  r1=0.9, cex=1)
  
  kpAxis(kp, ymax=ymax_axis, r0=0,  r1=0.23, cex=0.8)
  kpAxis(kp, ymax=ymax_axis, r0=0.27, r1=0.48, cex=0.8)
  kpAxis(kp, ymax=ymax_axis, r0=0.52,  r1=0.73, cex=0.8)
  kpAxis(kp, ymax=ymax_axis, r0=0.77,  r1=1, cex=0.8)
   
  dev.off()
  
}

```

<br>

**Possibility 2 : use CoveragePlot**    
Our problem here is that the chromosomes are too big to be represented in only one plot with the CoveragePlot function. It would require such a big amount of memory and end with an unreadable plot (see discussion : https://stackoverflow.com/questions/14629607/long-vector-plot-coverage-plot-in-r). We try to split the chromosomes plot in multiple windows but it's not really readable.    
Functional but extremely long and not very useful ! We should use it with a specific focus on a region to see something.   

```{r, CoveragePlot IFB, eval=FALSE}

list_plot_chr = list()

for (i in 1:nrow(chrom_gr)) {
  
  limits = chrom_gr[i,]
  chr = limits$seqnames
  chr_end = limits$end
  start = 0
  end = 0
  list_temp = list()
  
  for (j in 1:100) {
    
    end = end + round(chr_end/100)
    region = paste0(chr, "-", start, "-", end)

    list_temp[[j]] = CoveragePlot(object = seurat_obj,
                                  assay = "peaks",
                                  region = region,
                                  extend.upstream = 0,
                                  extend.downstream = 0,
                                  peaks = TRUE,
                                  tile = TRUE)
    
    start = end
    
  }
  
  list_plot_chr[[chr]] = list_temp
}

```



<br><br><br>



```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results = "hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

