---
title: "scATACseq_R_DWNS_ClustersInvestigation_Stat"
author: "Laëtitia Racine"
date: "2023-04-02"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(Signac)
library(knitr)
library(stringr)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(tidyverse)
library(ggrepel)
library(gridExtra)

```

```{r, Working directories and external script}

directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_ClustersInvestigation_Stat/"))
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_ClustersInvestigation_Stat/", current_date))
directory_output = paste0(directory,"exp/","scATACseq_R_DWNS_ClustersInvestigation_Stat/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scATACseq_R_NormalizationReduction/"))
seurat_merged = readRDS(paste0(dir, "/", "merged_seurat_qc_filter_annot_normreduc.rds"))

```


<br><br><br>


# Overview 

This code aims at exploring the clusters of cells and the repartition among conditions.

As a reminder : 

```{r, fig.width=16, fig.height=5}

seurat_merged = SetIdent(object = seurat_merged, value = "orig.ident")
ind_umap = list() 

for (i in 1:length(unique(seurat_merged$orig.ident))) {
  
  # Select condition to highlight
  highlight_cond = unique(seurat_merged$orig.ident)[i]
  color = unname(noquote(color_code[highlight_cond]))
  others_cond = setdiff(unique(seurat_merged$orig.ident), highlight_cond)
  others_cond_group = paste(others_cond, collapse=", ")
  
  # Sort cell according to their condition
  cell_highlight = unlist(CellsByIdentities(object = seurat_merged, idents = highlight_cond))
  cell_nohighlight = unlist(CellsByIdentities(object = seurat_merged, idents = others_cond))
  
  # Plot 
  plot_umap = DimPlot(object = seurat_merged,
                      group.by = "orig.ident",
                      reduction ="umap",
                      pt.size = 0.1,
                      cells.highlight = list(cell_highlight, cell_nohighlight),
                      cols.highlight = c(color,"grey")) +
    ggtitle(highlight_cond) +
    theme(legend.position = "none")
  
  ind_umap[[i]] = plot_umap
  
}

clust_umap = DimPlot(seurat_merged, 
                     reduction = "umap", 
                     group.by = "peaks_snn_res.0.25") + 
  ggtitle("Resolution_0.25") 

clust_umap | ind_umap[[1]] | ind_umap[[2]] | ind_umap[[3]] | ind_umap[[4]]

```


<br><br><br>


# Cluster's cell composition

First, we want to know how many cells of each conditions we have in each cluster.    
Documentation : https://github.com/satijalab/seurat/issues/738   

```{r}

# Cells' number in cluster per condition
tab_number = as.data.frame(table(seurat_merged@meta.data$peaks_snn_res.0.25, 
                                 seurat_merged@meta.data$orig.ident)) %>%
  dplyr::rename("cluster" = "Var1", "condition" = "Var2", "nbCells" = "Freq") %>%
  tidyr::pivot_wider(names_from = "cluster", values_from = "nbCells")
colnames(tab_number) = c("condition", paste0("cluster_", colnames(tab_number)[2:ncol(tab_number)]))
tab_number$total_cell_per_condition = rowSums(tab_number[,-1])

# Add total per cluster
temp = as.data.frame(t(data.frame(colSums(tab_number[,-1]))))
rownames(temp) = "total_cell_per_cluster"
temp = temp %>% tibble::rownames_to_column(var = "condition")
tab_number = rbind(tab_number, temp)

# Display data frame
tab_number %>%
  kable(caption = "Not normalized cells' distribution") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

<br>

We don't start with the same number of cells in each condition. A **normalization step** based on the initial number of cells per condition is required to be able to compare the conditions.
Documentation : https://stackoverflow.com/questions/54030049/divide-all-columns-by-a-chosen-column-using-mutate-all  

```{r}

# Divide per the number of cells, multiply by a 1000 factor and round
tab_number_norm = tab_number[-nrow(tab_number),]%>%
  dplyr::mutate_at(vars(-condition, -total_cell_per_condition), 
                   ~ round((./ total_cell_per_condition)*1000)) %>%
  dplyr::select(-total_cell_per_condition)
tab_number_norm$total_cell_per_condition = rowSums(tab_number_norm[,-1])

# Add total per cluster
temp = as.data.frame(t(data.frame(colSums(tab_number_norm[,-1]))))
rownames(temp) = "total_cell_per_cluster"
temp = temp %>% tibble::rownames_to_column(var = "condition")
tab_number_norm = rbind(tab_number_norm, temp)

# Display dataframe
tab_number_norm %>%
  kable(caption = "Normalized cells' distribution based on the initial number of cells in each condition.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

<br>

Cluster 0 : CTRL and DON same / 2DG a little less / AOA very less  
Cluster 1 : CTRL, DON and 2DG same / AOA less  
Cluster 2 : all close  
Cluster 3 : CTRL and 2DG same / DON very less / AOA very much   
Cluster 4 : CTRL and DON same / 2DG little much / AOA very much   
Cluster 5 : CTRL and DON same / 2DG and AOA same    
Cluster 6 : all close   
Cluster 7 : CTRL and 2DG same / DON little less / AOA little much   
Cluster 8 : CTRL and 2DG same / DON little less / AOA very much   
Cluster 9 : CTRL, DON and 2DG same / AOA much   
Cluster 10 : 2DG < DON < CTRL < AOA  
Cluster 11 : CTRL, 2DG and AOA close / DON little less   
Cluster 12 : 2DG and DON close / CTRL less / AOA much  
Cluster 13 : CTRL and 2DG close / DON less / AOA much  
Cluster 14 : CTRL, DON and 2DG same / AOA much  
Cluster 15 : DON < 2DG < CTRL < AOA  
  
<br>

Then we study the data in two alternative ways :  
- the condition composition : percentages of each cluster for a condition (tab_percent_cond)  
- the cluster composition : percentages of cells from each condition for a cluster (tab_percent_clust)    
    
```{r}

# Cells' percentage in cluster per condition
tab_percent_cond = tab_number_norm %>% dplyr::filter(condition != "total_cell_per_cluster")
tab_percent_cond = tab_percent_cond %>%
  dplyr::mutate_at(vars(-condition, -total_cell_per_condition), 
                   ~ round((./ total_cell_per_condition)*100)) %>%
  dplyr::select(-total_cell_per_condition)

tab_percent_cond %>%
  kable(caption = "Normalized cells' percentage distribution per condition.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

# Cells' percentage in condition per cluster 
tab_percent_clust = tab_number_norm %>% dplyr::select(-total_cell_per_condition)
tab_percent_clust = as.data.frame(t(tab_percent_clust))
colnames(tab_percent_clust) = tab_percent_clust[1,]
tab_percent_clust = tab_percent_clust[-1,] %>% 
  mutate_if(is.character, ~as.numeric(.)) %>%
  rownames_to_column(var = "cluster")
tab_percent_clust = tab_percent_clust %>%
  dplyr::mutate_at(vars(-cluster, -total_cell_per_cluster), 
                   ~ round((./ total_cell_per_cluster)*100)) %>%
  dplyr::select(-total_cell_per_cluster)

tab_percent_clust %>%
  kable(caption = "Normalized cells' percentage distribution per cluster.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```
  
<br>

We represent the table on plots.
  
```{r, bp plots, fig.width = 14, fig.height = 8}

###########
# Cluster #
###########

temp_clust = tab_percent_clust %>% tidyr::pivot_longer(cols = 2:ncol(tab_percent_clust),
                                                       names_to = "condition",
                                                       values_to = "percentage")
temp_clust$cluster =  factor(temp_clust$cluster,
                             levels = c("cluster_0", "cluster_1", "cluster_2", 
                                        "cluster_3", "cluster_4", "cluster_5",
                                        "cluster_6", "cluster_7", "cluster_8",
                                        "cluster_9", "cluster_10", "cluster_11", 
                                        "cluster_12", "cluster_13", "cluster_14",
                                        "cluster_15"))

bp_clust = ggplot(temp_clust, aes(x=cluster, y=percentage, fill=condition))+
  geom_bar(width = 0.8, stat = "identity") +
  scale_fill_manual(values = color_code, limits = force) +
  coord_flip()
bp_clust

#############
# Condition #
#############

temp_cond = tab_percent_cond %>% tidyr::pivot_longer(cols = 2:ncol(tab_percent_cond),
                                    names_to = "cluster",
                                    values_to = "percentage")
temp_cond$cluster =  factor(temp_cond$cluster,
                            levels = c("cluster_0", "cluster_1", "cluster_2", 
                                       "cluster_3", "cluster_4", "cluster_5",
                                       "cluster_6", "cluster_7", "cluster_8",
                                       "cluster_9", "cluster_10", "cluster_11", 
                                       "cluster_12", "cluster_13", "cluster_14",
                                       "cluster_15"))

bp_cond = ggplot(temp_cond, aes(x=condition, y=percentage, fill=cluster))+
  geom_bar(width = 0.8, stat = "identity", color="#737373") +
  coord_flip()
bp_cond

```

```{r, pie plots1, fig.width = 14, fig.height = 24}

###########
# Cluster #
###########

clust_list = tab_percent_clust$cluster
list_clust_pie = list()

for (i in 1:length(clust_list)) {
  
  temp2 = temp_clust %>% 
    dplyr::filter(cluster == clust_list[i]) %>% 
    dplyr::mutate(csum = rev(cumsum(rev(percentage))), 
                  pos = percentage/2 + lead(csum, 1),
                  pos = if_else(is.na(pos), percentage/2, pos))
  
  pie = ggplot(temp2, aes(x="", y=percentage, fill=condition))+
    geom_col(width = 1, color = 1) +
    coord_polar("y", start=0) + 
    scale_fill_manual(values = color_code, limits = force) +
    geom_label_repel(data = temp2,
                     aes(y = pos, label = paste0(percentage, "%")),
                     size = 4, nudge_x = 1, show.legend = FALSE, 
                     segment.color = "#999999",
                     col = "black") +
    theme_minimal() +
    ggtitle(label = clust_list[i]) +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(),
          panel.grid=element_blank(),
          axis.ticks = element_blank(),
          plot.title=element_text(size=14, face="bold")
    )
  
  list_clust_pie[[i]] = pie
}

do.call("grid.arrange", c(list_clust_pie, ncol = 3))

```

```{r, pie plots2, fig.width = 14, fig.height = 12}

#############
# Condition #
#############

cond_list = tab_percent_cond$condition
list_cond_pie = list()

for (i in 1:length(cond_list)) {

  temp2 = temp_cond %>% 
    dplyr::filter(condition == cond_list[i]) %>%
    dplyr::mutate(cluster = sort(cluster)) %>%
    dplyr::mutate(csum = rev(cumsum(rev(percentage))),
                  pos = percentage/2 + lead(csum, 1),
                  pos = if_else(is.na(pos), percentage/2, pos))

  pie = ggplot(temp2, aes(x="", y=percentage, fill=cluster))+
    geom_col(width = 1, color = 1) +
    coord_polar("y", start=0) +
    geom_label_repel(data = temp2,
                     aes(y = pos, label = paste0(percentage, "%")),
                     size = 4, nudge_x = 1, show.legend = FALSE,
                     segment.color = "#999999",
                     col = "black") +
    theme_minimal() +
    ggtitle(label = cond_list[i]) +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(),
          panel.grid=element_blank(),
          axis.ticks = element_blank(),
          plot.title=element_text(size=14, face="bold")
    )

  list_cond_pie[[i]] = pie

}

do.call("grid.arrange", c(list_cond_pie, ncol = 2))

```

```{r, fig.show="hide"}

write.csv2(tab_number, paste0(directory_output, "nbcell_perclust_percond.csv"))
write.csv2(tab_number_norm, paste0(directory_output, "nbcell_perclust_percond_norm.csv"))
write.csv2(tab_percent_clust, paste0(directory_output, "percent_cond_inclust.csv"))
write.csv2(tab_percent_cond, paste0(directory_output, "percent_clust_incond.csv"))

ggsave(plot = bp_clust | bp_cond, width = 12, height = 6,
       filename = paste0(directory_output, "barplot_clust_cellcomp.svg"))
ggsave(plot = do.call("grid.arrange", c(list_clust_pie, ncol = 3)), width = 16, 
       height=20, filename = paste0(directory_output, "pieplot1_clust_cellcomp.svg"))
ggsave(plot = do.call("grid.arrange", c(list_cond_pie, ncol = 2)), width=10, 
       height=10, filename = paste0(directory_output, "pieplot2_clust_cellcomp.svg"))

```

```{r}

list_to_keep = c("directory_output", "start_time", "color_code",
                 "seurat_merged")
rm(list = setdiff(ls(), list_to_keep))
gc()

```



<br><br><br>



# Cluster's peaks composition

Documentation :    
https://github.com/satijalab/seurat/issues/1053    
   

```{r, Change DefaultAssay and choose resolution}

# Change back to working with peaks instead of gene activities
DefaultAssay(seurat_merged) <- 'peaks'

# Choose working resolution
res = "res.0.25"
meta_col = paste0("peaks_snn_", res)
Idents(seurat_merged) = (seurat_merged@meta.data)[,meta_col]

```

```{r, Subset seurat_merged based on cluster}

# Subset seurat obj to create one object per cluster
for (i in 1:length(levels((seurat_merged@meta.data)[,meta_col])) ) {

  # Extract cluster number
  clust = levels((seurat_merged@meta.data)[,meta_col])[i]
  print(paste("Cells extraction for clust ", clust))
  # Create subset of seurat object with cells from clust x
  expr = FetchData(object = seurat_merged, vars = meta_col)
  seurat_clust = seurat_merged[, which(expr == clust)]                       # ligne qui prend du temps
  # Check if the subset was correctly done
  print(table((seurat_clust@meta.data)[,meta_col]))

  # Extract peak count matrix of clust x from general seurat object
  print("Matrix count extraction")
  tab_clust = as.data.frame(GetAssayData(seurat_clust, slot = "data"))    # ligne qui prend du temps

  # Save object
  saveRDS(tab_clust, file = paste0(directory_output, "/tab_dim2-50_", res, 
                                   "_matrix_count_clust", clust, ".rds"))
  saveRDS(seurat_clust, file = paste0(directory_output,  "/seurat_clust", 
                                      clust, "_dim2-50_", res, ".rds"))

}

```

```{r, Peaks distribution per cell across clusters - tabs creation}

# Create dataframe with summary of peaks per cell and/or per annotation
tab_distripeaks_clust = data.frame()
tab_bulkpeaks_clust = data.frame()
  
for ( i in 1:length(levels((seurat_merged@meta.data)[,meta_col])) ) {
  
  # Extract cluster number
  clust = levels((seurat_merged@meta.data)[,meta_col])[i]
  print(paste("Cells extraction for clust ", clust))
  
  # Load subset of seurat object with cells from clust x and corresponding matrix count
  seurat_clust = readRDS(paste0(directory_output, "seurat_clust", clust, 
                                "_dim2-50_", res, ".rds"))
  range_clust = seurat_clust@assays$peaks@ranges
  print("Load matrix count")
  tab_clust = readRDS(paste0(directory_output, "tab_dim2-50_", res,
                             "_matrix_count_clust", clust, ".rds"))
  # Transform matrix count into logical
  tab_clust = tab_clust %>% mutate_all(as.logical)
  
  # Extract peaks_name according to genomic annotation
  print("Calcul peaks vs genomic annotation")
  peaks_name_list = list(
    total_peaks = range_clust$peak_name,
    peaks_promoter = (range_clust[range_clust$hg38_genes_promoters == TRUE])$peak_name,
    peaks_UTR3P = (range_clust[range_clust$hg38_genes_3UTRs == TRUE])$peak_name,
    peaks_UTR5P = (range_clust[range_clust$hg38_genes_5UTRs == TRUE])$peak_name,
    peaks_CpG = (range_clust[range_clust$hg38_cpg_islands == TRUE])$peak_name,
    peaks_enhancer = (range_clust[range_clust$hg38_enhancers_fantom == TRUE])$peak_name,
    peaks_gene = (range_clust[range_clust$hg38_genes_1to5kb == TRUE])$peak_name,
    peaks_cds = (range_clust[range_clust$hg38_genes_cds == TRUE])$peak_name,
    peaks_exons = (range_clust[range_clust$hg38_genes_exons  == TRUE])$peak_name,
    peaks_introns = (range_clust[range_clust$hg38_genes_introns == TRUE])$peak_name,
    peaks_intergenic = (range_clust[range_clust$hg38_genes_intergenic == TRUE])$peak_name)

  # Calculate number of peaks total per annotation and per cell per annotation
  temp_distripeaks = lapply(peaks_name_list, function(x) {
    tab_filtered = tab_clust %>% dplyr::filter(row.names(tab_clust) %in% unlist(x))
    tab_calc = data.frame(colSums(tab_filtered)) 
    tab_calc = tab_calc %>% tibble::rownames_to_column(var = "cell")})
  temp_distripeaks = bind_rows(temp_distripeaks, .id = "peaks_category") %>% 
    dplyr::mutate(clust = clust)
  colnames(temp_distripeaks) = c("peaks_category", "cell", "total_nbpeaks_percell", "clust")
  
  temp_bulkpeaks = tibble(annotation = names(peaks_name_list), 
                              total_nbpeaks = as.numeric(lapply(peaks_name_list, length))) %>% 
    dplyr::mutate(clust = clust, .before = "total_nbpeaks")
  
  # Bind in general tab
  tab_bulkpeaks_clust = rbind(tab_bulkpeaks_clust, temp_bulkpeaks)
  tab_distripeaks_clust = rbind(tab_distripeaks_clust, temp_distripeaks)

}

tab_distripeaks_clust$clust = factor(tab_distripeaks_clust$clust,
                                    levels = c("0", "1", "2", "3", "4", "5",
                                               "6", "7", "8", "9", "10", "11", 
                                               "12", "13", "14", "15"))
# Add statistical information
tab_distripeaks_stat = tab_distripeaks_clust %>%
  dplyr::group_by(clust, peaks_category) %>%
  dplyr::mutate(nbcells_clust = n(),
                median_nbpeaks_percell = median(total_nbpeaks_percell),
                mean_nbpeaks_percell = round(mean(total_nbpeaks_percell)),
                max_nbpeaks_percell = max(total_nbpeaks_percell),
                min_nbpeaks_percell = min(total_nbpeaks_percell),
                var_nbpeaks_percell = round(sd(total_nbpeaks_percell))) %>%
  dplyr::mutate(percent_cell_clust = round((nbcells_clust/ncol(seurat_merged))*100)) %>%
  dplyr::ungroup() %>%
  dplyr::select(peaks_category, clust, nbcells_clust, percent_cell_clust, median_nbpeaks_percell, 
                mean_nbpeaks_percell, max_nbpeaks_percell, min_nbpeaks_percell, var_nbpeaks_percell) %>%
  dplyr::distinct()


```

```{r, fig.width = 16, fig.height = 16}

med_plot = ggplot(tab_distripeaks_stat, aes(x=clust, y = median_nbpeaks_percell, fill = clust)) +
  geom_col() +
  facet_wrap(facets = .~peaks_category) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

mean_plot = ggplot(tab_distripeaks_stat, aes(x=clust, y = mean_nbpeaks_percell, fill = clust)) +
  geom_col() +
  facet_wrap(facets = .~peaks_category) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

max_plot = ggplot(tab_distripeaks_stat, aes(x=clust, y = max_nbpeaks_percell, fill = clust)) +
  geom_col() +
  facet_wrap(facets = .~peaks_category) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

min_plot = ggplot(tab_distripeaks_stat, aes(x=clust, y = min_nbpeaks_percell, fill = clust)) +
  geom_col() +
  facet_wrap(facets = .~peaks_category) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

var_plot = ggplot(tab_distripeaks_stat, aes(x=clust, y = var_nbpeaks_percell, fill = clust)) +
  geom_col() +
  facet_wrap(facets = .~peaks_category) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

(med_plot | mean_plot) /
  (max_plot | min_plot | var_plot)

```

```{r, fig.width = 20, fig.height = 18}

plot = ggplot(tab_distripeaks_clust, aes(x = total_nbpeaks_percell)) +
  geom_histogram(binwidth = 100, aes(col = clust)) +
  ylab("Number of cells") +
  xlab("Number of peaks") +
  facet_grid(peaks_category~clust, scales = "free_y") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90))
plot

```

```{r, fig.show="hide"}

write_csv2(tab_bulkpeaks_clust, 
           file = paste0(directory_output, "/tab_dim2-50", res, "_bulkpeaks.csv"))
write_csv2(tab_distripeaks_clust, 
           file = paste0(directory_output,"/tab_dim2-50", res, "_distripeaks.csv"))
write_csv2(tab_distripeaks_stat, 
           file = paste0(directory_output, "/tab_dim2-50", res, "_distripeaks_stat.csv"))
ggsave(plot = (med_plot | mean_plot) / (max_plot | min_plot | var_plot), width = 16,
       height = 10, filename = paste0(directory_output, "stat_peakscomp.svg"))
ggsave(plot, width=16, height=14, filename=paste0(directory_output, "distri_peakscomp.svg"))

```



<br><br><br>



```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results = "hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

