---
title: "scATACseq_R_DWNS_ClustersInvestigation_DA"
author: "Laëtitia Racine"
date: "2023-04-07"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(Signac)
library(knitr)
library(stringr)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(GenomicRanges)
library(tibble)
library(pheatmap)
# motif analysis
library(JASPAR2020)
library(TFBSTools)
library(BSgenome.Hsapiens.UCSC.hg38)
library(gridExtra)
# GO analysis
library(org.Hs.eg.db)
library(clusterProfiler)

```

```{r, Working directories and external script}

directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_ClustersInvestigation_DA/"))
dir.create(path = paste0(directory,"exp/","scATACseq_R_DWNS_ClustersInvestigation_DA/", 
                         current_date))
directory_output = paste0(directory,"exp/","scATACseq_R_DWNS_ClustersInvestigation_DA/", 
                          current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scATACseq_R_NormalizationReduction/"))
seurat_obj = readRDS(paste0(dir, "/", "merged_seurat_qc_filter_annot_normreduc.rds"))

```



<br><br><br>



# Overview 

This code aims at exploring the clusters of cells by performing **differential accessibility analysis**.  
"To find differentially accessible regions between clusters of cells, we can perform a differential accessibility (DA) test. We utilize logistic regression for DA, as suggested by Ntranos et al. 2018 for scRNA-seq data, and add the total number of fragments as a latent variable to mitigate the effect of differential sequencing depth on the result. For sparse data (such as scATAC-seq), we find it is often necessary to lower the min.pct threshold in FindMarkers() from the default (0.1, which was designed for scRNA-seq data)."  
  
As a reminder : 

```{r, fig.width=16, fig.height=4}

seurat_obj = SetIdent(object = seurat_obj, value = "orig.ident")
ind_umap = list() 

for (i in 1:length(unique(seurat_obj$orig.ident))) {
  
  # Select condition to highlight
  highlight_cond = unique(seurat_obj$orig.ident)[i]
  color = unname(noquote(color_code[highlight_cond]))
  others_cond = setdiff(unique(seurat_obj$orig.ident), highlight_cond)
  others_cond_group = paste(others_cond, collapse=", ")
  
  # Sort cell according to their condition
  cell_highlight = unlist(CellsByIdentities(object = seurat_obj, idents = highlight_cond))
  cell_nohighlight = unlist(CellsByIdentities(object = seurat_obj, idents = others_cond))
  
  # Plot 
  plot_umap = DimPlot(object = seurat_obj,
                      group.by = "orig.ident",
                      reduction ="umap",
                      pt.size = 0.1,
                      cells.highlight = list(cell_highlight, cell_nohighlight),
                      cols.highlight = c(color,"grey")) +
    ggtitle(highlight_cond) +
    theme(legend.position = "none")
  
  ind_umap[[i]] = plot_umap
  
}

clust_umap = DimPlot(seurat_obj, 
                     reduction = "umap", 
                     group.by = "peaks_snn_res.0.25") + 
  ggtitle("Resolution_0.25") 

clust_umap | ind_umap[[1]] | ind_umap[[2]] | ind_umap[[3]] | ind_umap[[4]]

```


<br><br><br>


# Extract list of DA peaks for each cluster compare to all others

Compare each cluster against all others and output the peaks that are differentially expressed/present.  
This part of the code require a huge amount of memory and is extremely slow. We can only perform it on external cluster and it takes really long time. We did it on IFB (slurm job on long partition with memory 100GB accessible) and we saved it in the data folder.     
"For sparse data (such as scATAC-seq), we find it is often necessary to lower the min.pct threshold in FindMarkers() from the default (0.1, which was designed for scRNA-seq data)."  
```{r, Markers on IFB, eval=FALSE}

################################
# Possibility 1 : FindAllMarkers
################################

DefaultAssay(seurat_obj) <- 'peaks'
Idents(seurat_obj) = "peaks_snn_res.0.25"

clust_markers = FindAllMarkers(object = seurat_obj,
                      assay = "peaks",
                      features = NULL, #use all genes
                      logfcthreshold = 0.25,
                      test.use = "LR", # not default
                      slot = "data",
                      min.pct = 0.05, # not default
                      min.diff.pct = -Inf,
                      node = NULL,
                      verbose = TRUE,
                      only.pos = TRUE,
                      max.cells.per.ident = Inf,
                      random.seed = 42, # not default
                      latent.vars = NULL,
                      min.cells.feature = 3,
                      min.cells.group = 3,
                      mean.fxn = NULL,
                      fc.name = NULL,
                      base = 2,
                      return.thresh = 2,
                      densify = FALSE)

saveRDS(object = clust_markers, paste0(directory_output,
                              "clust_markers_FAM_",
                              current_date,
                              ".rds"))

#############################
# Possibility 2 : FindMarkers
#############################

## Example with cluster 0 => do it for each cluster

DefaultAssay(seurat_obj) <- 'peaks'
Idents(seurat_obj) = "peaks_snn_res.0.25"

clust_markers_choice = FindMarkers(object = seurat_obj,
                                   ident.1 = 0,
                                   ident.2 = setdiff(seq(0,15), 0),
                                   group.by = NULL,
                                   subset.ident = NULL,
                                   assay = "peaks",
                                   slot = "data",
                                   reduction = NULL,
                                   features = NULL, #use all genes
                                   logfc.threshold = 0.25,
                                   test.use = "LR",
                                   min.pct = 0.05,
                                   min.diff.pct = -Inf,
                                   verbose = TRUE,
                                   only.pos = TRUE,
                                   max.cells.per.ident = Inf,
                                   random.seed = 42,
                                   latent.vars = NULL,
                                   min.cells.feature = 3,
                                   min.cells.group = 3,
                                   mean.fxn = NULL,
                                   fc.name = NULL,
                                   base = 2,
                                   densify = FALSE)

saveRDS(clust_markers_choice, paste0(directory_output,
                                  "clust_markers_",
                                  current_date,
                                  ".rds"))

```

<br>

We decided to use FindMarkers instead of FindAllMarkers to be able to parrallelize on multiple cores on the cluster. We showed previoulsy that with the right parameters, they give the same results.  
```{r}

# Load the clust_markers from IFBcluster
files_clust_markers = list.files(paste0(directory, "data/scATACseq/"), pattern = "clust_markers_", full.names = TRUE)
list_clust_markers = lapply(files_clust_markers, readRDS)
list_clust_markers = lapply(list_clust_markers, function(x) { x = x %>% rownames_to_column("peak_name")})
names(list_clust_markers) = paste0("cluster_", str_extract(files_clust_markers, pattern = "(?<=markers_)[:digit:]+(?=_2023)"))
clust_markers = bind_rows(list_clust_markers, .id = 'cluster')

cat("Number of DA peaks for each cluster : \n")
table(clust_markers$cluster)

clust_markers = clust_markers %>%
  tidyr::separate(col = "peak_name", into = c("seqnames", "start", "end"), remove=FALSE)

saveRDS(clust_markers, paste0(directory_output, "clust_markers"))

```

<br>

See branch FAM_FM_IFBcluster on git project to access .R and .sh scripts used for this part.  

```{r, results="hide"}

list_to_keep = c("directory_output", "start_time", "seurat_obj", 
                 "clust_markers", "pic_last_dir", "directory")
rm(list = setdiff(ls(), list_to_keep))
gc()

```



<br><br><br>



# Associated gene for each peak

https://www.biostars.org/p/478325/  
Note : the gene_id and symbol column of annot_hg38 table contains NA when the peak doesn't correspond to a gene (enhancer, intergenic...). However, the overlap doesn't work when there are NA in the initial annot grange. We decide to replace them by the annotation type.  
```{r, results="hide"}

# Extract localisation of hg38 annotation (see annotationsFiles_createGrange.Rmd)
dir = pic_last_dir(paste0(directory, "exp/AnnotationsFiles_createGrange/"))
annot_hg38 = readRDS(paste0(dir, "/", "hg38_annotations_gr.rds"))

annot_df = data.frame(annot_hg38) %>%
  dplyr::select(seqnames, start, end, gene_id, symbol, type) 
annot_df = annot_df %>%
  dplyr::mutate(gene_id = ifelse(is.na(gene_id), type, gene_id)) %>%
  dplyr::mutate(symbol = ifelse(is.na(symbol), type, symbol))

rm(annot_hg38)
gc()

```

<br>

Then, we transform tables into dataframes to use findOverlaps function and we correlate the table to have the cluster information and the gene/annotation information in the same final dataframe.
```{r}

# transform tables into granges
annot_gr = makeGRangesFromDataFrame(annot_df, keep.extra.columns = TRUE)
clust_gr = makeGRangesFromDataFrame(clust_markers, keep.extra.columns = TRUE)

# find overlaps between granges 
overlaps = findOverlaps(query = annot_gr, subject = clust_gr, ignore.strand = TRUE)

# add corresponding columns from query granges into subject granges
mcols(clust_gr)$gene_id = NA
mcols(clust_gr)$symbol = NA
mcols(clust_gr)$type = NA
mcols(clust_gr)$gene_id[subjectHits(overlaps)] <- mcols(annot_gr)$gene_id[queryHits(overlaps)]
mcols(clust_gr)$symbol[subjectHits(overlaps)] <- mcols(annot_gr)$symbol[queryHits(overlaps)]
mcols(clust_gr)$type[subjectHits(overlaps)] <- mcols(annot_gr)$type[queryHits(overlaps)]

# final table
tab_clust = data.frame(clust_gr)
table(tab_clust$type, tab_clust$cluster)

# save 
write.csv2(tab_clust, paste0(directory_output, "tab_clust_DApeaks_annotation.csv"), row.names = FALSE)

```

```{r, results="hide"}

list_to_keep = c("directory_output", "start_time", "seurat_obj", 
                 "tab_clust", "directory")
rm(list = setdiff(ls(), list_to_keep))
gc()

```



<br><br><br>



# Select relevant peaks in the list (top 20) to plot on heatmap

```{r, Select relevant markers}

# filter on p value adjusted
cat("We remove markers with p_val_adj superior to 0.05.")
tab_clust_relevant = tab_clust %>% dplyr::filter(p_val_adj < 0.05)

# Extract top 20 of positive markers based on avg_log2FC only 
cat("We keep the positive markers with the biggest average_log2FC for each cluster.")
top20_pos = tab_clust_relevant %>%
  dplyr::filter(avg_log2FC >= 0) %>%
  dplyr::group_by(cluster) %>%
  top_n(n=20, wt = avg_log2FC) %>%
  dplyr::select(cluster,symbol, p_val_adj, everything()) %>%
  ungroup()

# positive markers : calcul the difference between pct1 and pct2
cat("We filter the positive markers to keep those with a difference > to 0.25 between pct.1 and pct.2.")
top20_pos_filtered = tab_clust_relevant %>% 
  dplyr::filter(avg_log2FC >= 0) %>%
  dplyr::mutate(diff = pct.1 - pct.2) %>%
  dplyr::filter(diff > 0.25) %>%
  dplyr::group_by(cluster) %>%
  top_n(n=20, wt = avg_log2FC) %>%
  dplyr::select(cluster, symbol, p_val_adj, everything()) %>%
  dplyr::select(-diff) %>%
  ungroup()

# Summary markers
summary_clust = data.frame(cluster = 0:15, 
                           nb_pos_markers = as.vector(table(top20_pos$cluster)),
                           nb_pos_markers_filtered = as.vector(table(top20_pos_filtered$cluster)))

summary_clust %>% 
  kable() %>% 
  kable_styling()


# save 
write.csv2(top20_pos, paste0(directory_output, "top20_pos.csv"), row.names = FALSE)
write.csv2(top20_pos_filtered, paste0(directory_output, "top20_pos_filtered.csv"), row.names = FALSE)

```

<br>

As we don't have data in scale.data slot, we can't use DoHeatmap function.  
https://davetang.org/muse/2018/05/15/making-a-heatmap-in-r-with-the-pheatmap-package/  
https://github.com/tidyverse/stringr/issues/103  
  
```{r, Heatmap,  fig.height = 26, fig.width = 14}

# Randomly downsample 200 cells of each cluster
cells_df = as.data.frame(seurat_obj@meta.data)  %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::select(cell, peaks_snn_res.0.25) %>%
  dplyr::rename("cluster" = "peaks_snn_res.0.25")
# table(cells_df$cluster)

cells_to_keep = c()
for(i in 1:length(unique(cells_df$cluster))) {
  temp = (cells_df %>% dplyr::filter(cluster==unique(cells_df$cluster)[i]))$cell
  if(length(temp) > 100){ temp = sample(temp,100) }
  cells_to_keep = c(cells_to_keep, temp)
}

cells_df = cells_df %>% 
  dplyr::filter(cell %in% cells_to_keep) %>%
  tibble::column_to_rownames("cell")

# Extract names of peaks to keep
peaks_df = top20_pos_filtered %>% 
  dplyr::select(symbol, peak_name) %>%
  dplyr::distinct()
peaks_to_keep = peaks_df$peak_name

# Filter matrix to keep only wanted cells and peaks
mat = as.data.frame(seurat_obj@assays$peaks@data[peaks_to_keep, cells_to_keep])
mat = mat %>% rownames_to_column("peak_name")
mat$peak_name = str_replace_all(string = mat$peak_name, pattern = "[.]", replacement = "-" )
mat = left_join(mat, peaks_df, by = "peak_name") 

mat_ht = mat %>% 
  dplyr::mutate(peak_name = paste0(symbol, " - ", peak_name)) %>%
  dplyr::select(-symbol) %>% 
  tibble:::column_to_rownames("peak_name") 

# scale the rows
cal_z_score <- function(x){(x - mean(x)) / sd(x)}
mat_ht <- t(apply(mat_ht, 1, cal_z_score))

# Draw plot
ht_all = pheatmap(mat = mat_ht, 
                  annotation_col = cells_df,
                  show_colnames = FALSE,
                  cluster_cols = FALSE,
                  cluster_rows = FALSE,
                  fontsize_rows = 6)

mat_filt = mat %>%
  dplyr::filter(!symbol %in% str_subset(symbol, pattern = "hg38")) %>% 
  dplyr::mutate(peak_name = paste0(symbol, " - ", peak_name)) %>%
  dplyr::select(-symbol) %>% 
  tibble:::column_to_rownames("peak_name") 
mat_filt <- t(apply(mat_filt, 1, cal_z_score))

ht_filt = pheatmap(mat = mat_filt, 
                   annotation_col = cells_df,
                   show_colnames = FALSE,
                   cluster_cols = FALSE,
                   cluster_rows = FALSE,
                   fontsize_rows = 6)

```

```{r, results="hide"}

save_pheatmap_png <- function(x, filename, width=1200, height=1200, res = 150) {
  png(filename, width = width, height = height, res = res)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}
 
save_pheatmap_png(ht_all, paste0(directory_output, "heatmap_DApeaks_clust_all.png"))
save_pheatmap_png(ht_filt, paste0(directory_output, "heatmap_DApeaks_clust_genes.png"))

```

```{r, results="hide"}

list_to_keep = c("directory_output", "start_time", "seurat_obj", 
                 "tab_clust_relevant", "directory")
rm(list = setdiff(ls(), list_to_keep))
gc()

```


<br><br><br>


# GO on list of peaks differentially accessible between clusters

```{r, fig.show = "hide"}

markers = tab_clust_relevant %>% dplyr::select(cluster, symbol, gene_id)
markers = markers %>% dplyr::filter(!symbol %in% str_subset(symbol, "hg38"))

list_GO_analysis = list()

for (i in 1:length(unique(tab_clust_relevant$cluster))) {

  cat("Cluster", i-1, "\n")
  markers_clust = (markers %>% dplyr::filter(cluster == i-1))$gene_id

  ego = enrichGO(gene = markers_clust,
                 OrgDb = "org.Hs.eg.db",
                 ont = "ALL",
                 pAdjustMethod = "BH",
                 pvalueCutoff = 0.05,
                 qvalueCutoff = 0.2,
                 readable = TRUE,
                 keyType = "ENTREZID")

  ego_df = ego@result %>% dplyr::mutate(cluster = i-1)
  
  list_GO_analysis[[paste0("clust_", i-1)]]$ego = ego_df
  
  if (nrow(ego_df) !=0) { 
    
    plot = dotplot(ego, showCategory = 25) +
      facet_grid(ONTOLOGY ~ ., scales="free") +
      ggtitle(paste0("DApeaks - Closest genes enrichment - Clust", i-1)) +
      theme(axis.text.y = element_text(size = 8))
    
    list_GO_analysis[[paste0("clust_", i-1)]]= list(dotplot = plot,
                                                    ego = ego_df)
    
  } else {
      
    list_GO_analysis[[paste0("clust_", i-1)]] = list(ego = ego_df)
    
  }

}

saveRDS(list_GO_analysis, paste0(directory_output, "list_GO_clust.rds"))

```

```{r, fig.width=16, fig.height=40}

list = unlist(list_GO_analysis, recursive = FALSE)
list_dotplot = list[str_subset(names(list), "dotplot")]
do.call("grid.arrange", c(list_dotplot, ncol = 3))

list_tab = list[str_subset(names(list), "ego")]
tab_ego = do.call("rbind", list_tab)
tab_ego %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height="250px")

```

```{r, fig.show="hide"}

ggsave(plot = do.call("grid.arrange", c(list_dotplot, ncol=4)), width=40,
       height=30, filename=paste0(directory_output, "GO_DApeaks_clust.svg"))

write.csv2(tab_ego, paste0(directory_output, "tab_GO_DApeaks_clust.csv"), row.names=FALSE)

```

```{r, results="hide"}

list_to_keep = c("directory_output", "start_time", "seurat_obj", 
                 "tab_clust_relevant", "directory")
rm(list = setdiff(ls(), list_to_keep))
gc()

```



<br><br><br>



# Motif analysis on list of peaks differentially accessible between clusters

Documentation :   
https://stuartlab.org/signac/articles/motif_vignette.html  
https://www.rdocumentation.org/packages/TFBSTools/versions/1.10.3/topics/getMatrixSet  
https://stuartlab.org/signac/reference/addmotifs  

```{r}

# Get a list of motif position frequency matrices from the JASPAR database
pfm <- getMatrixSet(
  x = JASPAR2020,
  opts = list(collection = "CORE", species = "Homo sapiens", 
              tax_group = 'vertebrates', all_versions = FALSE)
)

# add motif information
seurat_obj <- AddMotifs(
  object = seurat_obj,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  pfm = pfm
)

```

<br>

This part of the code with ChromVar is quite long. 
```{r, eval=FALSE}

# Compute a per-cell motif activity score !!extremely slow!!
seurat_obj <- RunChromVAR(
  object = seurat_obj,
  genome = BSgenome.Hsapiens.UCSC.hg38
)

saveRDS(seurat_obj, paste0(directory_output, "seurat_motif.rds"))

```

```{r}

# To gain time
seurat_obj = readRDS(paste0(directory, "exp/scATACseq_R_DWNS_ClustersInvestigation_DA/seurat_motif.rds"))

```

<br>

Search for DNA motifs that are overrepresented in a set of peaks that are differentially accessible between clusters.  
```{r}

list_enrich_clust = list()

for (i in 1:length(unique(tab_clust_relevant$cluster))) {
  
  # extract list of peaks
  peaks_clust = tab_clust_relevant %>% dplyr::filter(cluster == i-1)
  peaks_clust = peaks_clust$peak_name

  # test enrichment
  enriched.motifs <- FindMotifs(object = seurat_obj,
                                features = peaks_clust)
  enriched.motifs$clust = paste0("clust_", i-1)
    
  # plot more enriched motifs
  plot = MotifPlot(object = seurat_obj,
                   motifs = head(rownames(enriched.motifs))) +
    ggtitle(paste0("clust_", i-1))
  
  # save in a list 
  list_enrich_clust[[paste0("cluster_", i-1)]] = list(tab_motif = enriched.motifs,
                                                      plot_motif = plot)
  
}

```

```{r, fig.width=20, fig.height=22}

list = unlist(list_enrich_clust, recursive = FALSE)
list_tab = list[str_detect(names(list), "tab")]
list_plot = list[str_detect(names(list), "plot")]

glob_enrich_tab = do.call("rbind", list_tab)
glob_enrich_tab[1:10,] %>%
  kable(caption = "First 10 rows of the table") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "250px")

do.call("grid.arrange", c(list_plot, ncol = 2))

```

```{r, results="hide", fig.show="hide"}

ggsave(plot = do.call("grid.arrange", c(list_plot, ncol = 3)),
       filename = paste0(directory_output, "enriched_motifs_clust_plot.png"),
       width = 40, height = 40)

write.csv2(glob_enrich_tab, paste0(directory_output, "enriched_motifs_clust_tab.csv"))

```

<br>

Visualize motif activities per cell and provide an alternative method of identifying differentially-active motifs between cell clusters. 
```{r}

# identify interesting and/or common motifs

motif_df = glob_enrich_tab %>% dplyr::filter(p.adjust < 0.01)
common_motif_df = motif_df[motif_df$motif %in% motif_df$motif[duplicated(motif_df$motif)],]
common_motif_list = unique(common_motif_df$motif)
unique_motif_df = motif_df[!motif_df$motif %in% motif_df$motif[duplicated(motif_df$motif)],]
unique_motif_list = unique(unique_motif_df$motif)
  
# Display table 

common_motif_df %>%
  kable(caption = "Enriched motif in clusters with padj < 0.01 - shared motifs") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "250px")
  
unique_motif_df %>%
  kable(caption = "Enriched motif in clusters with padj < 0.01 - unique motifs") %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "250px")

# Plot the results

DefaultAssay(seurat_obj) <- 'chromvar'

unique_motif_plot = list()
common_motif_plot = list()

for(i in 1:length(unique_motif_list)){
  unique_motif_plot[[i]] = FeaturePlot(
    object = seurat_obj,
    features = unique_motif_list[i],
    min.cutoff = 'q10',    
    max.cutoff = 'q90',
    pt.size = 0.1)
}

for(j in 1:length(common_motif_list)){
  common_motif_plot[[j]] = FeaturePlot(
    object = seurat_obj,
    features = common_motif_list[j],
    min.cutoff = 'q10',    
    max.cutoff = 'q90',
    pt.size = 0.1)
}

```

```{r, fig.width=40, fig.height=45, fig.show="hold"}

cat("Motifs enriched in only one cluster.")
do.call("grid.arrange", c(unique_motif_plot[1:36], ncol = 6))
do.call("grid.arrange", c(unique_motif_plot[37:73], ncol = 6))

```

```{r, fig.width=40, fig.height=45, fig.show="hold"}

cat("Sample of motifs enriched in multiple clusters.")
do.call("grid.arrange", c(common_motif_plot[1:36], ncol = 6))
do.call("grid.arrange", c(common_motif_plot[37:74], ncol = 6))

```

```{r, fig.show="hide"}

write.csv2(unique_motif_df, paste0(directory_output, "unique_enriched_motif_clust.csv"))
write.csv2(common_motif_df, paste0(directory_output, "common_enriched_motif_clust.csv"))

ggsave(plot= do.call("grid.arrange", c(unique_motif_plot[1:36], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "unique_enriched_motif_clust_plot1.png"))
ggsave(plot= do.call("grid.arrange", c(unique_motif_plot[37:73], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "unique_enriched_motif_clust_plot2.png"))

ggsave(plot= do.call("grid.arrange", c(common_motif_plot[1:36], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot1.png"))
ggsave(plot= do.call("grid.arrange", c(common_motif_plot[37:74], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot2.png"))
ggsave(plot= do.call("grid.arrange", c(common_motif_plot[75:111], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot3.png"))
ggsave(plot= do.call("grid.arrange", c(common_motif_plot[112:148], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot4.png"))
ggsave(plot= do.call("grid.arrange", c(common_motif_plot[149:185], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot5.png"))
ggsave(plot= do.call("grid.arrange", c(common_motif_plot[186:222], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot6.png"))
ggsave(plot= do.call("grid.arrange", c(common_motif_plot[223:259], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot7.png"))
ggsave(plot= do.call("grid.arrange", c(unique_motif_plot[260:296], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot8.png"))
ggsave(plot= do.call("grid.arrange", c(common_motif_plot[297:337], ncol=6)), width=40,
       height=40, filename=paste0(directory_output, "common_enriched_motif_clust_plot9.png"))

```



<br><br><br>



```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results = "hide"}

# Clean working space and memory 
rm(list = ls())
gc()

```

