---
title: "scRNAseq_CITEseq_R_BatchEffectCorrection"
author: "Laëtitia Racine"
date: "2022-11-01"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(dplyr)
library(Seurat)
library(ggplot2)
library(stringr)
library(knitr)

```

# Overview

This code aims at determining if the dataset needs to be corrected for batch effect bias before applying the clustering algorithm.  
Depending on the results, the following steps are different :    
- if there is no batch effect, we can just merge the data  
- if there is a batch effect, we need to integrate both batches in one object  
As it's better to normalize on individual object before merging or integrating, we work here with the output objects of NormalizeData script.  
We decided here to work with the normalized object without cell cycle regression.  
    
Based on documentations :       
- https://satijalab.org/seurat/articles/integration_introduction.html    
- https://satijalab.org/seurat/articles/integration_large_datasets.html      
- https://github.com/satijalab/seurat/issues/4753      
- https://github.com/satijalab/seurat/issues/4372      
- https://github.com/satijalab/seurat/discussions/3998      
- https://satijalab.org/seurat/archive/v3.1/integration.html      
- https://blog.bioturing.com/2022/03/24/batch-effect-in-single-cell-rna-seq-frequently-asked-questions-and-answers/      
Question ask on the github forum : https://github.com/satijalab/seurat/issues/6616    
    
```{r, Working directories and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_BatchEffectCorrection/"))
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_BatchEffectCorrection/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_CITEseq_R_BatchEffectCorrection/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_CITEseq_R_NormalizeData/"))
list_seurat_norm = readRDS(paste0(dir, "/", "list_seurat_obj_qc_norm.rds"))

```

```{r, Add batch information}

# Add experiment information 
list_seurat_norm[["CTRL"]]$batch = "batch1"
list_seurat_norm[["DON"]]$batch = "batch1"
list_seurat_norm[["DONaK"]]$batch = "batch1"
list_seurat_norm[["2DG"]]$batch = "batch1"
list_seurat_norm[["2DGaK"]]$batch = "batch1"
list_seurat_norm[["VPA"]]$batch = "batch1"
list_seurat_norm[["CTRLaK"]]$batch = "batch1"

list_seurat_norm[["CTRL2"]]$batch = "batch2"
list_seurat_norm[["AOA"]]$batch = "batch2"
list_seurat_norm[["AOAaK"]]$batch = "batch2"

```


# Process two experiments separately and visualize

Merge data from the same batch.    
Make UMAP plots where you can see how different samples group.    
https://github.com/satijalab/seurat/issues/5761    

```{r, Merge same batch}

# Merged the conditions from the same experiment and select variable features
merged_batch1 = merge(x = list_seurat_norm[["CTRL"]], 
                      y = c(list_seurat_norm[["CTRLaK"]], 
                            list_seurat_norm[["DON"]],
                            list_seurat_norm[["DONaK"]], 
                            list_seurat_norm[["2DG"]],
                            list_seurat_norm[["2DGaK"]], 
                            list_seurat_norm[["VPA"]]))
batch1 = c("CTRL", "CTRLaK", "DON", "DONaK", "2DG", "2DGaK", "VPA")
list_batch1 = list_seurat_norm[batch1]
features_batch1 <- SelectIntegrationFeatures(object.list = list_batch1, nfeatures = 3000)

merged_batch2 = merge(x = list_seurat_norm[["CTRL2"]], 
                      y = c(list_seurat_norm[["AOAaK"]], 
                            list_seurat_norm[["AOA"]]))
batch2 = c("CTRL2", "AOA", "AOAaK")
list_batch2 = list_seurat_norm[batch2]
features_batch2 <- SelectIntegrationFeatures(object.list = list_batch2, nfeatures = 3000)

```

## Based on SCT normalisation

```{r, All conditions - Ind batch sct, fig.width = 20, fig.height = 14, results = 'hold'}

VariableFeatures(merged_batch1, assay = "SCT") <- features_batch1
VariableFeatures(merged_batch2, assay = "SCT") <- features_batch2

# Run PCA and UMAP on each experiment
merged_sct_batch1 = RunPCA(object = merged_batch1,
                           assay = "SCT",
                           reduction.name = "pca",
                           npcs = 100,
                           seed.use = 42,
                           verbose = TRUE)
merged_sct_batch2 = RunPCA(object = merged_batch2,
                           assay = "SCT",
                           reduction.name = "pca",
                           npcs = 100,
                           seed.use = 42,
                           verbose = TRUE)

merged_sct_batch1 = RunUMAP(object = merged_sct_batch1, 
                            assay = "SCT",
                            reduction =  "pca",
                            dims = 1:30,
                            reduction.name = "umap", 
                            set.seed = 42,
                            verbose = TRUE)
merged_sct_batch2 = RunUMAP(object = merged_sct_batch2, 
                            assay = "SCT",
                            reduction =  "pca",
                            dims = 1:30,
                            reduction.name = "umap",
                            set.seed = 42,
                            verbose = TRUE)


# Show all conditions independently
batch1_sct_umap_plot = DimPlot(object = merged_sct_batch1, 
                               group.by = "orig.ident", 
                               reduction = "umap",
                               cols = color_code)
batch1_sct_umap_split_plot = DimPlot(object = merged_sct_batch1, 
                                     group.by = "orig.ident", 
                                     split.by = "orig.ident", 
                                     reduction = "umap",
                               cols = color_code)

batch2_sct_umap_plot = DimPlot(object = merged_sct_batch2, 
                               group.by = "orig.ident", 
                               reduction = "umap",
                               cols = color_code)
batch2_sct_umap_split_plot = DimPlot(object = merged_sct_batch2, 
                                     group.by = "orig.ident", 
                                     split.by = "orig.ident", 
                                     reduction = "umap",
                               cols = color_code)

batch1_sct_umap_plot / batch1_sct_umap_split_plot
batch2_sct_umap_plot / batch2_sct_umap_split_plot

```

```{r, Save outputs - Ind batch sct, results ='hide'}

ggsave(batch1_sct_umap_plot / batch1_sct_umap_split_plot,
       file = paste0(directory_output, "umap_sct_batch1.png"),
       width = 20, height = 14)
ggsave(batch2_sct_umap_plot / batch2_sct_umap_split_plot,
       file = paste0(directory_output, "umap_sct_batch2.png"),
       width = 20, height = 14)

rm(batch1, batch2, list_batch1, list_batch2)
rm(merged_sct_batch1, merged_sct_batch2)
rm(batch1_sct_umap_plot, batch1_sct_umap_split_plot, batch2_sct_umap_plot, batch2_sct_umap_split_plot)
gc()

```

We observe on the global UMAP of SCT normalized data that all the conditions overlapped correctly for batch1. For batch2, as expected the conditions are less similar. The global shape of the population are similar between CTRL2 and AOAaK but AOA is very distinct.  


## Based on Log normalization

```{r, All conditions - Ind batch log, fig.width = 20, fig.height = 14, results = 'hold'}

VariableFeatures(merged_batch1, assay = "RNA") = features_batch1
VariableFeatures(merged_batch2, assay = "RNA") = features_batch2

merged_batch1 = ScaleData(merged_batch1, assay = "RNA")
merged_batch2 = ScaleData(merged_batch2, assay = "RNA")

merged_log_batch1 = RunPCA(object = merged_batch1,
                           assay = "RNA",
                           reduction.name = "pca",
                           npcs = 100,
                           seed.use = 42,
                           verbose = TRUE)
merged_log_batch2 = RunPCA(object = merged_batch2,
                           assay = "RNA",
                           reduction.name = "pca",
                           npcs = 100,
                           seed.use = 42,
                           verbose = TRUE)

merged_log_batch1 = RunUMAP(object = merged_log_batch1, 
                            assay = "RNA",
                            reduction =  "pca",
                            dims = 1:30,
                            reduction.name = "umap",
                            set.seed = 42,
                            verbose = TRUE)
merged_log_batch2 = RunUMAP(object = merged_log_batch2, 
                            assay = "RNA",
                            reduction =  "pca",
                            dims = 1:30,
                            reduction.name = "umap",
                            set.seed = 42,
                            verbose = TRUE)

# Show all conditions independently
batch1_log_umap_plot = DimPlot(object = merged_log_batch1, 
                               group.by = "orig.ident", 
                               reduction = "umap",
                               cols = color_code)
batch1_log_umap_split_plot = DimPlot(object = merged_log_batch1, 
                                     group.by = "orig.ident", 
                                     split.by = "orig.ident", 
                                     reduction = "umap",
                                     cols = color_code)

batch2_log_umap_plot = DimPlot(object = merged_log_batch2, 
                               group.by = "orig.ident", 
                               reduction = "umap",
                               cols = color_code)
batch2_log_umap_split_plot = DimPlot(object = merged_log_batch2, 
                                     group.by = "orig.ident", 
                                     split.by = "orig.ident", 
                                     reduction = "umap",
                                     cols = color_code)

batch1_log_umap_plot / batch1_log_umap_split_plot
batch2_log_umap_plot / batch2_log_umap_split_plot

```

```{r, Save outputs - Ind batch log, results ='hide'}

ggsave(batch1_log_umap_plot / batch1_log_umap_split_plot,
       file = paste0(directory_output, "umap_log_batch1.png"),
       width = 20, height = 14)
ggsave(batch2_log_umap_plot / batch2_log_umap_split_plot,
       file = paste0(directory_output, "umap_log_batch2.png"),
       width = 20, height = 14)

rm(merged_batch1, merged_batch2, features_batch1, features_batch2)
rm(merged_log_batch1, merged_log_batch2)
rm(batch1_log_umap_plot, batch1_log_umap_split_plot, batch2_log_umap_plot, batch2_log_umap_split_plot)
gc()

```

The conditions from the same batch overlap less with the log normalization. 

# Process two experiments together and visualize

Merge all the data without regard to the batch.        
Make same UMAP plots where you can see individual samples and experiment groups.    
  
```{r, Merge all conditions}

# Merged all the conditions from the two experiments and select variable features
merged_all = merge(x = list_seurat_norm[["CTRL"]], 
                   y = c(list_seurat_norm[["CTRLaK"]], 
                         list_seurat_norm[["DON"]],
                         list_seurat_norm[["DONaK"]], 
                         list_seurat_norm[["2DG"]],
                         list_seurat_norm[["2DGaK"]], 
                         list_seurat_norm[["VPA"]],
                         list_seurat_norm[["CTRL2"]],
                         list_seurat_norm[["AOA"]],
                         list_seurat_norm[["AOAaK"]]))
features_all <- SelectIntegrationFeatures(object.list = list_seurat_norm, nfeatures = 3000)

```

## Based on SCT normalisation

```{r, All conditions - All batches sct, fig.width = 16, fig.height = 14, results = 'hold'}

VariableFeatures(merged_all, assay = "SCT") = features_all

# Run PCA and UMAP
merged_all = RunPCA(object = merged_all,
                    assay = "SCT",
                    reduction.name = "pca",
                    seed.use = 42,
                    npcs = 100,
                    verbose = TRUE)
merged_all = RunUMAP(object = merged_all, 
                     assay = "SCT",
                     reduction =  "pca",
                     dims = 1:30,
                     reduction.name = "umap",
                     set.seed = 42,
                     verbose = TRUE)

# Show the conditions
umap_sct_plot = DimPlot(object = merged_all, 
                        group.by = "orig.ident", 
                        reduction = "umap",
                        cols = color_code)
umap_sct_plot_batch = DimPlot(object = merged_all, 
                              group.by = "batch", 
                              reduction = "umap")
umap_sct_split_plot = DimPlot(object = merged_all, 
                              group.by = "orig.ident", 
                              split.by = "orig.ident", 
                              reduction = "umap",
                              cols = color_code,
                              ncol = 5)

umap_sct_plot / umap_sct_plot_batch / umap_sct_split_plot

```

```{r, Focus on CTRL - All batches sct, fig.width = 10, fig.height = 7}

merged_ctrl = subset(merged_all, subset = orig.ident %in% c("CTRL", "CTRL2"))

umap_sct_ctrl =  DimPlot(object = merged_ctrl, 
                         group.by = "orig.ident", 
                         reduction = "umap")
umap_sct_split_ctrl = DimPlot(object = merged_ctrl, 
                              group.by = "orig.ident", 
                              split.by = "orig.ident", 
                              reduction = "umap")

umap_sct_ctrl / umap_sct_split_ctrl

```

```{r, Save outputs - All batches sct, results ='hide'}

ggsave(umap_sct_plot / umap_sct_plot_batch, 
       file = paste0(directory_output, "umap_sct_bothbatch.png"),
       width = 12, height = 12)
ggsave(umap_sct_split_plot, 
       file = paste0(directory_output, "umap_sct_bothbatch_split.png"),
       width = 20, height = 8)
ggsave(umap_sct_ctrl / umap_sct_split_ctrl, 
       file = paste0(directory_output, "umap_sct_bothbatch_ctrl.png"),
       width = 12, height = 12)

rm(merged_ctrl, umap_sct_ctrl, umap_sct_split_ctrl)
rm(umap_sct_plot, umap_sct_plot_batch, umap_sct_split_plot)
gc()

```

Cells from batch1 and batch2 overlap very well except for AOA cells (expected) for the SCT normalization.


## Based on Log normalization

```{r, All conditions - All batches log, fig.width = 20, fig.height = 14, results = 'hold'}

VariableFeatures(merged_all, assay = "RNA") = features_all
merged_all = ScaleData(merged_all, assay = "RNA")

# Run PCA and UMAP
merged_all = RunPCA(object = merged_all,
                    assay = "RNA",
                    reduction.name = "pca",
                    npcs = 100,
                    seed.use = 42,
                    verbose = TRUE)
merged_all = RunUMAP(object = merged_all, 
                     assay = "RNA",
                     reduction =  "pca",
                     dims = 1:30,
                     reduction.name = "umap",
                     set.seed = 42,
                     verbose = TRUE)

# Show the conditions
umap_log_plot = DimPlot(object = merged_all, 
                        group.by = "orig.ident", 
                        reduction = "umap",
                        cols = color_code)
umap_log_plot_batch = DimPlot(object = merged_all, 
                              group.by = "batch", 
                              reduction = "umap")
umap_log_split_plot = DimPlot(object = merged_all, 
                              group.by = "orig.ident", 
                              split.by = "orig.ident", 
                              reduction = "umap",
                              cols = color_code,
                              ncol = 5)

umap_log_plot / umap_log_plot_batch / umap_log_split_plot

```

```{r, Focus on CTRL - All batches log, fig.width = 10, fig.height = 7}

merged_ctrl = subset(merged_all, subset = orig.ident %in% c("CTRL", "CTRL2"))

umap_log_ctrl =  DimPlot(object = merged_ctrl, 
                         group.by = "orig.ident", 
                         reduction = "umap")
umap_log_split_ctrl = DimPlot(object = merged_ctrl, 
                              group.by = "orig.ident", 
                              split.by = "orig.ident", 
                              reduction = "umap")

umap_log_ctrl / umap_log_split_ctrl

```

```{r, Save outputs - All batches log, results ='hide'}

ggsave(umap_log_plot / umap_log_plot_batch, 
       file = paste0(directory_output, "umap_log_bothbatch.png"),
       width = 12, height = 12)
ggsave(umap_log_split_plot, 
       file = paste0(directory_output, "umap_log_bothbatch_split.png"),
       width = 20, height = 8)
ggsave(umap_log_ctrl / umap_log_split_ctrl, 
       file = paste0(directory_output, "umap_log_bothbatch_ctrl.png"),
       width = 12, height = 12)

rm(merged_ctrl, features_all)
rm(umap_log_ctrl, umap_log_split_ctrl, umap_log_plot)
rm(umap_log_plot_batch, umap_log_split_plot)
gc()

```

The batches overlap less with log normalization. We will preferentially work with SCT normalization for latter analyses.   

# Conclusion 

=> No need for batch effect correction (=integration) if we work with sct normalization after the merge for downstream analysis.

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Clean working space and memory 
rm(list = ls())
gc()

# Show package version
sessionInfo()

```
