---
title: "scRNAseq_CITEseq_GenomePosition_GeneNameCountMatrix"
author: "Laëtitia Racine"
date: "2023-02-20"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(dplyr)
library(biomaRt)
library(stringr)
library(tidyr)
library(tibble)

```

```{r, Working directory}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_GenomePosition/"))
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_GenomePosition/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_CITEseq_R_GenomePosition/", current_date, "/")

```

<br>

**Create a correspondence tab between Ensembl ID and Gene Symbol**  
```{r}

# We open a features.tsv.gz file to have the list of the features of the experiment.

features_tab = read.delim(paste0(directory, "data/scRNAseq/", 
                          "scRNAseq_CTRL_sample_feature_bc_matrix_features.tsv.gz"),
                          header = FALSE, col.names=c("ensembl_gene_id", 
                                                      "gene_symbol", "category"))

features_tab = features_tab %>% 
  dplyr::filter(category == "Gene Expression") %>%
  dplyr::select(-category) %>%
  dplyr::group_by(gene_symbol) %>%
  dplyr::mutate(dup = if_else(duplicated(gene_symbol) == TRUE, TRUE, FALSE)) %>%
  dplyr::ungroup() %>% 
  tibble::rownames_to_column("pos") %>%
  dplyr::mutate(gene_symbol_dupl = if_else(dup == FALSE, gene_symbol, paste0(gene_symbol, ".1")))

```

<br>

**Add Gene position (molecular information) to the correspondence tab**      
https://www.ensembl.org/info/data/biomart/biomart_r_package.html    
```{r} 

ensembl = useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl") # GRCh38
head(listAttributes(ensembl), 10)

chr_genes <- getBM(attributes=c('ensembl_gene_id',
                                'hgnc_id',
                                'chromosome_name',
                                'start_position',
                                'end_position'), 
                   mart = ensembl)

features_tab = left_join(features_tab, chr_genes, by = "ensembl_gene_id")

```

<br>

**Add Gene position (cytogenetic information) to the correspondence tab**    
https://support.bioconductor.org/p/9142452/       
ftp.ncbi.nlm.nih.gov/gene/DATA/GENE_INFO/Mammalia/Homo_sapiens.gene_info.gz      
https://medlineplus.gov/genetics/understanding/howgeneswork/genelocation/      
```{r}

db_list = read.delim(paste0(directory, "data/Homo_sapiens.gene_info.gz"))

db_list = db_list %>% 
   dplyr::mutate(mim_id = str_extract(dbXrefs, "(?<=MIM:)[:digit:]+")) %>%
   dplyr::mutate(hgnc_id = str_extract(dbXrefs, "(?<=HGNC:)HGNC:[:digit:]+")) %>%
   dplyr::mutate(ensembl_gene_id = str_extract(dbXrefs, "(?<=Ensembl:)ENSG[:digit:]+")) %>%
   dplyr::select(Symbol, Synonyms, chromosome, map_location, mim_id, hgnc_id, ensembl_gene_id) %>%
   dplyr::rename("gene_symbol" = Symbol,
                 "gene_symbol_synonyms" = Synonyms,
                 "chromosome_name" = chromosome)

features_tab = left_join(features_tab, db_list, 
                         by = c("chromosome_name", "hgnc_id", 
                                "gene_symbol", "ensembl_gene_id"))

```

```{r}

features_tab = features_tab %>%
  dplyr::select(gene_symbol, gene_symbol_dupl, gene_symbol_synonyms,
                ensembl_gene_id, hgnc_id, mim_id,
                chromosome_name, start_position, end_position, map_location)

write.csv2(x = features_tab, 
           file = paste0(directory_output, "scRNAseq_correspondence_tab_EnsemblID_geneName_genomePos.csv"),
           row.names = FALSE)

```

<br>

200 genes don't have the position information (chromosome_name, start, end) when we use biomaRt and org.Hs.eg.db.    
We can find those genes with a manual research on ensembl.org website. It's GRcH37 information but we can also see the corresponding GRcH38 coordinates in the description (ex : ENSG00000271895 => This gene maps to 20,184,242-20,186,486 in GRCh38 coordinates). To complete the tab, a manual research has to be done.     
http://grch37.ensembl.org/Homo_sapiens/Gene/Summary?g=ENSG00000271895;r=1:11089716-11090585;t=ENST00000607145    
http://grch37.ensembl.org/Homo_sapiens/Gene/Summary?g=ENSG00000225986;r=1:20510735-20512979;t=ENST00000442226    
  
```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results='hide'}

# Clean working space and memory 
rm(list = ls())
gc()

```
