---
title: "scRNAseq_CITEseq_R_DWNS_ClustersStatistics"
author: "Laëtitia Racine"
date: "2022-10-03"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(dplyr)
library(stringr)
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(kableExtra)
library(ggrepel)

```

# Overview 

This code aims to analyze cell clusters' composition (which drug condition in which cluster and in which proportion) and to identify interesting clusters for further investigation. 
  
Based on documentations for plots :         
- http://www.sthda.com/english/wiki/ggplot2-pie-chart-quick-start-guide-r-software-and-data-visualization  
- https://r-charts.com/part-whole/pie-chart-labels-outside-ggplot2/       
        
```{r, Working directories and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_ClustersStatistics/"))
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_ClustersStatistics/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_ClustersStatistics/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_CITEseq_R_ReductionDimension/"))
merged_all = readRDS(file = paste0(dir, "/", "merged_all_1-40dim.rds"))

```

As a reminder, here are the clusters we identified in the dataset.

```{r, Show clusters, fig.width = 6, fig.height = 4}

DimPlot(merged_all, 
        label = TRUE, 
        reduction = "umap", 
        group.by = "clusters")

```

And the cells' distribution according culture condition (control or drug medium). 

```{r, Show conditions, fig.width = 22, fig.height = 18}

umap_highlight = function(seurat_obj, color_code) {
  
  ind_umap = list()
  
  for (i in 1:length(unique(seurat_obj$orig.ident))) {
  
  # Select condition to highlight
  highlight_cond = unique(seurat_obj$orig.ident)[i]
  color = unname(noquote(color_code[highlight_cond]))
  others_cond = setdiff(unique(seurat_obj$orig.ident), highlight_cond)
  others_cond_group = paste(others_cond, collapse=", ")
  
  # Sort cell according to their condition
  cell_highlight = unlist(CellsByIdentities(object = seurat_obj, idents = highlight_cond))
  cell_nohighlight = unlist(CellsByIdentities(object = seurat_obj, idents = others_cond))
  
  # Plot 
  plot_umap = DimPlot(object = seurat_obj,
                      group.by = "orig.ident",
                      reduction ="umap",
                      pt.size = 0.05,
                      cells.highlight = list(cell_highlight, cell_nohighlight),
                      cols.highlight = c(color,"grey")) +
    ggtitle(highlight_cond) +
    theme(legend.position = "none")
  
  ind_umap[[i]] = plot_umap
  
  }
  
  # function's output 
  return(ind_umap)
  
}

merged_all = SetIdent(object = merged_all, value = "orig.ident")
umap_hl = umap_highlight(merged_all, color_code)
do.call("grid.arrange", c(umap_hl, ncol = 3))

```


# Cells' distribution in clusters 

As we decided there were no batch effect on the data, we use the two control samples as one. Then, the CTRL sample (orig.ident) describe all the kind of cells we can find in standard culture condition without any drug influence. 

```{r, Clusters statistics - tab number}

clusters = as.integer(levels(unique(merged_all@meta.data$clusters)))
merged_all = SetIdent(object = merged_all, value = "clusters")

# Cells' number in cluster per condition
tab_number = data.frame(condition = character())
for (i in 1:length(clusters)) {
  cat("Extract cluster", clusters[i], "information.")
  clust = subset(merged_all, idents = clusters[i])
  print(table(clust@meta.data$clusters))
  tab = as.data.frame(table(clust@meta.data$orig.ident)) 
  colnames(tab) = c("condition", paste0("cluster_", clusters[i]))
  tab_number = full_join(tab_number, tab, by = "condition")
}
tab_number[is.na(tab_number)] = 0
tab_number$total_cell_per_condition = rowSums(tab_number[,-1])

# Add total per cluster
temp = as.data.frame(t(data.frame(colSums(tab_number[,-1]))))
rownames(temp) = "total_cell_per_cluster"
temp = temp %>% tibble::rownames_to_column(var = "condition")
tab_number = rbind(tab_number, temp)

# Display data frame
tab_number %>%
  kable(caption = "Not normalized cells' distribution") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

We don't start with the same number of cells in each condition. A normalization step based on the initial number of cells per condition is required to be able to compare the conditions.

```{r, Clusters statistics - tab number normalized}

# Divide per the number of cells, multiply by a 1000 factor and round
tab_number_norm = tab_number[-nrow(tab_number),]
tab_number_norm[,-1] = round((tab_number_norm[,-1]/tab_number_norm[ ,ncol(tab_number_norm)])*1000)

# Add total per cluster
temp = as.data.frame(t(data.frame(colSums(tab_number_norm[,-1]))))
rownames(temp) = "total_cell_per_cluster"
temp = temp %>% tibble::rownames_to_column(var = "condition")
tab_number_norm = rbind(tab_number_norm, temp)

# Display dataframe
tab_number_norm %>%
  kable(caption = "Normalized cells' distribution based on the initial number of cells in each condition.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

Then we study the data in two alternative ways :  
- the condition composition : percentages of each cluster for a condition (tab_percent_cond)  
- the cluster composition : percentages of cells from each condition for a cluster (tab_percent_clust)    
    
```{r, Clusters statistics - tab percent normalized}

# Cells' percentage in cluster per condition
tab_percent_cond = tab_number_norm %>% dplyr::filter(condition != "total_cell_per_cluster")
tab_percent_cond[ ,-1] = round((tab_percent_cond[ ,-1]/tab_percent_cond[ ,ncol(tab_percent_cond)])*100, 1)

tab_percent_cond %>%
  kable(caption = "Normalized cells' percentage distribution per condition.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

# Cells' percentage in condition per cluster 
tab_percent_clust = tab_number_norm %>% dplyr::select(-total_cell_per_condition)
tab_percent_clust = as.data.frame(t(tab_percent_clust))
colnames(tab_percent_clust) = tab_percent_clust[1,]
tab_percent_clust = tab_percent_clust[-1,] %>% 
  mutate_if(is.character, ~as.numeric(.)) %>%
  rownames_to_column(var = "cluster")
tab_percent_clust[,-1] = round((tab_percent_clust[ ,-1]/tab_percent_clust[ ,ncol(tab_percent_clust)])*100, 1)

tab_percent_clust %>%
  kable(caption = "Normalized cells' percentage distribution per cluster.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```
  
  
# Graphic representations  
  
```{r, Clusters statistics - bp plots, fig.width = 14, fig.height = 8}

###########
# Cluster #
###########

temp_clust = tab_percent_clust %>% dplyr::select(-total_cell_per_cluster)
temp_clust = temp_clust %>% tidyr::pivot_longer(cols = 2:ncol(temp_clust),
                                    names_to = "condition",
                                    values_to = "percentage")
temp_clust$cluster =  factor(temp_clust$cluster,
                             levels = c("cluster_0", "cluster_1", "cluster_2", 
                                        "cluster_3", "cluster_4", "cluster_5",
                                        "cluster_6", "cluster_7", "cluster_8",
                                        "cluster_9", "cluster_10", "cluster_11", 
                                        "cluster_12", "cluster_13", "cluster_14",
                                        "cluster_15", "cluster_16", "cluster_17"))

bp_clust = ggplot(temp_clust, aes(x=cluster, y=percentage, fill=condition))+
  geom_bar(width = 0.8, stat = "identity") +
  scale_fill_manual(values = color_code, limits = force) +
  coord_flip()
bp_clust

#############
# Condition #
#############

temp_cond = tab_percent_cond %>% dplyr::select(-total_cell_per_condition)
temp_cond = temp_cond %>% tidyr::pivot_longer(cols = 2:ncol(temp_cond),
                                    names_to = "cluster",
                                    values_to = "percentage")
temp_cond$cluster =  factor(temp_cond$cluster,
                            levels = c("cluster_0", "cluster_1", "cluster_2", 
                                       "cluster_3", "cluster_4", "cluster_5",
                                       "cluster_6", "cluster_7", "cluster_8",
                                       "cluster_9", "cluster_10", "cluster_11", 
                                       "cluster_12", "cluster_13", "cluster_14",
                                       "cluster_15", "cluster_16", "cluster_17"))

bp_cond = ggplot(temp_cond, aes(x=condition, y=percentage, fill=cluster))+
  geom_bar(width = 0.8, stat = "identity", color="#737373") +
  coord_flip()
bp_cond

```

```{r, Clusters statistics - pie plots, fig.width = 14, fig.height = 24}

###########
# Cluster #
###########

clust_list = tab_percent_clust$cluster
list_clust_pie = list()

for (i in 1:length(clust_list)) {
  
  temp2 = temp_clust %>% 
    dplyr::filter(cluster == clust_list[i]) %>% 
    dplyr::mutate(csum = rev(cumsum(rev(percentage))), 
                  pos = percentage/2 + lead(csum, 1),
                  pos = if_else(is.na(pos), percentage/2, pos))
  
  pie = ggplot(temp2, aes(x="", y=percentage, fill=condition))+
    geom_col(width = 1, color = 1) +
    coord_polar("y", start=0) + 
    scale_fill_manual(values = color_code, limits = force) +
    geom_label_repel(data = temp2,
                     aes(y = pos, label = paste0(percentage, "%")),
                     size = 4, nudge_x = 1, show.legend = FALSE, 
                     segment.color = "#999999",
                     col = "black") +
    theme_minimal() +
    ggtitle(label = clust_list[i]) +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(),
          panel.grid=element_blank(),
          axis.ticks = element_blank(),
          plot.title=element_text(size=14, face="bold")
    )
  
  list_clust_pie[[i]] = pie
}

do.call("grid.arrange", c(list_clust_pie, ncol = 3))

#############
# Condition #
#############

cond_list = tab_percent_cond$condition
list_cond_pie = list()

for (i in 1:length(cond_list)) {

  temp2 = temp_cond %>% 
    dplyr::filter(condition == cond_list[i]) %>%
    dplyr::mutate(cluster = sort(cluster)) %>%
    dplyr::mutate(csum = rev(cumsum(rev(percentage))),
                  pos = percentage/2 + lead(csum, 1),
                  pos = if_else(is.na(pos), percentage/2, pos))

  pie = ggplot(temp2, aes(x="", y=percentage, fill=cluster))+
    geom_col(width = 1, color = 1) +
    coord_polar("y", start=0) +
    geom_label_repel(data = temp2,
                     aes(y = pos, label = paste0(percentage, "%")),
                     size = 4, nudge_x = 1, show.legend = FALSE,
                     segment.color = "#999999",
                     col = "black") +
    theme_minimal() +
    ggtitle(label = cond_list[i]) +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(),
          panel.grid=element_blank(),
          axis.ticks = element_blank(),
          plot.title=element_text(size=14, face="bold")
    )

  list_cond_pie[[i]] = pie

}

do.call("grid.arrange", c(list_cond_pie, ncol = 3))

```

```{r, Clusters number - bp plots}

temp = tab_number_norm[-nrow(tab_number_norm),] %>% dplyr::select(-total_cell_per_condition)
temp = temp %>% tidyr::pivot_longer(cols = 2:ncol(temp),
                                    names_to = "cluster",
                                    values_to = "cell_number")

temp$cluster =  factor(temp$cluster,
                       levels = c("cluster_0", "cluster_1", "cluster_2", 
                                  "cluster_3", "cluster_4", "cluster_5",
                                  "cluster_6", "cluster_7", "cluster_8",
                                  "cluster_9", "cluster_10", "cluster_11", 
                                  "cluster_12", "cluster_13", "cluster_14",
                                  "cluster_15", "cluster_16", "cluster_17"))

# All conditions
bp_number = ggplot(temp, aes(x=cluster, y=cell_number, fill=condition))+
  geom_bar(width = 0.8, stat = "identity") +
  scale_fill_manual(values = color_code, limits = force) +
  coord_flip()
bp_number

# Drugs conditions only
temp_drug = temp %>% dplyr::filter(condition %in% c("CTRL", "2DG", "DON", "AOA"))

bp_number_drugs = ggplot(temp_drug, aes(x=cluster, y=cell_number, fill=condition))+
  geom_bar(width = 0.8, stat = "identity") +
  scale_fill_manual(values = color_code, limits = force) +
  coord_flip()
bp_number_drugs

```
  
Interesting clusters :  
- AOA : 15,14, 8, 12, 16  
- 2DG : 2, 11    
  
```{r, Save outputs, fig.show ='hide'}

write.csv2(x = tab_number, 
           file = paste0(directory_output, "tab_number_cells_clusters.csv"), 
           row.names = FALSE)
write.csv2(x = tab_number_norm, 
           file = paste0(directory_output, "tab_number_norm_cells_clusters.csv"), 
           row.names = FALSE)

ggsave(plot = bp_clust, filename = paste0(directory_output, "bp_clust.svg"))
ggsave(plot = bp_cond, filename = paste0(directory_output, "bp_cond.svg"))
ggsave(plot = bp_number, filename = paste0(directory_output, "bp_number.svg"))

ggsave(plot = do.call("grid.arrange", c(list_cond_pie, ncol = 3)),
       filename = paste0(directory_output, "pie_cond.svg"),
       height = 18, width = 20)
ggsave(plot = do.call("grid.arrange", c(list_clust_pie, ncol = 3)),
       filename = paste0(directory_output, "pie_clust.svg"),
       height = 18, width = 20)

```

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Clean working space and memory 
rm(list = ls())
gc()

# Show package version
sessionInfo()

```