---
title: "scRNAseq_CITEseq_R_DWNS_ClustersStatistics"
author: "Laëtitia Racine"
date: "2022-10-03"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(dplyr)
library(stringr)
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(kableExtra)
library(ggrepel)

```

# Overview 

This code aims to analyze cell clusters' composition (which drug condition in which cluster and in which proportion) and to identify interesting clusters for further investigation. 
  
Based on documentations for plots :         
- http://www.sthda.com/english/wiki/ggplot2-pie-chart-quick-start-guide-r-software-and-data-visualization  
- https://r-charts.com/part-whole/pie-chart-labels-outside-ggplot2/       
        
```{r, Working directories and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_ClustersStatistics/"))
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_ClustersStatistics/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_ClustersStatistics/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_CITEseq_R_ReductionDimension/"))
merged_all = readRDS(file = paste0(dir, "/", "merged_all_1-40dim.rds"))

```

As a reminder, here are the clusters we identified in the dataset.

```{r, Show clusters, fig.width = 6, fig.height = 4}

DimPlot(merged_all, 
        label = TRUE, 
        reduction = "umap", 
        group.by = "clusters")

```

And the cells' distribution according culture condition (control or drug medium). 

```{r, Show conditions, fig.width = 22, fig.height = 20}

umap_highlight = function(seurat_obj, color_code) {
  
  ind_umap = list()
  
  for (i in 1:length(unique(seurat_obj$orig.ident))) {
  
  # Select condition to highlight
  highlight_cond = unique(seurat_obj$orig.ident)[i]
  color = unname(noquote(color_code[highlight_cond]))
  others_cond = setdiff(unique(seurat_obj$orig.ident), highlight_cond)
  others_cond_group = paste(others_cond, collapse=", ")
  
  # Sort cell according to their condition
  cell_highlight = unlist(CellsByIdentities(object = seurat_obj, idents = highlight_cond))
  cell_nohighlight = unlist(CellsByIdentities(object = seurat_obj, idents = others_cond))
  
  # Plot 
  plot_umap = DimPlot(object = seurat_obj,
                      group.by = "orig.ident",
                      reduction ="umap",
                      pt.size = 0.05,
                      cells.highlight = list(cell_highlight, cell_nohighlight),
                      cols.highlight = c(color,"grey")) +
    ggtitle(highlight_cond) +
    theme(legend.position = "none")
  
  ind_umap[[i]] = plot_umap
  
  }
  
  # function's output 
  return(ind_umap)
  
}

merged_all = SetIdent(object = merged_all, value = "orig.ident")
umap_hl = umap_highlight(merged_all, color_code)
do.call("grid.arrange", c(umap_hl, ncol = 3))

```



# Option 1 : clusters statistics with separated controls

```{r, Clusters statistics - tab number - separated ctrl}

clusters = as.integer(levels(unique(merged_all@meta.data$clusters)))
merged_all = SetIdent(object = merged_all, value = "clusters")

# Cells' number in cluster per condition
tab_number = data.frame(condition = character())
for (i in 1:length(clusters)) {
  cat("Extract cluster", clusters[i], "information.")
  clust = subset(merged_all, idents = clusters[i])
  print(table(clust@meta.data$clusters))
  tab = as.data.frame(table(clust@meta.data$orig.ident)) 
  colnames(tab) = c("condition", paste0("cluster_", clusters[i]))
  tab_number = full_join(tab_number, tab, by = "condition")
}
tab_number[is.na(tab_number)] = 0
tab_number$total_cell_per_condition = rowSums(tab_number[,-1])

# Add total per cluster
temp = as.data.frame(t(data.frame(colSums(tab_number[,-1]))))
rownames(temp) = "total_cell_per_cluster"
temp = temp %>% tibble::rownames_to_column(var = "condition")
tab_number = rbind(tab_number, temp)

# Display dataframe
tab_number %>%
  kable(caption = "Not normalized cells' distribution") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

We don't start with the same number of cells in each condition. A normalization step based on the initial number of cells per condition is required to be able to compare the conditions.
    
```{r, Clusters statistics - tab number normalized - separated ctrl}

# Divide per the number of cells, multiply by a 1000 factor and round
tab_number_norm = tab_number[-nrow(tab_number),]
tab_number_norm[,-1] = round((tab_number_norm[,-1]/tab_number_norm[ ,ncol(tab_number_norm)])*1000)

# Add total per cluster
temp = as.data.frame(t(data.frame(colSums(tab_number_norm[,-1]))))
rownames(temp) = "total_cell_per_cluster"
temp = temp %>% tibble::rownames_to_column(var = "condition")
tab_number_norm = rbind(tab_number_norm, temp)

# Display dataframe
tab_number_norm %>%
  kable(caption = "Normalized cells' distribution based on the initial number of cells in each condition.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

```{r, Clusters statistics - tab percent normalized - separated ctrl}

# Cells' percentage in cluster per condition
tab_percent_cond = tab_number_norm %>% dplyr::filter(condition != "total_cell_per_cluster")
tab_percent_cond[ ,-1] = round((tab_percent_cond[ ,-1]/tab_percent_cond[ ,ncol(tab_percent_cond)])*100, 1)

tab_percent_cond %>%
  kable(caption = "Normalized cells' percentage distribution per condition.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

# Cells' percentage in condition per cluster 
tab_percent_clust = tab_number_norm %>% dplyr::select(-total_cell_per_condition)
tab_percent_clust = as.data.frame(t(tab_percent_clust))
colnames(tab_percent_clust) = tab_percent_clust[1,]
tab_percent_clust = tab_percent_clust[-1,] %>% 
  mutate_if(is.character, ~as.numeric(.)) %>%
  rownames_to_column(var = "cluster")
tab_percent_clust[,-1] = round((tab_percent_clust[ ,-1]/tab_percent_clust[ ,ncol(tab_percent_clust)])*100, 1)

tab_percent_clust %>%
  kable(caption = "Normalized cells' percentage distribution per cluster.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

```{r, Clusters statistics - bp plots - separated ctrl, fig.width = 14, fig.height = 8}

###########
# Cluster #
###########

# Barplot with all clusters
temp_clust = tab_percent_clust %>% dplyr::select(-total_cell_per_cluster)
temp_clust = temp_clust %>% tidyr::pivot_longer(cols = 2:ncol(temp_clust),
                                    names_to = "condition",
                                    values_to = "percentage")

bp_clust = ggplot(temp_clust, aes(x=cluster, y=percentage, fill=condition))+
  geom_bar(width = 0.8, stat = "identity") +
   scale_fill_manual(values = color_code, limits = force) +
  theme(axis.text.x=element_text(angle=90))
bp_clust

#############
# Condition #
#############

temp_cond = tab_percent_cond %>% dplyr::select(-total_cell_per_condition)
temp_cond = temp_cond %>% tidyr::pivot_longer(cols = 2:ncol(temp_cond),
                                    names_to = "cluster",
                                    values_to = "percentage")

bp_cond = ggplot(temp_cond, aes(x=condition, y=percentage, fill=cluster))+
  geom_bar(width = 0.8, stat = "identity") +
  theme(axis.text.x=element_text(angle=90))
bp_cond

```
  
```{r, Clusters statistics - pie plots - separated ctrl, fig.width = 14, fig.height = 24}

###########
# Cluster #
###########

clust_list = tab_percent_clust$cluster
list_clust_pie = list()

for (i in 1:length(clust_list)) {
  
  temp2 = temp_clust %>% 
    dplyr::filter(cluster == clust_list[i]) %>% 
    dplyr::mutate(csum = rev(cumsum(rev(percentage))), 
                  pos = percentage/2 + lead(csum, 1),
                  pos = if_else(is.na(pos), percentage/2, pos))
  
  pie = ggplot(temp2, aes(x="", y=percentage, fill=condition))+
    geom_col(width = 1, color = 1) +
    coord_polar("y", start=0) + 
    scale_fill_manual(values = color_code, limits = force) +
    geom_label_repel(data = temp2,
                     aes(y = pos, label = paste0(percentage, "%")),
                     size = 4, nudge_x = 1, show.legend = FALSE, 
                     segment.color = "#999999",
                     col = "black") +
    theme_minimal() +
    ggtitle(label = clust_list[i]) +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(),
          panel.grid=element_blank(),
          axis.ticks = element_blank(),
          plot.title=element_text(size=14, face="bold")
    )
  
  list_clust_pie[[i]] = pie
  
}

do.call("grid.arrange", c(list_clust_pie, ncol = 3))

#############
# Condition #
#############

cond_list = tab_percent_cond$condition
list_cond_pie = list()

for (i in 1:length(cond_list)) {

  temp2 = temp_cond %>% 
    dplyr::filter(condition == cond_list[i]) %>%
    dplyr::mutate(cluster = sort(cluster)) %>%
    dplyr::mutate(csum = rev(cumsum(rev(percentage))),
                  pos = percentage/2 + lead(csum, 1),
                  pos = if_else(is.na(pos), percentage/2, pos))

  pie = ggplot(temp2, aes(x="", y=percentage, fill=cluster))+
    geom_col(width = 1, color = 1) +
    coord_polar("y", start=0) +
    geom_label_repel(data = temp2,
                     aes(y = pos, label = paste0(percentage, "%")),
                     size = 4, nudge_x = 1, show.legend = FALSE,
                     segment.color = "#999999",
                     col = "black") +
    theme_minimal() +
    ggtitle(label = cond_list[i]) +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(),
          panel.grid=element_blank(),
          axis.ticks = element_blank(),
          plot.title=element_text(size=14, face="bold")
    )

  list_cond_pie[[i]] = pie
}

do.call("grid.arrange", c(list_cond_pie, ncol = 3))

```

  
  
# Option 2 : clusters statistics with merged controls 

As we decided there were no batch effect on the data, we can merge the two control samples. Then, the new control describe all the kind of cells we can find in standard culture condition without any dug influence. 

```{r, Clusters statistics - tab number - merged ctrl}

# Merge CTRL and CTRL2
temp = tab_number %>% dplyr::filter(condition %in% c("CTRL", "CTRL2"))
temp = as.data.frame(t(data.frame(colSums(temp[,-1]))))
rownames(temp) = "CTRL_CTRL2"
temp  = temp %>% tibble::rownames_to_column(var = "condition")
tab_number_merge = rbind(tab_number %>% dplyr::filter(condition != c("CTRL", "CTRL2")), temp)
tab_number_merge = tab_number_merge[c(1:4, 10, 5:9),]

# Display dataframe
tab_number_merge %>%
  kable(caption = "Not normalized cells' distribution with merged controls") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

```{r, Clusters statistics - tab number normalized - merged ctrl}

# Divide per the number of cells, multiply by a 1000 factor and round
tab_number_merge_norm = tab_number_merge[-nrow(tab_number_merge),]
tab_number_merge_norm[,-1] = round((tab_number_merge_norm[,-1]/tab_number_merge_norm[ ,ncol(tab_number_merge_norm)])*1000)

# Add total per cluster
temp = as.data.frame(t(data.frame(colSums(tab_number_merge_norm[,-1]))))
rownames(temp) = "total_cell_per_cluster"
temp = temp %>% tibble::rownames_to_column(var = "condition")
tab_number_merge_norm = rbind(tab_number_merge_norm, temp)

# Display dataframe
tab_number_merge_norm %>%
  kable(caption = "Normalized cells' distribution based on the initial number of cells in each condition with merged controls.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

```{r, Clusters statistics - tab percent normalized - merged ctrl}

# Cells' percentage in cluster per condition
tab_percent_cond_merge = tab_number_merge_norm %>% dplyr::filter(condition != "total_cell_per_cluster")
tab_percent_cond_merge[ ,-1] = round((tab_percent_cond_merge[ ,-1]/tab_percent_cond_merge[ ,ncol(tab_percent_cond_merge)])*100, 1)

tab_percent_cond_merge %>%
  kable(caption = "Normalized cells' percentage distribution per condition with merged controls.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

# Cells' percentage in condition per cluster 
tab_percent_clust_merge = tab_number_merge_norm %>% dplyr::select(-total_cell_per_condition)
tab_percent_clust_merge = as.data.frame(t(tab_percent_clust_merge))
colnames(tab_percent_clust_merge) = tab_percent_clust_merge[1,]
tab_percent_clust_merge = tab_percent_clust_merge[-1,] %>% 
  mutate_if(is.character, ~as.numeric(.)) %>%
  rownames_to_column(var = "cluster")
tab_percent_clust_merge[,-1] = round((tab_percent_clust_merge[ ,-1]/tab_percent_clust_merge[ ,ncol(tab_percent_clust_merge)])*100, 1)

tab_percent_clust_merge %>%
  kable(caption = "Normalized cells' percentage distribution per clsuter with merged controls.") %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```
  
```{r, Clusters statistics - bp plots - merged ctrl, fig.width = 14, fig.height = 8}

###########
# Cluster #
###########

temp_clust = tab_percent_clust_merge %>% dplyr::select(-total_cell_per_cluster)
temp_clust = temp_clust %>% tidyr::pivot_longer(cols = 2:ncol(temp_clust),
                                    names_to = "condition",
                                    values_to = "percentage")

bp_clust_merge = ggplot(temp_clust, aes(x=cluster, y=percentage, fill=condition))+
  geom_bar(width = 0.8, stat = "identity") +
  scale_fill_manual(values = color_code, limits = force) +
  theme(axis.text.x=element_text(angle=90))
bp_clust_merge

#############
# Condition #
#############

temp_cond = tab_percent_cond_merge %>% dplyr::select(-total_cell_per_condition)
temp_cond = temp_cond %>% tidyr::pivot_longer(cols = 2:ncol(temp_cond),
                                    names_to = "cluster",
                                    values_to = "percentage")

bp_cond_merge = ggplot(temp_cond, aes(x=condition, y=percentage, fill=cluster))+
  geom_bar(width = 0.8, stat = "identity") +
  theme(axis.text.x=element_text(angle=90))
bp_cond_merge

```

```{r, Clusters statistics - pie plots - merged ctrl, fig.width = 14, fig.height = 24}

###########
# Cluster #
###########

clust_list = tab_percent_clust_merge$cluster
list_clust_merge_pie = list()

for (i in 1:length(clust_list)) {
  
  temp2 = temp_clust %>% 
    dplyr::filter(cluster == clust_list[i]) %>% 
    dplyr::mutate(csum = rev(cumsum(rev(percentage))), 
                  pos = percentage/2 + lead(csum, 1),
                  pos = if_else(is.na(pos), percentage/2, pos))
  
  pie = ggplot(temp2, aes(x="", y=percentage, fill=condition))+
    geom_col(width = 1, color = 1) +
    coord_polar("y", start=0) + 
    scale_fill_manual(values = color_code, limits = force) +
    geom_label_repel(data = temp2,
                     aes(y = pos, label = paste0(percentage, "%")),
                     size = 4, nudge_x = 1, show.legend = FALSE, 
                     segment.color = "#999999",
                     col = "black") +
    theme_minimal() +
    ggtitle(label = clust_list[i]) +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(),
          panel.grid=element_blank(),
          axis.ticks = element_blank(),
          plot.title=element_text(size=14, face="bold")
    )
  
  list_clust_merge_pie[[i]] = pie
}

do.call("grid.arrange", c(list_clust_merge_pie, ncol = 3))

#############
# Condition #
#############

cond_list = tab_percent_cond_merge$condition
list_cond_merge_pie = list()

for (i in 1:length(cond_list)) {

  temp2 = temp_cond %>% 
    dplyr::filter(condition == cond_list[i]) %>%
    dplyr::mutate(cluster = sort(cluster)) %>%
    dplyr::mutate(csum = rev(cumsum(rev(percentage))),
                  pos = percentage/2 + lead(csum, 1),
                  pos = if_else(is.na(pos), percentage/2, pos))

  pie = ggplot(temp2, aes(x="", y=percentage, fill=cluster))+
    geom_col(width = 1, color = 1) +
    coord_polar("y", start=0) +
    geom_label_repel(data = temp2,
                     aes(y = pos, label = paste0(percentage, "%")),
                     size = 4, nudge_x = 1, show.legend = FALSE,
                     segment.color = "#999999",
                     col = "black") +
    theme_minimal() +
    ggtitle(label = cond_list[i]) +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.y = element_blank(),
          panel.border = element_blank(),
          panel.grid=element_blank(),
          axis.ticks = element_blank(),
          plot.title=element_text(size=14, face="bold")
    )

  list_cond_merge_pie[[i]] = pie

}

do.call("grid.arrange", c(list_cond_merge_pie, ncol = 3))

```



```{r, Save outputs, fig.show='hide'}

write.csv2(x = tab_number, 
           file = paste0(directory_output, "tab_number_cells_clusters.csv"), 
           row.names = FALSE)
write.csv2(x = tab_number_merge, 
           file = paste0(directory_output, "tab_number_cells_clusters_mergectrl.csv"), 
           row.names = FALSE)

write.csv2(x = tab_number_norm, 
           file = paste0(directory_output, "tab_number_norm_cells_clusters.csv"), 
           row.names = FALSE)
write.csv2(x = tab_number_merge_norm, 
           file = paste0(directory_output, "tab_number_norm_cells_clusters_mergectrl.csv"), 
           row.names = FALSE)

write.csv2(x = tab_percent_cond_merge, 
           file = paste0(directory_output, "tab_norm_percent_cells_conditions_mergectrl.csv"), 
           row.names = FALSE)
write.csv2(x = tab_percent_clust_merge, 
           file = paste0(directory_output, "tab_norm_percent_cells_clusters_mergectrl.csv"), 
           row.names = FALSE)

ggsave(plot = bp_clust, filename = paste0(directory_output, "bp_clust.svg"))
ggsave(plot = bp_cond, filename = paste0(directory_output, "bp_cond.svg"))
ggsave(plot = bp_clust_merge, filename = paste0(directory_output, "bp_clust_mergectrl.svg"))
ggsave(plot = bp_cond_merge, filename = paste0(directory_output, "bp_cond_merge_ctrl.svg"))

ggsave(plot = do.call("grid.arrange", c(list_cond_pie, ncol = 3)),
       filename = paste0(directory_output, "pie_cond.svg"),
       height = 18, width = 20)
ggsave(plot = do.call("grid.arrange", c(list_clust_pie, ncol = 3)),
       filename = paste0(directory_output, "pie_clust.svg"),
       height = 18, width = 20)
ggsave(plot = do.call("grid.arrange", c(list_cond_merge_pie, ncol = 3)),
       filename = paste0(directory_output, "pie_cond_mergectrl.svg"),
       height = 18, width = 20)
ggsave(plot = do.call("grid.arrange", c(list_clust_merge_pie, ncol = 3)),
       filename = paste0(directory_output, "pie_clust_mergectrl.svg"),
       height = 18, width = 20)

```

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Clean working space and memory 
rm(list = ls())
gc()

# Show package version
sessionInfo()

```