---
title: "scRNAseq_CITEseq_R_ReductionDimension"
author: "Laëtitia Racine"
date: "2022-10-03"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(dplyr)
library(ggplot2)
library(stringr)
library(cowplot)
library(gridExtra)
library(tibble)
library(clustree)

```

```{r, Working directories and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_ReductionDimension/"))
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_ReductionDimension/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_CITEseq_R_ReductionDimension/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

# Overview 

This code aims at :            
- choosing the best dimensionality for the dataset    
- performing linear dimensional reduction (PCA)    
- performing non-linear dimensional reduction (UMAP)  
- finding the best clustering resolution for the dataset  
  
Based on :    
- https://satijalab.org/seurat/articles/pbmc3k_tutorial.html#perform-linear-dimensional-reduction-1    
- https://satijalab.org/seurat/articles/dim_reduction_vignette.html    
- https://sib-swiss.github.io/single-cell-training/2022.7/day2/clustering/    
- https://github.com/satijalab/seurat/issues/1676  
- https://github.com/satijalab/seurat/issues/2814  
- https://github.com/satijalab/seurat/issues/5135 
- https://github.com/satijalab/seurat/issues/5761  
  
As output file, we obtain a seurat object with merged conditions, normalized and reduced with the optimal number of dimensions and resolution (merged_all_1-40dim). We also have the same object prepared for SCT FindMarkers function (merged_all_prep_1-40dim). We finally create lots of plots with dimension choice, PCA or UMAP. Each time, we worked with the SCT normalized object with cell cycle regression and without cell cycle regression to keep the choice for downstream analysis. 
  
# Analysis on the SCT normalized object WITHOUT cell cycle regression

<br>

This part includes function definition of the analysis and the application on the normalized object without cell cycle regression.  
If we want to study the normalized object with cell cycle regression, see part II.   

## Merge conditions into one seurat object

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_CITEseq_R_NormalizeData/"))

# Objects without cc regression
list_seurat = readRDS(paste0(dir, "/", "list_seurat_obj_qc_norm.rds"))
merged_all = merge(x = list_seurat[["CTRL"]], 
                   y = c(list_seurat[["CTRLaK"]], list_seurat[["DON"]], 
                         list_seurat[["DONaK"]], list_seurat[["2DG"]], 
                         list_seurat[["2DGaK"]], list_seurat[["VPA"]], 
                         list_seurat[["CTRL2"]], list_seurat[["AOA"]], 
                         list_seurat[["AOAaK"]]),
                   add.cell.ids = c("CTRL", "CTRLaK", "DON", "DONaK", "DDG", 
                                    "2DGaK", "VPA", "CTRL2", "AOA", "AOAaK"),
                   merge.data = TRUE)

features_scale = rownames(merged_all[["SCT"]]@scale.data)
VariableFeatures(merged_all, assay = "SCT") = features_scale
rm(list_seurat, features_scale)

```

As we decided there were no batch effect, we can consider CTRL and CTRL2 as one unique sample. We update meta.data with a new orig.ident. The original one is already stored in the column old.ident. 

```{r, Merged control ident}

metadata = data.frame(merged_all@meta.data) %>%
  dplyr::mutate(orig.ident = ifelse(old.ident %in% c("CTRL", "CTRL2"), "CTRL", old.ident))

merged_all@meta.data = metadata

```


## Perform linear dimensional reduction and determine the "dimensionality" (PCA)

Jackstraw method is time consuming for large dataset. We choose to use only the heuristic elbow plot method to determine the dimensionality. Dimensions are ordered according to their implication in the global variance. To keep only the dimensions that explain the most the variance, we identify the bend on the elbow plot.  
https://github.com/satijalab/seurat/issues/2295  
https://satijalab.org/seurat/reference/runpca  
RunPCA uses the variable features only with features = NULL (default).   
```{r, PCA and dimensionality, fig.width = 20}

dim_choice_fun = function(obj, name_assay) {
  
  # Choose default assay
  DefaultAssay(obj) = name_assay

  # PCA calculation
  obj = RunPCA(object = obj,
               assay = name_assay,
               npcs = 100, # default 50
               reduction.name = "pca",
               features = NULL,
               seed.use = 42,
               verbose = FALSE)

  # Plots
  pca_plot = DimPlot(object = obj, 
                     group.by = "orig.ident", 
                     reduction = "pca") + NoLegend() 
  
  pca_split_plot = DimPlot(object = obj, 
                           group.by = "orig.ident", 
                           split.by = "orig.ident", 
                           reduction = "pca", 
                           ncol = 3)
   
  elb_plot = ElbowPlot(object = obj, ndims = 100, reduction = "pca") +
    geom_vline(xintercept = 10, color = "red", size = 1) +
    geom_vline(xintercept = 20, color = "red", size = 1) +
    geom_vline(xintercept = 30, color = "red", size = 1) +
    geom_vline(xintercept = 40, color = "red", size = 1) +
    theme(plot.background = element_rect(fill = "white"))

  pca_dim = VizDimLoadings(object = obj,
                           dims = 1:30, 
                           reduction = "pca")

  # Outputs
  results = list(obj = obj,
                 pca_plot = pca_plot,
                 pca_split_plot = pca_split_plot,
                 elb_plot = elb_plot,
                 pca_dim = pca_dim)
  return(results)
  
}

dim_seurat_obj = dim_choice_fun(merged_all, "SCT")
merged_all = dim_seurat_obj$obj

```

```{r, PCA plot, fig.width = 20, fig.height=18}

(dim_seurat_obj$elb_plot | dim_seurat_obj$pca_plot) / dim_seurat_obj$pca_split_plot
dim_seurat_obj$pca_dim

```

```{r, fig.width = 20, fig.height=22}

dim_seurat_obj$pca_dim

```

```{r, Dimension choice}

print("We chose dimensions 1 --> 40 for downstream analysis.")
dim_choice = c(1:40)

```


## Perform non-linear dimensional reduction (UMAP)

https://satijalab.org/seurat/reference/runumap

```{r, UMAP}

umap_fun = function(obj, assay_name, dim_choice, color_code) {
  
  obj = RunUMAP(object = obj, 
                assay = name_assay,
                dims = dim_choice, 
                reduction =  "pca", 
                reduction.name = "umap",
                set.seed = 42,
                verbose = FALSE)

  umap_plot = DimPlot(object = obj, 
                      group.by = "orig.ident", 
                      reduction = "umap",
                      cols = color_code) 
  
  umap_split_plot = DimPlot(object = obj, 
                            group.by = "orig.ident", 
                            split.by = "orig.ident", 
                            reduction = "umap",
                            cols = color_code,
                            ncol = 3) + NoLegend()
  
  results = list(obj = obj,
                 umap_plot = umap_plot,
                 umap_split_plot = umap_split_plot)
  
  return(results)
  
}

umap_highlight = function(seurat_obj, color_code) {
  
  ind_umap = list()
  
  for (i in 1:length(unique(seurat_obj$orig.ident))) {
  
  # Select condition to highlight
  highlight_cond = unique(seurat_obj$orig.ident)[i]
  color = unname(noquote(color_code[highlight_cond]))
  others_cond = setdiff(unique(seurat_obj$orig.ident), highlight_cond)
  others_cond_group = paste(others_cond, collapse=", ")
  
  # Sort cell according to their condition
  cell_highlight = unlist(CellsByIdentities(object = seurat_obj, idents = highlight_cond))
  cell_nohighlight = unlist(CellsByIdentities(object = seurat_obj, idents = others_cond))
  
  # Plot 
  plot_umap = DimPlot(object = seurat_obj,
                      group.by = "orig.ident",
                      reduction ="umap",
                      pt.size = 0.05,
                      cells.highlight = list(cell_highlight, cell_nohighlight),
                      cols.highlight = c(color,"grey")) +
    ggtitle(highlight_cond) +
    theme(legend.position = "none")
  
  ind_umap[[i]] = plot_umap
  
  }
  
  # function's output 
  return(ind_umap)
  
}

umap_seurat_obj = umap_fun(merged_all, "SCT", dim_choice, color_code)
merged_all = umap_seurat_obj$obj
merged_all = SetIdent(object = merged_all, value = "orig.ident")
umap_hl = umap_highlight(merged_all, color_code)

```

```{r, UMAP plot1, fig.width = 14, fig.height = 14}

umap_seurat_obj$umap_plot / umap_seurat_obj$umap_split_plot

```

NB : Not the same graph as in BatchEffectCorrection script because we didn't choose the same dimensions. If we change the dimensions in 1:30, we obtain the same graph.  
**For latter downstream analysis, we choose to use the merged object without cc and the SCT normalization with 1-40 dimensions.**  

```{r, UMAP plot2, fig.width = 22, fig.height = 20}

do.call("grid.arrange", c(umap_hl, ncol = 3))

```


## Determine clusters resolution

We are looking for the clustering resolution that describes the best the dataset and highlights the differences between conditions.  
https://satijalab.org/seurat/reference/findneighbors  
https://satijalab.org/seurat/reference/findclusters  

<br>

```{r, Resolution Calcul}

resolution_possibility = function(seurat_obj, dim_choice) {

  seurat_obj = FindNeighbors(object = seurat_obj, 
                             assay = "SCT",
                             dims = dim_choice,
                             verbose = FALSE)
  
  # Try different resolution to find the best granularity
  for (res in c(0.01, 0.02, 0.03, 0.04, 0.05, 0.06, 0.07, 0.08, 0.09, 
                0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1)) {
    seurat_obj <- FindClusters(
      object = seurat_obj, 
      resolution = res, 
      algorithm = 3,
      verbose = FALSE)
      # 1 = original Louvain algorithm
      # 2 = Louvain algorithm with multilevel refinement
      # 3 = SLM algorithm => SLM, Blondel et al., Journal of Statistical Mechanics
      # 4 = Leiden algorithm
  }
  return(seurat_obj)
  
}

merged_all = resolution_possibility(merged_all, dim_choice)

```

```{r, Resolution plot, fig.width = 20, fig.height = 26}

plot_resolution_fun = function(seurat_obj) {
  
  plot_resolution = plot_grid(ncol = 3, 
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.01") + ggtitle("Resolution_0.01"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.02") + ggtitle("Resolution_0.02"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.03") + ggtitle("Resolution_0.03"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.04") + ggtitle("Resolution_0.04"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.05") + ggtitle("Resolution_0.05"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.06") + ggtitle("Resolution_0.06"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap",
            group.by = "SCT_snn_res.0.07") + ggtitle("Resolution_0.07"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.08") + ggtitle("Resolution_0.08"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.09") + ggtitle("Resolution_0.09"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.1") + ggtitle("Resolution_0.1"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.2") + ggtitle("Resolution_0.2"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.3") + ggtitle("Resolution_0.3"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.4") + ggtitle("Resolution_0.4"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.5") + ggtitle("Resolution_0.5"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.6") + ggtitle("Resolution_0.6"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.7") + ggtitle("Resolution_0.7"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.8") + ggtitle("Resolution_0.8"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.0.9") + ggtitle("Resolution_0.9"),
    DimPlot(seurat_obj, label = TRUE, reduction = "umap", 
            group.by = "SCT_snn_res.1") + ggtitle("Resolution_1"))

  return(plot_resolution)
  
}

resolution_merged_all = plot_resolution_fun(merged_all)
print(resolution_merged_all)

```

<br>
To help in the decision, we used the clustree package : https://cran.r-project.org/web/packages/clustree/vignettes/clustree.html.  
<br>

```{r, Resolution choice, fig.height=16, fig.width=10}

clustree(merged_all@meta.data, prefix = "SCT_snn_res.")

print("Based on the conditions we want to analyze and the confirmation of the tree, resolution 0.6 seems the best for our study. However, clusters 15 and 17 from resolutions 0.7 are also interesting  because they allow to separate conditions. We are going to add them manually to the clustering by extracting the cells from cluster 11 and do a reclustering.")

```

```{r, Reclustering cluster 12, fig.width = 20, fig.height = 10}

# extract cells from cluster 12
merged_all = SetIdent(object = merged_all, value = "SCT_snn_res.0.6")
obj = subset(merged_all, idents = "11")
obj = FindNeighbors(object = obj, 
                    assay = "SCT",
                    dims = dim_choice,
                    verbose = FALSE)
# reclusterise the cells in order to have two clusters
for (res in c(0.01, 0.05, 0.1, 0.5, 1)) {
    obj <- FindClusters(
      object = obj, 
      resolution = res, 
      algorithm = 3,
      verbose = FALSE)
  }

# visualize the different resolutions to choose the best one
plot = plot_grid(ncol = 3, 
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.0.01") + ggtitle("Resolution_0.01"),
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.0.05") + ggtitle("Resolution_0.05"),
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.0.1") + ggtitle("Resolution_0.1"),
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.0.5") + ggtitle("Resolution_0.5"),
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.1") + ggtitle("Resolution_1"))
plot
print("Resolution 0.05 is the best to reclusterize those cells.")

```

NB : if the reclustering had failed, we would have worked with the raw data => Extract the cells we want to recluster, normalize and find their specific variable features and then apply a new clustering.  

```{r, Add reclustering info to seurat object}

# Create new column of clustering
new_clust = obj@meta.data %>% 
  dplyr::select("SCT_snn_res.0.05") %>% 
  dplyr::rename("new_clusters" = "SCT_snn_res.0.05") %>%
  dplyr::mutate(new_clusters = ifelse(new_clusters == 0, 11, 16))

old_clust = merged_all@meta.data %>% dplyr::select(SCT_snn_res.0.6)
old_clust$SCT_snn_res.0.6 = as.numeric(levels(old_clust$SCT_snn_res.0.6))[old_clust$SCT_snn_res.0.6]
# Attention ! utiliser as.integer() change la valeur dans le tableau...

new_clust = full_join(x = old_clust %>% tibble::rownames_to_column(var = "cell"), 
                      y = new_clust %>% tibble::rownames_to_column(var = "cell"), 
                      by = "cell") %>%
  dplyr::mutate(clusters = ifelse(is.na(new_clusters), SCT_snn_res.0.6, new_clusters)) %>%
  dplyr::select(-new_clusters, -SCT_snn_res.0.6)

# Add new clustering information in seurat object metadata
metadata_seurat = merged_all@meta.data %>%
  dplyr::select(-SCT_snn_res.0.01,-SCT_snn_res.0.02,-SCT_snn_res.0.03,
                -SCT_snn_res.0.04,-SCT_snn_res.0.05,-SCT_snn_res.0.06, 
                -SCT_snn_res.0.07,-SCT_snn_res.0.08,-SCT_snn_res.0.09,
                -SCT_snn_res.0.1,-SCT_snn_res.0.2,-SCT_snn_res.0.3,
                -SCT_snn_res.0.4,-SCT_snn_res.0.5,-SCT_snn_res.0.6, 
                -SCT_snn_res.0.7,-SCT_snn_res.0.8,-SCT_snn_res.0.9,
                -SCT_snn_res.1,-seurat_clusters) %>%
  dplyr::mutate(clusters = new_clust$clusters)
metadata_seurat$clusters = as.factor(metadata_seurat$clusters)

merged_all@meta.data = metadata_seurat
res_choice = DimPlot(merged_all, label = TRUE, reduction = "umap", group.by = "clusters")
print(res_choice)

```

```{r, Prepare the object for markers identification}

merged_all = SetIdent(merged_all, value = "clusters")
DefaultAssay(merged_all) = "SCT"
merged_prep = PrepSCTFindMarkers(merged_all)

```

```{r, Save outputs, fig.show='hide'}

# Without cc regression
ggsave(plot = (dim_seurat_obj$elb_plot | dim_seurat_obj$pca_plot) / dim_seurat_obj$pca_split_plot, 
       filename = paste0(directory_output, "dimension_choice_elb_pca.svg"),
       height = 18, width = 20)
ggsave(plot = dim_seurat_obj$pca_dim, 
       filename = paste0(directory_output, "dimension_choice_pca_genes.svg"),
       height = 26, width = 20)
ggsave(plot = umap_seurat_obj$umap_plot,
       filename = paste0(directory_output, "umap_1-40dim.svg"),
       height = 8, width = 12)
ggsave(plot = umap_seurat_obj$umap_split_plot,
       filename = paste0(directory_output, "umap_1-40dim_split.svg"),
       height = 12, width = 20)
ggsave(plot = do.call("grid.arrange", c(umap_hl, ncol = 3)),
       filename = paste0(directory_output, "umap_1-40dim_split_highlight.svg"),
       height = 18, width = 20)
ggsave(plot = resolution_merged_all,
       filename = paste0(directory_output, "umap_clusters_resolution.svg"),
       height = 24, width = 24)
ggsave(plot = res_choice,
       filename = paste0(directory_output, "umap_clusters_res_choice.svg"),
       height = 8, width = 24)

saveRDS(merged_all, paste0(directory_output, "merged_all_1-40dim.rds"))
saveRDS(merged_prep, paste0(directory_output, "merged_all_prep_1-40dim.rds"))

```

The object *merged_all_prep_1-40dim.rds* is ready to use for further downstream analysis. It contains merged controls (with same ident) and merged drugs condition (with different ident). The normalized matrix can be found in the SCT assay and the clusters are the optimal ones. The cluster information is available in the *cluster metadata column* for each cell.   
  
```{r, for article}
res_choice_light = DimPlot(merged_all, label = FALSE, reduction = "umap", group.by = "clusters")
ggsave(plot = res_choice_light,
       filename = paste0(directory_output, "umap_clusters_res_choice_light.svg"),
       height = 8, width = 6)
```
  
# Analysis on the SCT normalized object WITH cell cycle regression

```{r}

# Object with cc regression
list_seurat_cc = readRDS(paste0(dir, "/", "list_seurat_obj_qc_norm_ccr.rds"))
merged_cc = merge(x = list_seurat_cc[["CTRL"]],
                  y = c(list_seurat_cc[["CTRLaK"]], list_seurat_cc[["DON"]],
                        list_seurat_cc[["DONaK"]], list_seurat_cc[["2DG"]],
                        list_seurat_cc[["2DGaK"]], list_seurat_cc[["VPA"]],
                        list_seurat_cc[["CTRL2"]], list_seurat_cc[["AOA"]],
                        list_seurat_cc[["AOAaK"]]),
                  add.cell.ids = c("CTRL", "CTRLaK", "DON", "DONaK", "DDG", 
                                    "2DGaK", "VPA", "CTRL2", "AOA", "AOAaK"),
                  merge.data = TRUE)
features_all_cc = rownames(merged_cc[["SCT"]]@scale.data)
VariableFeatures(merged_cc, assay = "SCT") = features_all_cc
rm(list_seurat_cc, features_all_cc)

dim_seurat_obj_cc = dim_choice_fun(merged_cc, "SCT")
merged_cc = dim_seurat_obj_cc$obj

```

```{r, fig.height = 20, fig.width = 18}

(dim_seurat_obj_cc$elb_plot | dim_seurat_obj_cc$pca_plot) / dim_seurat_obj_cc$pca_split_plot

```

```{r, fig.height = 20, fig.width = 22}

dim_seurat_obj_cc$pca_dim

print("We chose dimensions 1 --> 40 for downstream analysis.")
dim_choice = c(1:40)

```

```{r}

umap_seurat_obj_cc = umap_fun(merged_cc, "SCT", dim_choice, color_code)
merged_cc = umap_seurat_obj_cc$obj
merged_cc = SetIdent(object = merged_cc, value = "orig.ident")
umap_hl_cc = umap_highlight(merged_cc, color_code)

```

```{r, fig.width = 14, fig.height = 14}

umap_seurat_obj_cc$umap_plot / umap_seurat_obj_cc$umap_split_plot

```

```{r, fig.width = 22, fig.height = 20}

do.call("grid.arrange", c(umap_hl_cc, ncol = 3))

```

```{r, fig.height=16, fig.width=10}

merged_cc = resolution_possibility(merged_cc, dim_choice)
resolution_merged_cc = plot_resolution_fun(merged_cc)
print(resolution_merged_cc)

```

```{r, fig.height=16, fig.width=10}

clustree(merged_cc@meta.data, prefix = "SCT_snn_res.")

print("Based on the conditions we want to analyze, resolution 0.6 seems the best. Clusters 14 and 15 from resolutions 0.7 are also interesting so we are going to add them manually to the clustering by extracting the cells from cluster 12 and do a reclustering.")

```

```{r, fig.width = 20, fig.height = 10}

# extract cells from cluster 12
merged_cc = SetIdent(object = merged_cc, value = "SCT_snn_res.0.6")
obj = subset(merged_cc, idents = "12")
obj = FindNeighbors(object = obj, 
                    assay = "SCT",
                    dims = dim_choice)
# reclusterise the cells in order to have two clusters
for (res in c(0.01, 0.05, 0.1, 0.5, 1)) {
    obj <- FindClusters(
      object = obj, 
      resolution = res, 
      algorithm = 3)
  }

# visualize the different resolutions to choose the best one
plot = plot_grid(ncol = 3, 
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.0.01") + ggtitle("Resolution_0.01"),
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.0.05") + ggtitle("Resolution_0.05"),
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.0.1") + ggtitle("Resolution_0.1"),
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.0.5") + ggtitle("Resolution_0.5"),
                 DimPlot(obj, label = TRUE, reduction = "umap", group.by = "SCT_snn_res.1") + ggtitle("Resolution_1"))
plot
print("Resolution 0.05 is the best to reclusterize those cells.")

```

```{r}
# Create new column of clusterings
new_clust = obj@meta.data %>% 
  dplyr::select("SCT_snn_res.0.05") %>% 
  dplyr::rename("new_clusters" = "SCT_snn_res.0.05") %>%
  dplyr::mutate(new_clusters = ifelse(new_clusters == 0, 12, 16))

old_clust = merged_cc@meta.data %>% dplyr::select(SCT_snn_res.0.6)
old_clust$SCT_snn_res.0.6 = as.numeric(levels(old_clust$SCT_snn_res.0.6))[old_clust$SCT_snn_res.0.6]
# Attention ! utiliser as.integer() change la valeur dans le tableau...

new_clust = full_join(x = old_clust %>% tibble::rownames_to_column(var = "cell"), 
                      y = new_clust %>% tibble::rownames_to_column(var = "cell"), 
                      by = "cell") %>%
  dplyr::mutate(clusters = ifelse(is.na(new_clusters), SCT_snn_res.0.6, new_clusters)) %>%
  dplyr::select(-new_clusters, -SCT_snn_res.0.6)

# Add new clustering information in seurat object metadata
metadata_seurat = merged_cc@meta.data %>%
  dplyr::select(-SCT_snn_res.0.01,-SCT_snn_res.0.02,-SCT_snn_res.0.03,
                -SCT_snn_res.0.04,-SCT_snn_res.0.05,-SCT_snn_res.0.06, 
                -SCT_snn_res.0.07,-SCT_snn_res.0.08,-SCT_snn_res.0.09,
                -SCT_snn_res.0.1,-SCT_snn_res.0.2,-SCT_snn_res.0.3,
                -SCT_snn_res.0.4,-SCT_snn_res.0.5,-SCT_snn_res.0.6, 
                -SCT_snn_res.0.7,-SCT_snn_res.0.8,-SCT_snn_res.0.9,
                -SCT_snn_res.1,-seurat_clusters) %>%
  dplyr::mutate(clusters = new_clust$clusters)
metadata_seurat$clusters = as.factor(metadata_seurat$clusters)

merged_cc@meta.data = metadata_seurat
res_choice_cc = DimPlot(merged_cc, label = TRUE, reduction = "umap", group.by = "clusters")
print(res_choice_cc)

```

```{r}

merged_cc = SetIdent(merged_cc, value = "clusters")
DefaultAssay(merged_cc) = "SCT"
merged_cc_prep = PrepSCTFindMarkers(merged_cc)

```

```{r, fig.show='hide'}

ggsave(plot = (dim_seurat_obj_cc$elb_plot | dim_seurat_obj_cc$pca_plot) / dim_seurat_obj_cc$pca_split_plot,
       filename = paste0(directory_output, "dimension_choice_elb_pca_cc.svg"),
       height = 18, width = 20)
ggsave(plot = dim_seurat_obj_cc$pca_dim,
       filename = paste0(directory_output, "dimension_choice_pca_genes_cc.svg"),
       height = 26, width = 20)
ggsave(plot = umap_seurat_obj_cc$umap_plot,
       filename = paste0(directory_output, "umap_1-40dim_cc.svg"),
       height = 8, width = 12)
ggsave(plot = umap_seurat_obj_cc$umap_split_plot,
       filename = paste0(directory_output, "umap_1-40dim_split_cc.svg"),
       height = 12, width = 20)
ggsave(plot = do.call("grid.arrange", c(umap_hl_cc, ncol = 3)),
       filename = paste0(directory_output, "umap_1-40dim_split_highlight_cc.svg"),
       height = 18, width = 20)
ggsave(plot = resolution_merged_cc,
       filename = paste0(directory_output, "umap_clusters_resolution_cc.svg"),
       height = 24, width = 24)
ggsave(plot = res_choice_cc,
       filename = paste0(directory_output, "umap_clusters_res_choice_cc.svg"),
       height = 8, width = 24)

saveRDS(merged_cc, paste0(directory_output, "merged_cc_1-40dim.rds"))
saveRDS(merged_cc_prep, paste0(directory_output, "merged_cc_prep_1-40dim.rds"))

```


```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results='hide'}

# Clean working space and memory 
rm(list = ls())
gc()

```