---
title: "scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA"
author: "Laëtitia Racine"
date: "2022-10-03"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r, Setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(dplyr)
library(ggplot2)
library(stringr)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(scCustomize) # function Split_FeatureScatter
library(kableExtra)
library(viridis)

```

```{r, Working directories and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA/"))
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_CorrelationADTRNA/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```



<br><br><br>



# Overview 

This code aims at :    
- visualizing the correlation between RNA and protein for CD34 and CD133    
- identifying groups of cells with different protein expression if they exist    
- observing protein expression among cells clusters    
    
It is based on :    
https://satijalab.org/seurat/articles/multimodal_vignette.html    
https://github.com/satijalab/seurat/issues/1080    
  
We work from the count matrix of the pre-processed seurat object (merged_all_1-40dim from scRNAseq_CITEseq_R_ReductionDimension). The **"count"** slot contains raw data whereas the **"data"** slot contains normalized data with CLR algorithm. We use here the normalized data.   

As **output files**, we obtain a seurat object with proteins groups in the meta.data table (merged_all_CITEgroups.rds), tables with cells repartition in groups and cluster and conditions and plots.  
  
```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_CITEseq_R_ReductionDimension/"))
merged_all = readRDS(file = paste0(dir, "/", "merged_all_1-40dim.rds"))

```

```{r}

cd34_expr = full_join(x = data.frame(merged_all@assays$ADT@counts[1,]) %>%
                        rownames_to_column(var = "cell"),
                      y = data.frame(merged_all@assays$ADT@data[1,]) %>%
                        rownames_to_column(var = "cell"),
                      by = "cell")
colnames(cd34_expr) = c("cell", "cd34_raw_value", "cd34_normalized_value")

cd133_expr = full_join(x = data.frame(merged_all@assays$ADT@counts[2,]) %>%
                        rownames_to_column(var = "cell"),
                      y = data.frame(merged_all@assays$ADT@data[2,]) %>%
                        rownames_to_column(var = "cell"),
                      by = "cell")
colnames(cd133_expr) = c("cell", "cd133_raw_value", "cd133_normalized_value")

prot_expr = full_join(cd34_expr, cd133_expr, by = "cell")

prot_expr %>%
  kable() %>%
  kable_styling() %>%
  scroll_box(height = "250px", width = "100%")

```

<br>

We depicted the data on plots to visualize the proteins' distribution in the population. We can identify outliers (i.e. cells expressing the protein much more than the rest of the cell population).
```{r, fig.height = 5, fig.width = 12}

print("Statistical distribution of CD34 protein expression : ")
summary(prot_expr$cd34_normalized_value)
print("Statistical distribution of CD133 protein expression : ")
summary(prot_expr$cd133_normalized_value)

cd34_expr_distri = ggplot(prot_expr, aes(x=cd34_normalized_value, y=0)) +
  geom_jitter(height = 10, width = 0, size = 1, color = "black") +
  ggplot2::annotate("segment", x=0, xend=5, y=0, yend=0, size=2, color = "grey") +
  ggplot2::annotate("segment", x=0, xend=0, y=-4,yend=4, size=2,  color = "grey") +
  ggplot2::annotate("segment", x=5, xend=5, y=-4,yend=4, size=2,  color = "grey") +
  ggplot2::annotate("segment", x=0.4539, xend=0.4539, y=-10,yend=10, size=1,  color = "red", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=0.4539, y=11, label="Q1", color="red", angle = 90) +
  ggplot2::annotate("segment", x=0.7093, xend=0.7093,y=-10,yend=10, size=1,  color = "red", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=0.7093, y=11, label="Med", color="red", angle = 90) +
  ggplot2::annotate("segment", x=1.0123, xend=1.0123, y=-10,yend=10, size=1,  color = "red", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=1.0123, y=11, label="Q3", color="red", angle = 90) +
  ggplot2::annotate("segment", x=1.75, xend=1.75, y=-10,yend=10, size=1,  color = "blue", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=1.75, y=11, label="1.75", color="blue", angle = 90) +
  ggplot2::annotate("segment", x=2.6, xend=2.6, y=-10,yend=10, size=1,  color = "blue", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=2.6, y=11, label="2.6", color="blue", angle = 90) +
  scale_y_continuous(limits = c(-11,11)) +
  theme(legend.title = element_blank(), 
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

cd133_expr_distri = ggplot(prot_expr, aes(x=cd133_normalized_value, y=0)) +
  geom_jitter(height = 10, width = 0, size = 1, color = "black") +
  ggplot2::annotate("segment", x=0, xend=8, y=0, yend=0, size=2, color = "grey") +
  ggplot2::annotate("segment", x=0, xend=0, y=-4,yend=4, size=2,  color = "grey") +
  ggplot2::annotate("segment", x=8, xend=8, y=-4,yend=4, size=2,  color = "grey") +
  ggplot2::annotate("segment", x=0.5031, xend=0.5031, y=-10,yend=10, size=1,  color = "red", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=0.5031, y=11, label="Q1", color="red", angle = 90) +
  ggplot2::annotate("segment", x=0.6649, xend=0.6649, y=-10,yend=10, size=1,  color = "red", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=0.6649, y=11, label="Med", color="red", angle = 90) +
  ggplot2::annotate("segment", x=0.8652, xend=0.8652, y=-10,yend=10, size=1,  color = "red", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=0.8652, y=11, label="Q3", color="red", angle = 90) +
  ggplot2::annotate("segment", x=1.75, xend=1.75, y=-10,yend=10, size=1,  color = "blue", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=1.75, y=11, label="1.75", color="blue", angle = 90) +
  ggplot2::annotate("segment", x=2.2, xend=2.2, y=-10,yend=10, size=1,  color = "blue", linetype = "longdash") +
  ggplot2::annotate(geom="text", x=2.2, y=11, label="2.2", color="blue", angle = 90) +
  scale_y_continuous(limits = c(-11,11)) +
  theme(legend.title = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())

cd34_expr_distri | cd133_expr_distri

```

<br><br><br>

# Membrane protein vs RNA in each culture condition

<br>

For UMAP representation of protein expression (ADT), we used the outliers representation above to choose threshold values. We set a max cutoff at 2.6 for CD34 and 2.2 for CD133 to have a more readable scale. This means that all values above those cutoffs have the same color on the scale. For RNA expression, we use the predefined max and min cutoff (95% and 5%) because we don't have precise value to provide.

```{r, UMAP ADT RNA}

DefaultAssay(merged_all) = "ADT"
p1 = FeaturePlot(object = merged_all,
                 features = "cite_CD34.1",
                 cols = viridis_magma_dark_high,
                 max.cutoff = "2.6",
                 split.by = "orig.ident",
                 combine = FALSE)                   
p2 = FeaturePlot(object = merged_all,
                 features = "cite_CD133",
                  cols = viridis_magma_dark_high,
                 max.cutoff = "2.2",
                 split.by = "orig.ident",
                 combine = FALSE)

DefaultAssay(merged_all) = "SCT"
p3 = FeaturePlot(object = merged_all,
                 features = "sct_CD34",
                  cols = viridis_magma_dark_high,
                 max.cutoff = "q95",
                 min.cutoff = "q05",
                 split.by = "orig.ident",
                 combine = FALSE)
p4 = FeaturePlot(object = merged_all,
                 features = "sct_PROM1",
                  cols = viridis_magma_dark_high,
                 max.cutoff = "q95",
                 min.cutoff = "q05",
                 split.by = "orig.ident",
                 combine = FALSE)

```

**CD34 membrane protein detection and RNA expression in each condition.**

```{r, results='hold', fig.width=18, fig.height=8, fig.fullwidth=TRUE}

do.call("grid.arrange", c(p1, ncol=3))
do.call("grid.arrange", c(p3, ncol=3))

```

**CD133 membrane protein detection and RNA expression in each condition.**

```{r, results='hold', fig.width=18, fig.height=8, fig.fullwidth=TRUE}

do.call("grid.arrange", c(p2, ncol=3))
do.call("grid.arrange", c(p4, ncol=3))

```

**Scatter plot RNA CD34 vs CD133 (PROM1) and CD34 protein (CD34.1) vs CD133 protein.**

```{r, Scatter ADT RNA, results='hold', fig.width=18, fig.height=16, fig.fullwidth=TRUE}

Idents(merged_all) = "orig.ident"

DefaultAssay(merged_all) = "SCT"
sc1 = Split_FeatureScatter(merged_all,
                     feature1 = "CD34",
                     feature2 = "PROM1",
                     split.by = "orig.ident",
                     group.by = "orig.ident",
                     num_columns = 3) & NoLegend()

DefaultAssay(merged_all) = "ADT"
sc2 = Split_FeatureScatter(merged_all,
                     feature1 = "CD34.1",
                     feature2 = "CD133",
                     split.by = "orig.ident",
                     group.by = "orig.ident",
                     num_columns = 3) & NoLegend()

sc1/sc2

```


<br><br><br>


# Membrane protein groups in each culture condition

We work with the normalized matrix stored in the *data* slot of ADT assay.   
We decided to use the *kmean algorithm* to define groups of expression in the cells. We tried with two groups (high/low) and three groups (high/medium/low).  

```{r, Choose assay}
DefaultAssay(merged_all) = "ADT"
```

**Two groups : high/low**

```{r, CD34 two groups, fig.width=12, fig.height=8}

# # Organize tab to calcul kmean
protein_tab_cd34 = as.data.frame(merged_all@assays$ADT@data)["CD34.1",]
protein_tab_cd34 = tidyr::pivot_longer(data = protein_tab_cd34,
                                  cols = 1:ncol(protein_tab_cd34),
                                  names_to = "cell",
                                  values_to = "protein")
protein_tab_cd34 = protein_tab_cd34 %>% tibble::column_to_rownames(var="cell")

# Kmean calcul for two clusters
set.seed(42) # !!! important for repeatability
fit = kmeans(protein_tab_cd34, 2)
protein_tab_cd34 = data.frame(protein_tab_cd34, fit$cluster)
table(protein_tab_cd34$fit.cluster)
# Limits of each cluster
lim_clust1 = c(min(protein_tab_cd34[protein_tab_cd34$fit.cluster==1,1]),
               max(protein_tab_cd34[protein_tab_cd34$fit.cluster==1,1]))
lim_clust2 = c(min(protein_tab_cd34[protein_tab_cd34$fit.cluster==2,1]),
               max(protein_tab_cd34[protein_tab_cd34$fit.cluster==2,1]))
# Ascending order
limits = sort(c(lim_clust1, lim_clust2))

# Threshold between the two clusters (use max of first cluster and min of second one)
var_kmean = mean(limits[2:3])
# Add the new groups into metadata of seurat object
cat("Limits cluster 1 :", lim_clust1, "=> high", "\n",
    "Limits cluster 2 :", lim_clust2, "=> low", "\n",
    "Kmean value = ", var_kmean)
# Add kmean information in the tab
protein_tab_cd34 = protein_tab_cd34 %>%
  dplyr::mutate(CD34_two_groups = ifelse(fit.cluster==1, "CD34_high", "CD34_low")) %>%
  tibble::rownames_to_column(var = "cell") %>%
  dplyr::select(cell, CD34_two_groups)

# Add the new groups into metadata of seurat object
meta_tab = as.data.frame(merged_all@meta.data) %>% rownames_to_column(var = "cell")
meta_tab = full_join(x = meta_tab, y = protein_tab_cd34, by = "cell")
meta_tab = meta_tab %>% column_to_rownames(var = "cell")
merged_all@meta.data = meta_tab

# Show the groups
cd34_umap_two = DimPlot(object = merged_all,
             split.by = "orig.ident",
             group.by = "CD34_two_groups",
             reduction = "umap",
             ncol = 3)
cd34_umap_two

table(merged_all$CD34_two_groups)
table(merged_all$orig.ident, merged_all$CD34_two_groups)

```

```{r, CD133 two groups, fig.width=12, fig.height=8}

# Organize tab to calcul kmean
protein_tab_cd133 = as.data.frame(merged_all@assays$ADT@data)["CD133",]
protein_tab_cd133 = tidyr::pivot_longer(data = protein_tab_cd133,
                                  cols = 1:ncol(protein_tab_cd133),
                                  names_to = "cell",
                                  values_to = "protein")
protein_tab_cd133 = protein_tab_cd133 %>% tibble::column_to_rownames(var="cell")

# Kmean calcul for two clusters
set.seed(42) # !!! important
fit = kmeans(protein_tab_cd133, 2)
protein_tab_cd133 = data.frame(protein_tab_cd133, fit$cluster)
table(protein_tab_cd133$fit.cluster)
# Limits of each cluster
lim_clust1 = c(min(protein_tab_cd133[protein_tab_cd133$fit.cluster==1,1]),
               max(protein_tab_cd133[protein_tab_cd133$fit.cluster==1,1]))
lim_clust2 = c(min(protein_tab_cd133[protein_tab_cd133$fit.cluster==2,1]),
               max(protein_tab_cd133[protein_tab_cd133$fit.cluster==2,1]))
# Ascending order
limits = sort(c(lim_clust1, lim_clust2))
# Threshold between the two clusters (use max of first cluster and min of second one)
var_kmean = mean(limits[2:3])
# Add the new groups into metadata of seurat object
cat("Limits cluster 1 :", lim_clust1, "=> low", "\n",
    "Limits cluster 2 :", lim_clust2, "=> high", "\n",
    "Kmean value = ", var_kmean)
# Add kmean information in the tab
protein_tab_cd133 = protein_tab_cd133 %>%
  dplyr::mutate(CD133_two_groups = ifelse(fit.cluster==1, "CD133_low", "CD133_high")) %>%
  tibble::rownames_to_column(var = "cell") %>%
  dplyr::select(cell, CD133_two_groups)

# Add the new groups into metadata of seurat object
meta_tab = as.data.frame(merged_all@meta.data) %>% rownames_to_column(var = "cell")
meta_tab = full_join(x = meta_tab, y = protein_tab_cd133, by = "cell")
meta_tab = meta_tab %>% column_to_rownames(var = "cell")
merged_all@meta.data = meta_tab

# Show the groups
cd133_umap_two = DimPlot(object = merged_all,
             split.by = "orig.ident",
             group.by = "CD133_two_groups",
             reduction = "umap",
             ncol = 3)
cd133_umap_two

table(merged_all$CD133_two_groups)
table(merged_all$orig.ident, merged_all$CD133_two_groups)

```

```{r, CD34 CD133 two groups tab}

# Tab number
tab_34 = data.frame(table(merged_all@meta.data$CD34_two_groups, 
                          merged_all@meta.data$orig.ident)) %>%
  tidyr::pivot_wider(names_from = Var2, values_from = Freq) %>%
  dplyr::rename("Surface_Protein" = "Var1")

tab_133 = data.frame(table(merged_all@meta.data$CD133_two_groups, 
                           merged_all@meta.data$orig.ident)) %>%
  tidyr::pivot_wider(names_from = Var2, values_from = Freq) %>%
  dplyr::rename("Surface_Protein" = "Var1")

tab_raw_cond = rbind(tab_34, tab_133)

tab_raw_cond %>%
  kable(caption = "Raw cell distribution in conditions") %>%
  kable_styling()

# Tab number normalized
tot_nb_cells = tab_raw_cond[1,-1]+tab_raw_cond[2,-1]
tab_norm_cond = as.data.frame(round(mapply('/', tab_raw_cond[,-1] , tot_nb_cells)*1000))
tab_norm_cond$Surface_Protein = tab_raw_cond$Surface_Protein
tab_norm_cond = tab_norm_cond %>% dplyr::relocate(Surface_Protein, .before = "2DG")

tab_norm_cond %>%
  kable(caption = "Normalized cell distribution in conditions per number of cell per condition.") %>%
  kable_styling()

# Tab percent 
tab_percent_cond = as.data.frame(mapply('/', tab_norm_cond[,-1], 10))
tab_percent_cond$Surface_Protein = tab_norm_cond$Surface_Protein
tab_percent_cond = tab_percent_cond %>% dplyr::relocate(Surface_Protein, .before = "2DG")
tab_percent_cond %>%
  kable(caption = "Percentage of cells with high/low surface protein for each condition (from normalized per number of cells per condition numbers).") %>%
  kable_styling()

# save tables  
write.csv2(x = tab_raw_cond, row.names = FALSE,
           file = paste0(directory_output, "tab_raw_cond_two_groups.csv"))
write.csv2(x = tab_norm_cond, row.names = FALSE,
           file = paste0(directory_output, "tab_norm_cond_two_groups.csv"))
write.csv2(x = tab_percent_cond, row.names = FALSE,
           file = paste0(directory_output, "tab_percent_cond_two_groups.csv"))

```

**Three groups : high/medium/low** 

```{r, CD34 three groups, fig.width=12, fig.height=8}

# Organize tab to calcul kmean
protein_tab_cd34 = as.data.frame(merged_all@assays$ADT@data)["CD34.1",]
protein_tab_cd34 = tidyr::pivot_longer(data = protein_tab_cd34,
                                  cols = 1:ncol(protein_tab_cd34),
                                  names_to = "cell",
                                  values_to = "protein")
protein_tab_cd34 = protein_tab_cd34 %>% tibble::column_to_rownames(var="cell")

# Kmean calcul for three clusters
set.seed(42) # !!! important for repeatability
fit = kmeans(protein_tab_cd34, 3)
protein_tab_cd34 = data.frame(protein_tab_cd34, fit$cluster)
table(protein_tab_cd34$fit.cluster)
# Limits of each cluster
lim_clust1 = c(min(protein_tab_cd34[protein_tab_cd34$fit.cluster==1,1]),
               max(protein_tab_cd34[protein_tab_cd34$fit.cluster==1,1]))
lim_clust2 = c(min(protein_tab_cd34[protein_tab_cd34$fit.cluster==2,1]),
               max(protein_tab_cd34[protein_tab_cd34$fit.cluster==2,1]))
lim_clust3 = c(min(protein_tab_cd34[protein_tab_cd34$fit.cluster==3,1]),
               max(protein_tab_cd34[protein_tab_cd34$fit.cluster==3,1]))
# Ascending order
limits = sort(c(lim_clust1, lim_clust2, lim_clust3))
# Threshold between the clusters (use max of first cluster and min of second one)
var_kmean = c(mean(limits[2:3]), mean(limits[4:5]))
# Add the new groups into metadata of seurat object
cat("Limits cluster 1 :", lim_clust1, "=> high", "\n",
    "Limits cluster 2 :", lim_clust2, "=> low", "\n",
    "Limits cluster 3 :", lim_clust3, "=> medium", "\n",
    "Kmean values = ", var_kmean)
# Add kmean information in the tab
protein_tab_cd34 = protein_tab_cd34 %>%
  dplyr::mutate(CD34_three_groups = ifelse(fit.cluster==1, "CD34_high", 
                                           ifelse(fit.cluster==2, "CD34_low", "CD34_medium"))) %>%
  tibble::rownames_to_column(var = "cell") %>%
  dplyr::select(cell, CD34_three_groups)

# Add the new groups into metadata of seurat object
meta_tab = as.data.frame(merged_all@meta.data) %>% rownames_to_column(var = "cell")
meta_tab = full_join(x = meta_tab, y = protein_tab_cd34, by = "cell")
meta_tab = meta_tab %>% column_to_rownames(var = "cell")
merged_all@meta.data = meta_tab

# Show the groups
cd34_umap_three = DimPlot(object = merged_all,
             split.by = "orig.ident",
             group.by = "CD34_three_groups",
             reduction = "umap",
             ncol = 3)
cd34_umap_three

table(merged_all$CD34_three_groups)
table(merged_all$orig.ident, merged_all$CD34_three_groups)

```

```{r, CD133 three groups, fig.width=12, fig.height=8}

# Organize tab to calcul kmean
protein_tab_cd133 = as.data.frame(merged_all@assays$ADT@data)["CD133",]
protein_tab_cd133 = tidyr::pivot_longer(data = protein_tab_cd133,
                                  cols = 1:ncol(protein_tab_cd133),
                                  names_to = "cell",
                                  values_to = "protein")
protein_tab_cd133 = protein_tab_cd133 %>% tibble::column_to_rownames(var="cell")

# Kmean calcul for three clusters
set.seed(42) # !!! important
fit = kmeans(protein_tab_cd133, 3)
protein_tab_cd133 = data.frame(protein_tab_cd133, fit$cluster)
table(protein_tab_cd133$fit.cluster)
# Limits of each cluster
lim_clust1 = c(min(protein_tab_cd133[protein_tab_cd133$fit.cluster==1,1]),
               max(protein_tab_cd133[protein_tab_cd133$fit.cluster==1,1]))
lim_clust2 = c(min(protein_tab_cd133[protein_tab_cd133$fit.cluster==2,1]),
               max(protein_tab_cd133[protein_tab_cd133$fit.cluster==2,1]))
lim_clust3 = c(min(protein_tab_cd133[protein_tab_cd133$fit.cluster==3,1]),
               max(protein_tab_cd133[protein_tab_cd133$fit.cluster==3,1]))
# Ascending order
limits = sort(c(lim_clust1, lim_clust2, lim_clust3))
# Threshold between the  clusters (use max of first cluster and min of second one)
var_kmean = c(mean(limits[2:3]), mean(limits[4:5]))
# Add the new groups into metadata of seurat object
cat("Limits cluster 1 :", lim_clust1, "=> low", "\n",
    "Limits cluster 2 :", lim_clust2, "=> high", "\n",
    "Limits cluster 3 :", lim_clust3, "=> medium", "\n",
    "Kmean values = ", var_kmean)
# Add kmean information in the tab
protein_tab_cd133 = protein_tab_cd133 %>%
  dplyr::mutate(CD133_three_groups = ifelse(fit.cluster==1, "CD133_low", 
                                           ifelse(fit.cluster==2, "CD34_high", "CD34_medium"))) %>%
  tibble::rownames_to_column(var = "cell") %>%
  dplyr::select(cell, CD133_three_groups)

# Add the new groups into metadata of seurat object
meta_tab = as.data.frame(merged_all@meta.data) %>% rownames_to_column(var = "cell")
meta_tab = full_join(x = meta_tab, y = protein_tab_cd133, by = "cell")
meta_tab = meta_tab %>% column_to_rownames(var = "cell")
merged_all@meta.data = meta_tab

# Show the groups
cd133_umap_three = DimPlot(object = merged_all,
             split.by = "orig.ident",
             group.by = "CD133_three_groups",
             reduction = "umap",
             ncol = 3)
cd133_umap_three

table(merged_all$CD133_three_groups)
table(merged_all$orig.ident, merged_all$CD133_three_groups)

```

```{r, CD34 CD133 three groups tab}

# Tab number
tab_three_34 = data.frame(table(merged_all@meta.data$CD34_three_groups, 
                          merged_all@meta.data$orig.ident)) %>%
  tidyr::pivot_wider(names_from = Var2, values_from = Freq) %>%
  dplyr::rename("Surface_Protein" = "Var1")

tab_three_133 = data.frame(table(merged_all@meta.data$CD133_three_groups, 
                           merged_all@meta.data$orig.ident)) %>%
  tidyr::pivot_wider(names_from = Var2, values_from = Freq) %>%
  dplyr::rename("Surface_Protein" = "Var1")

tab_raw_cond = rbind(tab_three_34, tab_three_133)

tab_raw_cond %>%
  kable(caption = "Raw cell distribution in conditions") %>%
  kable_styling()

# Tab number normalized
tot_nb_cells = tab_raw_cond[1,-1]+tab_raw_cond[2,-1]
tab_norm_cond = as.data.frame(round(mapply('/', tab_raw_cond[,-1] , tot_nb_cells)*1000))
tab_norm_cond$Surface_Protein = tab_raw_cond$Surface_Protein
tab_norm_cond = tab_norm_cond %>% dplyr::relocate(Surface_Protein, .before = "2DG")

tab_norm_cond %>%
  kable(caption = "Normalized cell distribution in conditions per number of cell per condition.") %>%
  kable_styling()

# Tab percent 
tab_percent_cond = as.data.frame(mapply('/', tab_norm_cond[,-1], 10))
tab_percent_cond$Surface_Protein = tab_norm_cond$Surface_Protein
tab_percent_cond = tab_percent_cond %>% dplyr::relocate(Surface_Protein, .before = "2DG")
tab_percent_cond %>%
  kable(caption = "Percentage of cells with high/low surface protein for each condition (from normalized per number of cells per condition numbers).") %>%
  kable_styling()

# save tables  
write.csv2(x = tab_raw_cond, row.names = FALSE,
           file = paste0(directory_output, "tab_raw_cond_three_groups.csv"))
write.csv2(x = tab_norm_cond, row.names = FALSE,
           file = paste0(directory_output, "tab_norm_cond_three_groups.csv"))
write.csv2(x = tab_percent_cond, row.names = FALSE,
           file = paste0(directory_output, "tab_percent_cond_three_groups.csv"))

```



# Membrane protein groups repartition in clusters

By now, we do not look at the data in terms of conditions but in terms of clusters.

```{r, ADT default}
DefaultAssay(merged_all) = "ADT"
```

```{r, UMAP, fig.width = 24, fig.height = 6}

umap_cluster = DimPlot(merged_all, 
                       label = TRUE, 
                       reduction = "umap", 
                       group.by = "clusters")

umap_two_cd34 = DimPlot(merged_all,
                    reduction = "umap", 
                    group.by = "CD34_two_groups")

umap_two_cd133 = DimPlot(merged_all,
                     reduction = "umap", 
                     group.by = "CD133_two_groups")

umap_cluster|umap_two_cd34|umap_two_cd133


umap_three_cd34 = DimPlot(merged_all,
                    reduction = "umap", 
                    group.by = "CD34_three_groups")

umap_three_cd133 = DimPlot(merged_all,
                     reduction = "umap", 
                     group.by = "CD133_three_groups")

umap_cluster|umap_three_cd34|umap_three_cd133

```

```{r, Vln, fig.width = 15, fig.height = 8}

vln_cd34 = VlnPlot(merged_all, features = "CD34.1",
                   pt.size = 0, group.by = "clusters") + 
  theme(legend.position = "top")

vln_cd133 = VlnPlot(merged_all, features = "CD133",
                    pt.size = 0, group.by = "clusters") + 
  theme(legend.position = "none")

vln_cd34/vln_cd133

```

```{r, tab two groups}

tab_34 = data.frame(table(merged_all@meta.data$CD34_two_groups, 
                          merged_all@meta.data$clusters)) %>%
  tidyr::pivot_wider(names_from = Var2, values_from = Freq) %>%
  dplyr::rename("Surface_Protein" = "Var1")

tab_133 = data.frame(table(merged_all@meta.data$CD133_two_groups, 
                           merged_all@meta.data$clusters)) %>%
  tidyr::pivot_wider(names_from = Var2, values_from = Freq) %>%
  dplyr::rename("Surface_Protein" = "Var1")

tab_raw_clusters = rbind(tab_34, tab_133)

tab_raw_clusters %>%
  kable(caption = "Raw cell distribution in clusters") %>%
  kable_styling()

# Tab number normalized
tot_nb_cells = tab_raw_clusters[1,-1]+tab_raw_clusters[2,-1]
tab_norm_clusters = as.data.frame(round(mapply('/', tab_raw_clusters[,-1] , tot_nb_cells)*1000))
tab_norm_clusters$Surface_Protein = tab_raw_clusters$Surface_Protein
tab_norm_clusters = tab_norm_clusters %>% dplyr::relocate(Surface_Protein, .before = "0")

tab_norm_clusters %>%
  kable(caption = "Normalized cell distribution in clusters") %>%
  kable_styling()

# Tab percent
tab_percent_clusters = as.data.frame(mapply('/', tab_norm_clusters[,-1], 10))
tab_percent_clusters$Surface_Protein = tab_norm_clusters$Surface_Protein
tab_percent_clusters = tab_percent_clusters %>% dplyr::relocate(Surface_Protein, .before = "0")
tab_percent_clusters %>%
  kable(caption = "Percentage of cells with high/low surface protein for each cluster.") %>%
  kable_styling()

# save tables
write.csv2(x = tab_raw_clusters, row.names = FALSE,
           file = paste0(directory_output, "tab_raw_clusters_two_groups.csv"))
write.csv2(x = tab_norm_clusters, row.names = FALSE,
           file = paste0(directory_output, "tab_norm_clusters_two_groups.csv"))
write.csv2(x = tab_percent_clusters, row.names = FALSE,
           file = paste0(directory_output, "tab_percent_clusters_two_groups.csv"))

```

```{r, tab three groups}

tab_34 = data.frame(table(merged_all@meta.data$CD34_three_groups, 
                          merged_all@meta.data$clusters)) %>%
  tidyr::pivot_wider(names_from = Var2, values_from = Freq) %>%
  dplyr::rename("Surface_Protein" = "Var1")

tab_133 = data.frame(table(merged_all@meta.data$CD133_three_groups, 
                           merged_all@meta.data$clusters)) %>%
  tidyr::pivot_wider(names_from = Var2, values_from = Freq) %>%
  dplyr::rename("Surface_Protein" = "Var1")

tab_raw_clusters = rbind(tab_34, tab_133)

tab_raw_clusters %>%
  kable(caption = "Raw cell distribution in clusters") %>%
  kable_styling()

# Tab number normalized
tot_nb_cells = tab_raw_clusters[1,-1]+tab_raw_clusters[2,-1]
tab_norm_clusters = as.data.frame(round(mapply('/', tab_raw_clusters[,-1] , tot_nb_cells)*1000))
tab_norm_clusters$Surface_Protein = tab_raw_clusters$Surface_Protein
tab_norm_clusters = tab_norm_clusters %>% dplyr::relocate(Surface_Protein, .before = "0")

tab_norm_clusters %>%
  kable(caption = "Normalized cell distribution in clusters") %>%
  kable_styling()

# Tab percent
tab_percent_clusters = as.data.frame(mapply('/', tab_norm_clusters[,-1], 10))
tab_percent_clusters$Surface_Protein = tab_norm_clusters$Surface_Protein
tab_percent_clusters = tab_percent_clusters %>% dplyr::relocate(Surface_Protein, .before = "0")
tab_percent_clusters %>%
  kable(caption = "Percentage of cells with high/low surface protein for each cluster.") %>%
  kable_styling()

# save tables
write.csv2(x = tab_raw_clusters, row.names = FALSE,
           file = paste0(directory_output, "tab_raw_clusters_three_groups.csv"))
write.csv2(x = tab_norm_clusters, row.names = FALSE,
           file = paste0(directory_output, "tab_norm_clusters_three_groups.csv"))
write.csv2(x = tab_percent_clusters, row.names = FALSE,
           file = paste0(directory_output, "tab_percent_clusters_three_groups.csv"))

```


# Conclusion

We tried heatmaps but as downsampling is a mandatory step for those, it's not very good representation of all the cells from the condition.   

```{r, CD34 and CD133 heatmaps, fig.height = 26, eval = FALSE}

# Random genes
conditions = unique(merged_all$orig.ident)
var_genes = merged_all@assays$SCT@var.features
random_genes = sample(var_genes, 200)

# CD34 heatmap
for (i in 1:length(conditions)) {

  obj = subset(merged_all, subset = orig.ident == conditions[i])
  plot = DoHeatmap(object = subset(obj, downsample = 1000), # Lost information but no choice here
          features = random_genes,
          group.by = "CD34_group",
          group.bar = TRUE,
          assay = "SCT",
          slot = "scale.data") +
    ggtitle(conditions[i])
  print(plot)
}

# CD133 heatmap
for (i in 1:length(conditions)) {

  obj = subset(merged_all, subset = orig.ident == conditions[i])
  plot = DoHeatmap(object = subset(obj, downsample = 1000), # Lost information but no choice here
          features = random_genes,
          group.by = "CD133_group",
          group.bar = TRUE,
          assay = "SCT",
          slot = "scale.data") +
    ggtitle(conditions[i])
  print(plot)
}

```

<br>

Further possible analysis :   
- choose a group resolution (two groups or three groups)  
- extract cells for each group  
- perform a GO on gene expressed in the cells of each group  
- see if some interesting categories are linked to the groups  

```{r, Save graphs, fig.show="hide", results="hide"}

# save plots
ggsave(plot = cd34_expr_distri | cd133_expr_distri, width = 14, height = 6,
       filename = paste0(directory_output, "adt_distri_expr.svg"))

ggsave(plot = do.call("grid.arrange", c(p1, ncol=3)),
       filename = paste0(directory_output, "umap_adt_cd34.svg"),
        width = 2, height = 16)
ggsave(plot = do.call("grid.arrange", c(p2, ncol=3)),
       filename = paste0(directory_output, "umap_adt_cd133.svg"),
       width = 22, height = 16)
ggsave(plot = do.call("grid.arrange", c(p3, ncol=3)),
       filename = paste0(directory_output, "umap_rna_cd34.svg"),
       width = 20, height = 16)
ggsave(plot = do.call("grid.arrange", c(p4, ncol=3)),
       filename = paste0(directory_output, "umap_rna_cd133.svg"),
       width = 20, height = 16)

ggsave(plot = cd34_umap_two, width = 20, height = 16,
       filename = paste0(directory_output, "umap_adt_cd34_twogroups.svg"))
ggsave(plot = cd133_umap_two, width = 20, height = 16,
       filename = paste0(directory_output, "umap_adt_cd133_twogroups.svg"))
ggsave(plot = cd34_umap_three, width = 20, height = 16,
       filename = paste0(directory_output, "umap_adt_cd34_threegroups.svg"))
ggsave(plot = cd133_umap_three, width = 20, height = 16,
       filename = paste0(directory_output, "umap_adt_cd133_threegroups.svg"))

ggsave(plot = sc1/sc2, width = 12, height = 16,
       filename = paste0(directory_output, "scatter_plot_adt.svg"))
ggsave(plot = umap_cluster|umap_two_cd34|umap_two_cd133, width = 24, height = 6,
       filename = paste0(directory_output, "umap_adt_clusters_twogroups.svg"))
ggsave(plot = umap_cluster|umap_three_cd34|umap_three_cd133, width = 24, height = 6,
       filename = paste0(directory_output, "umap_adt_clusters_threegroups.svg"))
ggsave(plot = vln_cd34/vln_cd133, width = 12, height = 8,
       filename = paste0(directory_output, "vln_adt_clusters.svg"))

saveRDS(object = merged_all, paste0(directory_output, "merged_all_CITEgroups.rds"))

```

```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, fig.show="hide", results="hide"}

# Clean working space and memory
rm(list = ls())
gc()

```

