---
title: "scRNAseq_CITEseq_R_DWNS_CompareConditions_GO"
author: "Laëtitia Racine"
date: "2022-12-15"
subtitle: "Last modification : `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: "hide"
    toc: true
    toc_float: false
    theme: journal
---

<style>
body {text-align: justify}
</style>

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```

```{r, Dependencies}

library(Seurat)
library(dplyr)
library(stringr)
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(ggrepel)
library(org.Hs.eg.db) # GO functions
library(clusterProfiler) # GO functions
library(kableExtra)
# library(xlsx)
library(XLConnect)

```


# Overview

This script aims at comparing conditions (CTRL vs drug) in terms of gene expression. We extract list of differentially expressed genes (on/off analysis or differential analysis) and we perform gene ontology on those list to see if we find something relevant.  

Based on :  
- https://satijalab.org/seurat/articles/de_vignette.html  
- https://satijalab.org/seurat/articles/sctransform_v2_vignette.html  
- https://ycl6.github.io/GO-Enrichment-Analysis-Demo/  
- https://satijalab.org/seurat/archive/v3.1/immune_alignment.html  
- https://stackoverflow.com/questions/27713310/easy-way-to-export-multiple-data-frame-to-multiple-excel-worksheets  
- https://stackoverflow.com/questions/66646394/dplyr-mutate-grouped-data-without-using-exact-column-name  
- https://rdrr.io/cran/XLConnect/man/writeWorksheet-methods.html  

```{r, Working directories and external script}

# Load working directories
directory = str_extract(string = getwd(), pattern = "[:graph:]+(?=bin)")
start_time = Sys.time()

# Create a unique folder for output corresponding to the date of the day
current_date = format(Sys.time(), "%Y%m%d")
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_CompareConditions/"))
dir.create(path = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_CompareConditions/", current_date))
directory_output = paste0(directory, "exp/scRNAseq_CITEseq_R_DWNS_CompareConditions/", current_date, "/")

# Load external script with functions and constants
source(file = paste0(directory, "bin/", "functions_constants.R"))

```

```{r, Input loading}

dir = pic_last_dir(paste0(directory, "exp/", "scRNAseq_CITEseq_R_ReductionDimension/"))
merged_all = readRDS(file = paste0(dir, "/", "merged_all_prep_1-40dim.rds"))

dir_corr = pic_last_dir(paste0(directory, "exp/", "scRNAseq_CITEseq_R_GenomePosition/"))
tab_corr = read.csv2(paste0(dir_corr, "/scRNAseq_correspondence_tab_EnsemblID_geneName_genomePos.csv")) %>% 
    dplyr::select(gene_symbol_dupl, ensembl_gene_id) %>%
    dplyr::rename(gene = "gene_symbol_dupl") %>%
    dplyr::distinct()  # to remove duplicated genes

gene_universe = read.delim(paste0(directory, "data/scRNAseq/", 
                                  "scRNAseq_CTRL_sample_feature_bc_matrix_features.tsv.gz"), 
                                  header = FALSE)[,1]

```

<br>

Choose the combinations we want to analyse.
```{r, Combinations}

combination = list(c("CTRL","AOA"), c("CTRL","2DG"), c("CTRL","DON"),
                   c("CTRL", "VPA"), c("CTRL","CTRLaK"), c("2DG","2DGaK"),
                   c("DON","DONaK"), c("AOA","AOAaK"), c("CTRLaK","2DGaK"),
                   c("CTRLaK","AOAaK"), c("CTRLaK","DONaK"))

```



# Qualitative analysis : expressed / no expressed

<br>

On/Off analysis : we created a binary matrix expressed/no expressed to remove the variation of the number of UMI. 

```{r}

list_quali_analysis = list()

```

## Binary matrix 

```{r, Binary matrix expressed/no expressed}

# extract count matrix from SCT assay and transform into data frame
# we can work with counts slot because it's not a quantitative analysis 
# just expressed/no expressed, no need for normalized data
# Split into list of expressed and no expressed genes
gene_expr_fun = function(seurat_obj, condition) {
  
  tab = subset(seurat_obj, subset = orig.ident %in% condition)
  tab = as.data.frame(tab@assays$SCT@counts) %>% mutate_all(as.logical)
  tab = data.frame(gene = rownames(tab), expression = rowSums(tab))
  tab$expression = as.logical(tab$expression)
  tab_exp = tab %>% dplyr::filter(expression == TRUE)
  tab_noexp = tab %>% dplyr::filter(expression == FALSE)
  return(list(tab_exp = tab_exp, tab_noexp = tab_noexp))
  
}

list_bin_mat = list("CTRL" = gene_expr_fun(merged_all, c("CTRL")),
                    "AOA" = gene_expr_fun(merged_all, "AOA"),
                    "2DG" = gene_expr_fun(merged_all, "2DG"),
                    "DON" = gene_expr_fun(merged_all, "DON"),
                    "CTRLaK" = gene_expr_fun(merged_all, "CTRLaK"),
                    "AOAaK" = gene_expr_fun(merged_all, "AOAaK"),
                    "2DGaK" = gene_expr_fun(merged_all, "2DGaK"),
                    "DONaK" = gene_expr_fun(merged_all, "DONaK"),
                    "VPA" = gene_expr_fun(merged_all, "VPA"))

```

## Extract list of genes 

```{r}

# Loop for combination of comparisons
for (i in 1:length(combination)) {
  
  cond = paste(combination[[i]], collapse = "_")
  name_cond1 = combination[[i]][1]
  name_cond2 = combination[[i]][2]
  expr_cond1 = list_bin_mat[[name_cond1]]$tab_exp
  expr_cond2 = list_bin_mat[[name_cond2]]$tab_exp
  
  #################################
  ### Genes lost across conditions
  #################################
  
  # elements in cond1 but not in cond2
  diff_cond1_cond2 = setdiff(expr_cond1$gene, expr_cond2$gene)
  # elements in cond2 but not in cond1
  diff_cond2_cond1 = setdiff(expr_cond2$gene, expr_cond1$gene)
  # save in the list 
  list_quali_analysis[[paste0("diff_", name_cond1, "_", name_cond2)]]$gene_list = diff_cond1_cond2
  list_quali_analysis[[paste0("diff_", name_cond2, "_", name_cond1)]]$gene_list = diff_cond2_cond1
  
  #####################################
  ### Genes conserved across conditions
  #####################################
  
  # common elements between cond1 and cond2
  inter_cond1_cond2 = intersect(expr_cond1$gene, expr_cond2$gene)
  # save in the list
  list_quali_analysis[[paste0("inter_", name_cond1, "_", name_cond2)]]$gene_list = inter_cond1_cond2
  
}

```

```{r}

# Summarize in a dataframe
tab_diff = data.frame(comparison = names(list_quali_analysis),
                      nbgenes_comparison = lengths(unlist(list_quali_analysis, recursive=FALSE))) %>%
  tidyr::separate(col = comparison, into = c("comparison", "cond1", "cond2"), sep = "_", remove = TRUE) %>%
  dplyr::select(cond1, cond2, comparison, nbgenes_comparison)

tab_genes_expressed = as.data.frame(sapply(unlist(list_bin_mat, recursive=FALSE), nrow)) %>%
  tibble::rownames_to_column("condition") %>%
  dplyr::filter(str_detect(condition, "tab_exp")) %>%
  tidyr::separate(col = condition, into = c("condition", "trash"), remove = TRUE)
colnames(tab_genes_expressed) = c("condition", "trash", "nbgenes_expressed")
tab_genes_expressed = tab_genes_expressed %>% dplyr::select(-trash)

tab_diff = tab_diff %>% dplyr::rename(condition = "cond1")
tab_diff = full_join(tab_diff, tab_genes_expressed, by = "condition") %>%
  dplyr::rename(cond1 = "condition", nbgenes_expressed_cond1 = "nbgenes_expressed") %>%
  dplyr::rename(condition = "cond2")
tab_diff = full_join(tab_diff, tab_genes_expressed, by = "condition") %>%
  dplyr::rename(cond2 = "condition", nbgenes_expressed_cond2 = "nbgenes_expressed") %>%
  dplyr::select(cond1, nbgenes_expressed_cond1, cond2, nbgenes_expressed_cond2, comparison, nbgenes_comparison)

tab_diff %>%
  kable() %>%
  kable_styling()

```

```{r}

#####################################################################
### Genes expressed or not expressed in all conditions simultaneously
#####################################################################
  
all_cond = unlist(list_bin_mat,recursive=FALSE)

all_cond_noexpr = all_cond[str_detect(names(all_cond), "tab_noexp")]
all_cond_noexpr = unlist(all_cond_noexpr, recursive=FALSE)
all_cond_noexpr = all_cond_noexpr[str_detect(names(all_cond_noexpr), "gene")]
genes_absent = Reduce(intersect, all_cond_noexpr)
list_quali_analysis[["Never_Expressed"]]$gene_list = genes_absent

all_cond_expr = all_cond[str_detect(names(all_cond), "tab_exp")]
all_cond_expr = unlist(all_cond_expr, recursive=FALSE)
all_cond_expr = all_cond_expr[str_detect(names(all_cond_expr), "gene")]
genes_all = Reduce(intersect, all_cond_expr)
list_quali_analysis[["Always_Expressed"]]$gene_list = genes_all

cat("Number of genes absent in all conditions (even CTRL) :", length(genes_absent), "\n",
    "Number of genes present in all conditions (even CTRL) :", length(genes_all), "\n")

###########################################################
### Genes expressed in CTRL but not in any other conditions 
###########################################################

# Genes expressed in drug conditions
others = unlist(list_bin_mat[names(list_bin_mat) != "CTRL"], recursive=FALSE)
others = unlist(others, recursive=FALSE)
others = others[str_detect(names(others), "tab_exp.gene")]
others = unique(unlist(others, recursive=FALSE))

# Genes expressed in ctrl condition
ctrl = list_bin_mat[["CTRL"]]$tab_exp$gene

# Genes expressed in ctrl but not in others
diff_ctrl = setdiff(ctrl, others)
list_quali_analysis[["CTRL_only"]]$gene_list = diff_ctrl

cat("Number of genes expressed in CTRL but absent in every other conditions :", 
    length(diff_ctrl), "\n")

###########################################################
### Genes expressed in AOA but not in any other conditions
###########################################################

# Genes expressed in AOA condition
aoa = list_bin_mat[["AOA"]]$tab_exp$gene

# Genes expressed in others conditions
others = unlist(list_bin_mat[names(list_bin_mat) != "AOA"], recursive=FALSE)
others = unlist(others, recursive=FALSE)
others = others[str_detect(names(others), "tab_exp.gene")]
others = unique(unlist(others, recursive=FALSE))

# Genes expressed in ctrl but not in others
diff_aoa = setdiff(aoa, others)
list_quali_analysis[["AOA_only"]]$gene_list = diff_aoa

cat("Number of genes expressed in AOA but absent in every other conditions :", 
    length(diff_aoa), "\n")

###########################################################
### Genes expressed in 2DG but not in any other conditions
###########################################################

# Genes expressed in 2DG condition
ddg = list_bin_mat[["2DG"]]$tab_exp$gene

# Genes expressed in others conditions
others = unlist(list_bin_mat[names(list_bin_mat) != "2DG"], recursive=FALSE)
others = unlist(others, recursive=FALSE)
others = others[str_detect(names(others), "tab_exp.gene")]
others = unique(unlist(others, recursive=FALSE))

# Genes expressed in ctrl but not in others
diff_ddg = setdiff(ddg, others)
list_quali_analysis[["2DG_only"]]$gene_list = diff_ddg

cat("Number of genes expressed in 2DG but absent in every other conditions :", 
    length(diff_ddg), "\n")

```

```{r}

as.data.frame(lengths(unlist(list_quali_analysis, recursive=FALSE))) %>%
  kable() %>%
  kable_styling()

```

## Perform GO on list of genes

```{r}

for (i in 1:length(list_quali_analysis)) {
  
  print(names(list_quali_analysis[i]))
  tab_gene = data.frame(list_quali_analysis[[i]]$gene_list)

  if(nrow(tab_gene) == 0) { 
    print("No gene in the list.") 
    
  } else {
    GO_results = GO_clusterProfiler_fun(tab_corr = tab_corr,
                                        list_gene = tab_gene,
                                        gene_universe = gene_universe,
                                        title_plot = names(list_quali_analysis[i]))
    list_quali_analysis[[names(list_quali_analysis[i])]]$GO_results = GO_results
  }
  
} 

```

```{r, fig.height = 8, fig.width = 8}

list_GO_plot_quali = unlist(list_quali_analysis,recursive=FALSE)
list_GO_plot_quali = list_GO_plot_quali[str_detect(names(list_GO_plot_quali), "GO_results")]
list_GO_plot_quali = unlist(list_GO_plot_quali,recursive=FALSE)
list_GO_plot_quali = list_GO_plot_quali[str_detect(names(list_GO_plot_quali), "plot")]
list_GO_plot_quali

```

```{r, Save outputs qualitative, fig.show='hide'}

# Save binary matrix and GO list of genes for further R analysis
saveRDS(list_bin_mat, paste0(directory_output, "list_binary_mat.rds"))
saveRDS(list_quali_analysis, paste0(directory_output, "list_compCond_quali.rds"))

# Save excel file (list of genes and GO tables)
# wb = createWorkbook()  
# for (i in 1:length(list_quali_analysis)) {
#   print(i)
#   name = names(list_quali_analysis[i])
#   sheet = createSheet(wb, name)
#   addDataFrame(list_quali_analysis[[name]]$gene_list, sheet=sheet, startColumn=1, row.names=FALSE)
#   addDataFrame(list_quali_analysis[[name]]$GO_results$ggo, sheet=sheet, startColumn=3, row.names=FALSE)
#   addDataFrame(list_quali_analysis[[name]]$GO_results$ego, sheet=sheet, startColumn=10, row.names=FALSE)
# }
# saveWorkbook(wb, paste0(directory_output, "list_quali_analysis.xlsx"))

# wb = loadWorkbook(filename =  paste0(directory_output, "list_quali_analysis.xlsx"), create = TRUE)
# for (i in 1:length(list_quali_analysis)) {
#   print(i)
#   name = names(list_quali_analysis[i])
#   createSheet(wb, name = name)
#   writeWorksheet(wb, list_quali_analysis[[name]]$gene_list, sheet=name, startRow=1, startCol=1)
#   writeWorksheet(wb, list_quali_analysis[[name]]$GO_results$ggo, sheet=name, startRow=1, startCol=3)
#   writeWorksheet(wb, list_quali_analysis[[name]]$GO_results$ego, sheet=name, startRow=1, startCol=10)
# }
# saveWorkbook(wb)

# Save plots
for (i in 1:length(list_GO_plot_quali)) {
  plot = list_GO_plot_quali[[i]]
  if (str_detect(names(list_GO_plot_quali[i]), "barplot")) { suffix = "barplot.svg" }
  if (str_detect(names(list_GO_plot_quali[i]), "dotplot")) { suffix = "dotplot.svg" }
  prefix = str_extract(names(list_GO_plot_quali[i]), "[:graph:]+(?=.[:alpha:]{3}plot_25)")
  ggsave(plot = plot,
         filename = paste0(directory_output, prefix, "_GO_plots_quali_", suffix),
         width = 8, height = 12)
}

```

```{r}

rm(list=setdiff(ls(), c("directory_output", "current_date", "start_time", "color_code",  
                        "merged_all", "tab_corr", "gene_universe", "combination",
                        "GO_clusterProfiler_fun")))

```


# Quantitative analysis : differential analysis

<br>

Number of UMI taken in account.  

```{r}

list_quanti_analysis = list()

```

## List of genes : compare conditions two by two 

### Visual outliers = average gene expression comparison

One way to look broadly at these changes is to plot the average expression of both the stimulated and control cells and look for genes that are visual outliers on a scatter plot. 

```{r, fig.width = 8, fig.height = 6}

Idents(merged_all) = "orig.ident"

for (i in 1:length(combination)) {
  
  cond = paste(combination[[i]], collapse = "_")
  
  # subset to keep cells from those two conditions
  temp = subset(merged_all, subset = orig.ident %in% combination[[i]])
  print(table(temp$orig.ident))
  
  # calcul average expression
  avg = as.data.frame(log1p(AverageExpression(temp)$SCT)) 
  
  # identify interesting genes
  avg = avg %>%
    dplyr::mutate(pos_threshold = cur_data()[[1]]+0.5) %>%
    dplyr::mutate(neg_threshold = cur_data()[[1]]-0.5) %>%
    dplyr::mutate(sign = ifelse(cur_data()[[2]] > neg_threshold & cur_data()[[2]] < pos_threshold, FALSE, TRUE)) %>%
    tibble::rownames_to_column(var = "gene") 
  
  # plot the result
  plot = ggplot(avg, aes(x = avg[,2], y = avg[,3], label = gene)) + 
    geom_point() + 
    geom_abline(slope = 1, intercept = 0, linetype = 2, col = "grey") +
    geom_abline(slope = 1, intercept = 0.5, linetype = 2, col = "red") +
    geom_abline(slope = 1, intercept = -0.5, linetype = 2, col = "red") +
    xlab(colnames(avg)[2]) +
    ylab(colnames(avg)[3]) +
    ggtitle(cond) +
    geom_label_repel(aes(label = ifelse(sign == TRUE, gene, "")),
                     segment.color = "black", max.overlaps = Inf, 
                     min.segment.length = 0.5, size = 3, direction ="both",
                     box.padding = 0.5, force = 2)
  print(plot)
  
  # significatives markers 
  avg_sign = avg %>% dplyr::filter(sign == TRUE)
  pos_sign = avg_sign %>% dplyr::filter(avg_sign[,2] > avg_sign[,3])
  neg_sign = avg_sign %>% dplyr::filter(avg_sign[,2] < avg_sign[,3])
  
  list_quanti_analysis[[cond]]$avg_outliers = list(
    gene_all = avg,
    gene_all_sign = avg_sign,
    gene_pos_sign = pos_sign,
    gene_neg_sign = neg_sign
  )
  
}

```

### Differential Expression between two conditions with FindMarkers

```{r}

for (i in 1:length(combination)) {
  
  cond1 = combination[[i]][1]
  cond2 = combination[[i]][2]
  cond = paste(combination[[i]], collapse = "_")
  
  markers = FindMarkers(object = merged_all, 
                        ident.1 = cond1, # all cells from cond1
                        ident.2 = cond2, # all cells from cond2
                        features = NULL, # all genes
                        logfc.threshold = 0.25,
                        test.use = "wilcox",
                        only.pos = FALSE,
                        assay = "SCT",
                        slot = "data") %>%
    dplyr::filter(p_val_adj <= 0.05) %>%
    tibble::rownames_to_column("gene")  %>% 
    dplyr::mutate(conditions = cond)
  
  pos_markers = markers %>% 
    dplyr::filter(avg_log2FC >= 0) %>%
    dplyr::arrange(p_val_adj) %>%
    dplyr::select(conditions, gene, p_val_adj, everything())
  
  neg_markers = markers %>%
    dplyr::filter(avg_log2FC < 0) %>%
    dplyr::arrange(p_val_adj) %>%
    dplyr::select(conditions, gene, p_val_adj, everything())
  
  list_quanti_analysis[[cond]]$FM_DE = list(
    gene_all_sign = markers,
    gene_pos_sign = pos_markers,
    gene_neg_sign = neg_markers
  )
  
}

```

### Compare lists obtained with the two methods 

```{r}

for( i in 1:length(combination)) {
  
  cond = paste(combination[[i]], collapse = "_")
  method_1 = list_quanti_analysis[[cond]]$avg_outliers$gene_all_sign$gene
  method_2 = list_quanti_analysis[[cond]]$FM_DE$gene_all_sign$gene
  diff = setdiff(method_1, method_2)
  common = intersect(method_1, method_2)
  cat("For", cond, ", we have found", length(method_1), 
      "genes with the method of average expression outliers and",
      length(method_2), "genes with the method of FindMarkers (Wilcoxon DE).",
      length(common), "/", length(method_1), "genes are common with both method and",
      length(diff), "/", length(method_1), "genes are found as outliers but not significant in DE analysis.",
      "\n", "Common genes :", common, "\n", "Different genes :", diff, "\n", "\n")
  
  list_quanti_analysis[[cond]]$comp = list(
    common_avg_FM = common,
    diff_avg_FM = diff
  )
  
}

```

## Gene ontology

We perform GO on list of genes from both method.

```{r}

for (i in 1:length(list_quanti_analysis)) {
  
  print(names(list_quanti_analysis[i]))
  
  # Average expression - Pos markers
  tab_gene_avgP = data.frame(list_quanti_analysis[[i]]$avg_outliers$gene_pos_sign$gene)

  if(nrow(tab_gene_avgP) == 0) { 
    print("No gene in the list.") 
    
  } else {
    GO_results_avgP = GO_clusterProfiler_fun(tab_corr = tab_corr,
                                          list_gene = tab_gene_avgP,
                                          gene_universe = gene_universe,
                                          title_plot = paste0("avg positive markers -", names(list_quanti_analysis[i])))
    
  }
  
  # Average expression - Neg markers
  tab_gene_avgN = data.frame(list_quanti_analysis[[i]]$avg_outliers$gene_neg_sign$gene)

  if(nrow(tab_gene_avgN) == 0) { 
    print("No gene in the list.") 
    
  } else {
    GO_results_avgN = GO_clusterProfiler_fun(tab_corr = tab_corr,
                                          list_gene = tab_gene_avgN,
                                          gene_universe = gene_universe,
                                          title_plot = paste0("avg negative markers -", names(list_quanti_analysis[i])))
    
  }
  
  # FindMarkers - positive markers (more expressed in cond1)
  tab_gene_FMp = data.frame(list_quanti_analysis[[i]]$FM_DE$gene_pos_sign$gene)

  if(nrow(tab_gene_FMp) == 0) { 
    print("No gene in the list.") 
    
  } else {
    GO_results_FMp = GO_clusterProfiler_fun(tab_corr = tab_corr,
                                        list_gene = tab_gene_FMp,
                                        gene_universe = gene_universe,
                                          title_plot = paste0("FM positive markers -", names(list_quanti_analysis[i])))
    
  }
  
  # FindMarkers - negative markers (more expressed in cond2)
  tab_gene_FMn = data.frame(list_quanti_analysis[[i]]$FM_DE$gene_neg_sign$gene)

  if(nrow(tab_gene_FMn) == 0) { 
    print("No gene in the list.") 
    
  } else {
    GO_results_FMn = GO_clusterProfiler_fun(tab_corr = tab_corr,
                                          list_gene = tab_gene_FMn,
                                          gene_universe = gene_universe,
                                          paste0("FM negative markers -", names(list_quanti_analysis[i])))
    
  }

   list_quanti_analysis[[names(list_quanti_analysis[i])]]$GO_results = list(
      GO_avg_pos = GO_results_avgP,
      GO_avg_neg = GO_results_avgN,
      GO_FM_pos = GO_results_FMp,
      GO_FM_neg = GO_results_FMn
  )
  
} 

```

```{r, fig.height = 10, fig.width = 10}

list_GO_plot_quanti = unlist(list_quanti_analysis,recursive=FALSE)
list_GO_plot_quanti = list_GO_plot_quanti[str_detect(names(list_GO_plot_quanti), "GO_results")]
list_GO_plot_quanti = unlist(unlist(list_GO_plot_quanti,recursive=FALSE), recursive=FALSE)
list_GO_plot_quanti = list_GO_plot_quanti[str_detect(names(list_GO_plot_quanti), "plot")]
list_GO_plot_quanti

```

```{r, Save outputs quantitative}

# Save list of genes and tables for further R analysis
saveRDS(list_quanti_analysis, paste0(directory_output, "list_compCond_quanti.rds"))

# Save plots
for (i in 1:length(list_GO_plot_quanti)) {
  plot = list_GO_plot_quanti[[i]]
  if (str_detect(names(list_GO_plot_quanti[i]), "barplot")) { suffix = "barplot.svg" }
  if (str_detect(names(list_GO_plot_quanti[i]), "dotplot")) { suffix = "dotplot.svg" }
  prefix = str_extract(names(list_GO_plot_quanti[i]), "[:graph:]+(?=.[:alpha:]{3}plot_25)")
  ggsave(plot = plot,
         filename = paste0(directory_output, prefix, "_GO_plots_quanti_", suffix),
         width = 8, height = 12)
}

# Save excel file (list of genes and GO tables)
# for (i in 1:length(list_quanti_analysis)) {
#   # create file
#   name = names(list_quanti_analysis[i])
#   wb = createWorkbook()
#   # sheet 1 for avg pos tables
#   sheet = createSheet(wb, paste0(name, "_avg_pos"))
#   addDataFrame(list_quanti_analysis[[name]]$avg_outliers$gene_pos_sign, sheet=sheet, startColumn=1, row.names=FALSE)
#   addDataFrame(list_quanti_analysis[[name]]$GO_results$GO_avg_pos$ggo, sheet=sheet, startColumn=8, row.names=FALSE)
#   addDataFrame(list_quanti_analysis[[name]]$GO_results$GO_avg_pos$ego, sheet=sheet, startColumn=14, row.names=FALSE)
#   # sheet 2 for avg neg tables
#   sheet = createSheet(wb, paste0(name, "_avg_neg"))
#   addDataFrame(list_quanti_analysis[[name]]$avg_outliers$gene_neg_sign, sheet=sheet, startColumn=1, row.names=FALSE)
#   addDataFrame(list_quanti_analysis[[name]]$GO_results$GO_avg_neg$ggo, sheet=sheet, startColumn=8, row.names=FALSE)
#   addDataFrame(list_quanti_analysis[[name]]$GO_results$GO_avg_neg$ego, sheet=sheet, startColumn=14, row.names=FALSE)
#   # sheet 3 for FM pos markers tables
#   sheet = createSheet(wb, paste0(name, "_FM_pos"))
#   addDataFrame(list_quanti_analysis[[name]]$FM_DE$gene_pos_sign, sheet=sheet, startColumn=1, row.names=FALSE)
#   addDataFrame(list_quanti_analysis[[name]]$GO_results$GO_FM_pos$ggo, sheet=sheet, startColumn=9, row.names=FALSE)
#   addDataFrame(list_quanti_analysis[[name]]$GO_results$GO_FM_pos$ego, sheet=sheet, startColumn=15, row.names=FALSE)
#   # sheet 4 for FM neg markers tables
#   sheet = createSheet(wb, paste0(name, "_FM_neg"))
#   addDataFrame(list_quanti_analysis[[name]]$FM_DE$gene_neg_sign, sheet=sheet, startColumn=1, row.names=FALSE)
#   addDataFrame(list_quanti_analysis[[name]]$GO_results$GO_FM_neg$ggo, sheet=sheet, startColumn=9, row.names=FALSE)
#   addDataFrame(list_quanti_analysis[[name]]$GO_results$GO_FM_neg$ego, sheet=sheet, startColumn=15, row.names=FALSE)
#   # save file
#   saveWorkbook(wb, paste0(directory_output, name, "_quali_analysis.xlsx"))
# }

# for (i in 1:length(list_quanti_analysis)) {
#   # create file
#   name = names(list_quanti_analysis[i])
#   wb = loadWorkbook(filename =  paste0(directory_output, name, "_quanti_analysis.xlsx"), create = TRUE)
#   # sheet 1 for avg pos tables
#   sheetname = paste0(name, "_avg_pos")
#   createSheet(wb, name = sheetname)
#   writeWorksheet(wb, list_quanti_analysis[[name]]$avg_outliers$gene_pos_sign, sheet=sheetname, startRow=1, startCol=1)
#   writeWorksheet(wb, list_quanti_analysis[[name]]$GO_results$GO_avg_pos$ggo, sheet=sheetname, startRow=1, startCol=8)
#   writeWorksheet(wb, list_quanti_analysis[[name]]$GO_results$GO_avg_pos$ego, sheet=sheetname, startRow=1, startCol=14)
#   # sheet 2 for avg neg tables
#   sheetname = paste0(name, "_avg_neg")
#   createSheet(wb, name = sheetname)
#   writeWorksheet(list_quanti_analysis[[name]]$avg_outliers$gene_neg_sign, sheet=sheetname, startRow=1, startCol=1)
#   writeWorksheet(list_quanti_analysis[[name]]$GO_results$GO_avg_neg$ggo, sheet=sheetname, startRow=1, startCol=8)
#   writeWorksheet(list_quanti_analysis[[name]]$GO_results$GO_avg_neg$ego, sheet=sheetname, startRow=1, startCol=14)
#   # sheet 3 for FM pos markers tables
#   sheetname = paste0(name, "_FM_pos")
#   createSheet(wb, name = sheetname)
#   writeWorksheet(list_quanti_analysis[[name]]$FM_DE$gene_pos_sign, sheet=sheetname, startRow=1, startCol=1)
#   writeWorksheet(list_quanti_analysis[[name]]$GO_results$GO_FM_pos$ggo, sheet=sheetname, startRow=1, startCol=9)
#   writeWorksheet(list_quanti_analysis[[name]]$GO_results$GO_FM_pos$ego, sheet=sheetname, startRow=1, startCol=15)
#   # sheet 4 for FM neg markers tables
#   sheetname = paste0(name, "_FM_neg")
#   createSheet(wb, name = sheetname)
#   writeWorksheet(list_quanti_analysis[[name]]$FM_DE$gene_neg_sign, sheet=sheetname, startRow=1, startCol=1)
#   writeWorksheet(list_quanti_analysis[[name]]$GO_results$GO_FM_neg$ggo, sheet=sheetname, startRow=1, startCol=9)
#   writeWorksheet(list_quanti_analysis[[name]]$GO_results$GO_FM_neg$ego, sheet=sheetname, startRow=1, startCol=15)
#   # save file
#   saveWorkbook(wb)
# }

```


```{r, Rsession}

end_time = Sys.time()
cat("Total execution time : ", as.numeric (end_time - start_time, units = "mins"), "minutes")

# Show package version
sessionInfo()

```

```{r, results = "hide"}

# Clean working space and memory
rm(list = ls())
gc()

```
